### 2020-07-07 manuscript.tex
Initial Overleaf Import

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article]{achemso}

%\documentclass[english,aps,preprint,pre,floatfix,nofootinbib,showpacs,showkeys]{revtex4-1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later. Do NOT use any
%% packages which require e-TeX (for example etoolbox): the e-TeX
%% extensions are not currently available on the ACS conversion
%% servers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}

\SectionNumbersOn

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{CJK}https://www.overleaf.com/project/5e820b28d741e00001b1ccac
\usepackage{framed}
\usepackage[hidelinks]{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\makeatletter
    \setlength\@fptop{0\p@}
\makeatother

\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}

%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauwr}[1]{#1}

%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}

%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
\newcommand*{\noter}[1]{}					% notes off

%\usepackage[draft]{todonotes}   % notes shown
\usepackage[disable]{todonotes} % notes hidden

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Mike Jones}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  University of Chicago, %
  Chicago, Illinois 60637%
}

\author{Andrew L. Ferguson}
\email{andrewferguson@uchicago.edu}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  University of Chicago, %
  Chicago, Illinois 60637%
}
%\author{I. Ken Groupleader}
%\altaffiliation{A shared footnote}
%\email{i.k.groupleader@unknown.uu}
%\phone{+123 (0)123 4445556}
%\fax{+123 (0)123 4445557}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%\author{Susanne K. Laborator}
%\email{s.k.laborator@bigpharma.co}
%\affiliation[BigPharma]
%{Lead Discovery, BigPharma, Big Town, USA}

%\author{Kay T. Finally}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]{TBA}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\abbreviations{IR -- infrared, NMR -- nuclear magnetic resonance, UV -- ultraviolet}
%\keywords{Takens}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.  The document should begin with an
%% abstract, if appropriate.  If one is given and should not be, the
%% contents will be gobbled.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\begin{abstract}

\noindent

\end{abstract}
https://www.overleaf.com/project/5e9e5110c524b8000192c548
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{\label{sec:intro}Introduction}

DNA motivation
\begin{enumerate}
	\item Connection to Tokmakoff results
	\item Some origami, sensor, other nanotech appications
\end{enumerate}

SRV motivation
\begin{enumerate}
	\item First time studying a two body problem
	\item Compare with other TICA-MSM approaches to DNA
	\item Compare with Noe nucleotide paper
\end{enumerate}

\section{\label{sec:methods}Methods}

Simulation set up

Featurization
\begin{enumerate}
	\item Vamp-2 score for permutable distances is slightly lower but should not contain degenerate information (SI data)
	\item Inverse distances have higher vamp-2 score than regular distances (SI data)
\end{enumerate}

SRV validation
\begin{enumerate}
	\item Plot loss scores and self-consistency check for each sequence (fig 1 and 2)
	\item Select number of slow modes (different for each) and lag time (same for each) based on this analysis.
	\item Only use off diagonals and add sequence visuals along axes (fig 3)
\end{enumerate}

MSM construction
\begin{enumerate}
	\item hyperparameters include: number of microstates, lag (use same as SRV), number of macrostates (usually n+1) but we use n states for GC-core and GC-mid. I have not done a full optimization sweep over these parameters, but can explain the motivation for choosing these.
    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) (fig 4)
\end{enumerate}

\section{\label{sec:methods}Discussion}

Brief description and diagrams for collective variables:
\begin{enumerate}
	\item Inverse average shifting distance
	\item Fourth basepair distances
\end{enumerate}

AT-all
\begin{enumerate}
	\item Dynamics dominated by slow shifting modes
	\item sources to verify this would be the case
	\item Minimal data from Tokmakoff on this, but can look at relative hybridization rates
	\item SI Calculations to prove preferred thermo stability of 5’ shifted state
\end{enumerate}

GC-end
\begin{enumerate}
	\item Similar dynamics to AT-all, but with overall lower stability and timescales
	\item Difficult to make any claims because the thermodynamics of external mismatches with dangling ends have not been measured experimentally, and 3spn2 does not paramertize these mismatches explicitly
	\item According to SantaLucia external mismatches and dangling ends tend to have weakly stabilizing effect, and there are 4-6 intact basepairs for these conformations
	\item Might account for unknown signal in 2018 Tokmakoff paper with timescale between the fraying and dehybrization signal, but not sure... 
\end{enumerate}

GC-core
\begin{enumerate}
	\item Frayed states requires the fourth base pair to be broken, most partially frayed states are not included in metastable clustering
	\item 2nd slow mode correlates well with expected fraying behavior, can sample along this mode and generally find increasing frayed states (although the SRV coords are spiky)
	\item 3rd mode is reflective of changes in orientation in dissociated behavior. This could be interesting for the SI, but this state does not inform the macrostate clustering
\end{enumerate}  
GC-mid
\begin{enumerate}
	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
\end{enumerate}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/n_events.PNG}
        \caption{Number of events per sequence. I would only included the four sequences in the actual pub. (this could be supplemental}
        \label{fig:n_events}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_validation.PNG}
        \caption{For Supplemental info: Selection process for AT-all slow modes}
        \label{fig:AT-all_val}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_validation.PNG}
        \caption{For Supplemental info: Selection process for GC-core slow modes}
        \label{fig:GC-core_val}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/all_modes_correlations.PNG}
        \caption{For Supplemental info: Correlations between all leading modes}
        \label{fig:all_modes}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/tica_comparison_GC-core.PNG}
        \caption{SRV vs. TICA MSMs and resulting clsuter memberships}
        \label{fig:tica_comparison}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends}
        \label{fig:shifting_distributions}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/sample_fray.PNG}
        \caption{Sample along 2nd GC-core SRV mode and show that these have increasing amounts of fraying}
        \label{fig:sample_fray}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_vis.png}
        \caption{AT-all transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
        \label{fig:sample_fray}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_vis.png}
        \caption{GC-core transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
        \label{fig:sample_fray}
	\end{center}
\end{figure}

\clearpage
\newpage

\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\end{document}

### 2020-07-14 manuscript.tex
First pull
more detailed outline

@@
 \section{\label{sec:intro}Introduction}
+Should be 4-5 paragraphs with similar structure to other papers in the field.
+-Funnel shaped to highlight general DNA technologies (oligomers and
 
-DNA motivation
-\begin{enumerate}
-	\item Connection to Tokmakoff results
-	\item Some origami, sensor, other nanotech appications
-\end{enumerate}
+\subsection{\label{sec:intro}DNA motivation}
+\subsubsection{\label{sec:intro}Main Question: how does sequence determine the kinetics of DNA oligonucleotides}
 
-SRV motivation
-\begin{enumerate}
-	\item First time studying a two body problem
-	\item Compare with other TICA-MSM approaches to DNA
-	\item Compare with Noe nucleotide paper
-\end{enumerate}
+\subsubsection{\label{sec:intro}What is known about the system and what remains an open question -- reference Tokmakoff frayed states here}
+
+\subsubsection{\label{sec:intro}Thermodynamics are well established and some MSM based (high resolution kinetic) models have been employed to study kinetics}
+
+\subsubsection{\label{sec:intro}Circle back to applications after discussing the open question that we are seeking to solve -- how does this expand on existing technolgies}
+
+\subsection{\label{sec:intro}SRV motivation}
+\subsubsection{\label{sec:intro}First time studying a two body problem}
+\subsubsection{\label{sec:intro}Compare with other TICA-MSM approaches to DNA}
 
@@
 
-Simulation set up
+-don't confuse methods plots with results
+-not necessary to plot vamp-2 scores (or can include comaprisons in the SI)
+-follow basic outline of Trp-Cage methods in the order sim, srv, msm (a lot of the msm will be in results, but can go through basic assumptions and pytemma pipeline)
 
-Featurization
-\begin{enumerate}
-	\item Vamp-2 score for permutable distances is slightly lower but should not contain degenerate information (SI data)
-	\item Inverse distances have higher vamp-2 score than regular distances (SI data)
-\end{enumerate}
-
-SRV validation
-\begin{enumerate}
-	\item Plot loss scores and self-consistency check for each sequence (fig 1 and 2)
-	\item Select number of slow modes (different for each) and lag time (same for each) based on this analysis.
-	\item Only use off diagonals and add sequence visuals along axes (fig 3)
-\end{enumerate}
-
-MSM construction
-\begin{enumerate}
-	\item hyperparameters include: number of microstates, lag (use same as SRV), number of macrostates (usually n+1) but we use n states for GC-core and GC-mid.
-I have not done a full optimization sweep over these parameters, but can explain the motivation for choosing these.
-    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) (fig 4)
-\end{enumerate}
-
-\section{\label{sec:methods}Discussion}
-
-Brief description and diagrams for collective variables:
-\begin{enumerate}
-	\item Inverse average shifting distance
-	\item Fourth basepair distances
-\end{enumerate}
-
-AT-all
-\begin{enumerate}
-	\item Dynamics dominated by slow shifting modes
-	\item sources to verify this would be the case
-	\item Minimal data from Tokmakoff on this, but can look at relative hybridization rates
-	\item SI Calculations to prove preferred thermo stability of 5’ shifted state
-\end{enumerate}
-
-GC-end
-\begin{enumerate}
-	\item Similar dynamics to AT-all, but with overall lower stability and timescales
-	\item Difficult to make any claims because the thermodynamics of external mismatches with dangling ends have not been measured experimentally, and 3spn2 does not paramertize these mismatches explicitly
-	\item According to SantaLucia external mismatches and dangling ends tend to have weakly stabilizing effect, and there are 4-6 intact basepairs for these conformations
-	\item Might account for unknown signal in 2018 Tokmakoff paper with timescale between the fraying and dehybrization signal, but not sure... 
-\end{enumerate}
-
-GC-core
-\begin{enumerate}
-	\item Frayed states requires the fourth base pair to be broken, most partially frayed states are not included in metastable clustering
-	\item 2nd slow mode correlates well with expected fraying behavior, can sample along this mode and generally find increasing frayed states (although the SRV coords are spiky)
-	\item 3rd mode is reflective of changes in orientation in dissociated behavior. This could be interesting for the SI, but this state does not inform the macrostate clustering
-\end{enumerate}  
-GC-mid
-\begin{enumerate}
-	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
-	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
-\end{enumerate}
+\subsection{\label{sec:methods}Simulation set up}
+\subsubsection{\label{sec:methods}Tmelt selection etc}
 
@@
 
+\subsection{\label{sec:methods}SRV}
+\subsubsection{\label{sec:methods}Featurization}
+\begin{enumerate}
+	\item Vamp-2 score for permutable distances is slightly lower but should not contain degenerate information (SI data)
+	\item Inverse distances have higher vamp-2 score than regular distances (SI data)
+\end{enumerate}
+
+\subsubsection{\label{sec:methods}Select number of slow modes (different for each) and lag time (same for each) based on this analysis}
+
+\subsection{\label{methods}MSM construction}
+\begin{enumerate}
+	\item hyperparameters include: number of microstates, lag (use same as SRV), number of macrostates (usually n+1) but we use n states for GC-core and GC-mid.
+I have not done a full optimization sweep over these parameters, but can explain the motivation for choosing these.
+    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) (fig 4)
+\end{enumerate}
+
+\section{\label{sec:Results}Results}
+- lead with same kind of diagram for each sequence, add additional figures when necessary for a given sequence
+
+\subsection{\label{sec:Results}AT-all}
+\subsubsection{\label{sec:Results}Explanation of state selection and slow modes}
+\subsubsection{\label{sec:Results}Brief description and diagrams for collective variables}
+\begin{enumerate}
+	\item Inverse average shifting distance
+	\item Fourth basepair distances
+\end{enumerate}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_vis.png}
+        \caption{AT-all transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
+        \label{fig:sample_fray}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}Observations and key takeaways}
+\begin{enumerate}
+	\item Dynamics dominated by slow shifting modes
+	\item sources to verify this would be the case
+	\item Minimal data from Tokmakoff on this, but can look at relative hybridization rates
+	\item SI Calculations to prove preferred thermo stability of 5’ shifted state
+\end{enumerate}
+
+\subsection{\label{sec:Results}GC-end}
+\begin{enumerate}
+	\item Similar dynamics to AT-all, but with overall lower stability and timescales
+	\item Difficult to make any claims because the thermodynamics of external mismatches with dangling ends have not been measured experimentally, and 3spn2 does not paramertize these mismatches explicitly
+	\item According to SantaLucia external mismatches and dangling ends tend to have weakly stabilizing effect, and there are 4-6 intact basepairs for these conformations
+	\item Might account for unknown signal in 2018 Tokmakoff paper with timescale between the fraying and dehybrization signal, but not sure... 
+\end{enumerate}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
+        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends}
+        \label{fig:shifting_distributions}
+	\end{center}
+\end{figure}
+
+\subsection{\label{sec:Results}GC-core}
+\begin{enumerate}
+	\item Frayed states requires the fourth base pair to be broken, most partially frayed states are not included in metastable clustering
+	\item 2nd slow mode correlates well with expected fraying behavior, can sample along this mode and generally find increasing frayed states (although the SRV coords are spiky)
+	\item 3rd mode is reflective of changes in orientation in dissociated behavior. This could be interesting for the SI, but this state does not inform the macrostate clustering
+\end{enumerate}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/tica_comparison_GC-core.PNG}
+        \caption{SRV vs. TICA MSMs and resulting clsuter memberships}
+        \label{fig:tica_comparison}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_vis.png}
+        \caption{GC-core transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
+        \label{fig:sample_fray}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/sample_fray.PNG}
+        \caption{Sample along 2nd GC-core SRV mode and show that these have increasing amounts of fraying}
+        \label{fig:sample_fray}
+	\end{center}
+\end{figure}
+
+\subsection{\label{sec:Results}GC-mid}
+\begin{enumerate}
+	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
+	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
+\end{enumerate}
+
+\subsection{\label{sec:Results}Compare all results together}
+\subsubsection{\label{sec:Results}Figure showing Pearson correlations between leading modes on common basis}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/all_modes_correlations.PNG}
+        \caption{For Supplemental info: Correlations between all leading modes}
+        \label{fig:all_modes}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}Figure showing all sequences plotted on a common low dimensional space (probably the 2nd TICA mode and the 2nd GC-core mode)}
+
+\begin{enumerate}
+	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
+	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
+\end{enumerate}
+
+\section{\label{sec:conc}Conclusion}
+\subsection{\label{sec:Results}Looks a lot like the abstract}
+
+\section{\label{sec:SI}Potential SI figures}
 \begin{figure}[ht!]
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/all_modes_correlations.PNG}
-        \caption{For Supplemental info: Correlations between all leading modes}
-        \label{fig:all_modes}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/tica_comparison_GC-core.PNG}
-        \caption{SRV vs. TICA MSMs and resulting clsuter memberships}
-        \label{fig:tica_comparison}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
-        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends}
-        \label{fig:shifting_distributions}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/sample_fray.PNG}
-        \caption{Sample along 2nd GC-core SRV mode and show that these have increasing amounts of fraying}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_vis.png}
-        \caption{AT-all transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_vis.png}
-        \caption{GC-core transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
-
 \clearpage
@@
 
-\bibliography{references}
+%\bibliography{references}
+Test cite: \citep{Zhang2018DeepMechanics}
+\bibliography{refs_mendeley}
 

### 2020-07-16 manuscript.tex
Updates from Overleaf

@@
 
+Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
+From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology} (cite more).
+Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales \citep{MhatreV.HoJi-AnnLee2012NIHAccess}, and dynamic DNA nanotechnology has been used to perform basic calculation \citep{Bui2018} and to probe single molecules via temporal DNA signatures \citep{Shah2019}.
+Both technologies rely on the hybridization reaction between complementary (or similar) DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
+Although many experimental and computational studies have been performed to investigate hybridization and dissociation, the sequence-dependent mechanisms of hybridization dynamics are not fully understood (cite).
+Moreover, it is unclear whether these processes occur in an "all-or-nothing" fashion or if some metastable states can facilitate the process.
+The stability of out of register or "slipped" base pairing in repetitive sequences has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}) and are referred to as slip-strand DNA for longer seqeunces in vivo (cite, maybe too unrelated).
+Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by oligonucleotide sequence \citep{Sanstead2016}.
+In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.
+
+Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, \citep{Hoefert2011MolecularOligonucleotides}  and forward flux sampling  \citep{Phys2014} (cite more).
+In this work, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk}--  to run many unbiased coarse-grained simulations and combine these independent simulations to form an understanding of the kinetics and thermodynamics.
+MSMs have been used recently to study mechanisms of DNA hybridization  \citep{Jin2019} \citep{Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
+Pinamonti et al.
+used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}.
+We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
+Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed.
+We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall MSM quality when compared to more conventional methods such as Time-structure independent components analysis (tICA).
+
+Our analysis reflects similar results to previous computational and experimental DNA work, while elucidating some new insights into sequences dependent dynamics and relative timescales.
+(something here about how this could motivate future experimental work or applications?) By employing SRVs to generate an optimized low dimensional basis, we can access higher resolution MSMs (shorter lag time) and generate a more detailed model.
+The slow dynamical modes learned by SRVs can be directly compared between sequences to determine otherwise subtle trends.
+Furthermore, this work represents the first application of SRVs to a two-body system -- in particular a system that includes permutable distances.
+Our results suggest that this will be a promising tool to study future two-body or many-body molecular dynamics systems.
+
 \section{\label{sec:methods}Methods}
+
+\subsection{\label{sec:methods}Previous feedback and advice}
 
@@
 -follow basic outline of Trp-Cage methods in the order sim, srv, msm (a lot of the msm will be in results, but can go through basic assumptions and pytemma pipeline)
+-less of a narrative and more of an assertive manner
+
+\subsection{\label{sec:methods}Notes and questions} 
+Trp-Cage paper did not really have a dedicated methods section, so basing this more on other papers.
 
@@
 \subsubsection{\label{sec:methods}Tmelt selection etc}
+
+We initialized four sequences previously investigated by Sandstead et al \citep{Sanstead2016} -- AT-all, GC-core, GC-end, and GC-mid -- (small figure here) along with their complementary strands according to 3SPN.2 documentation \citep{Phys2014}.
+We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
+In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end to end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
+This translated to a box size of 77.74 A and an effective oligo concentration of 7 mM.
+We used an Ewald potential to calculate long range
+Coulombic interaction between DNA and ions.
+We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions.
+These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}.
+A 15 fs timestep was used for all simulations runs.
+
+Simulation temperature was determined by sequence specific melting temperature (Tmelt selection).
+Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps.
+For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation.
+An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state.
+In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
 
@@
 \subsubsection{\label{sec:methods}Featurization}
+
 \begin{enumerate}
@@
 
-\subsubsection{\label{sec:methods}Select number of slow modes (different for each) and lag time (same for each) based on this analysis}
+All intermolecular pairwise distances for both oligonucleotides were calculated at each frame.
+Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together.
+This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
+The VAMP-2 scoring method, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
+We compared the VAMP-2 score (SI ?) for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set.
+Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set.
+These features were normalized and passed into sequence-specific SRVs.
+
+\subsubsection{\label{sec:methods}Select number of slow modes (different for each) and lag (same for each) based on this analysis}
+
+Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set.
+In order to maintain consistency between sequences, we kept all training hyperparameters the same between SRVs with the exception of the number of outputed slow modes.
+An 80/20 validation split was used during training, and early stopping was implemented in Keras in order to halt training when validation loss was be minimized.
+SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
+(Should use cross-validation to prove with results were significant here).
+
+**Describe selection of slow modes for each sequence first, then can talk about how we ensure an appropriate lag time is selected.
+Lag time was determine by looking for the shortest time at which all relevant slow modes had converged.
+We found that the leading slow mode took substantially longer to converge than higher order modes and did not include this mode in our metric for convergence.
 
@@
 I have not done a full optimization sweep over these parameters, but can explain the motivation for choosing these.
-    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) (fig 4)
+    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) 
+    \item As we discussed, I think a lot of this could got into the results, so I'm not exactly sure what to include in the methods
 \end{enumerate}
+
+We employed the Pyemma Markov State Model (MSM) pipeline to analyze sequence-specific SRV coordinates.
+Again, we maintained as many hyperparameters consistent as possible between sequences in order to facilitate comparability.
+Using the SRV coordinates as a basis, kmeans clustering was used to group data into microstates where the number of microstates was optimized via VAMP-2 score.
+The score for each sequence leveled off around 200 microstates, and clustering in SRV space consistently yielded a high VAMP-2 score than clustering in TICA space (SI fig).
+Next, implied timescales were estimated over a range of lag times based on these discretized states.
+We then looked for an MSM lag time that was both Markovian and capable of resolving as much detail as possible.
+We found that SRV timescales converged consistently faster and to higher values than TICA.
+Because these timescales are underestimates for the true timescales of the system, we can assume that SRV coordinates are better at both capturing dynamics and enabling better resolution for MSM construction.
+Bayesian Markov Models were constructed from both SRV and TICA clusters, and PCCA+ clustering was performed for macrostates assignment.
+The procedure for determining the number of PCCA cluster centers was performed on a sequence specific basis and will be discussed in the results section.
+
+** The above paragraph is definitely blending into results, not sure how much exactly to include.
+It might be easier to to use AT-all as a case study for explaining the overall process, but the those results are pretty distinct from other sequences so I don't want it to be misleading.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/microstate_optimization.png}
+        \caption{Finding optimal amount of microstates for SRV and TICA clustering. VAMP-2 scores are consistently higher when clustering in SRV space}
+        \label{fig:sample_fray}
+	\end{center}
+\end{figure}
 

### 2020-08-05 manuscript.tex
Updates from Overleaf

@@
 \section{\label{sec:intro}Introduction}
-Should be 4-5 paragraphs with similar structure to other papers in the field.
--Funnel shaped to highlight general DNA technologies (oligomers and
-
-\subsection{\label{sec:intro}DNA motivation}
-\subsubsection{\label{sec:intro}Main Question: how does sequence determine the kinetics of DNA oligonucleotides}
-
-\subsubsection{\label{sec:intro}What is known about the system and what remains an open question -- reference Tokmakoff frayed states here}
-
-\subsubsection{\label{sec:intro}Thermodynamics are well established and some MSM based (high resolution kinetic) models have been employed to study kinetics}
-
-\subsubsection{\label{sec:intro}Circle back to applications after discussing the open question that we are seeking to solve -- how does this expand on existing technolgies}
-
-\subsection{\label{sec:intro}SRV motivation}
-\subsubsection{\label{sec:intro}First time studying a two body problem}
-\subsubsection{\label{sec:intro}Compare with other TICA-MSM approaches to DNA}
+Should be 4-5 paragraphs (funnel shaped) with similar structure to other papers in the field.
 
@@
 Although many experimental and computational studies have been performed to investigate hybridization and dissociation, the sequence-dependent mechanisms of hybridization dynamics are not fully understood (cite).
-Moreover, it is unclear whether these processes occur in an "all-or-nothing" fashion or if some metastable states can facilitate the process.
+Moreover, it is unclear whether these processes occur in an "all-or-nothing" fashion or if some metastable states facilitate the transition.
 The stability of out of register or "slipped" base pairing in repetitive sequences has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}) and are referred to as slip-strand DNA for longer seqeunces in vivo (cite, maybe too unrelated).
@@
 Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, \citep{Hoefert2011MolecularOligonucleotides}  and forward flux sampling  \citep{Phys2014} (cite more).
-In this work, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk}--  to run many unbiased coarse-grained simulations and combine these independent simulations to form an understanding of the kinetics and thermodynamics.
-MSMs have been used recently to study mechanisms of DNA hybridization  \citep{Jin2019} \citep{Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
+Other studies use dramatically elevated or lowered temperatures to induce one way dissociation and hybridization events, respectively, \citep{Wong2008TheSimulations} (cite more).
+Experimental studies often employ temperature jump or other perturtabitive methods to drive DNA our of equilibrium and monitor relaxation processes (cite early T-jump and others).
+Single molecule analysis facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales.
+In this work, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk}--  to run many unbiased coarse-grained simulations and combine these independent simulations to form an understanding of sequence-specific kinetics and thermodynamics.
+MSMs have recently been implemented to study mechanisms and microstates distribution of DNA hybridization  \citep{Jin2019} \citep{Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
 Pinamonti et al.
@@
 Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed.
-We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall MSM quality when compared to more conventional methods such as Time-structure independent components analysis (tICA).
+We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
 
@@
 The slow dynamical modes learned by SRVs can be directly compared between sequences to determine otherwise subtle trends.
-Furthermore, this work represents the first application of SRVs to a two-body system -- in particular a system that includes permutable distances.
-Our results suggest that this will be a promising tool to study future two-body or many-body molecular dynamics systems.
+More summarizing results here .....
 
@@
 
-\subsection{\label{sec:methods}Previous feedback and advice}
-
--don't confuse methods plots with results
--not necessary to plot vamp-2 scores (or can include comaprisons in the SI)
--follow basic outline of Trp-Cage methods in the order sim, srv, msm (a lot of the msm will be in results, but can go through basic assumptions and pytemma pipeline)
--less of a narrative and more of an assertive manner
-
-\subsection{\label{sec:methods}Notes and questions} 
-Trp-Cage paper did not really have a dedicated methods section, so basing this more on other papers.
+\subsection{\label{sec:methods}3spn2 Model}
+Quick summary of model, why we chose to use it, and interesting features.
+-Compatible with lammps
+-Enough resolution to deserve dynamical behavior, coarse enough to run equlibrium sims
+-Previous hybridization experiments show agreement with experiments
 
 \subsection{\label{sec:methods}Simulation set up}
-\subsubsection{\label{sec:methods}Tmelt selection etc}
 
-We initialized four sequences previously investigated by Sandstead et al \citep{Sanstead2016} -- AT-all, GC-core, GC-end, and GC-mid -- (small figure here) along with their complementary strands according to 3SPN.2 documentation \citep{Phys2014}.
+We initialized four sequences previously investigated by Sandstead et al \citep{Sanstead2016} -- AT-all, GC-core, GC-end, and GC-mix -- (small figure here) along with their complementary strands according a Beta helices according to 3SPN.2 documentation \citep{Phys2014}.
 We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
@@
 This translated to a box size of 77.74 A and an effective oligo concentration of 7 mM.
-We used an Ewald potential to calculate long range
-Coulombic interaction between DNA and ions.
+We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions.
 We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions.
 These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}.
+We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
+Simulation temperature was determined by sequence specific melting temperature such that strands were about equally likely to spend time in the hybridized or dissociated state.
+Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps.
 A 15 fs timestep was used for all simulations runs.
-
-Simulation temperature was determined by sequence specific melting temperature (Tmelt selection).
-Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps.
 For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation.
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/n_events.PNG}
-        \caption{Number of events per sequence. I would only included the four sequences in the actual pub. (this could be supplemental}
-        \label{fig:n_events}
-	\end{center}
-\end{figure}
+\subsection{\label{sec:methods}Featurization}
 
-\subsection{\label{sec:methods}SRV}
-\subsubsection{\label{sec:methods}Featurization}
-
-\begin{enumerate}
-	\item Vamp-2 score for permutable distances is slightly lower but should not contain degenerate information (SI data)
-	\item Inverse distances have higher vamp-2 score than regular distances (SI data)
-\end{enumerate}
-
-All intermolecular pairwise distances for both oligonucleotides were calculated at each frame.
+All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
 Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together.
@@
 
-\subsubsection{\label{sec:methods}Select number of slow modes (different for each) and lag (same for each) based on this analysis}
+\subsubsection{\label{sec:methods}SRVs}
+
+SRVs were first developed by Chen et al.
+as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}.
+The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features.
+This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?).
+The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct markov state models.
+Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}.
+SRVs provide robust nonlinear approximation and computation timescales linearly with the amount of input data.
+This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data.
+The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein (cite).
+For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
 
 Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set.
-In order to maintain consistency between sequences, we kept all training hyperparameters the same between SRVs with the exception of the number of outputed slow modes.
-An 80/20 validation split was used during training, and early stopping was implemented in Keras in order to halt training when validation loss was be minimized.
+In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
+We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
+In particular we looked for a diminishing increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data.
+We used a batch size of 50000 and ran each model for total of 30 training epochs.
+We used two hidden layers and set the size of each layer to 100.
+For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
 SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-(Should use cross-validation to prove with results were significant here).
 
-**Describe selection of slow modes for each sequence first, then can talk about how we ensure an appropriate lag time is selected.
-Lag time was determine by looking for the shortest time at which all relevant slow modes had converged.
-We found that the leading slow mode took substantially longer to converge than higher order modes and did not include this mode in our metric for convergence.
+\subsubsection{\label{sec:methods}SRV-MSMs}
 
-\subsection{\label{methods}MSM construction}
-\begin{enumerate}
-	\item hyperparameters include: number of microstates, lag (use same as SRV), number of macrostates (usually n+1) but we use n states for GC-core and GC-mid.
-I have not done a full optimization sweep over these parameters, but can explain the motivation for choosing these.
-    \item Can show comparison data with TICA -- timescales converge faster and state assignments tend to be more concrete (although this has varied) 
-    \item As we discussed, I think a lot of this could got into the results, so I'm not exactly sure what to include in the methods
-\end{enumerate}
-
-We employed the Pyemma Markov State Model (MSM) pipeline to analyze sequence-specific SRV coordinates.
-Again, we maintained as many hyperparameters consistent as possible between sequences in order to facilitate comparability.
-Using the SRV coordinates as a basis, kmeans clustering was used to group data into microstates where the number of microstates was optimized via VAMP-2 score.
-The score for each sequence leveled off around 200 microstates, and clustering in SRV space consistently yielded a high VAMP-2 score than clustering in TICA space (SI fig).
-Next, implied timescales were estimated over a range of lag times based on these discretized states.
-We then looked for an MSM lag time that was both Markovian and capable of resolving as much detail as possible.
-We found that SRV timescales converged consistently faster and to higher values than TICA.
-Because these timescales are underestimates for the true timescales of the system, we can assume that SRV coordinates are better at both capturing dynamics and enabling better resolution for MSM construction.
-Bayesian Markov Models were constructed from both SRV and TICA clusters, and PCCA+ clustering was performed for macrostates assignment.
-The procedure for determining the number of PCCA cluster centers was performed on a sequence specific basis and will be discussed in the results section.
-
-** The above paragraph is definitely blending into results, not sure how much exactly to include.
-It might be easier to to use AT-all as a case study for explaining the overall process, but the those results are pretty distinct from other sequences so I don't want it to be misleading.
+MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.
+The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time.
+The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
+To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}.
+This becomes crucial in our system given the large difference in timescales between leading modes.
+Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis.
+To prove this, SRV-MSM VAMP-2 scores were shown to be consistently higher than MSMs constructed from TICA coorinates (TICA-MSMs) \citep{Sidky}.
+Furthermore, SRV-MSM implied timescales converge faster than TICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.
+To build our SRV-MSM framework, we employed the Pyemma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
+After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ macrostate assignments.
+The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
+The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{\label{sec:Results}Results}
@@
 \subsubsection{\label{sec:Results}Explanation of state selection and slow modes}
-\subsubsection{\label{sec:Results}Brief description and diagrams for collective variables}
-\begin{enumerate}
-	\item Inverse average shifting distance
-	\item Fourth basepair distances
-\end{enumerate}
+
+In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes.
+For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step by step.
+Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
+We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns.
+We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation (*not sure if this warrants further discussion).
+It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
+Next we selected an optimal number of SRV components to include in our analysis.
+After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
+It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
+For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
+For the AT-all sequence, we generate SRVs with five components, representing the five slowest processes in the systems.
+Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
+The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by each process.
+For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal).
+Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the dynamics of the overall hybridization and dissociation process.
+The next four SRV components all show a relatively high correlation along different offset diagonals.
+These diagonals correspond to the intermolecular distances between shifted base pairs (SI figure for this?), and point to the existence of gradually faster sliding or "slithering" processes.
+Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than some  underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_vis.png}
-        \caption{AT-all transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
-        \label{fig:sample_fray}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_srv_correlations.png}
+        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
+        \label{fig:AT-all_srv_correlations}
 	\end{center}
@@
 
-\subsubsection{\label{sec:Results}Observations and key takeaways}
-\begin{enumerate}
-	\item Dynamics dominated by slow shifting modes
-	\item sources to verify this would be the case
-	\item Minimal data from Tokmakoff on this, but can look at relative hybridization rates
-	\item SI Calculations to prove preferred thermo stability of 5’ shifted state
-\end{enumerate}
+From this analysis we can determine that the slowest dynamics are fully characterized by the hybridization process and slithering behavior of out of register base pairs.
+Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
+Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
+We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
+Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
+In the figure (dynamical figure) we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes.
+This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis.
+Using an MSM lag time of 1.2 ns, we then built a Bayesian MSM to calculate transition probability matrix between each microstate.
+Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represent a collection of metastable structures.
+Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels}, \citep{PinamontiAnalyzedModels}.
+Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
+For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
+To visualize these six macrostates, we project the data into the two leading TICA coordinates.
+Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
+It is clear in figures (thermo fig) that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
+After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states.
+In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_dynamic.png}
+        \caption{SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns}
+        \label{fig:AT-all_dynamic}
+	\end{center}
+\end{figure}
+
+In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings.
+These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states.
+This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019}, \citep{Phys2014}.
+We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states.
+This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
+Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length.
+It has been shown that two base pair 5' dangling ends (or say one base pair inert tails) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}.
+This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends.
+Furthermore, Santalucia thermodynamic calculations reveal that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (NN calculations SI?).
+Taken together, these experimental results substantiate why we observe the single shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
+Moreover, it is expected that implied timescales corresponding to this process is on the same order of magnitude at the hybridization timescale.
+The stability of the single shifted 3' state is notably lower and on the same order of magnitude as the double shifted 5' state despite it having two less intact base pairs.
+The final state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
+Experimental work  \citep{Doktycz1990ThermodynamicATGC} studied four base pair dangling ends on DNA hairpins, and found that TATA ends are most stabilizing than ATAT.
+Although NN interactions paly a more important role in determining stabiity, this could explain why the relative free energy difference between single shifted 3' state and double shifted 3' is not as large as the difference between the single shifted 5' and double shifted 5' states.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_thermo.png}
+        \caption{SRV-MSM TICA projection, state free energies, and transitions probabilities}
+        \label{fig:AT-all_thermo}
+	\end{center}
+\end{figure}
+
+Beyond state probabilities and free energy approximations, the macrostate MSM also yields discrete kinetic information in the form of transition probabilities between states.
+Figure (thermo fig) show the probability of moving from one state to another within the MSM lag time (1.2 ns).
+All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is readily disrupted by complete dissociation.
+In particular, we observe that the transition probability from double shifted 3' state to the dissociated state is 6x higher than to the neighboring single shifted 3' state.
+From this we conclude that shifted states are important in facilitating the hybridization process, but that a higher degree of shiftedness (in particular when shifted in the 3' direction) is less likely to evolve into a fully aligned hybridized state than it is to dissociate in the process.
+
+For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
+Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates.
+The number of SRV coordinates and MSM macrostates are varied based on the kinetics and thermodynamics behavior of specific sequences.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \subsection{\label{sec:Results}GC-end}
-\begin{enumerate}
-	\item Similar dynamics to AT-all, but with overall lower stability and timescales
-	\item Difficult to make any claims because the thermodynamics of external mismatches with dangling ends have not been measured experimentally, and 3spn2 does not paramertize these mismatches explicitly
-	\item According to SantaLucia external mismatches and dangling ends tend to have weakly stabilizing effect, and there are 4-6 intact basepairs for these conformations
-	\item Might account for unknown signal in 2018 Tokmakoff paper with timescale between the fraying and dehybrization signal, but not sure... 
-\end{enumerate}
+
+Next we examine the GC-end sequence, which add GC caps to the same repetitive AT motifs.
+Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.
+These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some slithering dynamics.
+When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in figure (SI fig) by a distinct spectral gap.
+We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score.
+The clustered macrostates show a similar distribution where shifted populations are notably lower.
+The 3' double shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes).
+Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches decreases the number of out-of-register WC pairs by two relative to corresponding AT-all states.
+This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and lifetimes.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_thermo.PNG}
+        \caption{Shows Pearson correlations between three leading SRV mode and all 100 intermolecular distances (not sure why there are 6 color on the legend since there are only 5 states, need to fix this)}
+        \label{fig:shifting_distributions}
+	\end{center}
+\end{figure}
+
+For our analysis, we consider C:T and G:A base pairs in the GC-ends shifted states as non-interacting dangling ends.
+Although such internal mismatches of this kind may cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
+In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+The only interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch bond and can therefore not represent the state with highest fidelity.
+Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (NN calculations).
+Interestingly, we found that 5' single shifted stability might be elevated by a substantial portion of conformations retaing one intact GC base pair.
+Visualizations reveal that the shifted oligos -- particularly in the 5' shifted state -- have a tendency to sacrifice some helical conformational entropy in a way that facilitates an out of register GC base pairs.
+To compare these single shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 10000 conformations in both states.
+Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance (we use intermolecular distances) and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps}.
+The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
+This reveals that the mismatch GC-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+On y-axis we examine the second diffusion mode, which correlates highly with the average distance between 3' and 5' ends.
+Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
+These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations, and their existence in the simulations confirmed by molecular renderings of these regions.
+
+It is unclear the extent to which these interactions are physical given the strain required to form these shifted loops (need to discuss toehold loop experiments with Brennan).
+Single shifted 3' and double shifted 5' strands have much lower population of these shifted loops (need to perform calculations to see if these account for increased stabilization of single shifted 5' versus single shifted 3').
+Not sure how much more I should reall go into here...
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/diff_maps_full.png}
+        \caption{First three diffusion map coordinates built from 20000 single shifted 5' states, equally sampled from AT-all and GC-end. (need to add some molecular rendering of states).}
+        \label{fig:shifting_distributions}
+	\end{center}
+\end{figure}
+
+The dynamic picture looks fairly different for the GC-end macrostates compared to AT-all.
+We do not observe any significant flux between the double shifted to single shifted 5' states.
+While this could be due to insufficient sampling, it is likely that the double shifted states has too few contacts to directly transition into a single shifted state without first dissociating.
+This is further demonstated by the kinetic distance between this state and the aligned hybridized state (S0 and S4, respectively) shown in TICA projections (thermo figure).We observe some flux between the single shifted states and aligned hybridized state, but these events are more rare than their AT-all equivalents.
+This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
+In T-jump experiments, the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
+Further analysis showed a unknown kinetic response was recorded for GC-end sequence -- particularly at lower temperature -- but it was unclear which physical process this response corresponded to \citep{Sanstead2018DirectDehybridization}.
+It is possible that this signal represented some oligomer population shifting out of register as we observe in our MSMs, although it is difficult to say for certain given that the signal was in a congested spectroscopic range and we observe the process to be more active during the hybridization process than dissociation.
 
@@
         \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
-        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends}
+        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (probs better to grey out underlying free energy landscape and move this to the SI.}
         \label{fig:shifting_distributions}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\label{sec:Results}GC-core}
-\begin{enumerate}
-	\item Frayed states requires the fourth base pair to be broken, most partially frayed states are not included in metastable clustering
-	\item 2nd slow mode correlates well with expected fraying behavior, can sample along this mode and generally find increasing frayed states (although the SRV coords are spiky)
-	\item 3rd mode is reflective of changes in orientation in dissociated behavior. This could be interesting for the SI, but this state does not inform the macrostate clustering
-\end{enumerate}
+
+The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end.
+Based on SRV cross-validation (SI), we select the first three components for our analysis.
+When correlating these first three modes with intermolecular distances (heatmap figure) we observe a similar first mode to the previous sequences reflecting the overall hybridization and dissociation process.
+The second mode no longer shows shifted diagonal behavior, but rather positive correlations between 5'/3' ends, negative correlations between 3'/3' and 5'/5' ends, and correlations closest to zero for the central GC pairs.
+Qualitatively, this suggests that the mode is picking up motions where base pair ends drift apart and opposite ends come closer together, while the oligo interior remains bound.
+The third mode shows symmetry along the main diagonal, much like the first mode but with an inverse trend.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_srv_correlations.PNG}
+        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
+        \label{fig:GC-core_srv_correlations}
+	\end{center}
+\end{figure}
+
+(Not sure whether to include this section before or after building the MSM)
+We examined each slow mode with respect to physical coordinates with which it showed high correlation.
+Unsurprisingly, the first mode tracks closely with the inverse distance between complement G:C pairs.
+These distances are the best indicators for a hybridization/dissociation, and we see a sharp change in the first SRV signal as these bonds form or break.
+The second slow mode is most active when the GC core pair are bound but the adjacent AT pairs are not.
+There is a small signal for fraying at the outer basepairs, but the mode overwhelmingly learns about these neighboring AT/GC bonds.
+This behavior reflects a kinetic trap between the core binding event and the fully hybridized (or evenly partially frayed) state.
+Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward (cite).
+Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states.
+This occurs during the dissociation process as well, where one half of the AT base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
+The third mode is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
+We attribute this to the SRV learning about the slow diffusive motions of the two body system (not sure about this).
+In addition to picking up on dissociation behavior, the third mode seems to peak when the oligos are close together but configured in such a way that is not amenable hybridization.
+These misaligned conformations include inverse contacts where 5'/5' and 3'/3' ends meet, looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts, or some combination of the two (SI third mode screenshots).
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_tracking_modes.png}
+        \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.}
+        \label{fig:GC-core_tracking_modes}
+	\end{center}
+\end{figure}
+
+To further investigate these dynamics, we built an SRV-MSM using these first three SRV modes as a basis.
+We we found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "frayed" states.
+Again, we found transition probabilities between states and visualized representative molecular renderings.
+As we observed from the 2nd SRV mode analysis above, the "frayed" metatable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
+The MSM transition probabilities suggests that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
+We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
+This make sense considering that this state is more accessible from an already bound helix.
+Moreover, once oligo are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate.
+Thus once a dissociated to frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation.
+On the other hand, transitions from the hybridized to frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_thermo.PNG}
+        \caption{Full MSM output for GC-core}
+        \label{fig:GC-core_thermo}
+	\end{center}
+\end{figure}
+
+When examining these same four sequences using T-jump spectroscopy, Sandstead et al.
+found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
+As their analysis did not consider the previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
+While high amounts of 1-2 base pair fraying were observed for GC-mid and AT-all as well, lattice model predictions that GC-core was the only sequence to have multiple frayed pairs during the dissociation process.
+Further analysis speculated that AT-all termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
+We see the same rapid fraying in simulation data (which too fast to be attributed with an SRV mode), however we stipulate that this inter-conversion first relies on the formation of the the AT-bond nearest to the GC center.
+In visualizing trajectories, we observe that the frayed ends rotate away from each other and impede helix formation, further highlighting the importance of this 4th base pair formation.
+
+(free energy minimum at high temp but not low temp?).
+
+\citep{Sanstead2018DirectDehybridization} Our data support a picture in which the dimer ensemble can exist in a broad basin where the populated configurations are not separated by appreciable barriers, suggesting rapid interconver- sion can occur between all of the accessible frayed structures within the well.
+GC-end placement leads to all or nothing mechanism.
+
+Refs for DNA fraying:
+\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro).
+Use TPS to focus on kinetic pathway by which singl base pair bind/unbind.
+Focus on end base of 3-bp oligomer in explicit solvent.
+Finds order parameters tracking base pair binding as well as intra-strand stacking.
+(energy and distance for each).
+
+\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,
+
+6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
+37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
+38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbind- ing
+
+%%%%%% intro notes %%%%%%%
+\citep{Sanstead2018DirectDehybridization} Good intro with many repesimportance of dynamics "rich dynamics even for short oligos"
+
+%%%%%% Other MSM and timescales approaches  %%%%%%%%%%%
+## Good references:
+
+McGibbon, R.
+T., & Pande, V.
+S.
+(2015).
+Variational cross-validation of slow dynamical modes in molecular kinetics : Importance of not generating too many microstates and employting cross-validation to avoid artificially high scores or artificially slow dynamics.
+Both of these lead to models fitting statistical noise which is the root of boosted scores.
+
+Pinamonti, G., Paul, F., Rodriguez, A., & Bussi, G.
+(n.d.).
+analyzed with core-set Markov state models, 43: Good ideas on core-based clustering.
+Assigning microstates based on physical characteristics is interesting but didn't seem to work for me in practice (could try doing this based on sim.log data with number of base pairs bound or some energy metric).
+Lot's of notes on this paper regarding comparisons to timescales.
+Good notes on higher resolution fraying interactions that we likely can't resolve.
+
+Sua, E., Adelman, J. L., & Zuckerman, D. M. (2016). Accurate Estimation of Protein Folding and Unfolding Times : Beyond Markov State Models, 1. https://doi.org/10.1021/acs.jctc.6b00339:
+More accurate ways to adapt MFPT into a rate
 
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_vis.png}
-        \caption{GC-core transition diagram, implied timescales between SRV-MSM and TICA-MSM, free energy surface and metastable state assignment.}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/sample_fray.PNG}
-        \caption{Sample along 2nd GC-core SRV mode and show that these have increasing amounts of fraying}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
-
 \subsection{\label{sec:Results}GC-mid}
@@
 	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
+	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
+
 \end{enumerate}
@@
 
+Nucleation-zipper mechnasism is a fast process that can orginate anywhere along the strand making it difficult to track..
+
+\subsubsection{\label{sec:Results}Limitations} 
+- not capturing zippering
+One limititation of our appraoch that is that processes that are.
+For example, there are eight 3-bp sites on each oligo from which a zippering action coud nucleated.
+Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
+Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
+
 \section{\label{sec:conc}Conclusion}
 \subsection{\label{sec:Results}Looks a lot like the abstract}
-
-\section{\label{sec:SI}Potential SI figures}
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/AT-all_validation.PNG}
-        \caption{For Supplemental info: Selection process for AT-all slow modes}
-        \label{fig:AT-all_val}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/GC-core_validation.PNG}
-        \caption{For Supplemental info: Selection process for GC-core slow modes}
-        \label{fig:GC-core_val}
-	\end{center}
-\end{figure}
 

### 2020-08-05 manuscript.tex

@@
 \usepackage{color}
-\usepackage{CJK}https://www.overleaf.com/project/5e820b28d741e00001b1ccac
+\usepackage{CJK}
 \usepackage{framed}
@@
 
+TEST CITE \cite{Phys2014}
+
 \section{\label{sec:intro}Introduction}
@@
 \section{\label{sec:methods}Methods}
-
-\subsection{\label{sec:methods}3spn2 Model}
-Quick summary of model, why we chose to use it, and interesting features.
--Compatible with lammps
--Enough resolution to deserve dynamical behavior, coarse enough to run equlibrium sims
--Previous hybridization experiments show agreement with experiments
 

### 2020-08-05 manuscript.tex
Updates from Overleaf

@@
 \usepackage{color}
-\usepackage{CJK}
+\usepackage{CJK}https://www.overleaf.com/project/5e820b28d741e00001b1ccac
 \usepackage{framed}
@@
 
-TEST CITE \cite{Phys2014}
-
 \section{\label{sec:intro}Introduction}
-Should be 4-5 paragraphs (funnel shaped) with similar structure to other papers in the field.
 
@@
 \section{\label{sec:methods}Methods}
+
+\subsection{\label{sec:methods}3spn2 Model}
+Not sure if this is necessary, but can include a quick summary of model, why we chose to use it, and interesting features.
+-Compatible with lammps
+-Enough resolution to deserve dynamical behavior, coarse enough to run equlibrium sims
+-Previous hybridization experiments show good agreement with experiments
 
@@
 
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
+        \caption{Cross val procedure to selection number of SRV coordinates (probably will be SI, but could include AT-all cross-val in the walkthrough section)}
+        \label{fig:srv_crossval}
+	\end{center}
+\end{figure}
+
 \subsubsection{\label{sec:methods}SRV-MSMs}
@@
 The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/microstate_optimization.png}
-        \caption{Finding optimal amount of microstates for SRV and TICA clustering. VAMP-2 scores are consistently higher when clustering in SRV space}
-        \label{fig:sample_fray}
-	\end{center}
-\end{figure}
 
@@
 Single shifted 3' and double shifted 5' strands have much lower population of these shifted loops (need to perform calculations to see if these account for increased stabilization of single shifted 5' versus single shifted 3').
-Not sure how much more I should reall go into here...
+Not sure how much more much to go into here ...
 
@@
 
-When examining these same four sequences using T-jump spectroscopy, Sandstead et al.
+When examining these same four sequences using T-jump and 2D spectroscopy, Sandstead et al.
 found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
 As their analysis did not consider the previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
-While high amounts of 1-2 base pair fraying were observed for GC-mid and AT-all as well, lattice model predictions that GC-core was the only sequence to have multiple frayed pairs during the dissociation process.
-Further analysis speculated that AT-all termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
+While 1-2 base pair fraying was commonly observed for GC-mid and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs.
+Further analysis showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
 We see the same rapid fraying in simulation data (which too fast to be attributed with an SRV mode), however we stipulate that this inter-conversion first relies on the formation of the the AT-bond nearest to the GC center.
@@
 
-(free energy minimum at high temp but not low temp?).
-
-\citep{Sanstead2018DirectDehybridization} Our data support a picture in which the dimer ensemble can exist in a broad basin where the populated configurations are not separated by appreciable barriers, suggesting rapid interconver- sion can occur between all of the accessible frayed structures within the well.
-GC-end placement leads to all or nothing mechanism.
+%%%%%% intro notes %%%%%%%
+\citep{Sanstead2018DirectDehybridization} Good intro with many repesimportance of dynamics "rich dynamics even for short oligos"
 
@@
 37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
-38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbind- ing
+38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding
 
-%%%%%% intro notes %%%%%%%
-\citep{Sanstead2018DirectDehybridization} Good intro with many repesimportance of dynamics "rich dynamics even for short oligos"
+\subsection{\label{sec:Results}GC-mid}
+\begin{enumerate}
+	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
+	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
+	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
+
+\end{enumerate}
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/tica_comparison_GC-core.PNG}
-        \caption{SRV vs. TICA MSMs and resulting clsuter memberships}
-        \label{fig:tica_comparison}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_tica_cvs.PNG}
+        \caption{All sequences plotted on combined TICA axes}
+        \label{fig:all_seq_tica_cvs}
 	\end{center}
@@
 
-\subsection{\label{sec:Results}GC-mid}
-\begin{enumerate}
-	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
-	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
-	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
-
-\end{enumerate}
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_physical_cvs.PNG}
+        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes}
+        \label{fig:all_seq_physical_cvs}
+	\end{center}
+\end{figure}
 
@@
 \subsubsection{\label{sec:Results}Figure showing Pearson correlations between leading modes on common basis}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/all_modes_correlations.PNG}
-        \caption{For Supplemental info: Correlations between all leading modes}
-        \label{fig:all_modes}
-	\end{center}
-\end{figure}
 

### 2020-08-24 manuscript.tex
Updates 0823

@@
 \usepackage[disable]{todonotes} % notes hidden
+\usepackage{comment}
 
@@
 Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
-From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology} (cite more).
-Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales \citep{MhatreV.HoJi-AnnLee2012NIHAccess}, and dynamic DNA nanotechnology has been used to perform basic calculation \citep{Bui2018} and to probe single molecules via temporal DNA signatures \citep{Shah2019}.
-Both technologies rely on the hybridization reaction between complementary (or similar) DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
-Although many experimental and computational studies have been performed to investigate hybridization and dissociation, the sequence-dependent mechanisms of hybridization dynamics are not fully understood (cite).
-Moreover, it is unclear whether these processes occur in an "all-or-nothing" fashion or if some metastable states facilitate the transition.
-The stability of out of register or "slipped" base pairing in repetitive sequences has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}) and are referred to as slip-strand DNA for longer seqeunces in vivo (cite, maybe too unrelated).
-Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by oligonucleotide sequence \citep{Sanstead2016}.
+From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}.
+Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}.
+Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
+Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019}.
+Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition.
+Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}.
+Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
+The stability of certain intermediates states, such as out-of-register or "slipped" base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}).
+Furthermore, Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}.
 In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.
@@
 Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, \citep{Hoefert2011MolecularOligonucleotides}  and forward flux sampling  \citep{Phys2014} (cite more).
-Other studies use dramatically elevated or lowered temperatures to induce one way dissociation and hybridization events, respectively, \citep{Wong2008TheSimulations} (cite more).
-Experimental studies often employ temperature jump or other perturtabitive methods to drive DNA our of equilibrium and monitor relaxation processes (cite early T-jump and others).
-Single molecule analysis facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales.
-In this work, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk}--  to run many unbiased coarse-grained simulations and combine these independent simulations to form an understanding of sequence-specific kinetics and thermodynamics.
-MSMs have recently been implemented to study mechanisms and microstates distribution of DNA hybridization  \citep{Jin2019} \citep{Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
-Pinamonti et al.
-used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.
+Other computational studies use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Experimental studies often employ temperature jump, concentration jumps, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA}.
+Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales and data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices}.
+Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.
+
+In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior at each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
+MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
+\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}.
 We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
 Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed.
-We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
+We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
 
-Our analysis reflects similar results to previous computational and experimental DNA work, while elucidating some new insights into sequences dependent dynamics and relative timescales.
-(something here about how this could motivate future experimental work or applications?) By employing SRVs to generate an optimized low dimensional basis, we can access higher resolution MSMs (shorter lag time) and generate a more detailed model.
-The slow dynamical modes learned by SRVs can be directly compared between sequences to determine otherwise subtle trends.
-More summarizing results here .....
+We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
+By evaluating equilibrium trajectories we can study the relevance of meta-stable states during the hybridization and dissociation process.
+Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
+Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models.
+Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs.
+Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.
 
@@
 \subsection{\label{sec:methods}3spn2 Model}
-Not sure if this is necessary, but can include a quick summary of model, why we chose to use it, and interesting features.
--Compatible with lammps
--Enough resolution to deserve dynamical behavior, coarse enough to run equlibrium sims
--Previous hybridization experiments show good agreement with experiments
+
+Is some additional information warranted on the model itself before going into this specific setup?
 
@@
 
-We initialized four sequences previously investigated by Sandstead et al \citep{Sanstead2016} -- AT-all, GC-core, GC-end, and GC-mix -- (small figure here) along with their complementary strands according a Beta helices according to 3SPN.2 documentation \citep{Phys2014}.
+We initialized four sequences previously investigated by Sandstead et al.
+-- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
 We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
-In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end to end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
-This translated to a box size of 77.74 A and an effective oligo concentration of 7 mM.
+In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
+This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM.
 We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions.
@@
 We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
-Simulation temperature was determined by sequence specific melting temperature such that strands were about equally likely to spend time in the hybridized or dissociated state.
+Simulation temperature was determined by sequence specific melting temperature such that strands were equally likely to spend time in the hybridized or dissociated state.
 Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps.
@@
 
-All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
+All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
 Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together.
@@
 The VAMP-2 scoring method, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
-We compared the VAMP-2 score (SI ?) for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set.
+We compared the VAMP-2 score for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set.
 Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set.
 These features were normalized and passed into sequence-specific SRVs.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
+        \caption{Cross val procedure to selection number of SRV coordinates (probably will be SI, but could include AT-all cross-val in the walkthrough section)}
+        \label{fig:srv_crossval}
+	\end{center}
+\end{figure}
 
@@
 This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?).
-The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct markov state models.
+The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.
 Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}.
@@
 This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data.
-The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein (cite).
+The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen}, \citep{Sidky}.
 For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
@@
 SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
-        \caption{Cross val procedure to selection number of SRV coordinates (probably will be SI, but could include AT-all cross-val in the walkthrough section)}
-        \label{fig:srv_crossval}
-	\end{center}
-\end{figure}
+SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com} \citep{Abadi2016TensorFlow:Systems}.
 
@@
 \subsection{\label{sec:Results}AT-all}
-\subsubsection{\label{sec:Results}Explanation of state selection and slow modes}
+
+\subsubsection{\label{sec:Results}SRV optimization and analysis}
 
 In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes.
-For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step by step.
+For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step-by-step.
 Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
 We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns.
-We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation (*not sure if this warrants further discussion).
+We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation.
 It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
@@
 For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-For the AT-all sequence, we generate SRVs with five components, representing the five slowest processes in the systems.
+As shown in (Figure \ref{fig:srv_crossval}), we see a clear convergence of the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
 Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
-The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by each process.
+The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by (or effective on) each process.
 For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal).
@@
 The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs (SI figure for this?), and point to the existence of gradually faster sliding or "slithering" processes.
-Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than some  underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.
+These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes.
+Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_srv_correlations.png}
-        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
-        \label{fig:AT-all_srv_correlations}
-	\end{center}
-\end{figure}
+\subsubsection{\label{sec:Results}SRV-MSM construction and optimization}
 
@@
 Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
-In the figure (dynamical figure) we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes.
+In Figure \ref{fig:AT-all_dynamic} we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes.
 This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis.
@@
 Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
-It is clear in figures (thermo fig) that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
+It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
 After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_dynamic.png}
-        \caption{SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns}
+        \caption{AT-all SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns. Leading slow mode correlations to all 100 intermolecular distances between oligos.}
         \label{fig:AT-all_dynamic}
@@
 \end{figure}
+
+\subsubsection{\label{sec:Results}Implied timescales for TICA-MSMs for }
 
@@
 Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length.
-It has been shown that two base pair 5' dangling ends (or say one base pair inert tails) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}.
+It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}.
 This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends.
-Furthermore, Santalucia thermodynamic calculations reveal that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (NN calculations SI?).
-Taken together, these experimental results substantiate why we observe the single shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
-Moreover, it is expected that implied timescales corresponding to this process is on the same order of magnitude at the hybridization timescale.
-The stability of the single shifted 3' state is notably lower and on the same order of magnitude as the double shifted 5' state despite it having two less intact base pairs.
+Furthermore, Santalucia thermodynamic calculations (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability).
+Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
+Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average lifetimes of these states.
+The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state.
+In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having more less intact base pairs.
+This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below.
 The final state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
-Experimental work  \citep{Doktycz1990ThermodynamicATGC} studied four base pair dangling ends on DNA hairpins, and found that TATA ends are most stabilizing than ATAT.
-Although NN interactions paly a more important role in determining stabiity, this could explain why the relative free energy difference between single shifted 3' state and double shifted 3' is not as large as the difference between the single shifted 5' and double shifted 5' states.
+This might be attributed to the asymptotic effects of inert tail destabilitization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
+Accordingly, the longer inert tails in the double-shifted states have less of an effect on relative 5' vs.
+3' stability compared to the shorter tails in the single-shifted states.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_thermo.png}
-        \caption{SRV-MSM TICA projection, state free energies, and transitions probabilities}
-        \label{fig:AT-all_thermo}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+        \label{fig:AT-all_in_one}
 	\end{center}
@@
 
+\subsubsection{\label{sec:Results}Transition probabilities between states}
+
 Beyond state probabilities and free energy approximations, the macrostate MSM also yields discrete kinetic information in the form of transition probabilities between states.
-Figure (thermo fig) show the probability of moving from one state to another within the MSM lag time (1.2 ns).
+Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns).
 All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is readily disrupted by complete dissociation.
-In particular, we observe that the transition probability from double shifted 3' state to the dissociated state is 6x higher than to the neighboring single shifted 3' state.
-From this we conclude that shifted states are important in facilitating the hybridization process, but that a higher degree of shiftedness (in particular when shifted in the 3' direction) is less likely to evolve into a fully aligned hybridized state than it is to dissociate in the process.
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation.
+In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state.
+Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states.
+These asymmetric kinetic effects may have some role in the substantially lower state populations we see above.
+From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state.
 
@@
 
+\subsubsection{\label{sec:Results}SRV-MSM construction and states}
+
 Next we examine the GC-end sequence, which add GC caps to the same repetitive AT motifs.
@@
 These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some slithering dynamics.
-When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in figure (SI fig) by a distinct spectral gap.
-We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score.
+When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in SI Figure \ref{fig:GC-end_dynamic}) by a distinct spectral gap.
+We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score (SI Figure \ref{fig:srv_crossval}).
 The clustered macrostates show a similar distribution where shifted populations are notably lower.
-The 3' double shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes).
+The 3' double-shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes).
 Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches decreases the number of out-of-register WC pairs by two relative to corresponding AT-all states.
-This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and lifetimes.
+This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and shorter timescales.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_thermo.PNG}
-        \caption{Shows Pearson correlations between three leading SRV mode and all 100 intermolecular distances (not sure why there are 6 color on the legend since there are only 5 states, need to fix this)}
-        \label{fig:shifting_distributions}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dynamic.png}
+        \caption{(SI) GC-end SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns. Leading slow mode correlations to all 100 intermolecular distances between oligos.}
+        \label{fig:GC-end_dynamic}
 	\end{center}
@@
 
-For our analysis, we consider C:T and G:A base pairs in the GC-ends shifted states as non-interacting dangling ends.
-Although such internal mismatches of this kind may cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
+\subsubsection{\label{sec:Results}Comparison between GC-end and AT-all shifted states}
+
+For our analysis, we consider C:T and G:A base pairs in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends and two inert tails.
+Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
 In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
-The only interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch bond and can therefore not represent the state with highest fidelity.
-Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (NN calculations).
-Interestingly, we found that 5' single shifted stability might be elevated by a substantial portion of conformations retaing one intact GC base pair.
-Visualizations reveal that the shifted oligos -- particularly in the 5' shifted state -- have a tendency to sacrifice some helical conformational entropy in a way that facilitates an out of register GC base pairs.
-To compare these single shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 10000 conformations in both states.
-Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance (we use intermolecular distances) and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps}.
-The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch GC-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-On y-axis we examine the second diffusion mode, which correlates highly with the average distance between 3' and 5' ends.
-Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
-These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations, and their existence in the simulations confirmed by molecular renderings of these regions.
-
-It is unclear the extent to which these interactions are physical given the strain required to form these shifted loops (need to discuss toehold loop experiments with Brennan).
-Single shifted 3' and double shifted 5' strands have much lower population of these shifted loops (need to perform calculations to see if these account for increased stabilization of single shifted 5' versus single shifted 3').
-Not sure how much more much to go into here ...
+The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch interaction.
+Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
+Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic effects from inert tails.
+Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
+Visualizations reveal that the shifted oligos -- particularly in the 5' shifted state -- have a tendency to sacrifice some helical conformational entropy in a way that facilitates G:C termini bonding even when all available A:T bonds are formed out-of-register.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/diff_maps_full.png}
-        \caption{First three diffusion map coordinates built from 20000 single shifted 5' states, equally sampled from AT-all and GC-end. (need to add some molecular rendering of states).}
-        \label{fig:shifting_distributions}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.PNG}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+        \label{fig:GC-end_in_one}
 	\end{center}
@@
 
-The dynamic picture looks fairly different for the GC-end macrostates compared to AT-all.
-We do not observe any significant flux between the double shifted to single shifted 5' states.
-While this could be due to insufficient sampling, it is likely that the double shifted states has too few contacts to directly transition into a single shifted state without first dissociating.
-This is further demonstated by the kinetic distance between this state and the aligned hybridized state (S0 and S4, respectively) shown in TICA projections (thermo figure).We observe some flux between the single shifted states and aligned hybridized state, but these events are more rare than their AT-all equivalents.
+\subsubsection{\label{sec:Results}Diffusion maps show sequence differences in shifted states}
+
+To compare these single-shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 5000 conformations from 5' shifted state of both sequences.
+Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps}.
+We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
+This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands.
+In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}.
+The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
+This reveals that the mismatch GC-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
+Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
+These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions.
+Although AT-all shifted ends tend to stay bound out-of-register, we find the second diffusion map coordinate to be indicative of inert tails that fold back onto these terminal out-of-register pairings.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps.png}
+        \caption{First two diffusion map coordinates built from 10000 single-shifted 5' states, equally sampled from AT-all and GC-end.
+Color maps show inverse distances between out-of-register ends and complementary ends.
+}
+        \label{fig:GC-end_dmaps}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_full.PNG}
+        \caption{SI fig }
+        \label{fig:GC-end_dmaps_full}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}GC-end dynamics are largely two-state}
+
+The kinetic behavior between GC-end macrostates is also distinct compared to AT-all.
+We do not observe any significant flux between the double-shifted to single-shifted 5' states.
+While this could be due to insufficient sampling, it is likely that the double-shifted states has too few contacts to directly transition into a single-shifted state without first dissociating.
+This is further demonstrated by the kinetic distance between this state and the aligned hybridized state (S0 and S4, respectively) shown in TICA projections (Figure \ref{fig:GC-end_in_one}).
+We observe some flux between the single-shifted states and aligned hybridized state, but these events are more rare than their AT-all equivalents.
 This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
 In T-jump experiments, the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
-Further analysis showed a unknown kinetic response was recorded for GC-end sequence -- particularly at lower temperature -- but it was unclear which physical process this response corresponded to \citep{Sanstead2018DirectDehybridization}.
-It is possible that this signal represented some oligomer population shifting out of register as we observe in our MSMs, although it is difficult to say for certain given that the signal was in a congested spectroscopic range and we observe the process to be more active during the hybridization process than dissociation.
+Futhermore, 2D spectroscopy analysis showed a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the physical processes under investigation.
+\citep{Sanstead2018DirectDehybridization}.
+Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some oligomer population shifting out-of-register as we observe in our MSMs.
+This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence.
 
@@
 
+\subsubsection{\label{sec:Results}Building SRV-MSM}
+
 The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end.
-Based on SRV cross-validation (SI), we select the first three components for our analysis.
-When correlating these first three modes with intermolecular distances (heatmap figure) we observe a similar first mode to the previous sequences reflecting the overall hybridization and dissociation process.
-The second mode no longer shows shifted diagonal behavior, but rather positive correlations between 5'/3' ends, negative correlations between 3'/3' and 5'/5' ends, and correlations closest to zero for the central GC pairs.
-Qualitatively, this suggests that the mode is picking up motions where base pair ends drift apart and opposite ends come closer together, while the oligo interior remains bound.
-The third mode shows symmetry along the main diagonal, much like the first mode but with an inverse trend.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_srv_correlations.PNG}
-        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
-        \label{fig:GC-core_srv_correlations}
-	\end{center}
-\end{figure}
-
-(Not sure whether to include this section before or after building the MSM)
-We examined each slow mode with respect to physical coordinates with which it showed high correlation.
-Unsurprisingly, the first mode tracks closely with the inverse distance between complement G:C pairs.
-These distances are the best indicators for a hybridization/dissociation, and we see a sharp change in the first SRV signal as these bonds form or break.
-The second slow mode is most active when the GC core pair are bound but the adjacent AT pairs are not.
-There is a small signal for fraying at the outer basepairs, but the mode overwhelmingly learns about these neighboring AT/GC bonds.
-This behavior reflects a kinetic trap between the core binding event and the fully hybridized (or evenly partially frayed) state.
-Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward (cite).
+Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis.
+Next, we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
+We found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "4bp-frayed" states.
+Again, we found transition probabilities between states and visualized representative molecular renderings.
+As we observed from the 2nd SRV mode analysis above, the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
+Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence}, \citep{Yin2011KineticsHybridization}.
 Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states.
-This occurs during the dissociation process as well, where one half of the AT base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
-The third mode is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
-We attribute this to the SRV learning about the slow diffusive motions of the two body system (not sure about this).
-In addition to picking up on dissociation behavior, the third mode seems to peak when the oligos are close together but configured in such a way that is not amenable hybridization.
-These misaligned conformations include inverse contacts where 5'/5' and 3'/3' ends meet, looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts, or some combination of the two (SI third mode screenshots).
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_tracking_modes.png}
-        \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.}
-        \label{fig:GC-core_tracking_modes}
-	\end{center}
-\end{figure}
-
-To further investigate these dynamics, we built an SRV-MSM using these first three SRV modes as a basis.
-We we found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "frayed" states.
-Again, we found transition probabilities between states and visualized representative molecular renderings.
-As we observed from the 2nd SRV mode analysis above, the "frayed" metatable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
+This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
 The MSM transition probabilities suggests that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
 We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
-This make sense considering that this state is more accessible from an already bound helix.
-Moreover, once oligo are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate.
-Thus once a dissociated to frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation.
-On the other hand, transitions from the hybridized to frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.
+This is expected considering that this state is more accessible from an already bound helix.
+Moreover, once oligos are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate.
+Thus once a dissociated to 4bp-frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation.
+On the other hand, transitions from the hybridized to 4bp-frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.
 
@@
 
+\subsubsection{\label{sec:Results}Diffusion maps show ensemble of frayed states}
+
+To further examine this 4bp-frayed state, we build diffusion maps using 10000 frames sampled from the macrostate.
+This time, we set out distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
+Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
+We found that a larger distance between the third and fourth A:T pairs increased the likelihood that a configuration was clustered into the meta-stable macrostate.
+This distance also correlated closely with the second diffusion map mode.
+Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core.
+In these conformations, one of the strands maintains some helical character while the other twist out of place, resulting in WC bonds being obstructed by the oligo backbone.
+These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
+We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_dmaps.png}
+        \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
+Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the the overlap between adjacent AT basepairs and the GC core.}
+        \label{fig:GC-core_dmaps}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}SRV correlations to physical coordinates?}
+
+Having constructed and interpreted our SRV-MSM, we revisited our original SRV basis to analyze what physical correlations could be interpreted from the model.
+We found these GC-core modes to be of particular interest as they reveal the hierarchical nature of the dynamical encoding.
+We examined a collection of 1000-frame "trimmed" trajectories centered on both hybridization or dissociation events.
+For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they showed strong correlation.
+Two representative trajectories are shown in figure: (\label{fig:GC-core_tracking_modes}).
+Complementary G:C pairs are the best indicators for a hybridization/dissociation event, and we see a sharp change in the first SRV mode (SRV1) as these bonds form or break.
+The second slow mode (SRV2) is most active when G:C pairs are bound but the adjacent AT pairs are not.
+There is a small signal for fraying at the outer base pairs, but the mode overwhelmingly learns about these neighboring A:T/G:C bonds.
+This behavior reflects the above-mentioned kinetic trap between core binding and the fully hybridized state.
+SRV3 is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
+We attribute this to the SRV learning about the diffusive motions of the two body system.
+In addition to picking up on dissociation behavior, the third mode peaks when the oligos are close together but configured in such a way that is not amenable hybridization.
+These misaligned conformations include inverse contacts where 5'/5' and 3'/3' ends meet and looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts.
+
+Despite the strong qualitative trends we observe between physical coordinates and leading SRV modes, we only find high Pearson correlations for SRV1.
+For the next two SRV modes, the sign of their correlation switches depending on whether the oligos are in the hybridized or dissociated state.
+This shows that these modes are providing support on top of the first mode -- which determines hybridization vs.
+dissociation -- and thus can display very different behaviors in either state.
+With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in states 0 and 1 -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state 2.
+Accordingly, we recalculated Pearson correlations between each SRV mode and all distances in states where the modes are active.
+Figure \label{fig:GC-core_srv_correlations} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each other during 4bp-fraying.
+We also observed a highly symmetric correlation between SRV3 and central base pairs distances, indicative of overall diffusive behavior.
+Taken together, these analyses reveal how the SRV learns and represents the dynamical space, leading to the MSM results we observe above.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_tracking_modes.png}
+        \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.
+Should probably go to SI if we also include GC-core diffusion maps}
+        \label{fig:GC-core_tracking_modes}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_srv_correlations_by_state.png}
+        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
+        \label{fig:GC-core_srv_correlations}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}Comparison to Experiment?}
+
 When examining these same four sequences using T-jump and 2D spectroscopy, Sandstead et al.
@@
 As their analysis did not consider the previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mid and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs.
+While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs.
 Further analysis showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
-We see the same rapid fraying in simulation data (which too fast to be attributed with an SRV mode), however we stipulate that this inter-conversion first relies on the formation of the the AT-bond nearest to the GC center.
-In visualizing trajectories, we observe that the frayed ends rotate away from each other and impede helix formation, further highlighting the importance of this 4th base pair formation.
+We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
+Our diffusion map analysis further shows how the ensemble of "macrofrayed trajectories" impede helix formation.
+Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\label{sec:Results}GC-mix}
+
+The GC-mix sequence shows a similar implied timescale distribution to GC-core, however we no longer see a converged slow mode corresponding to frayed behavior.
+Although we observe substantial fraying of the two AT termini in the simulation data, these dynamics are too fast to converge as distinct frayed modes.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior the strands while dissociated.
+These correlate closely with the first and third GC-core SRV modes.
+Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
+As such, we designated this transitions as an effectively two-state process with no long-lived or kinetically relevant metastable states.
+Previous work has attributed AT-all termini fraying prior to dissociation as a deviation from two-state behavior \citep{Sanstead2016}.
+Here we consider these AT-frayed states to be part of the overall hybridization state.
+While AT-termini fraying is surely a prerequisite for dissociation, we find these states to be so common and fleeting that very few evolve initiate dissociation.
+Furthermore, they do not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
+
+\subsubsection{\label{sec:Results}Definition of two-state and nucleation}
+
+In the simulation data, we do observe sustained regions of partial base pairing during hybridization.
+These states reflect initial contact that form between base pairs prior to a fast zippering mechanism that binds the oligos into a fully hybridized state.
+Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization} \citep{} (cite earlier porske 1971).
+Our featurization method, however, was designed to elucidate sequence-dependent -- thus location-dependent -- dynamics, and does not naturally group these kinds of state.
+This limitation also does not allow us to pick up on proceeding zippering mechanisms, as these would each appear as separate dynamics per the independent nucleation site and need to be sampled accordingly.
+
+Other ways to investigate GC-mix:
+\begin{enumerate}
+	\item Can try building simplified MSM here: (use number of base pairs bound as features)
+    \item Timescale analysis shows that GC-mix is reversible two-state process.
+We note that it takes about 40 percent longer for a fully hybridized oligomer to fully dissociate (once > 2 bp are borken) for GC-core compared to GC-mix.
+	\item Given that GC-mix and GC-core are run at very similar temperature, we would expect the extra basepairs broken during GC-mix dissociation process to slow dissociation, however in the opposite effect was observed in the slow mode data
+	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
+	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
+	\item Discuss misaligned configurations here instead of in the GC-core section
+\end{enumerate}
+
+\subsection{\label{sec:Results}Project results onto shared basis}
+
+Here I've constructed a shared basis from physical coordinates informed by the AT-all SRV2 and GC-core SRV2 (Figure \ref{fig:all_seq_physical_cvs_labels}) and TICA coordinates learned on a combination of all four sequence (Figure \ref{fig:all_seq_tica_cvs}).
+I'm leaning towards presenting the first in the main text and including the other in the SI, but I'm not sure how clear either of these are.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_physical_cvs_labels.PNG}
+        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes, Not sure how helpful these axes drawings are -- also need to switch x and y here.}
+        \label{fig:all_seq_physical_cvs_labels}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_tica_cvs.PNG}
+        \caption{All sequences plotted on combined TICA axes. Can add similar macrostate distribution as shown in the above figure.}
+        \label{fig:all_seq_tica_cvs}
+	\end{center}
+\end{figure}
+
+\subsubsection{\label{sec:Results}Limitations}
+
+\begin{enumerate}
+	\item Does not explicitly pick up on nucleation/zippering events
+	\item Coarse grain-models makes it difficult to compare timescales.
+Furthermore our implied timescales function more as an average frequency of a process rather than a value that can be directly compared to experiment.
+	\item Limited by memory (number of frames) required to capture equilibrium trajectories at a save rate that enabled adequate resolution for the dynamics of interest.
+
+\end{enumerate}
+
+\section{\label{sec:conc}Conclusion}
+\subsection{\label{sec:Results}Will look a lot like the last paragraph of the intro / the abstract}
+
+\begin{comment}
+Classic model has hybridization forming in three distinct step:
+A nucleation step involves bringing two random- coil ssDNAs together to form an intermediate state, in which a few base pairs have formed, and this intermediate pro- ceeds further to form the fully hybridized dsDNA.
+
+-
+
+- A smoother free energy surface minimizes the time spent in these intermediate states, and our featurization method
+
+- not capturing zippering
+One limititation of our appraoch that is that processes that are.
+For example, there are eight 3-bp sites on each oligo from which a zippering action coud nucleated.
+Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
+Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
 
@@
 
-Refs for DNA fraying:
-\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro).
-Use TPS to focus on kinetic pathway by which singl base pair bind/unbind.
-Focus on end base of 3-bp oligomer in explicit solvent.
-Finds order parameters tracking base pair binding as well as intra-strand stacking.
-(energy and distance for each).
-
-\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,
-
-6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
-37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
-38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding
-
-\subsection{\label{sec:Results}GC-mid}
-\begin{enumerate}
-	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
-	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
-	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
-
-\end{enumerate}
-
 %%%%%% Other MSM and timescales approaches  %%%%%%%%%%%
-## Good references:
 
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_tica_cvs.PNG}
-        \caption{All sequences plotted on combined TICA axes}
-        \label{fig:all_seq_tica_cvs}
-	\end{center}
-\end{figure}
+Refs for DNA fraying:
+\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro).
+Use TPS to focus on kinetic pathway by which singl base pair bind/unbind.
+Focus on end base of 3-bp oligomer in explicit solvent.
+Finds order parameters tracking base pair binding as well as intra-strand stacking.
+(energy and distance for each).
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_physical_cvs.PNG}
-        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes}
-        \label{fig:all_seq_physical_cvs}
-	\end{center}
-\end{figure}
+\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,
 
-\subsection{\label{sec:Results}Compare all results together}
-\subsubsection{\label{sec:Results}Figure showing Pearson correlations between leading modes on common basis}
+6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
+37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
+38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding
 
-\subsubsection{\label{sec:Results}Figure showing all sequences plotted on a common low dimensional space (probably the 2nd TICA mode and the 2nd GC-core mode)}
-
-\begin{enumerate}
-	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
-	\item Treat this as a two-states model, where fraying is an expected feature of the hybridized state but not does not characterize a metastable state
-\end{enumerate}
-
-Nucleation-zipper mechnasism is a fast process that can orginate anywhere along the strand making it difficult to track..
-
-\subsubsection{\label{sec:Results}Limitations} 
-- not capturing zippering
-One limititation of our appraoch that is that processes that are.
-For example, there are eight 3-bp sites on each oligo from which a zippering action coud nucleated.
-Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
-Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
-
-\section{\label{sec:conc}Conclusion}
-\subsection{\label{sec:Results}Looks a lot like the abstract}
+\end{comment}
 
@@
 %\bibliography{references}
-Test cite: \citep{Zhang2018DeepMechanics}
-\bibliography{refs_mendeley}
+\bibliography{refs_mendeley2}
 

### 2020-08-25 manuscript.tex
Andy pull

@@
 
-%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
-\newcommand*{\noter}[1]{}					% notes off
+\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
+%\newcommand*{\noter}[1]{}					% notes off
 
@@
 
-\noindent
+\noindent \url{https://www.overleaf.com/project/5e9e5110c524b8000192c548}
 
 \end{abstract}
-https://www.overleaf.com/project/5e9e5110c524b8000192c548
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}.
-Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
+\noter{Please correct reference ``MhatreV.HoJi-AnnLee2012NIHAccess''} Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
 Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019}.

### 2020-08-25 manuscript.tex
Updates from Overleaf

@@
   Pritzker School of Molecular Engineering, %
-  University of Chicago, %
-  Chicago, Illinois 60637%
+  The University of Chicago, %
+  929 East 57th Street, Chicago, Illinois 60637, United States%
+}
+
+\author{Brennan Ashwood}
+\affiliation{%
+  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
+  The University of Chicago, %
+  929 East 57th Street, Chicago, Illinois 60637, United States%
+}
+
+\author{Andrei Tokmakoff}
+\affiliation{%
+  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
+  The University of Chicago, %
+  929 East 57th Street, Chicago, Illinois 60637, United States%
 }
@@
   Pritzker School of Molecular Engineering, %
-  University of Chicago, %
-  Chicago, Illinois 60637%
+  The University of Chicago, %
+  929 East 57th Street, Chicago, Illinois 60637, United States%
 }
+
 %\author{I. Ken Groupleader}
@@
 \noter{Please correct reference ``MhatreV.HoJi-AnnLee2012NIHAccess''} Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
-Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019}.
+Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
 Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition.
@@
 Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
-The stability of certain intermediates states, such as out-of-register or "slipped" base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014}, \citep{Xiao2019}).
-Furthermore, Sandstead et al suggested the existence of frayed metastable states during duplex dissociation, where the stability of these states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}.
+The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019}).
+Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
+Sandstead et al.
+highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}.
 In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.
 
-Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
+\textcolor{purple}{I would suggest re-organizing this paragraph to emphasize how the dynamic insight from various experiments is limited and therefore high-level simulations are required to understand DNA dynamics in detail.
+Then, you can summarize why simulating DNA hybridization is difficult and the previously employed techniques} Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
 Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.
 Other computational studies use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
-Experimental studies often employ temperature jump, concentration jumps, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA}.
-Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by long timescales and data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices}.
+Experimental studies often employ temperature-jump, salt-jump, pH-jump, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}.
 Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.
 
-In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior at each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
 We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
@@
 -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
-We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
+We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 \textcolor{purple}{So these are initially in the bulk and then may interact with the DNA?
+Are these the effective concentrations after you shrink the box?} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
 In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
@@
 An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state.
-In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
+In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed  \textcolor{purple}{Why 1.5 microseconds?
+Is this arbitrary or is there a good reason to remove 1.5?}, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
 
@@
 This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
-The VAMP-2 scoring method, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
+The VAMP-2 scoring method \textcolor{purple}{Since the VAMP-2 scoring comes up a lot, can you maybe another sentence or two on how exactly this works?
+Citing the Noe paper helps, but I think experimentalists may get a bit lost with it}, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
 We compared the VAMP-2 score for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
-        \caption{Cross val procedure to selection number of SRV coordinates (probably will be SI, but could include AT-all cross-val in the walkthrough section)}
+        \caption{5-fold cross validation procedure to select number of SRV coordinates.
+We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
+(probably will be SI, but could include AT-all cross-val in the walkthrough section)}
         \label{fig:srv_crossval}
@@
 We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular we looked for a diminishing increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data.
+In particular we looked for a diminishinghttps://www.overleaf.com/project/5eff37185f425e0001a920ca increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data.
 We used a batch size of 50000 and ran each model for total of 30 training epochs.
@@
 We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns.
-We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation.
+We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation \textcolor{purple}{Not sure how to describe this or if it warrants further discussion?} .
 It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
@@
 The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes.
-Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics \citep{Markegard2015}, \citep{Xiao2019}.
+These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes \textcolor{purple}{I suppose the GCGCGC sequence in the Wales paper provides an interesting comparison to AT-all
+}.
+Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
 
@@
 To visualize these six macrostates, we project the data into the two leading TICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
+Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky} \textcolor{purple}{Still not sure why you use tICA here, does not nonlinear nature make them harder to include physically?
+Should these be in SI}.
 It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
@@
 This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length.
-It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes}, \citep{Dickman2012ThermodynamicDNAs}.
+Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length \textcolor{purple}{The effect of inert tails seems sequence dependent and not always destabilizing (Benight), especially at high ionic strength}.
+It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs} \textcolor{purple}{The Benight paper also shows that 3'-dangling ends and tails alter the change in heat capacity (dCp) upon hybridization much more than 5'.
+Additionally, their analysis suggests 3'-dangling ends lead to a reduction in bound counterions relative to 5'.
+Taken together, it would seem that 3'-dangling ends and tails may perturb the duplex structure/solvation environment in some manner.
+}.
 This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends.
-Furthermore, Santalucia thermodynamic calculations (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \citep{Allawi1998NearestDNA} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability).
+Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \textcolor{purple}{Also, it may be worth stating that NN parameters agree with experimental melting thermodynamics for the studied sequences (cite Paul's JACS paper).
+The NN parameters are occasionally quite bad for some sequences.} \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability) \textcolor{purple}{You may want to be more specific about how counterions impact inert tail thermodynamics  (more destabilizing at low counterion) and that your simulations are at fairly high counterion concentration}.
 Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
@@
 The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state.
-In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having more less intact base pairs.
-This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below.
-The final state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
+In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having less intact base pairs.
+This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below \textcolor{purple}{another inaccuracy of DNA NN is that they assume an all-or-nothing hybridization, so if the melting of each of these isolated conformations contained varying deviation from all-or-nothing behavior, then there would be another factor unaccounted for.
+}.
+The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
 This might be attributed to the asymptotic effects of inert tail destabilitization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{purple}{Consider getting rid of enthalpy and entropy columns (can add to SI).
+Need to add State numbers next to names}}
         \label{fig:AT-all_in_one}
@@
 All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation.
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation \textcolor{purple}{This is consisting with the GCGCGC sequence of Wales, right?
+They saw a large barrier to association from the double shifted state}.
 In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state.
@@
 These asymmetric kinetic effects may have some role in the substantially lower state populations we see above.
-From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state.
+From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state \textcolor{purple}{Based on the thermo studies of dangling ends and inert tails, it seems like the role of shifted states may depend on salt conditions, this could be an interesting future direction.}.
 
 For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
-Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates.
+Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{purple}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
 The number of SRV coordinates and MSM macrostates are varied based on the kinetics and thermodynamics behavior of specific sequences.
@@
 Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
-In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization} \textcolor{purple}{You may want to be more clear about the excluded volume for each of the nucleobases since this could have a large impact on the GC-ends shifted states.
+}.
 The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch interaction.
 Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
-Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic effects from inert tails.
+Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalipic (electrostatic) effects from inert tails \textcolor{Purple}{Message here should be that inert tails, looping, other effects have more substantial contribution than two-state NN calculation alone}.
 Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{Purple}{Need to fix 5' double shifted sequence, make same adjustments as for AT-all above}}
         \label{fig:GC-end_in_one}
@@
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch GC-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction \textcolor{purple}{state explicitly that C and T have greater excluded volume than G and A in 3spn.2}-- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+\textcolor{purple}{Will include 3' shifting comparision in SI}.
 The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
@@
 This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
-In T-jump experiments, the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
-Futhermore, 2D spectroscopy analysis showed a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the physical processes under investigation.
+In T-jump experiments, \textcolor{purple}{In addition, there was less deviation from all-or-nothing behavior in the FTIR and 2D IR temperature series} the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
+Futhermore, there was a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the fraying processes under investigation.
 \citep{Sanstead2018DirectDehybridization}.
 Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some oligomer population shifting out-of-register as we observe in our MSMs.
-This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence.
+This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence \textcolor{purple}{I think the real problem is that its unclear if IR spectroscopy (or most probes for that matter) can distinguish between fraying and shifting.
+Both will appear as the loss of AT and GC base pairs.
+The only thing I can think of is if we could see cross-peak between C and T or G and A features in a T-jump 2D IR surface.
+This is where spectral congestion could be a real problem.
+}.
 
@@
         \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
-        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (probs better to grey out underlying free energy landscape and move this to the SI.}
+        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (Not sure if this adds anything, if anything can grey out underlying free energy landscape and move this to the SI.}
         \label{fig:shifting_distributions}
@@
 
-The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end.
+The GC-core sequence represents a departure from the dominant slithering \textcolor{purple}{Need to pick a a slither/shift term to stick with.
+"Slithering" has been used most in the past, but "slithered " state doesn't sound right} dynamics observed for AT-all and GC-end.
 Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis.
@@
 This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
+We observe these events to happen with equal probability on either permutable side of the helix.
 The MSM transition probabilities suggests that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
@@
 These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
-We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart.
+We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart \textcolor{purple}{This seems like a very small difference.
+Could be due to undersampling or another excluded volume effect}.
 
@@
 For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they showed strong correlation.
-Two representative trajectories are shown in figure: (\label{fig:GC-core_tracking_modes}).
+Two representative trajectories are shown in figure: (\ref{fig:GC-core_tracking_modes}).
 Complementary G:C pairs are the best indicators for a hybridization/dissociation event, and we see a sharp change in the first SRV mode (SRV1) as these bonds form or break.
@@
         \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.
-Should probably go to SI if we also include GC-core diffusion maps}
+Should probably go to SI if we also include GC-core diffusion maps \textcolor{purple}{These single trajectory figures have proven useful for visualizing SRV dynamics, could be employed for GC-mix comparisons as well}}
         \label{fig:GC-core_tracking_modes}
@@
 
-When examining these same four sequences using T-jump and 2D spectroscopy, Sandstead et al.
+When examining these same four sequences using T-jump IR and 2D IR spectroscopy, Sandstead et al.
 found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
-As their analysis did not consider the previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs.
-Further analysis showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
+As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
+While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
+Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
 We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
@@
 Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as an effectively two-state process with no long-lived or kinetically relevant metastable states.
+As such, we designated this transitions as an effectively two-state \textcolor{purple}{I think there needs to be better clarity for the difference between "two-state" and "all-or-nothing".Your model gives only 2 clustered states, but the faster fraying motions still indicate that the dissociation is not all-or-nothing} process with no long-lived or kinetically relevant metastable states.
 Previous work has attributed AT-all termini fraying prior to dissociation as a deviation from two-state behavior \citep{Sanstead2016}.
@@
 These states reflect initial contact that form between base pairs prior to a fast zippering mechanism that binds the oligos into a fully hybridized state.
-Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization} \citep{} (cite earlier porske 1971).
+Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition} \textcolor{purple}{It may be a good idea to briefly discuss which sequences follow a nucleation-zipper model which show deviations.
+The nucleation-zipper is the most kinetic model for interpreting DNA hybridization, but it is probably a serious oversimplification in many cases!}.
 Our featurization method, however, was designed to elucidate sequence-dependent -- thus location-dependent -- dynamics, and does not naturally group these kinds of state.

### 2020-08-25 manuscript.tex
Updates from Overleaf

@@
 Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
-The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019}).
+The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}) \textcolor{purple}{No experimental has been done to hint at these dynamics (difficult to differentiate from fraying), but they've been discuss in many computational works and seem likely to occur}.
 Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
@@
 \section{\label{sec:Results}Results}
-- lead with same kind of diagram for each sequence, add additional figures when necessary for a given sequence
 

### 2020-08-25 manuscript.tex
Andy pull

--- 2020-08-25 manuscript.tex
@@
 Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
-The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019}).
+The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}) \textcolor{magenta}{No experimental has been done to hint at these dynamics (difficult to differentiate from fraying), but they've been discuss in many computational works and seem likely to occur}.
 Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
@@
 
-\textcolor{purple}{I would suggest re-organizing this paragraph to emphasize how the dynamic insight from various experiments is limited and therefore high-level simulations are required to understand DNA dynamics in detail.
+\textcolor{magenta}{I would suggest re-organizing this paragraph to emphasize how the dynamic insight from various experiments is limited and therefore high-level simulations are required to understand DNA dynamics in detail.
 Then, you can summarize why simulating DNA hybridization is difficult and the previously employed techniques} Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
@@
 -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
-We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 \textcolor{purple}{So these are initially in the bulk and then may interact with the DNA?
+We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 \textcolor{magenta}{So these are initially in the bulk and then may interact with the DNA?
 Are these the effective concentrations after you shrink the box?} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
@@
 An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state.
-In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed  \textcolor{purple}{Why 1.5 microseconds?
+In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed  \textcolor{magenta}{Why 1.5 microseconds?
 Is this arbitrary or is there a good reason to remove 1.5?}, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
@@
 This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
-The VAMP-2 scoring method \textcolor{purple}{Since the VAMP-2 scoring comes up a lot, can you maybe another sentence or two on how exactly this works?
+The VAMP-2 scoring method \textcolor{magenta}{Since the VAMP-2 scoring comes up a lot, can you maybe another sentence or two on how exactly this works?
 Citing the Noe paper helps, but I think experimentalists may get a bit lost with it}, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
@@
 \section{\label{sec:Results}Results}
-- lead with same kind of diagram for each sequence, add additional figures when necessary for a given sequence
 
-\subsection{\label{sec:Results}AT-all}
+\subsection{AT-all}
 
-\subsubsection{\label{sec:Results}SRV optimization and analysis}
+\subsubsection{SRV optimization and analysis}
 
@@
 We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns.
-We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation \textcolor{purple}{Not sure how to describe this or if it warrants further discussion?} .
+We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation \textcolor{magenta}{Not sure how to describe this or if it warrants further discussion?} .
 It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
@@
 The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes \textcolor{purple}{I suppose the GCGCGC sequence in the Wales paper provides an interesting comparison to AT-all
+These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes \textcolor{magenta}{I suppose the GCGCGC sequence in the Wales paper provides an interesting comparison to AT-all
 }.
@@
 
-\subsubsection{\label{sec:Results}SRV-MSM construction and optimization}
+\subsubsection{SRV-MSM construction and optimization}
 
@@
 To visualize these six macrostates, we project the data into the two leading TICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky} \textcolor{purple}{Still not sure why you use tICA here, does not nonlinear nature make them harder to include physically?
+Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky} \textcolor{magenta}{Still not sure why you use tICA here, does not nonlinear nature make them harder to include physically?
 Should these be in SI}.
@@
 
-\subsubsection{\label{sec:Results}Implied timescales for TICA-MSMs for }
+\subsubsection{Implied timescales for TICA-MSMs for }
 
@@
 This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length \textcolor{purple}{The effect of inert tails seems sequence dependent and not always destabilizing (Benight), especially at high ionic strength}.
-It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs} \textcolor{purple}{The Benight paper also shows that 3'-dangling ends and tails alter the change in heat capacity (dCp) upon hybridization much more than 5'.
+Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length \textcolor{magenta}{The effect of inert tails seems sequence dependent and not always destabilizing (Benight), especially at high ionic strength}.
+It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs} \textcolor{magenta}{The Benight paper also shows that 3'-dangling ends and tails alter the change in heat capacity (dCp) upon hybridization much more than 5'.
 Additionally, their analysis suggests 3'-dangling ends lead to a reduction in bound counterions relative to 5'.
@@
 This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends.
-Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \textcolor{purple}{Also, it may be worth stating that NN parameters agree with experimental melting thermodynamics for the studied sequences (cite Paul's JACS paper).
-The NN parameters are occasionally quite bad for some sequences.} \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability) \textcolor{purple}{You may want to be more specific about how counterions impact inert tail thermodynamics  (more destabilizing at low counterion) and that your simulations are at fairly high counterion concentration}.
+Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \textcolor{magenta}{Also, it may be worth stating that NN parameters agree with experimental melting thermodynamics for the studied sequences (cite Paul's JACS paper).
+The NN parameters are occasionally quite bad for some sequences.} \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability) \textcolor{magenta}{You may want to be more specific about how counterions impact inert tail thermodynamics  (more destabilizing at low counterion) and that your simulations are at fairly high counterion concentration}.
 Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
@@
 In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having less intact base pairs.
-This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below \textcolor{purple}{another inaccuracy of DNA NN is that they assume an all-or-nothing hybridization, so if the melting of each of these isolated conformations contained varying deviation from all-or-nothing behavior, then there would be another factor unaccounted for.
+This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below \textcolor{magenta}{another inaccuracy of DNA NN is that they assume an all-or-nothing hybridization, so if the melting of each of these isolated conformations contained varying deviation from all-or-nothing behavior, then there would be another factor unaccounted for.
 }.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{purple}{Consider getting rid of enthalpy and entropy columns (can add to SI).
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{magenta}{Consider getting rid of enthalpy and entropy columns (can add to SI).
 Need to add State numbers next to names}}
@@
 
-\subsubsection{\label{sec:Results}Transition probabilities between states}
+\subsubsection{Transition probabilities between states}
 
@@
 All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation \textcolor{purple}{This is consisting with the GCGCGC sequence of Wales, right?
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation \textcolor{magenta}{This is consisting with the GCGCGC sequence of Wales, right?
 They saw a large barrier to association from the double shifted state}.
@@
 These asymmetric kinetic effects may have some role in the substantially lower state populations we see above.
-From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state \textcolor{purple}{Based on the thermo studies of dangling ends and inert tails, it seems like the role of shifted states may depend on salt conditions, this could be an interesting future direction.}.
+From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state \textcolor{magenta}{Based on the thermo studies of dangling ends and inert tails, it seems like the role of shifted states may depend on salt conditions, this could be an interesting future direction.}.
 
 For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
-Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{purple}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
+Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{magenta}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
 The number of SRV coordinates and MSM macrostates are varied based on the kinetics and thermodynamics behavior of specific sequences.
@@
 
-\subsection{\label{sec:Results}GC-end}
+\subsection{GC-end}
 
-\subsubsection{\label{sec:Results}SRV-MSM construction and states}
+\subsubsection{SRV-MSM construction and states}
 
@@
 
-\subsubsection{\label{sec:Results}Comparison between GC-end and AT-all shifted states}
+\subsubsection{Comparison between GC-end and AT-all shifted states}
 
@@
 Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
-In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization} \textcolor{purple}{You may want to be more clear about the excluded volume for each of the nucleobases since this could have a large impact on the GC-ends shifted states.
+In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization} \textcolor{magenta}{You may want to be more clear about the excluded volume for each of the nucleobases since this could have a large impact on the GC-ends shifted states.
 }.
@@
 Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
-Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalipic (electrostatic) effects from inert tails \textcolor{Purple}{Message here should be that inert tails, looping, other effects have more substantial contribution than two-state NN calculation alone}.
+Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalipic (electrostatic) effects from inert tails \textcolor{magenta}{Message here should be that inert tails, looping, other effects have more substantial contribution than two-state NN calculation alone}.
 Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{Purple}{Need to fix 5' double shifted sequence, make same adjustments as for AT-all above}}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{magenta}{Need to fix 5' double shifted sequence, make same adjustments as for AT-all above}}
         \label{fig:GC-end_in_one}
@@
 
-\subsubsection{\label{sec:Results}Diffusion maps show sequence differences in shifted states}
+\subsubsection{Diffusion maps show sequence differences in shifted states}
 
@@
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction \textcolor{purple}{state explicitly that C and T have greater excluded volume than G and A in 3spn.2}-- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-\textcolor{purple}{Will include 3' shifting comparision in SI}.
+This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction \textcolor{magenta}{state explicitly that C and T have greater excluded volume than G and A in 3spn.2}-- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+\textcolor{magenta}{Will include 3' shifting comparision in SI}.
 The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
@@
 
-\subsubsection{\label{sec:Results}GC-end dynamics are largely two-state}
+\subsubsection{GC-end dynamics are largely two-state}
 
@@
 This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
-In T-jump experiments, \textcolor{purple}{In addition, there was less deviation from all-or-nothing behavior in the FTIR and 2D IR temperature series} the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
+In T-jump experiments, \textcolor{magenta}{In addition, there was less deviation from all-or-nothing behavior in the FTIR and 2D IR temperature series} the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
 Futhermore, there was a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the fraying processes under investigation.
@@
 Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some oligomer population shifting out-of-register as we observe in our MSMs.
-This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence \textcolor{purple}{I think the real problem is that its unclear if IR spectroscopy (or most probes for that matter) can distinguish between fraying and shifting.
+This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence \textcolor{magenta}{I think the real problem is that its unclear if IR spectroscopy (or most probes for that matter) can distinguish between fraying and shifting.
 Both will appear as the loss of AT and GC base pairs.
@@
 
-\subsection{\label{sec:Results}GC-core}
+\subsection{GC-core}
 
-\subsubsection{\label{sec:Results}Building SRV-MSM}
+\subsubsection{Building SRV-MSM}
 
-The GC-core sequence represents a departure from the dominant slithering \textcolor{purple}{Need to pick a a slither/shift term to stick with.
+The GC-core sequence represents a departure from the dominant slithering \textcolor{magenta}{Need to pick a a slither/shift term to stick with.
 "Slithering" has been used most in the past, but "slithered " state doesn't sound right} dynamics observed for AT-all and GC-end.
@@
 
-\subsubsection{\label{sec:Results}Diffusion maps show ensemble of frayed states}
+\subsubsection{Diffusion maps show ensemble of frayed states}
 
@@
 These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
-We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart \textcolor{purple}{This seems like a very small difference.
+We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart \textcolor{magenta}{This seems like a very small difference.
 Could be due to undersampling or another excluded volume effect}.
@@
 
-\subsubsection{\label{sec:Results}SRV correlations to physical coordinates?}
+\subsubsection{SRV correlations to physical coordinates?}
 
@@
         \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.
-Should probably go to SI if we also include GC-core diffusion maps \textcolor{purple}{These single trajectory figures have proven useful for visualizing SRV dynamics, could be employed for GC-mix comparisons as well}}
+Should probably go to SI if we also include GC-core diffusion maps \textcolor{magenta}{These single trajectory figures have proven useful for visualizing SRV dynamics, could be employed for GC-mix comparisons as well}}
         \label{fig:GC-core_tracking_modes}
@@
 
-\subsubsection{\label{sec:Results}Comparison to Experiment?}
+\subsubsection{Comparison to Experiment?}
 
@@
 
-\subsection{\label{sec:Results}GC-mix}
+\subsection{GC-mix}
 
@@
 Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as an effectively two-state \textcolor{purple}{I think there needs to be better clarity for the difference between "two-state" and "all-or-nothing".Your model gives only 2 clustered states, but the faster fraying motions still indicate that the dissociation is not all-or-nothing} process with no long-lived or kinetically relevant metastable states.
+As such, we designated this transitions as an effectively two-state \textcolor{magenta}{I think there needs to be better clarity for the difference between "two-state" and "all-or-nothing".Your model gives only 2 clustered states, but the faster fraying motions still indicate that the dissociation is not all-or-nothing} process with no long-lived or kinetically relevant metastable states.
 Previous work has attributed AT-all termini fraying prior to dissociation as a deviation from two-state behavior \citep{Sanstead2016}.
@@
 
-\subsubsection{\label{sec:Results}Definition of two-state and nucleation}
+\subsubsection{Definition of two-state and nucleation}
 
@@
 These states reflect initial contact that form between base pairs prior to a fast zippering mechanism that binds the oligos into a fully hybridized state.
-Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition} \textcolor{purple}{It may be a good idea to briefly discuss which sequences follow a nucleation-zipper model which show deviations.
+Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition} \textcolor{magenta}{It may be a good idea to briefly discuss which sequences follow a nucleation-zipper model which show deviations.
 The nucleation-zipper is the most kinetic model for interpreting DNA hybridization, but it is probably a serious oversimplification in many cases!}.
@@
 
-\subsection{\label{sec:Results}Project results onto shared basis}
+\subsection{Project results onto shared basis}
 
@@
 
-\subsubsection{\label{sec:Results}Limitations}
+\subsubsection{Limitations}
 
@@
 \section{\label{sec:conc}Conclusion}
-\subsection{\label{sec:Results}Will look a lot like the last paragraph of the intro / the abstract}
+\subsection{Will look a lot like the last paragraph of the intro / the abstract}
 

### 2020-09-23 manuscript.tex
Updates from Overleaf

--- 2020-08-25 manuscript.tex
@@
 \newcommand*{\blauwr}[1]{#1}
+\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
 
@@
 \usepackage{comment}
+
+\usepackage[normalem]{ulem}
+\usepackage{algorithm}
+\usepackage{subfig}
+\usepackage{algpseudocode}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author{Mike Jones}
+\author{Michael S. Jones}
 \affiliation{%
@@
 
+\begin{comment}
+%%%%%% intro notes %%%%%%%
+\citep{Sanstead2018DirectDehybridization} Good intro with refs and focus on importance of dynamics "rich dynamics even for short oligos" ...
+Many of the most important functions performed by DNA are highly dynamic, whether in the the biological roles of replication transcription, and gene regulation, or in the expanding field of DNA-based nanotechnology
+\end{comment}
+
 \section{\label{sec:intro}Introduction}
@@
 Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}.
-\noter{Please correct reference ``MhatreV.HoJi-AnnLee2012NIHAccess''} Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
+Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
 Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
@@
 Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
-The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019}).
+The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}).
 Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
 Sandstead et al.
-highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}.
+highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}.
 In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.
 
-\textcolor{purple}{I would suggest re-organizing this paragraph to emphasize how the dynamic insight from various experiments is limited and therefore high-level simulations are required to understand DNA dynamics in detail.
-Then, you can summarize why simulating DNA hybridization is difficult and the previously employed techniques} Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.
-Other computational studies use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
-Experimental studies often employ temperature-jump, salt-jump, pH-jump, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
-Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}.
-Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.
+Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}.
+Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.
+Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
+Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.
 
-In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+\textcolor{purple}{Could add more here about all-atom simulations?}
+
+In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
 We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
@@
 \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}.
-We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
+We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
 Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed.
@@
 We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
-By evaluating equilibrium trajectories we can study the relevance of meta-stable states during the hybridization and dissociation process.
+By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process.
 Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
@@
 Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
+Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics \textcolor{purple}{Not sure how much to say about this here, but I think these experiments should be mentioned in the intro}.
 Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.
@@
 
-Is some additional information warranted on the model itself before going into this specific setup?
+\textcolor{purple}{Is some additional information warranted on the model itself before going into this specific setup?}
 
@@
 -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
-We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 \textcolor{purple}{So these are initially in the bulk and then may interact with the DNA?
-Are these the effective concentrations after you shrink the box?} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
+We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
 In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
@@
 We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions.
-We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions.
-These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}.
-We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
-Simulation temperature was determined by sequence specific melting temperature such that strands were equally likely to spend time in the hybridized or dissociated state.
-Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps.
-A 15 fs timestep was used for all simulations runs.
+We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}.
+We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
+We set the simulation temperature was set about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states.
+We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time.
+Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes.
 For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation.
 An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state.
-In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed  \textcolor{purple}{Why 1.5 microseconds?
-Is this arbitrary or is there a good reason to remove 1.5?}, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
+In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
 
@@
 This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
-The VAMP-2 scoring method \textcolor{purple}{Since the VAMP-2 scoring comes up a lot, can you maybe another sentence or two on how exactly this works?
-Citing the Noe paper helps, but I think experimentalists may get a bit lost with it}, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}.
-We compared the VAMP-2 score for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set.
-Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set.
+The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}.
+The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models.
+In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
+The score $R_2$ can then be found for $\chi$ by applying the VAMP principle with cross-validation.
+
+\begin{align}
+ 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
+ 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
+ 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
+	\\
+ 	R_2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
+ 	%R_2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
+\end{align}\label{CK1}
+
+% for more general case with  \chi_0, \chi_1, but in practice this is not used
+\begin{comment} 
+\begin{align}
+ 	\mathscr{C}_{00}=&\Expect{\chi_0(x_t)\chi_0(x_t)^\intercal}_t\\
+ 	\mathscr{C}_{01}=&\Expect{\chi_0(x_t)\chi_1(x_{t+\tau})^\intercal}_t\\
+ 	\mathscr{C}_{11}=&\Expect{\chi_1(x_{t+\tau})\chi_1(x_{t+\tau})^\intercal}_{t+\tau}\\
+	\\
+ 	R_2[\chi_0, \chi_1]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2\\
+ 	R_2^{val}[\chi_0, \chi_1]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1\\
+\end{align}\label{CK2}
+\end{comment}
+
+We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set.
+We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates as a consistent feature set.
 These features were normalized and passed into sequence-specific SRVs.
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
-        \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
-(probably will be SI, but could include AT-all cross-val in the walkthrough section)}
-        \label{fig:srv_crossval}
-	\end{center}
-\end{figure}
+% include covariance, vamp-2  and validation equations, lag time equation from vampnet paper mardt2018
 
@@
 The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features.
-This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?).
+This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.
 The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.
-Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}.
-SRVs provide robust nonlinear approximation and computation timescales linearly with the amount of input data.
+Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}.
+SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data.
 This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data.
-The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen}, \citep{Sidky}.
+The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
 For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
 
-Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set.
+Using optimized hyperparameters and featurized trajectory data, we transform 55 reciprocal pairwise distances into a low dimensional SRV basis set.
 In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular we looked for a diminishinghttps://www.overleaf.com/project/5eff37185f425e0001a920ca increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data.
-We used a batch size of 50000 and ran each model for total of 30 training epochs.
+We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
+In particular, we looked for convergence in the vamp-2 score and inconsistency between cross validation scores -- suggesting that the model may be overfitting on artifacts in the training data.
+We used a batch size of 50000 and ran each model for a total of 30 training epochs.
 We used two hidden layers and set the size of each layer to 100.
@@
 SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com} \citep{Abadi2016TensorFlow:Systems}.
+SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
 
@@
 Furthermore, SRV-MSM implied timescales converge faster than TICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.
-To build our SRV-MSM framework, we employed the Pyemma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
-After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ macrostate assignments.
+To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
+After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.
 The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
@@
 Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that most implied timescales (with the exception of the leading timescale) converge at a lag time of 600 ns.
-We kept a looser constraint on the convergence of the leading mode as it reflects the less Markovian dynamic of strand association and dissociation \textcolor{purple}{Not sure how to describe this or if it warrants further discussion?} .
+We found that most implied timescales converge at an SRV lag time of 600 ps.
+We kept a looser constraint on the convergence of the leading mode -- corresponding to the overall hybridization/dissociation process -- as it tends to have a longer transit period and becomes more difficult to fulfill Markovian conditions.
 It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
@@
 For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-As shown in (Figure \ref{fig:srv_crossval}), we see a clear convergence of the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
+As shown in (Figure \ref{fig:srv_crossval-AT-all}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
 Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
@@
 The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes \textcolor{purple}{I suppose the GCGCGC sequence in the Wales paper provides an interesting comparison to AT-all
-}.
-Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
+These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes.
+Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=50mm, 
+        scale=0.5]{Figs/figs_0804/srv_crossval-AT-all.png}
+        \caption{5-fold cross validation procedure to select number of SRV coordinates.
+We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
+For AT-all we observed a spectral gap after the 5th coordinate.
+Cross-validation on other sequences is in the SI.
+(Add AT-all feature scores here as well?)}
+        \label{fig:srv_crossval-AT-all}
+	\end{center}
+\end{figure}
 
@@
 Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represent a collection of metastable structures.
-Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels}, \citep{PinamontiAnalyzedModels}.
+Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
 Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
@@
 To visualize these six macrostates, we project the data into the two leading TICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky} \textcolor{purple}{Still not sure why you use tICA here, does not nonlinear nature make them harder to include physically?
-Should these be in SI}.
+Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
 It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
@@
 These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states.
-This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019}, \citep{Phys2014}.
+This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019, Phys2014}.
 We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states.
 This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling 5' ends have a consistently stabilizing effect, and inert tails decrease stability as they increase in length \textcolor{purple}{The effect of inert tails seems sequence dependent and not always destabilizing (Benight), especially at high ionic strength}.
-It has been shown that two base pair 5' dangling ends (or dangling ends with one inert tail) have higher melting temperature and are enthalpically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs} \textcolor{purple}{The Benight paper also shows that 3'-dangling ends and tails alter the change in heat capacity (dCp) upon hybridization much more than 5'.
-Additionally, their analysis suggests 3'-dangling ends lead to a reduction in bound counterions relative to 5'.
-Taken together, it would seem that 3'-dangling ends and tails may perturb the duplex structure/solvation environment in some manner.
-}.
-This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends.
-Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state \textcolor{purple}{Also, it may be worth stating that NN parameters agree with experimental melting thermodynamics for the studied sequences (cite Paul's JACS paper).
-The NN parameters are occasionally quite bad for some sequences.} \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} (Note that these calculations are performed at higher salt concentration, and that salt seems to have a significant effect on the impact of inert tails on stability) \textcolor{purple}{You may want to be more specific about how counterions impact inert tail thermodynamics  (more destabilizing at low counterion) and that your simulations are at fairly high counterion concentration}.
-Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
-Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average lifetimes of these states.
+Dangling 5' ends have a consistently stabilizing effect, and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
+It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
+This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends and/or 3' ends binding few counterions and perturbing the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
+Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state.
+Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
+Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average "dwell" time between transition into and out of these states.
 The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state.
-In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having less intact base pairs.
-This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below \textcolor{purple}{another inaccuracy of DNA NN is that they assume an all-or-nothing hybridization, so if the melting of each of these isolated conformations contained varying deviation from all-or-nothing behavior, then there would be another factor unaccounted for.
-}.
+In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having two less intact base pairs.
+This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below.
 The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
-This might be attributed to the asymptotic effects of inert tail destabilitization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
+This might be attributed to the asymptotic effects of inert tail destabilization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
 Accordingly, the longer inert tails in the double-shifted states have less of an effect on relative 5' vs.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{purple}{Consider getting rid of enthalpy and entropy columns (can add to SI).
-Need to add State numbers next to names}}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
         \label{fig:AT-all_in_one}
@@
 
-Beyond state probabilities and free energy approximations, the macrostate MSM also yields discrete kinetic information in the form of transition probabilities between states.
+Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
 Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns).
 All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation \textcolor{purple}{This is consisting with the GCGCGC sequence of Wales, right?
-They saw a large barrier to association from the double shifted state}.
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation.
 In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state.
@@
 These asymmetric kinetic effects may have some role in the substantially lower state populations we see above.
-From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state \textcolor{purple}{Based on the thermo studies of dangling ends and inert tails, it seems like the role of shifted states may depend on salt conditions, this could be an interesting future direction.}.
+From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state.
+We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.
 
@@
 The 3' double-shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes).
-Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches decreases the number of out-of-register WC pairs by two relative to corresponding AT-all states.
+Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches replace two intact WC bonds in the corresponding AT-all states.
 This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and shorter timescales.
@@
 For our analysis, we consider C:T and G:A base pairs in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends and two inert tails.
-Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM}, \citep{DiMichele2014EffectHybridization}.
-In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization} \textcolor{purple}{You may want to be more clear about the excluded volume for each of the nucleobases since this could have a large impact on the GC-ends shifted states.
-}.
+Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
+In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
 The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch interaction.
-Nevertheless, we find these bonds to be stabilizing enough to account for the relatively small population of conformation we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
+Despite fewer intact base pairs, we find these structures stabilizing enough to account for the relatively small population of conformations we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
 Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalipic (electrostatic) effects from inert tails \textcolor{Purple}{Message here should be that inert tails, looping, other effects have more substantial contribution than two-state NN calculation alone}.
+However, state populations reveal that the 5' single-shifted is again significantly more stable than the 3'.
 Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
@@
 To compare these single-shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 5000 conformations from 5' shifted state of both sequences.
-Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps}.
-We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
+Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either identical end of the shifted conformation.
 This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands.
 In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}.
+Diffusion maps built from samples of the 3' shifted states are also shown in the SI.
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
 This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction \textcolor{purple}{state explicitly that C and T have greater excluded volume than G and A in 3spn.2}-- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-\textcolor{purple}{Will include 3' shifting comparision in SI}.
+This effect may be exaggerated given that C and T base pairs are assigning slightly higher exluded volume radii in 3spn2.
 The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
 Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
-These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions.
-Although AT-all shifted ends tend to stay bound out-of-register, we find the second diffusion map coordinate to be indicative of inert tails that fold back onto these terminal out-of-register pairings.
+These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
 
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_full.PNG}
-        \caption{SI fig }
-        \label{fig:GC-end_dmaps_full}
-	\end{center}
-\end{figure}
-
 \subsubsection{\label{sec:Results}GC-end dynamics are largely two-state}
@@
 This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
-In T-jump experiments, \textcolor{purple}{In addition, there was less deviation from all-or-nothing behavior in the FTIR and 2D IR temperature series} the GC-end sequences was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
-Futhermore, there was a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the fraying processes under investigation.
+In T-jump, FTIR, and 2D IR experiments, the GC-end sequence was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
+Furthermore, there was a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the fraying processes under investigation.
 \citep{Sanstead2018DirectDehybridization}.
-Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some oligomer population shifting out-of-register as we observe in our MSMs.
-This cannot be verified given that the signal was recorded in a congested spectroscopic range, however this dynamic should be taken into consideration when analyzing or utilizing oligos of a similar sequence \textcolor{purple}{I think the real problem is that its unclear if IR spectroscopy (or most probes for that matter) can distinguish between fraying and shifting.
-Both will appear as the loss of AT and GC base pairs.
-The only thing I can think of is if we could see cross-peak between C and T or G and A features in a T-jump 2D IR surface.
-This is where spectral congestion could be a real problem.
-}.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
-        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (Not sure if this adds anything, if anything can grey out underlying free energy landscape and move this to the SI.}
-        \label{fig:shifting_distributions}
-	\end{center}
-\end{figure}
+Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some shifted or related looping dynamics.
+This cannot be verified given that the signal was recorded in a congested spectroscopic range and could be hard to differentiate from aligned G:C fraying.
+We will discuss this more in our investigation of higher temperatures.
 
@@
 
-The GC-core sequence represents a departure from the dominant slithering \textcolor{purple}{Need to pick a a slither/shift term to stick with.
-"Slithering" has been used most in the past, but "slithered " state doesn't sound right} dynamics observed for AT-all and GC-end.
+The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end.
 Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis.
@@
 Again, we found transition probabilities between states and visualized representative molecular renderings.
-As we observed from the 2nd SRV mode analysis above, the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
-Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence}, \citep{Yin2011KineticsHybridization}.
+We observe that the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
+Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
 Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states.
 This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
-We observe these events to happen with equal probability on either permutable side of the helix.
-The MSM transition probabilities suggests that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
+We observe these events to occur with equal probability on either permutable side of the helix.
+The MSM transition probabilities suggest that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
 We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
@@
 
-To further examine this 4bp-frayed state, we build diffusion maps using 10000 frames sampled from the macrostate.
-This time, we set out distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
+To further examine this 4bp-frayed state, we build diffusion maps using 10000 frames sampled from the macrostate \ref{fig:GC-core_dmaps}.
+This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
 Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
-We found that a larger distance between the third and fourth A:T pairs increased the likelihood that a configuration was clustered into the meta-stable macrostate.
+We found that a larger distance between the third and fourth A:T pairs increased the PCCA+ probability of inclusion into macrostate.
 This distance also correlated closely with the second diffusion map mode.
@@
 These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
-We observed this mode to be mostly symmetric, however there is tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart \textcolor{purple}{This seems like a very small difference.
-Could be due to undersampling or another excluded volume effect}.
+We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart.
+This might be another consequence of differential excluded volume radii in the 3spn2 force field.
 
@@
 dissociation -- and thus can display very different behaviors in either state.
-With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in states 0 and 1 -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state 2.
+With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in states 0 and 1 -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state (S2).
 Accordingly, we recalculated Pearson correlations between each SRV mode and all distances in states where the modes are active.
-Figure \label{fig:GC-core_srv_correlations} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each other during 4bp-fraying.
+Figure \label{fig:GC-core_tracking_modes} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each other during 4bp-fraying.
 We also observed a highly symmetric correlation between SRV3 and central base pairs distances, indicative of overall diffusive behavior.
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_srv_correlations_by_state.png}
-        \caption{Shows Pearson correlations between leading SRV modes and all 100 intermolecular distances (needs labels for slow modes)}
-        \label{fig:GC-core_srv_correlations}
-	\end{center}
-\end{figure}
-
-\subsubsection{\label{sec:Results}Comparison to Experiment?}
+\subsubsection{\label{sec:Results}Comparison to Experiment}
 
@@
 While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
-Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
+Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
 We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
-Our diffusion map analysis further shows how the ensemble of "macrofrayed trajectories" impede helix formation.
+Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.
 Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
@@
 
-The GC-mix sequence shows a similar implied timescale distribution to GC-core, however we no longer see a converged slow mode corresponding to frayed behavior.
-Although we observe substantial fraying of the two AT termini in the simulation data, these dynamics are too fast to converge as distinct frayed modes.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior the strands while dissociated.
+The GC-mix sequence shows a similar implied timescale distribution to GC-core, however we no longer see a converged slow mode corresponding to multi-base fraying behavior.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
 These correlate closely with the first and third GC-core SRV modes.
 Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as an effectively two-state \textcolor{purple}{I think there needs to be better clarity for the difference between "two-state" and "all-or-nothing".Your model gives only 2 clustered states, but the faster fraying motions still indicate that the dissociation is not all-or-nothing} process with no long-lived or kinetically relevant metastable states.
-Previous work has attributed AT-all termini fraying prior to dissociation as a deviation from two-state behavior \citep{Sanstead2016}.
-Here we consider these AT-frayed states to be part of the overall hybridization state.
-While AT-termini fraying is surely a prerequisite for dissociation, we find these states to be so common and fleeting that very few evolve initiate dissociation.
-Furthermore, they do not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
+As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data.
+Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence attributing fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
+While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation.
+On the other hand, the 4bp-frayed state in GC-core had a substantial probability (~10\%) of fully dissociated rather reforming an intact helix.
+Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
 
-\subsubsection{\label{sec:Results}Definition of two-state and nucleation}
+Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
+For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
+During hybridization, we observe the formations of some key base pair contacts before the full duplex forms.
+Although it is statistically more likely for contact to form at the end of the duplex, we observe first contact at interior base pairs as well.
+This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition}.
+We obverse dissociation events proceeded by two base pair fraying on one side of the duplex followed by more rapid dissociation of the central base pairs.
+There tends to be a short-lived state composed of 2-4 base pair contacts shortly before fully dissociation occurs.
+In contrast to the 4-bps frayed state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or TICA space, nor do they tend to reform intact duplexes.
+As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}.
+We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
 
-In the simulation data, we do observe sustained regions of partial base pairing during hybridization.
-These states reflect initial contact that form between base pairs prior to a fast zippering mechanism that binds the oligos into a fully hybridized state.
-Qualitatively, these observations match well with the three-state assumption for hybridization that proceeds through a "nucleation" state of around 3 base pairs bound \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition} \textcolor{purple}{It may be a good idea to briefly discuss which sequences follow a nucleation-zipper model which show deviations.
-The nucleation-zipper is the most kinetic model for interpreting DNA hybridization, but it is probably a serious oversimplification in many cases!}.
-Our featurization method, however, was designed to elucidate sequence-dependent -- thus location-dependent -- dynamics, and does not naturally group these kinds of state.
-This limitation also does not allow us to pick up on proceeding zippering mechanisms, as these would each appear as separate dynamics per the independent nucleation site and need to be sampled accordingly.
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-mix_transitions.PNG}
+        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
+        \label{fig:GC-mix_transitions}
+	\end{center}
+\end{figure}
 
-Other ways to investigate GC-mix:
+\subsubsection{\label{sec:Results}Temperature-dependent timescale comparisons}
+
+Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone.
+To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence.
+We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core \citep{Sanstead2018DirectDehybridization}.
+Although there was no experimental data for AT-all and GC-mix, we used our relaxation fits to predict how these might compare with the other sequences.
+We measure the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
+We find the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the known thermodynamic influence on duplex stability.
+Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
+
+The fast response -- which Sanstead et.
+al attributed to single base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
+Furthermore, all-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen binding, and terminal stacked conformations\citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
+Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
+Instead, we measure the fast fraying response by observing how long it takes for either duplex terminal end to split beyond a given cutoff.
+This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
+This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events.
+For further comparison, we include fits over a range of cutoffs and repeat these calculations on the second and third interior base pairs.
+
+For GC-core, each base pair and cutoff combination appears linear with temperature, indicating a barrierless and diffusion-driven process.
+As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated 100-300x relative to experiments.
+In contrast, GC-end is distinctly exponential, especially at base pairs closer to the core of the duplex.
+This behavior makes sense given the thermodynamic barrier to G:C fraying, but it also deviates the GC-end experimental linear fit.
+In addition to fraying, we notice that the rate of slithering increases at higher temperature.
+Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex as temperature increases.
+For higher T-jumps, it is possible that a significant portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by slower "looping" G:C bonding that we observed in our diffusion map analysis above.
+It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
+
+The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
+Previous work has shows that kinetic association rates were accelerated by about 200x relative to experiment \citep{Hinckley2013AnHybridization}.
+This seems to be on the same order of magnitude as frayed acceleration, but dissociation is noticeably less accelerated when compared to experiment.
+It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not parameterized based on any dynamical properties \citep{?}.
+
+%We also did not find available data on other acceleration studies using explicit ions.
+
+\subsubsection{\label{sec:Results}Limitations}
+
 \begin{enumerate}
-	\item Can try building simplified MSM here: (use number of base pairs bound as features)
-    \item Timescale analysis shows that GC-mix is reversible two-state process.
-We note that it takes about 40 percent longer for a fully hybridized oligomer to fully dissociate (once > 2 bp are borken) for GC-core compared to GC-mix.
-	\item Given that GC-mix and GC-core are run at very similar temperature, we would expect the extra basepairs broken during GC-mix dissociation process to slow dissociation, however in the opposite effect was observed in the slow mode data
-	\item Evidence of fraying correlations in higher order modes, but these are not consistent and their inclusion does not generate a metastable third mode
-	\item Show 2nd slow mode correlations in SI plus some configuration snapshots
-	\item Discuss misaligned configurations here instead of in the GC-core section
+	\item Does not explicitly pick up on nucleation/zippering events.
+Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
+Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
+	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
+	\item Limited by memory (number of frames) required to capture equilibrium trajectories at a save rate that enabled adequate resolution for the dynamics of interest.
+	\item A smoother free energy surface minimizes the time spent in these intermediate states, and our featurization method is not sensitive to certain kinds of dynamics
+	\item Sensitive to 3spn2 artefacts.
+Slithering may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}.
+Base-pair mismatch interactions may be somewhat unphysical and alter state populations.
+Although we present slithering as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
+We hope that these data will present a
+
 \end{enumerate}
 
-\subsection{\label{sec:Results}Project results onto shared basis}
+\subsection{\label{sec:Results}SI figs}
 
-Here I've constructed a shared basis from physical coordinates informed by the AT-all SRV2 and GC-core SRV2 (Figure \ref{fig:all_seq_physical_cvs_labels}) and TICA coordinates learned on a combination of all four sequence (Figure \ref{fig:all_seq_tica_cvs}).
-I'm leaning towards presenting the first in the main text and including the other in the SI, but I'm not sure how clear either of these are.
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
+        \caption{5-fold cross validation procedure to select number of SRV coordinates.
+We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
+(probably will be SI, but could include AT-all cross-val in the walkthrough section)}
+        \label{fig:srv_crossval}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_full.PNG}
+        \caption{Full 5' diffusion maps including degenerate third diffusion map mode}
+        \label{fig:GC-end_dmaps_full}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
+        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (Not sure if this adds anything, if anything can grey out underlying free energy landscape and move this to the SI.}
+        \label{fig:shifting_distributions}
+	\end{center}
+\end{figure}
 
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_tica_cvs.PNG}
-        \caption{All sequences plotted on combined TICA axes. Can add similar macrostate distribution as shown in the above figure.}
+        \caption{All sequences plotted on combined TICA axes. Might include in SI or not at all.}
         \label{fig:all_seq_tica_cvs}
@@
 
-\subsubsection{\label{sec:Results}Limitations}
+\subsection{\label{sec:Results}Experimental comparisons across all sequences}
 
+\subsubsection{\label{sec:Results}Thermodynamics of fraying align with lattice model}
+Show similar distribution between lattice model and Tm data\citep{Phys2019}
+
+\subsubsection{\label{sec:Results}Temperature-dependent correlations to slow and fast timescales}
+Find approximate acceleration of model and find best physical comparison to fraying distances
+
+\begin{comment}
+
+Other ways to investigate GC-mix:
 \begin{enumerate}
-	\item Does not explicitly pick up on nucleation/zippering events
-	\item Coarse grain-models makes it difficult to compare timescales.
-Furthermore our implied timescales function more as an average frequency of a process rather than a value that can be directly compared to experiment.
-	\item Limited by memory (number of frames) required to capture equilibrium trajectories at a save rate that enabled adequate resolution for the dynamics of interest.
 
+	\item Track GC-mid trajectories along transition region
+	\item Discuss misaligned configurations here instead of in the GC-core section
+\end{enumerate}
+
+Compare frayed thermo to Paul's lattice model predictions.
+
+\section{\label{sec:Discussion}Discussion}
+
+\citep{Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}
+Long simulations shows that terminal fraying does not converge on the picosecond timescale
+
+\citep{Nonin1995TerminalFraying}
+Most simulations suppress fraying by placing GC pairs at termini.
+
+\citep{Andreatta2006UltrafastHelix}
+- Base opening process is slower than 40 ns but faster than 1ms
+- Open termini state shows new fraying mode of 5ps (I think shows free base motions)
+
+\subsubsection{\label{sec:Results}Timescale comparison with experiment} 
+SRV implied timescales pick up on an average time for the reversible transition A-->B-->A and thus describe the dwell time of kinetically relevant states discussed above.
+A more direct comparison to experiments can be made by examining the transition times of A-->B once the process has been initiated.
+In particular, we were interested in the fast and slow kinetic traces that were picked up by by Sandstead et al.
+To analyze the slow response -- which they attributed to overall strand dissociation -- we tried two methods.
+In the first, we used a variant of core sets where we defined "core regions" by PCCA+ probability of each microstate being clustered into a hybridized or dissociated macrostate \citep{PinamontiTheModels, (milestoning)}.
+We counted and averaged the number of frames between a 99 percent hybridized state and a 99 percent dissociated state, ensuring that the trajectory spend 10 consecutive frames in these cores before and after the transition, respectively.
+In the second method, we perform virtual relaxation experiments on each sequence-specific MSM (cite whoever built this method?).
+By assigning an initial probability distribution that favors microstates far from hybridized or dissociated regions, we can observe how this probability distribution shifts as the system relaxes to equilibrium.
+For each system there is fast response corresponding to the initial population finding its way to a more stable state, and a slow response showing the system reach global equilibrium.
+The fast response is providing information about the timescale of the transition itself and might be the best comparison.
+For GC-end and AT-all, a third signal is present representing equilibration from shifted states.
+There is some variation between sequences, but we find timescales of about 1-3 ns in simulation time.
+This is is about four orders of magnitude faster than the 13-32 (15-60 with updated data) us timescales reported by Sandstead et al.
+
+The fast response Sanstead et.
+al -- attributed to single base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
+Furthermore, all-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen binding, and terminal stacked conformations\citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
+Given that there is only one interaction site parameterized on each 3spn2 , we would not expect to resolve this diverse collection of states and dynamics.
+To obtain a rough metric for terminal base fraying, we ran shorter simulations with a more frequent save rate and calculated an averaged transit time from the closed to open state (SI).
+We observed substantial variability in fraying behavior, but overall we found transit times to be within 5-20 ps in simulation time.
+If we compare these to fast response timescales of Sanstead et al., we find that these rates are again accelerated by about four orders of magnitude.
+
+Previous studies with 3spn2 showed a 200x acceleration when looking at kinetic association rates alone \citep{Hinckley2013AnHybridization}.
+When we examine higher order processes, such at dissociation transit time and single-base fraying, we observed a more substantial acceleration of nearly four order of magnitude.
+This is not surprising given that a smoothened free energy landscape would have a more substantial effect on higher order dynamics than the diffusion-limited pathways that determine association.
+
+\end{comment}
+
+\subsubsection{\label{sec:Results}Code to provide on GitHub} 
+\begin{enumerate}
+    \item{Cross val data}
+    \item lammps run script and each sequence input files
+    \item Featurization with mdtraj (reindexing and sparse saving)
+    \item Dmaps for GC-core and GC-end
+    \item timescales comparison
+    \item MSM construction
 \end{enumerate}
@@
 \begin{comment}
-Classic model has hybridization forming in three distinct step:
-A nucleation step involves bringing two random- coil ssDNAs together to form an intermediate state, in which a few base pairs have formed, and this intermediate pro- ceeds further to form the fully hybridized dsDNA.
-
--
-
-- A smoother free energy surface minimizes the time spent in these intermediate states, and our featurization method
-
-- not capturing zippering
-One limititation of our appraoch that is that processes that are.
-For example, there are eight 3-bp sites on each oligo from which a zippering action coud nucleated.
-Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
-Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
-
-%%%%%% intro notes %%%%%%%
-\citep{Sanstead2018DirectDehybridization} Good intro with many repesimportance of dynamics "rich dynamics even for short oligos"
 

### 2020-09-23 manuscript.tex

@@
 
-\subsection{\label{sec:Results}AT-all}
+\subsection{AT-all}
 
-\subsubsection{\label{sec:Results}SRV optimization and analysis}
+\subsubsection{SRV optimization and analysis}
 
@@
         \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+
         \label{fig:AT-all_in_one}
@@
 For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
-Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{purple}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
+Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{magenta}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
 The number of SRV coordinates and MSM macrostates are varied based on the kinetics and thermodynamics behavior of specific sequences.
@@
 
-\subsection{\label{sec:Results}GC-core}
+\subsection{GC-core}
 
-\subsubsection{\label{sec:Results}Building SRV-MSM}
+\subsubsection{Building SRV-MSM}
 

### 2020-09-26 manuscript.tex
Updates from Overleaf

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[]{TBA}
+\title[]{Investigating DNA d}
 
@@
 
-\textcolor{purple}{Could add more here about all-atom simulations?}
+\textcolor{purple}{Could add more background on here about all-atom simulations, but these rarely explore fully hybridization/dissocition events}
 
@@
 
-\begin{align}
+\begin{align*}
  	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
@@
  	%R_2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
-\end{align}\label{CK1}
+\end{align*}\label{CK1}
 
@@
 
-We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set.
-We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates as a consistent feature set.
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=80mm,
+        scale=0.5]{Figs/figs_0804/all_permute_vamps_grouped.png}
+        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set.
+The inverse distances showed improvement across sequences, and the 55 reduced coordinate set performed about as well as larger features set.}
+        \label{fig:allseq_features_vamp2}
+	\end{center}
+\end{figure}
+
+We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set (\ref{fig:allseq_features_vamp2}).
+We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates to train the model.
+The smaller feature set enabled faster training times and better statistics over permutatable distances, without appearing to suffer a loss in generality or model resolution.
 These features were normalized and passed into sequence-specific SRVs.
@@
 Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that most implied timescales converge at an SRV lag time of 600 ps.
+We found that most implied timescales converge at an SRV lag time of 1.2 ns.
 We kept a looser constraint on the convergence of the leading mode -- corresponding to the overall hybridization/dissociation process -- as it tends to have a longer transit period and becomes more difficult to fulfill Markovian conditions.
@@
 For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-As shown in (Figure \ref{fig:srv_crossval-AT-all}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
+As shown in (Figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
 Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
@@
 The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster shifting or "slithering" processes.
+These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster "slithering" processes.
 Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
@@
 	\begin{center}
-        \includegraphics[width=50mm, 
-        scale=0.5]{Figs/figs_0804/srv_crossval-AT-all.png}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_0804/allseq_cross_val_scores_lag-40.png}
         \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
-For AT-all we observed a spectral gap after the 5th coordinate.
-Cross-validation on other sequences is in the SI.
-(Add AT-all feature scores here as well?)}
-        \label{fig:srv_crossval-AT-all}
+We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
+        \label{fig:allseq_srv_crossval}
 	\end{center}
@@
 
-From this analysis we can determine that the slowest dynamics are fully characterized by the hybridization process and slithering behavior of out of register base pairs.
+From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and slithering behavior of out of register base pairs.
 Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
@@
 This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis.
-Using an MSM lag time of 1.2 ns, we then built a Bayesian MSM to calculate transition probability matrix between each microstate.
-Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represent a collection of metastable structures.
-Previous work have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
+Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
+Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
+Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
 Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
@@
 To visualize these six macrostates, we project the data into the two leading TICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, TICA represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
+Although SRVs outperform these coordinates for the purpose of MSM construction, TICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
 It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
@@
 It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
-This is likely due to 5' ends preferentially stacking on the core duplex compared to 3' ends and/or 3' ends binding few counterions and perturbing the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
+This is might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
 Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state.
@@
 For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
-Although these might not represent fully optimized hyperparameters for each sequence, they preserve Markovian properties of the system and provide adequate resolution to evaluate macrostates \textcolor{magenta}{not sure if this undermines the point I make earlier about building unique microstates and macrostates for each sequence so as to optimize the model}.
-The number of SRV coordinates and MSM macrostates are varied based on the kinetics and thermodynamics behavior of specific sequences.
+For each sequence, these preserve Markovian properties of the system and provide some consistency to compare across models.
+The number of SRV coordinates and MSM macrostates are determined based on cross-validation and the kinetic behavior of specific sequences.
 
@@
 In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
-The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more simplistic than the true mismatch interaction.
+The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more 3 primestic than the true mismatch interaction.
 Despite fewer intact base pairs, we find these structures stabilizing enough to account for the relatively small population of conformations we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
-Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalipic (electrostatic) effects from inert tails \textcolor{Purple}{Message here should be that inert tails, looping, other effects have more substantial contribution than two-state NN calculation alone}.
-However, state populations reveal that the 5' single-shifted is again significantly more stable than the 3'.
+Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalpic effects from inert tails.
+However, state populations reveal that the 5' single-shifted state is again significantly more stable than the 3', indicating that dangling end and inert tail effects outweigh nearest neighbor effects alone.
 Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states \textcolor{magenta}{Need to fix 5' double shifted sequence, make same adjustments as for AT-all above}}
+        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions}
         \label{fig:GC-end_in_one}
@@
 
-To compare these single-shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 5000 conformations from 5' shifted state of both sequences.
+To compare these single-shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 5000 conformations from the 5' single-shifted state of both sequences.
 Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
@@
 This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands.
-In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}.
-Diffusion maps built from samples of the 3' shifted states are also shown in the SI.
+In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}).
+Diffusion maps built from samples of the 3' shifted states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction \textcolor{purple}{state explicitly that C and T have greater excluded volume than G and A in 3spn.2}-- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-This effect may be exaggerated given that C and T base pairs are assigning slightly higher exluded volume radii in 3spn2.
+This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
 The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
@@
 
-\begin{figure}[ht!]
+\begin{figure} %[ht!]
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_tracking_modes.png}
+        \includegraphics[width=120mm]{Figs/figs_0804/GC-core_tracking_modes.png}
         \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.
-Should probably go to SI if we also include GC-core diffusion maps \textcolor{magenta}{These single trajectory figures have proven useful for visualizing SRV dynamics, could be employed for GC-mix comparisons as well}}
+Should probably go to SI if we also include GC-core diffusion maps.}
         \label{fig:GC-core_tracking_modes}
@@
 \subsection{\label{sec:Results}GC-mix}
+
+\begin{figure} %[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-mix_transitions.PNG}
+        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
+        \label{fig:GC-mix_transitions}
+	\end{center}
+\end{figure}
 
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-mix_transitions.PNG}
-        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
-        \label{fig:GC-mix_transitions}
-	\end{center}
-\end{figure}
-
 \subsubsection{\label{sec:Results}Temperature-dependent timescale comparisons}
@@
 Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
+
+\begin{figure}  %[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/fast-slow_end-core.png}
+        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
+Both show an exponential relationship between dissociation and temperature, however fraying appears more linear for GC-core than for GC-end.}
+        \label{fig:relaxation-comparison}
+	\end{center}
+\end{figure}
 
@@
 For GC-core, each base pair and cutoff combination appears linear with temperature, indicating a barrierless and diffusion-driven process.
-As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated 100-300x relative to experiments.
+As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated by about two order of magnitude relative to experiments.
 In contrast, GC-end is distinctly exponential, especially at base pairs closer to the core of the duplex.
@@
 In addition to fraying, we notice that the rate of slithering increases at higher temperature.
-Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex as temperature increases.
-For higher T-jumps, it is possible that a significant portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by slower "looping" G:C bonding that we observed in our diffusion map analysis above.
+Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
+For higher T-jumps, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
 It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
@@
 It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not parameterized based on any dynamical properties \citep{?}.
+Furthermore, more data is needed for explicit ion model comparisons.
 
 %We also did not find available data on other acceleration studies using explicit ions.
+
+\section{\label{sec:conc}Conclusion}
+\subsection{\label{sec:Results}Will look a lot like the last paragraph of the intro/abstract}
 
@@
 	\item Does not explicitly pick up on nucleation/zippering events.
-Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding mode.
+Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding modes.
 Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
 	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
-	\item Limited by memory (number of frames) required to capture equilibrium trajectories at a save rate that enabled adequate resolution for the dynamics of interest.
-	\item A smoother free energy surface minimizes the time spent in these intermediate states, and our featurization method is not sensitive to certain kinds of dynamics
+	\item Limited by memory (number of frames) required to capture equilibrium trajectories and rarity of hybridization/dissociation events
+	\item A smoother free energy surface minimizes the time spent in these intermediate states
 	\item Sensitive to 3spn2 artefacts.
@@
 Although we present slithering as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
-We hope that these data will present a
-
 \end{enumerate}
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/srv_crossval.png}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/allseq_cross_val_scores_lag-40.png}
         \caption{5-fold cross validation procedure to select number of SRV coordinates.
@@
 	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_3prime.PNG}
+        \caption{Full 3' diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct "looping" region for intact GC bonds}
+        \label{fig:GC-end_dmaps_3prime}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
         \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
@@
         \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_physical_cvs_labels.PNG}
-        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes, Not sure how helpful these axes drawings are -- also need to switch x and y here.}
+        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes.}
         \label{fig:all_seq_physical_cvs_labels}
@@
 
-\subsection{\label{sec:Results}Experimental comparisons across all sequences}
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/fast-slow_all-mix.png}
+        \caption{Slow and Fast fits for GC-mix and AT-all, no acceleration values were calculated as there was not the same temperature-dependent data available for these sequences.
+GC-mix again shows barrierles fraying at the terminus, with a noticeable barrier for G:C bonds breaking at the second base pair in.
+AT-all shows a more exponential relationship for terminal fraying, likely because the simulations are run at lower relative temperatures.}
+        \label{fig:relaxation-comparison_mix_and_all}
+	\end{center}
+\end{figure}
 
-\subsubsection{\label{sec:Results}Thermodynamics of fraying align with lattice model}
-Show similar distribution between lattice model and Tm data\citep{Phys2019}
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/relaxation_core_and_end_allcuts.PNG}
+        \caption{Fits across multiple cutoffs and basepairs for GC-core and GC-end fast response. (This is a placeholder until I reformat these to match the main text).}
+        \label{fig:relaxation_core_and_end_allcuts}
+	\end{center}
+\end{figure}
 
-\subsubsection{\label{sec:Results}Temperature-dependent correlations to slow and fast timescales}
-Find approximate acceleration of model and find best physical comparison to fraying distances
+\subsubsection{\label{sec:Results}Code to provide on GitHub} 
+\begin{enumerate}
+    \item Cross val data
+    \item lammps run script and each sequence input files
+    \item Featurization with mdtraj (reindexing and sparse saving)
+    \item Dmaps for GC-core and GC-end
+    \item timescales comparison
+    \item SRV cross valss
+    \item MSM construction
+\end{enumerate}
 
@@
 
-\subsubsection{\label{sec:Results}Timescale comparison with experiment} 
-SRV implied timescales pick up on an average time for the reversible transition A-->B-->A and thus describe the dwell time of kinetically relevant states discussed above.
-A more direct comparison to experiments can be made by examining the transition times of A-->B once the process has been initiated.
-In particular, we were interested in the fast and slow kinetic traces that were picked up by by Sandstead et al.
-To analyze the slow response -- which they attributed to overall strand dissociation -- we tried two methods.
-In the first, we used a variant of core sets where we defined "core regions" by PCCA+ probability of each microstate being clustered into a hybridized or dissociated macrostate \citep{PinamontiTheModels, (milestoning)}.
-We counted and averaged the number of frames between a 99 percent hybridized state and a 99 percent dissociated state, ensuring that the trajectory spend 10 consecutive frames in these cores before and after the transition, respectively.
-In the second method, we perform virtual relaxation experiments on each sequence-specific MSM (cite whoever built this method?).
-By assigning an initial probability distribution that favors microstates far from hybridized or dissociated regions, we can observe how this probability distribution shifts as the system relaxes to equilibrium.
-For each system there is fast response corresponding to the initial population finding its way to a more stable state, and a slow response showing the system reach global equilibrium.
-The fast response is providing information about the timescale of the transition itself and might be the best comparison.
-For GC-end and AT-all, a third signal is present representing equilibration from shifted states.
-There is some variation between sequences, but we find timescales of about 1-3 ns in simulation time.
-This is is about four orders of magnitude faster than the 13-32 (15-60 with updated data) us timescales reported by Sandstead et al.
-
-The fast response Sanstead et.
-al -- attributed to single base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
-Furthermore, all-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen binding, and terminal stacked conformations\citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
-Given that there is only one interaction site parameterized on each 3spn2 , we would not expect to resolve this diverse collection of states and dynamics.
-To obtain a rough metric for terminal base fraying, we ran shorter simulations with a more frequent save rate and calculated an averaged transit time from the closed to open state (SI).
-We observed substantial variability in fraying behavior, but overall we found transit times to be within 5-20 ps in simulation time.
-If we compare these to fast response timescales of Sanstead et al., we find that these rates are again accelerated by about four orders of magnitude.
-
-Previous studies with 3spn2 showed a 200x acceleration when looking at kinetic association rates alone \citep{Hinckley2013AnHybridization}.
-When we examine higher order processes, such at dissociation transit time and single-base fraying, we observed a more substantial acceleration of nearly four order of magnitude.
-This is not surprising given that a smoothened free energy landscape would have a more substantial effect on higher order dynamics than the diffusion-limited pathways that determine association.
-
 \end{comment}
-
-\subsubsection{\label{sec:Results}Code to provide on GitHub} 
-\begin{enumerate}
-    \item{Cross val data}
-    \item lammps run script and each sequence input files
-    \item Featurization with mdtraj (reindexing and sparse saving)
-    \item Dmaps for GC-core and GC-end
-    \item timescales comparison
-    \item MSM construction
-\end{enumerate}
-
-\section{\label{sec:conc}Conclusion}
-\subsection{\label{sec:Results}Will look a lot like the last paragraph of the intro / the abstract}
 

### 2020-09-26 manuscript.tex
Updates from Overleaf

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[]{Investigating DNA d}
+\title[]{Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides}
 
@@
 As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data.
-Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence attributing fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
+Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
 While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation.
@@
 
-	\item Track GC-mid trajectories along transition region
+	\item Track GC-mix trajectories along transition region
 	\item Discuss misaligned configurations here instead of in the GC-core section

### 2020-10-01 manuscript.tex
Updates from Overleaf

@@
 
-%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
 \newcommand*{\rood}[1]{#1}
-%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
 \newcommand*{\blauw}[1]{#1}
-%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
 \newcommand*{\groen}[1]{#1}
-%\newcommand*{\blauwr}[1]{\textcolor{blue}{#1}}
 \newcommand*{\blauwr}[1]{#1}
@@
 Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
-From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}.
+From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}.
 Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}.
@@
 Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
-Sandstead et al.
+Sanstead et al.
 highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}.
-In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.
+In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.
 
@@
 Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
 Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.
 
-\textcolor{purple}{Could add more background on here about all-atom simulations, but these rarely explore fully hybridization/dissocition events}
-
-In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
-We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
+In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn.2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
+Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
 MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
@@
 By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process.
-Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
-Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models.
+Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models.
 Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs.
 Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
-Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics \textcolor{purple}{Not sure how much to say about this here, but I think these experiments should be mentioned in the intro}.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.
+Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, meta-stable structures, and relative timescales.
 
@@
 
-\subsection{\label{sec:methods}3spn2 Model}
+\subsection{\label{sec:methods}3spn.2 Model}
 
-\textcolor{purple}{Is some additional information warranted on the model itself before going into this specific setup?}
+The 3 site per nucleotide (3spn.2) coarse grain model was designed to accelerate DNA computation relative to all atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
+Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}.
+3spn.2 has been implemented to study DNA pack in viral capsids, protein-DNA binding, and nucleosome unwrapping, and recently served as a basis for the coarser 1cpn model \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping, Lequieu20191CPN:Chromatin}.
+The model is amenable to an implicit or explicit ion environment -- we chose the latter to include Mg2+ effects from experiment.
+We achieved better sampling of hybridization events with explicit ions, despite a more demanding computational cost.
 
@@
 
-We initialized four sequences previously investigated by Sandstead et al.
+We initialized four sequences previously investigated by Sanstead et al.
 -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
@@
 We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
-We set the simulation temperature was set about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states.
+We set the simulation about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states.
 We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time.
@@
 All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
-Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together.
+Based on the self-complementary nature of each sequence -- meaning that both the sense and anti-sense strands are identical to each other -- we averaged permutable distances (45 pairs in total) together.
 This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
@@
 In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
-The score $R_2$ can then be found for $\chi$ by applying the VAMP principle with cross-validation.
+The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
 
@@
 	\\
- 	R_2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
- 	%R_2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
+ 	VAMP2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
+ 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
 \end{align*}\label{CK1}
@@
 SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data.
-This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data.
+This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
 The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
@@
 \subsubsection{SRV optimization and analysis}
+
+%Throughout this manuscript, we adopt "slithering" from previous work to describe dynamics between out-of-register states. We describe the metastable states along these pathway as "shifted."
 
@@
 It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
-This is might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
-Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state.
-Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state.
-Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average "dwell" time between transition into and out of these states.
-The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state.
-In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having two less intact base pairs.
-This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below.
+These effects might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
+Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that conformations in the 5' shifted state are more energetically favorable than those in the 3' shifted state.
+Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics}.
+Therefore it makes sense that our PCCA+ free energies are all higher than thermodynamics prediction, due to some destabilizing effect of the inert tails.
+Notably, the 3' dangling states have a 6-8 kJ/mol higher free energy compared to theory, whereas the 5' dangling states deviate by less than 3 kJ/mol.
+In fact, the free energy of the 5'-double shifted state is slightly lower than that of the 3'-single shifted state, despite the former having two fewer intact base pairs.
 The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
 This might be attributed to the asymptotic effects of inert tail destabilization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
-Accordingly, the longer inert tails in the double-shifted states have less of an effect on relative 5' vs.
-3' stability compared to the shorter tails in the single-shifted states.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.png}
+        \caption{a) AT-all free energies calculated from nearest neighbor data and from MSM PCCA+ probabilities b) TICA projections of free energies and PCCA+ state clusterings c) Transitions probabilities between states within the 1.2 ns lag time}
 
@@
 Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states.
-These asymmetric kinetic effects may have some role in the substantially lower state populations we see above.
+These asymmetric kinetic effects, in addition to NN and inert tail contributions described above, may further lower the 3' shifted states populations relative to 5'.
 From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state.
@@
 
-Next we examine the GC-end sequence, which add GC caps to the same repetitive AT motifs.
+Next we examine the GC-end sequence, which add G:C caps to the same repetitive AT motifs.
 Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.
@@
 Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
-In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
-The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely much more 3 primestic than the true mismatch interaction.
+In the context of the 3spn.2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
 Despite fewer intact base pairs, we find these structures stabilizing enough to account for the relatively small population of conformations we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
-Contrary to AT-all thermodynamics, nearest neighbor calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalpic effects from inert tails.
+Contrary to AT-all thermodynamics, NN calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalpic effects from inert tails.
 However, state populations reveal that the 5' single-shifted state is again significantly more stable than the 3', indicating that dangling end and inert tail effects outweigh nearest neighbor effects alone.
+As in the AT-all case, we observe a smaller deviation between 5' shifted theoretical and simulated free energies (6.5-8.7 kJ/mol) compared to the 3' case (13 kJ/mole).
 Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.PNG}
-        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions}
+        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.png}
+        \caption{GC-end free energies calculated from nearest neighbor data and from MSM PCCA+ probabilities b) TICA projections of free energies and PCCA+ state clusterings c) Transitions probabilities between states within the 1.2 ns lag time}
         \label{fig:GC-end_in_one}
@@
 Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either identical end of the shifted conformation.
+We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
 This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands.
@@
 These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
+
+%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3spn2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
 
@@
 In T-jump, FTIR, and 2D IR experiments, the GC-end sequence was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
-Furthermore, there was a distinct kinetic response -- particularly at lower temperature -- that did not clearly match the fraying processes under investigation.
-\citep{Sanstead2018DirectDehybridization}.
-Given that this signal contained some A:T and G:C character and was slower than single base pair fraying, it is possible that it represented some shifted or related looping dynamics.
+Furthermore, GC-ends revealed a distinct T-jump response containing evidence for the loss of G:C and A:T base pairing prior to full dissociation of the duplex \citep{Sanstead2018DirectDehybridization}.
+Given that this response was slower than the A:T fraying response observed in other sequences, it is possible that it corresponds to shifting or related looping dynamics.
 This cannot be verified given that the signal was recorded in a congested spectroscopic range and could be hard to differentiate from aligned G:C fraying.
@@
 
-When examining these same four sequences using T-jump IR and 2D IR spectroscopy, Sandstead et al.
+When examining these same four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
 found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
@@
 During hybridization, we observe the formations of some key base pair contacts before the full duplex forms.
-Although it is statistically more likely for contact to form at the end of the duplex, we observe first contact at interior base pairs as well.
+We observe that first contacts tend to involve one of the G:C bonds, which is not surprising that these are more evenly spaced out than in GC-core and GC-end.
 This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition}.
@@
 We measure the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
-We find the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the known thermodynamic influence on duplex stability.
+We find the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
 Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
@@
 The fast response -- which Sanstead et.
-al attributed to single base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
+al attributed to base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
 Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
@@
 Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
-For higher T-jumps, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
+For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
 It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
@@
 The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
-Previous work has shows that kinetic association rates were accelerated by about 200x relative to experiment \citep{Hinckley2013AnHybridization}.
-This seems to be on the same order of magnitude as frayed acceleration, but dissociation is noticeably less accelerated when compared to experiment.
+Previous work has shows that kinetic association rates were accelerated by about one to two orders of magnitude relative to experiment \citep{Hinckley2013AnHybridization}.
 It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not parameterized based on any dynamical properties \citep{?}.
@@
         \label{fig:GC-end_dmaps_3prime}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
-        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (Not sure if this adds anything, if anything can grey out underlying free energy landscape and move this to the SI.}
-        \label{fig:shifting_distributions}
 	\end{center}

### 2020-10-01 manuscript.tex
Updates from Overleaf

@@
 
-%Throughout this manuscript, we adopt "slithering" from previous work to describe dynamics between out-of-register states. We describe the metastable states along these pathway as "shifted."
-
 In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes.
@@
 For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal).
-Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the dynamics of the overall hybridization and dissociation process.
-The next four SRV components all show a relatively high correlation along different offset diagonals.
-These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster "slithering" processes.
-Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
+Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
+The next four SRV components all show a relatively high correlation along offset diagonals.
+These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
+Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
 
@@
 
-From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and slithering behavior of out of register base pairs.
+From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and shifting behavior of out of register base pairs.
 Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
@@
 All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation.
+Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
 In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state.
@@
 Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.
-These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some slithering dynamics.
+These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some shifting dynamics.
 When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in SI Figure \ref{fig:GC-end_dynamic}) by a distinct spectral gap.
@@
 
-The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end.
+The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end.
 Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis.
@@
 This behavior makes sense given the thermodynamic barrier to G:C fraying, but it also deviates the GC-end experimental linear fit.
-In addition to fraying, we notice that the rate of slithering increases at higher temperature.
+In addition to fraying, we notice that the rate of shifting increases at higher temperature.
 Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
@@
 Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding modes.
-Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate.
+Furthermore these zippering dynamics occur significantly faster than shifiting or GC-core fraying and cannot be captured via our save rate.
 	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
@@
 	\item Sensitive to 3spn2 artefacts.
-Slithering may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}.
+Shifting may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}.
 Base-pair mismatch interactions may be somewhat unphysical and alter state populations.
-Although we present slithering as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
+Although we present shifting as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
 \end{enumerate}

### 2020-11-23 manuscript.tex
Andy markup
Updates from Overleaf

@@
 \SectionNumbersOn
+
+%include hyperlinks to figures
+\documentclass{article}
+\usepackage{graphicx}
+\usepackage[%  
+    colorlinks=false,
+    pdfborder={0 0 0},
+    linkcolor=blue
+]{hyperref}
 
@@
 
+Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood.
+In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement.
+We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Models (MSMs) construction.
+Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides.
+For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states.
+In contrast, G:C pairs near the center of the duplex induce more rapid fraying dynamics.
+In both cases, hybridization/dissociation mechanisms deviate from an “all-or-nothing” model.
+Our computational predictions are in excellent accord with experiment, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
+
+% revise "excellent accord" above?
+
 \noindent \url{https://www.overleaf.com/project/5e9e5110c524b8000192c548}
@@
 
-\begin{comment}
-%%%%%% intro notes %%%%%%%
-\citep{Sanstead2018DirectDehybridization} Good intro with refs and focus on importance of dynamics "rich dynamics even for short oligos" ...
-Many of the most important functions performed by DNA are highly dynamic, whether in the the biological roles of replication transcription, and gene regulation, or in the expanding field of DNA-based nanotechnology
-\end{comment}
-
 \section{\label{sec:intro}Introduction}
@@
 Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
-From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}.
-Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}.
-Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions.
-Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
-Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition.
-Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}.
-Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates.
-The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}).
-Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
+From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}.
+DNA dynamical behavior has become increasingly important in these technologies and a point of focus in fundamental biological processes such as transcription and gene regulation.
+\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}.
+Our understanding of DNA hybridization -- the formation of a duplex from two amenable single strands -- has been integral to these developments.
+The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for many mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
+Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}.
+Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
+Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
+
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags altering strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
+Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
+Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.
+Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
+Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.
+
+Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}.
+Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates.
+For example, metastable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, , Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}).
+The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequences dependent.
+Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
 Sanstead et al.
-highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}.
-In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.
+highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}.
+Together, these dynamics represent the most substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
 
-Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
-More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}.
-Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.
-Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
-Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
-Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.
-
-In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn.2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
+We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
 We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
@@
 We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
-Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed.
+Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed.
 We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
 
-We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
-By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process.
+We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
+By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
 Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models.
-Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs.
-Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
-Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, meta-stable structures, and relative timescales.
+Additionally, we can compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs.
+Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
+Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+
+%%%%%%%% Intro comments %%%%%%%%%%%%%%%%%
+
+%Progress in DNA science and engineering has been bolstered by a well-established understanding of sequence-dependent thermodynamics.
+Thousands of DNA and DNA-complex crystal structures have been captured, and the equilibrium structure of every tetranucleotide motif were simulated in a distributed computational effort \citep{Manuscript2014The2012}.
+Hybridization thermodynamics can predicted with high fidelity by nearest-neighbor stacking, and these calculations have been experimentally verified to account for mismatched pairs, dangling ends, and other non-native motifs \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
+
+% Secondary DNA structures such as hairpins and quadruplexes have also been studied in depth and leverage for nanotechnology applications.
+
+%Furthermore several computational models have been paramaterized to capture these thermodynamic characteristics.
+
+%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}.
+Certain dynamical models for hybridization -- such the nucleation-zipper mechanism and fraying-peeling mechanism for dissociation -- have been well-established for decades (\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition}).
+Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
+Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
+
+%Still, rich dynamics have been observed short nucleotides \citep{Sanstead2018DirectDehybridization} and the sequence-dependent effects on these dynamics are not well understood.
+
+%Progress has also been made in modeling kinetics for certain dynamic DNA systems, such as toehold exchanges and optical barcodes \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting}.
+Sequence effects on association and dissociation rates were predicted using combined experimental and machine learning approaches, however higher order dynamics are less well understood \citep{Zhang2018PredictingSequence}.
+Hairpin \citep{Woodside2006DirectAcid} and in duplex hybridization \citep{yer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization} studies have revealed sequence-dependent effects and varied differential deviations from "all or nothing" behavior.
+In temperature-jump spectroscopy, G:C placement alters the slow kinetic response attributed to dissociation and a fast response likely corresponding to terminal A:T fraying.
+Experimentally, however, it is difficult to pick up on dynamics that have similar spectroscopic signatures and to tease apart temporally close responses that might be combined into one stretched exponential signal.
+
+% include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior.
+The slow response was attributed to complete strand dissociation, while the fast response described base fraying.
+We perform longer equilibrium sampling near Tm to get a sense of sequence equilibrium behavior, as well shorter simulations at a range of higher temperatures to replicate T-jump data.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
 
@@
 
-\subsection{\label{sec:methods}3spn.2 Model}
+\subsection{\label{sec:methods}Simulation set up}
 
-The 3 site per nucleotide (3spn.2) coarse grain model was designed to accelerate DNA computation relative to all atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
+The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
 Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
 The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}.
-3spn.2 has been implemented to study DNA pack in viral capsids, protein-DNA binding, and nucleosome unwrapping, and recently served as a basis for the coarser 1cpn model \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping, Lequieu20191CPN:Chromatin}.
-The model is amenable to an implicit or explicit ion environment -- we chose the latter to include Mg2+ effects from experiment.
-We achieved better sampling of hybridization events with explicit ions, despite a more demanding computational cost.
+3spn2 has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
+\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.
 
-\subsection{\label{sec:methods}Simulation set up}
-
-We initialized four sequences previously investigated by Sanstead et al.
--- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}.
-We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
-In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A).
-This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM.
-We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions.
-We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}.
-We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
-We set the simulation about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states.
-We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time.
-Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes.
-For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation.
-An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state.
-In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.
+We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}.
+We specified a 240 mM implicit salt concentration and used a A Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
+We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
+Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states.
+We use a 20 fs time step and $13\times10^{9}$ steps in each simulation, saving frames every 100 ps.
+For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors for each simulations.
+Half of runs were initialized from the hybridized state and half from a dissociated state.
+The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence.
+Depending on sequence, we observed between 40-100 hybridizing and dissociating transitions across all runs.
 
@@
 All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
-Based on the self-complementary nature of each sequence -- meaning that both the sense and anti-sense strands are identical to each other -- we averaged permutable distances (45 pairs in total) together.
-This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
+Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other.
+To account for this, we averaged permutable distances (45 pairs in total) together.
+This permutation reduction follows a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
 The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}.
-The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models.
+The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=80mm,
+        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
+        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set.
+The inverse distances showed improvement across sequences, and the 55-dimension symmetrized coordinate set performed about as well as larger features set.}
+        \label{fig:allseq_features_vamp2}
+	\end{center}
+\end{figure}
+
+We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances and found small differences between the three that varied by sequences, but overall we did not observe a loss in kinetic information when using the symmetrized feature set (\ref{fig:allseq_features_vamp2}).
+We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal symmetrized coordinates to train the model.
+The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
+These features were normalized and passed into sequence-specific SRVs.
+
+\subsection{\label{sec:methods}SRVs}
+
+SRVs were first developed by Chen et al.
+as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}.
+The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis},and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}.
+The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}.
+The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}.
+SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
+This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
+The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
+For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
+
+Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
+In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
+We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
+In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
+We used a batch size of 50000 and ran each model for a total of 20 training epochs.
+We used two hidden layers and set the size of each layer to 100.
+For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
+SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
+SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
+
+% might want to include these training params in the SI?
+
+\subsection{\label{sec:methods}SRV-MSMs}
+
+MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}.
+The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time.
+The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
+To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}.
+Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis.
+To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
+In a similar approach to Sidky et al.
+we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}.
+The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
+The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.
+Additional step-by-step details on SRV-MSM construction for each sequence are provided in the supplemental information and on Github (https://github.com/mrjoness/...).
+
+%%%%%%%% Methods comments %%%%%%%%%%%%%%%%%
+
+%We found close agreement with explicit ion simulations performed with 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} (supplemental) but 7x slower simulation speed made adequate sampling more difficult to achieve under these conditions.
+
+%We initialized explicit ions such that 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
+We used a periodic box size of 7.774 nm and an effective oligo concentration of 7 mM.
+We used an Ewald summation to calculate long range Coulombic interaction between DNA and ions using a real space cutoff of 2 nm.
+
+% A Debye-Huckle screening potential was used to monitor account for ion concentration .
+
+\begin{comment} 
+% SI equations
 In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
@@
 
-% for more general case with  \chi_0, \chi_1, but in practice this is not used
-\begin{comment} 
-\begin{align}
- 	\mathscr{C}_{00}=&\Expect{\chi_0(x_t)\chi_0(x_t)^\intercal}_t\\
- 	\mathscr{C}_{01}=&\Expect{\chi_0(x_t)\chi_1(x_{t+\tau})^\intercal}_t\\
- 	\mathscr{C}_{11}=&\Expect{\chi_1(x_{t+\tau})\chi_1(x_{t+\tau})^\intercal}_{t+\tau}\\
-	\\
- 	R_2[\chi_0, \chi_1]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2\\
- 	R_2^{val}[\chi_0, \chi_1]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1\\
-\end{align}\label{CK2}
 \end{comment}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=80mm,
-        scale=0.5]{Figs/figs_0804/all_permute_vamps_grouped.png}
-        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set.
-The inverse distances showed improvement across sequences, and the 55 reduced coordinate set performed about as well as larger features set.}
-        \label{fig:allseq_features_vamp2}
-	\end{center}
-\end{figure}
-
-We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set (\ref{fig:allseq_features_vamp2}).
-We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates to train the model.
-The smaller feature set enabled faster training times and better statistics over permutatable distances, without appearing to suffer a loss in generality or model resolution.
-These features were normalized and passed into sequence-specific SRVs.
-
-% include covariance, vamp-2  and validation equations, lag time equation from vampnet paper mardt2018
-
-\subsubsection{\label{sec:methods}SRVs}
-
-SRVs were first developed by Chen et al.
-as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}.
-The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features.
-This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.
-The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.
-Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}.
-SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data.
-This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
-The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
-For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
-
-Using optimized hyperparameters and featurized trajectory data, we transform 55 reciprocal pairwise distances into a low dimensional SRV basis set.
-In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular, we looked for convergence in the vamp-2 score and inconsistency between cross validation scores -- suggesting that the model may be overfitting on artifacts in the training data.
-We used a batch size of 50000 and ran each model for a total of 30 training epochs.
-We used two hidden layers and set the size of each layer to 100.
-For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
-SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
-
-\subsubsection{\label{sec:methods}SRV-MSMs}
-
-MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.
-The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time.
-The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
-To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}.
-This becomes crucial in our system given the large difference in timescales between leading modes.
-Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis.
-To prove this, SRV-MSM VAMP-2 scores were shown to be consistently higher than MSMs constructed from TICA coorinates (TICA-MSMs) \citep{Sidky}.
-Furthermore, SRV-MSM implied timescales converge faster than TICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.
-To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
-After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.
-The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
-The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.
 
@@
 
-\subsection{AT-all}
+% Include analysis on overall hybridization rates here or at the end?
+% Show difference between between full hybridization and partial events (GC-end and AT-all)
 
-\subsubsection{SRV optimization and analysis}
-
-In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes.
-For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step-by-step.
-Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that most implied timescales converge at an SRV lag time of 1.2 ns.
-We kept a looser constraint on the convergence of the leading mode -- corresponding to the overall hybridization/dissociation process -- as it tends to have a longer transit period and becomes more difficult to fulfill Markovian conditions.
-It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency.
-Next we selected an optimal number of SRV components to include in our analysis.
-After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
-It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
-For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-As shown in (Figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
-Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
-The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by (or effective on) each process.
-For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal).
-Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
-The next four SRV components all show a relatively high correlation along offset diagonals.
-These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
-Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.
+\subsection{SRV-MSMs provide better resolution than tICA-MSMs}
 
@@
 	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_0804/allseq_cross_val_scores_lag-40.png}
-        \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
-        \label{fig:allseq_srv_crossval}
-	\end{center}
-\end{figure}
-
-\subsubsection{SRV-MSM construction and optimization}
-
-From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and shifting behavior of out of register base pairs.
-Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
-Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
-We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
-Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
-In Figure \ref{fig:AT-all_dynamic} we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes.
-This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis.
-Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
-Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
-Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
-Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
-For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
-To visualize these six macrostates, we project the data into the two leading TICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, TICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
-It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states.
-After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states.
-In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_dynamic.png}
-        \caption{AT-all SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns. Leading slow mode correlations to all 100 intermolecular distances between oligos.}
+        \includegraphics[width=400, scale=1]{Figs/figs_imp/AT-all_implied_timescales.png}
+        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
         \label{fig:AT-all_dynamic}
@@
 
-\subsubsection{Implied timescales for TICA-MSMs for }
+As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach.
+Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model \citep{Sidky}.
+Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
+Similar plots for the other sequences are included in the SI.
+We also note that the infrequency of transitions relative to the individual trajectory length leads to slower convergence of the leading mode.
+Because this mode corresponds the overall hybridization and dissociation process, we found that it displayed similar behavior across all sequences and that higher order modes were more informative for lag time selection.
 
-In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings.
-These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states.
-This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019, Phys2014}.
-We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states.
-This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling 5' ends have a consistently stabilizing effect, and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
-It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
-These effects might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}.
-Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that conformations in the 5' shifted state are more energetically favorable than those in the 3' shifted state.
-Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics}.
-Therefore it makes sense that our PCCA+ free energies are all higher than thermodynamics prediction, due to some destabilizing effect of the inert tails.
-Notably, the 3' dangling states have a 6-8 kJ/mol higher free energy compared to theory, whereas the 5' dangling states deviate by less than 3 kJ/mol.
-In fact, the free energy of the 5'-double shifted state is slightly lower than that of the 3'-single shifted state, despite the former having two fewer intact base pairs.
-The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states.
-This might be attributed to the asymptotic effects of inert tail destabilization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.
+%addition to timescale comparison, we found that tICA-MSMs often produced inconsistent PCCA+ clusterings between runs and tended to be less physically meaningfully
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.png}
-        \caption{a) AT-all free energies calculated from nearest neighbor data and from MSM PCCA+ probabilities b) TICA projections of free energies and PCCA+ state clusterings c) Transitions probabilities between states within the 1.2 ns lag time}
-
-        \label{fig:AT-all_in_one}
+        \includegraphics[width=160mm, 
+        scale=1]{Figs/figs_imp/allseq_states.png}
+        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates.
+The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
+        \label{fig:allseq_table}
 	\end{center}
@@
 
-\subsubsection{Transition probabilities between states}
-
-Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
-Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns).
-All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
-In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state.
-Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states.
-These asymmetric kinetic effects, in addition to NN and inert tail contributions described above, may further lower the 3' shifted states populations relative to 5'.
-From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state.
-We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.
-
-For all proceeding sequences, we use the same SRV and MSM lag times, number of microstates, and cross-validation procedures as in the AT-all case.
-For each sequence, these preserve Markovian properties of the system and provide some consistency to compare across models.
-The number of SRV coordinates and MSM macrostates are determined based on cross-validation and the kinetic behavior of specific sequences.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{GC-end}
-
-\subsubsection{SRV-MSM construction and states}
-
-Next we examine the GC-end sequence, which add G:C caps to the same repetitive AT motifs.
-Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.
-These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some shifting dynamics.
-When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in SI Figure \ref{fig:GC-end_dynamic}) by a distinct spectral gap.
-We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score (SI Figure \ref{fig:srv_crossval}).
-The clustered macrostates show a similar distribution where shifted populations are notably lower.
-The 3' double-shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes).
-Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches replace two intact WC bonds in the corresponding AT-all states.
-This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and shorter timescales.
-
-\begin{figure}[ht!]
+\begin{figure}[hb!]
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dynamic.png}
-        \caption{(SI) GC-end SRV-MSM timescales convergence and implied timescales using a lag time of 1.2 ns. Leading slow mode correlations to all 100 intermolecular distances between oligos.}
-        \label{fig:GC-end_dynamic}
+        \includegraphics[width=160mm, 
+        scale=1]{Figs/figs_imp/allseq_transitions.png}
+        \caption{Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions.
+Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each state.
+Arrows indicate the probability of transitioning out of the state within the lag time.
+Circle areas are proportional to state stability.}
+        \label{fig:allseq_transition}
 	\end{center}
@@
 
-\subsubsection{Comparison between GC-end and AT-all shifted states}
+\subsection{All SRV-MSMs}
 
-For our analysis, we consider C:T and G:A base pairs in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends and two inert tails.
-Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
-In the context of the 3spn.2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
-The only direct interaction between non-WC basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
-Despite fewer intact base pairs, we find these structures stabilizing enough to account for the relatively small population of conformations we observed in these shifted states (Figure \ref{fig:GC-end_in_one}).
-Contrary to AT-all thermodynamics, NN calculations predict the 3' shifted states to be slightly more stable than 5' shifted states, again without taking into account entropic or enthalpic effects from inert tails.
-However, state populations reveal that the 5' single-shifted state is again significantly more stable than the 3', indicating that dangling end and inert tail effects outweigh nearest neighbor effects alone.
-As in the AT-all case, we observe a smaller deviation between 5' shifted theoretical and simulated free energies (6.5-8.7 kJ/mol) compared to the 3' case (13 kJ/mole).
-Interestingly, we found that 5' single-shifted stability might be elevated by a substantial portion of conformations retaining one intact GC base pair.
-Visualizations reveal that the shifted oligos -- particularly in the 5' shifted state -- have a tendency to sacrifice some helical conformational entropy in a way that facilitates G:C termini bonding even when all available A:T bonds are formed out-of-register.
+Following the analysis pipeline described above, we generate SRV-MSMs for each sequence.
+We identified seven kinetically relevant states that are captured within the resolution of the model.
+These include the fully hybridized state (H) in which all native base pairings are intact, the fully dissociated state (D), four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, and a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound.
+Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs.
+3') and the number of shifted motifs relative the native state (1 vs.
+2).
+GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states.
+Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}.
+We also show that PCCA state assignments are capturing free energy minima in tICA space as independent states.
 
-\begin{figure}[ht!]
+% rewrite this into one paragraph
+\subsection{Shifted state stability}
+
+The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}.
+The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
+Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails.
+Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
+It is informative to compare our MSM predictions against NN calculations to understand how inert tail or kinetic effects cause deviation from these predictions.
+For AT-all, NN calculations (figure \ref{fig:NN_table}) predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
+We find qualitative agreement to our PCCA+ free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone.
+For GC-end, we consider C:T and G:A mismatches in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends.
+Based on this treatment, NN calculations yield higher overall free energies (due to fewer native base pair contacts), but, contrary to PCCA+ results, predict that 3' out-of-register states to be more stable than 5' states.
+This leads us to believe that 5' vs.
+3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.
+
+% and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
+It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
+These effects might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure \citep{Doktycz1990ThermodynamicATGC}.
+
+% Although NN calculations predict 5' vs.
+3' stabilities for AT-all, we observe that GC-end 5S1 state remains more populated than the 3S1 state contrary to NN predictions.
+Indeed, we do not observe the formation of a stable 3S2 state for GC-end, further indicating that these states 3' states are unfavorable.
+This leads us to believe that 5' vs.
+3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.
+
+\begin{figure}[ht!] % right place for NN table? could go to SI
 	\begin{center} 
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_in_one.png}
-        \caption{GC-end free energies calculated from nearest neighbor data and from MSM PCCA+ probabilities b) TICA projections of free energies and PCCA+ state clusterings c) Transitions probabilities between states within the 1.2 ns lag time}
-        \label{fig:GC-end_in_one}
+        \includegraphics[width=100mm, scale=1]{Figs/figs_imp/shifted_FE_table.png}
+        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations.
+All free energies are normalized such that that the native hybridized state is set to zero.
+Calculations included dangling end contributions but do not take inert tails effects.}
+
+        \label{fig:NN_table}
 	\end{center}
@@
 
-\subsubsection{Diffusion maps show sequence differences in shifted states}
+%if AT-all has highest frequency and GC-end has lowest then can describe role of metastable states as increasing rate for the former but decreasing rate for the later (for kinetic traps that do not terminate in native hybridization and delay the process as a whole).
 
-To compare these single-shifted 5' prime states with the corresponding AT-all state, we employ diffusion maps built on an equal sampling of 5000 conformations from the 5' single-shifted state of both sequences.
-Diffusion maps generate a low dimensional embedding based on some metric for diffusive distance and are well-suited to find subtle structural differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+\subsection{Shifting dynamics}
+%\subsubsection{Transition probabilities between states}
+
+Beyond state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
+Figure \ref{fig:allseq_transition}!!!change ref!!!
+shows the probability of moving from one state to another within the MSM lag time.
+For AT-all, we observe approximately equal probability of transitioning from D to any other state, indicating that this is primarily a diffusion-driven process.
+Once a transition has been made, however, the 5' vs.
+3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
+Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
+Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
+In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than 3S2 $\rightarrow$ 3S1.
+Indeed for GC-end, we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
+The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D.
+This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.
+
+%not sure to mention "inchworm" and "pseudoknot" since these were not directly observed/not in the scope of this work
+
+Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
+A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of H $\rightarrow$ D pathways pass through some out-of-register state.
+While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states.
+These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
+Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and subtle kinetic traces.
+To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}.
+Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}.
+Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
+It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs.
+This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
+
+\subsection{Shifting experimental comparisons}
+% how to discuss this without first introducing the fast response?
+
+In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses.
+For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale.
+We expect that fraying in out-of-register conformations would occur at different rates than intact fraying.
+Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contribute to the ensemble.
+Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, sliding mechanisms during the dissociation process, or some combination of the two.
+In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.
+We interrogate these differences by examining high temperature simulations later in this manuscript.
+
+%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
+%Although reproducing these stretching responses is not within the scope of this analysis, we take these
+
+%We emphasize that this does necessarily reflect shifting relaxation timescales themselves but more likely changes in the fast responses -- e.g.
+fraying out-of-register -- within the collection of metastable states.
+
+\subsection{GC-core fraying} % not sure if I should make this its own section?
+
+The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end.
+Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4).
+Previous studies suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
+Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived metastable state.
+This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
+We observe these events to occur with equal probability on either permutable side of the helix.
+The hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
+This is expected considering that this state is more accessible from an already bound helix.
+Moreover, once oligos are in F4, they are over 10x more likely to return to H than to D.
+Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed it.
+On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
+
+% try to split this up between an experimental section (all SRV-MSMs) and more detail on diffusion maps
+\subsection{GC-core Comparison to Experiment}
+
+% discuss computational comparisons needs work
+Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019}.
+Araque et al.
+defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}.
+When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
+As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
+While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
+Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
+We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
+Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
+
+%Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.
+
+\subsection{GC-mix displays more canonical hybridization and dissociation}
+Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
+As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data.
+Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
+While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation.
+On the other hand, the F4 state in GC-core had a substantial probability of fully dissociated rather reforming an intact helix.
+Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
+
+Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
+For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
+During hybridization, we observe the formations of some key base pair contacts before the full duplex forms.
+We observe that first contacts tend to involve one of the G:C bonds, which is not surprising that these are more evenly spaced out than in GC-core and GC-end.
+This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}.
+We obverse dissociation events proceeded by two base pair fraying on one side of the duplex followed by more rapid dissociation of the central base pairs.
+There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs.
+In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes.
+As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}.
+We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
+
+\begin{figure} %[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
+        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
+        \label{fig:GC-mix_transitions}
+	\end{center}
+\end{figure}
+
+%%%%%%% start of "interesting phenomena" section   %%%%% 
+\subsection{Diffusion maps show structural diversity within PCCA+ macrostates}
+
+% add new diffusion map section introduction
+Although our SRV-MSMs provide useful insight into slow processes, we are also interested in the ensemble of configurations within PCCA+ macrostates.
+These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population.
+Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most popluated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.
+
+% use this to substantiate the shifted fast response
+\subsubsection{5s1 state analysis}
+While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
+We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding.
+To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences.
 We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
@@
 Diffusion maps built from samples of the 3' shifted states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).
+
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch C:T-pairs are never bound -- a consequence of the excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+This reveals that the mismatch C:T-pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
 This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
@@
 
+% need to work on this if adding to this section (might belong somewhere else or not at all?
+Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
+In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
+
 %AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3spn2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps.png}
-        \caption{First two diffusion map coordinates built from 10000 single-shifted 5' states, equally sampled from AT-all and GC-end.
-Color maps show inverse distances between out-of-register ends and complementary ends.
-}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps.PNG}
+        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
+Color maps show inverse distances between out-of-register ends and complementary ends.}
         \label{fig:GC-end_dmaps}
@@
 
-\subsubsection{GC-end dynamics are largely two-state}
+\subsubsection{F4 state analysis}  % shift to include this in the overall diffusion map section
+%\subsubsection{Diffusion maps show ensemble of frayed states}
 
-The kinetic behavior between GC-end macrostates is also distinct compared to AT-all.
-We do not observe any significant flux between the double-shifted to single-shifted 5' states.
-While this could be due to insufficient sampling, it is likely that the double-shifted states has too few contacts to directly transition into a single-shifted state without first dissociating.
-This is further demonstrated by the kinetic distance between this state and the aligned hybridized state (S0 and S4, respectively) shown in TICA projections (Figure \ref{fig:GC-end_in_one}).
-We observe some flux between the single-shifted states and aligned hybridized state, but these events are more rare than their AT-all equivalents.
-This suggests that although these transitions are possible routes for hybridization/dissociation, it is substantially more common for the transition to proceed in a two-state manner.
-In T-jump, FTIR, and 2D IR experiments, the GC-end sequence was observed to have less deviation from the two-state dissociation model compared to oligonucleotides with GC pairs closer to the core \citep{Sanstead2016}.
-Furthermore, GC-ends revealed a distinct T-jump response containing evidence for the loss of G:C and A:T base pairing prior to full dissociation of the duplex \citep{Sanstead2018DirectDehybridization}.
-Given that this response was slower than the A:T fraying response observed in other sequences, it is possible that it corresponds to shifting or related looping dynamics.
-This cannot be verified given that the signal was recorded in a congested spectroscopic range and could be hard to differentiate from aligned G:C fraying.
-We will discuss this more in our investigation of higher temperatures.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{GC-core}
-
-\subsubsection{Building SRV-MSM}
-
-The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end.
-Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis.
-Next, we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
-We found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "4bp-frayed" states.
-Again, we found transition probabilities between states and visualized representative molecular renderings.
-We observe that the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not.
-Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
-Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states.
-This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
-We observe these events to occur with equal probability on either permutable side of the helix.
-The MSM transition probabilities suggest that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions.
-We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
-This is expected considering that this state is more accessible from an already bound helix.
-Moreover, once oligos are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate.
-Thus once a dissociated to 4bp-frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation.
-On the other hand, transitions from the hybridized to 4bp-frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_thermo.PNG}
-        \caption{Full MSM output for GC-core}
-        \label{fig:GC-core_thermo}
-	\end{center}
-\end{figure}
-
-\subsubsection{Diffusion maps show ensemble of frayed states}
-
-To further examine this 4bp-frayed state, we build diffusion maps using 10000 frames sampled from the macrostate \ref{fig:GC-core_dmaps}.
+We applied a similar approach to investigate the structural composition of the F4 macrostate.
+We build diffusion maps using 10000 frames sampled from the macrostate \ref{fig:GC-core_dmaps}.
 This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-core_dmaps.png}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
         \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
@@
 
-\subsubsection{SRV correlations to physical coordinates?}
+% not sure if this should directly follow the SRV-MSM GC-core description or directly follow the GC-core dmaps?
+\subsection{SRV correlations to physical coordinates}
 
-Having constructed and interpreted our SRV-MSM, we revisited our original SRV basis to analyze what physical correlations could be interpreted from the model.
-We found these GC-core modes to be of particular interest as they reveal the hierarchical nature of the dynamical encoding.
-We examined a collection of 1000-frame "trimmed" trajectories centered on both hybridization or dissociation events.
-For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they showed strong correlation.
+Having constructed and interpreted SRV-MSMs, we revisited our original SRV basis to investigate how slow modes are constructed from input features.
+We found these GC-core modes to be of particular interest as they reveal the hierarchical and nonlinear nature of the dynamical encoding.
+In particular, we examined a collection of "trimmed" trajectories centered on both hybridization or dissociation events.
+For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they shared behavior.
 Two representative trajectories are shown in figure: (\ref{fig:GC-core_tracking_modes}).
@@
 There is a small signal for fraying at the outer base pairs, but the mode overwhelmingly learns about these neighboring A:T/G:C bonds.
-This behavior reflects the above-mentioned kinetic trap between core binding and the fully hybridized state.
+This behavior reflects movement in and out of the F4 state in the corresponding SRV-MSM.
 SRV3 is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
-We attribute this to the SRV learning about the diffusive motions of the two body system.
-In addition to picking up on dissociation behavior, the third mode peaks when the oligos are close together but configured in such a way that is not amenable hybridization.
+We attribute this to the SRV learning about the diffusive motions of the two body system and evaluating the likelihood of an imminent hybridization event.
+The third mode also peaks when the oligos are close together but configured in such a way that is not amenable hybridization.
 These misaligned conformations include inverse contacts where 5'/5' and 3'/3' ends meet and looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts.
 
-Despite the strong qualitative trends we observe between physical coordinates and leading SRV modes, we only find high Pearson correlations for SRV1.
+Despite the strong trends we observe between physical coordinates and leading SRV modes, we only find high Pearson correlations between SRV1 and physical coordinates.
 For the next two SRV modes, the sign of their correlation switches depending on whether the oligos are in the hybridized or dissociated state.
-This shows that these modes are providing support on top of the first mode -- which determines hybridization vs.
-dissociation -- and thus can display very different behaviors in either state.
-With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in states 0 and 1 -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state (S2).
-Accordingly, we recalculated Pearson correlations between each SRV mode and all distances in states where the modes are active.
-Figure \label{fig:GC-core_tracking_modes} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each other during 4bp-fraying.
-We also observed a highly symmetric correlation between SRV3 and central base pairs distances, indicative of overall diffusive behavior.
-Taken together, these analyses reveal how the SRV learns and represents the dynamical space, leading to the MSM results we observe above.
+This shows that these modes are providing support on top of the first mode -- which serves as an indicator function for hybridized vs.
+dissociated state -- and thus can display very different behavior in either state.
+With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in the H and F4 states -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state.
+Accordingly, we calculated Pearson correlations between each SRV mode and all distances in states where the modes are active.
+Figure \label{fig:GC-core_tracking_modes} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each when the duplex is in the F4 state.
+We also observed a highly symmetric correlation between SRV3 and central base pairs distances, which indicative of overall diffusive behavior.
+Taken together, these analyses reveal how the SRV learns and represents the dynamical space in an hierarchical manner, providing enhanced resolution to a linear method like tICA and producing the MSM results we observe above.
+
+% note that similar analysis on AT-all and GC-end can be found in the SI (move both dynamics plots and descriptions there)
 
@@
 	\begin{center}
-        \includegraphics[width=120mm]{Figs/figs_0804/GC-core_tracking_modes.png}
-        \caption{Show how SRV coordinates correspond to physical features over time and during sample dissociation and hybridization events.
-Should probably go to SI if we also include GC-core diffusion maps.}
+        \includegraphics[width=120mm]{Figs/figs_imp/GC-core_tracking_modes.png}
+        \caption{Leading GC-core SRV coordinates correlated with identifiable physical coordinates during sample hybridization (a) and dissociation (b) events.
+(c) Pearson correlations between all 100 intermolecular base pair distances and each the three leading slow modes.
+SRV2 correlations were calculated in the H and F4 states; SRV correlations were calculated in the D state.}
         \label{fig:GC-core_tracking_modes}
@@
 
-\subsubsection{\label{sec:Results}Comparison to Experiment}
+\subsection{\label{sec:Results}Temperature-dependent timescale comparisons}
 
-When examining these same four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
-found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
-As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
-Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
-We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
-Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.
-Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
+Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone.
+To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence.
+We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core.
+Additionally, we repeated the protocol in Sandstead et al.
+to generate temperature series data for AT-all and GC-mix sequences \citep{Sanstead2018DirectDehybridization}.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a cutoff.
+We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
+Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
+After accounting for the acceleration factor, we saw agreement between the experimental data and simulated relaxation fits.
+GC-core showed the largest deviation experiment, with dissociation rate increasing much more quickly with temperature in simulations.
 
-\subsection{\label{sec:Results}GC-mix}
+% GC-core fast timescales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
+% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor.
+We maintained a consistent accerelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses.
+It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-mix_transitions.PNG}
-        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
-        \label{fig:GC-mix_transitions}
-	\end{center}
-\end{figure}
-
-The GC-mix sequence shows a similar implied timescale distribution to GC-core, however we no longer see a converged slow mode corresponding to multi-base fraying behavior.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
-These correlate closely with the first and third GC-core SRV modes.
-Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data.
-Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
-While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation.
-On the other hand, the 4bp-frayed state in GC-core had a substantial probability (~10\%) of fully dissociated rather reforming an intact helix.
-Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
-
-Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
-For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
-During hybridization, we observe the formations of some key base pair contacts before the full duplex forms.
-We observe that first contacts tend to involve one of the G:C bonds, which is not surprising that these are more evenly spaced out than in GC-core and GC-end.
-This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Yin2011KineticsHybridization, Porschke1971CooperativeTransition}.
-We obverse dissociation events proceeded by two base pair fraying on one side of the duplex followed by more rapid dissociation of the central base pairs.
-There tends to be a short-lived state composed of 2-4 base pair contacts shortly before fully dissociation occurs.
-In contrast to the 4-bps frayed state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or TICA space, nor do they tend to reform intact duplexes.
-As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}.
-We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
-
-\subsubsection{\label{sec:Results}Temperature-dependent timescale comparisons}
-
-Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone.
-To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence.
-We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core \citep{Sanstead2018DirectDehybridization}.
-Although there was no experimental data for AT-all and GC-mix, we used our relaxation fits to predict how these might compare with the other sequences.
-We measure the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
-We find the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
-Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
-
-\begin{figure}  %[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/fast-slow_end-core.png}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
         \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
-Both show an exponential relationship between dissociation and temperature, however fraying appears more linear for GC-core than for GC-end.}
+The effective simulation temperature is shifted by 4 degrees K to account for systematic differences in simulation and experimental melting temperature.
+Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
         \label{fig:relaxation-comparison}
@@
 
+% A documented short-coming for the 3spn2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
+This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
+
 The fast response -- which Sanstead et.
-al attributed to base pair fraying signatures -- was more difficult to compare given the limits our coarse-grained model.
+al attributed to base pair fraying signatures -- was more difficult to compare against simulation observables.
 Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
-Furthermore, all-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen binding, and terminal stacked conformations\citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
+All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
 Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
-Instead, we measure the fast fraying response by observing how long it takes for either duplex terminal end to split beyond a given cutoff.
+Instead, we measured the fast fraying response by counting frames until either duplex terminal end to split beyond a cutoff.
 This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
 This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events.
-For further comparison, we include fits over a range of cutoffs and repeat these calculations on the second and third interior base pairs.
+Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
+Experimental comparisons at high temperatures were limited due to mixing between the fast and slow responses.
 
-For GC-core, each base pair and cutoff combination appears linear with temperature, indicating a barrierless and diffusion-driven process.
-As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated by about two order of magnitude relative to experiments.
-In contrast, GC-end is distinctly exponential, especially at base pairs closer to the core of the duplex.
-This behavior makes sense given the thermodynamic barrier to G:C fraying, but it also deviates the GC-end experimental linear fit.
-In addition to fraying, we notice that the rate of shifting increases at higher temperature.
+For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
+In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater thermodynamic barrier associated with G:C fraying.
+We observe similar trends in the experimental data, especially after applying a consistent acceleration factor as shown in figure \ref{fig:relaxation-comparison}.
+We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about two order of magnitude relative to experiments.
+It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
+In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation.
+Although we should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
+and that 3spn2 can capture some sequence-dependent fraying effects.
+
+%Again, GC-core behavior deviates from simulations prediction with consistently lower than expected fast responses.
+We believe this could be due to slower contributions from internal A:T base pairs sequentially fraying.
+
+% Experiments fast responses are difficult to discern at higher temperatures due to mixing with slow responses. This is especially true for GC-end because to the signals are closer to begin with.
+% add new info here, not sure if shifting discussion is necessary
+% take away message is that these are good proxies for the responses
+% although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response
+
+%In addition to fraying, we notice that the rate of shifting increases at higher temperature.
 Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
-For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
+For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
 It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
 
-The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
+%The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
 Previous work has shows that kinetic association rates were accelerated by about one to two orders of magnitude relative to experiment \citep{Hinckley2013AnHybridization}.
-It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not parameterized based on any dynamical properties \citep{?}.
-Furthermore, more data is needed for explicit ion model comparisons.
+It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not extensively parameterized on dynamical properties properties \citep{Hinckley2013AnHybridization}.
 
@@
 \section{\label{sec:conc}Conclusion}
-\subsection{\label{sec:Results}Will look a lot like the last paragraph of the intro/abstract}
 
-\subsubsection{\label{sec:Results}Limitations}
+We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
+We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
+We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
+On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
+This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense.
+Furthermore, we replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
 
-\begin{enumerate}
-	\item Does not explicitly pick up on nucleation/zippering events.
-Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding modes.
-Furthermore these zippering dynamics occur significantly faster than shifiting or GC-core fraying and cannot be captured via our save rate.
-	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
-	\item Limited by memory (number of frames) required to capture equilibrium trajectories and rarity of hybridization/dissociation events
-	\item A smoother free energy surface minimizes the time spent in these intermediate states
-	\item Sensitive to 3spn2 artefacts.
-Shifting may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}.
-Base-pair mismatch interactions may be somewhat unphysical and alter state populations.
-Although we present shifting as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
-\end{enumerate}
+\subsection{\label{sec:conc}Limitations}
+
+Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge on relevant faster processes such as duplex nucleation and zippering.
+These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.
+Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process.
+We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations.
+Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence, and were working near memory limits when training SRVs and building SRV-MSMs.
+
+In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
+For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
+In general, coarse-grained models produce a smoother free-energy surface which can results in must faster motions between states.
+This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment.
+Furthermore, it may be easier to cross between states -- e.g.
+a 5S1 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
+
+\subsubsection{\label{sec:conc}Not sure whether to discuss explicit comparisons} 
+We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
+In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the exact ion conditions specified in the \citet{Sanstead2016} using the 3spn2 explicit implementation \citep{Hinckley2015}.
+These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities.
+Comparisons were limited by the higher computational demand required to simulate 5x the beads in the box, a shorter integration timestep, and reliance on Ewald summation to account for electrostatics.
+We caution over-interpretation of these results as the explicit model has not been as widely adopted or thoroughly validated compared to the implicit model.
+
+(Should add something else here or re-arrange above...) Going forward, we believe that this work will contribute novel insights to DNA design strategies, and can be leveraged as a prediction mechanism for the broader sequences spaces.
+
+\section{\label{sec:Results}SI Methods notes}
+
+% add secion on dmap and high temp methods?
+%\subsection{Additional methods}
+
+%In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff.
+
+\subsubsection{SRV optimization and analysis}
+In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the simplest dynamics and displayed a clear spectral gap between modes.
+For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step.
+Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
+We found that most implied timescales converge at an SRV lag time of 1.2 ns.
+Next we selected an optimal number of SRV components to include in our analysis.
+After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
+It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
+For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
+As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
+
+%\subsubsection{SRV-MSM construction and optimization}
+Although SRV coordinates alone provide some information, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
+Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
+We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
+Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
+This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous tICA basis.
+Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
+Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
+Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
+Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
+
+Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
+The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most relevant to each process.
+For example, the first slow mode shows a positive correlation to each distance and the strongest correlation with native base pair distances (shown along the main diagonal).
+Given these correlations and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
+The next four SRV components all show a relatively high correlation along offset diagonals.
+These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
+Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
+
+We found GC-end had a similar implied timescales distribution to AT-all, with a distinct spectral gap after the fourth mode.
+For AT-all and GC-end, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
+For GC-core we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
+We found that four macrostate clustering was unstable -- likely because the third mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized (H), dissociated (D), and 4 4 base pair frayed (F4) states.
+The GC-mix sequence showed a similar implied timescale distribution to GC-core, however we no longer saw a converged slow mode corresponding to multi-base fraying behavior.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
+These correlate closely with the first and third GC-core SRV modes.
+Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
+As such, we designated this transitions as effectively two-state within the resolution of our model.
+
+%For GC-core and GC-mix, we found this clustering convention to be unstable -- likely because the last converged mode was providing information about dissociation dynamics -- and performed PCCA+ with one less macrostate.
+
+To visualize macrostates, we project the data into the two leading tICA coordinates.
+Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
+Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
+After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment.
+Means and standard deviations were calculated from this ensemble.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/allseq_cross_val_scores_lag-40.png}
-        \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for both the converge of the VAMP-2 and consistent scores between folds to ensure that the model is not fitting on statistical noise.
-(probably will be SI, but could include AT-all cross-val in the walkthrough section)}
-        \label{fig:srv_crossval}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_imp/allseq_event_count.png}
+        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in dissociated state.
+The fraction of hybridized duplexes averaged across these sets is shown over time.
+For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive Tmelt difficult to identify.
+The number of hybridization/dissocation combined across trajectories is shown.
+Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics.
+AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.}
+        \label{fig:allseq_event_count}
 	\end{center}
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_full.PNG}
-        \caption{Full 5' diffusion maps including degenerate third diffusion map mode}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_imp/allseq_crossval.png}
+        \caption{5-fold cross validation procedure to select number of SRV coordinates.
+We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
+        \label{fig:allseq_srv_crossval}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=80mm, 
+        scale=0.5]{Figs/figs_imp/all_seq_implied_times.png}
+        \caption{SRV-MSM vs.
+tICA-MSM implied timescales convergence for GC-end, GC-core, and GC-mix.
+Progressively larger spectral gaps are observed between the first mode and higher order modes.
+We observe convegenve of resolvable higher order modes at a shared lag time of 1.2 ns.
+It is difficult to converge the leading mode in this regime due to the infrequency of hybridization/dissociation events.}
+        \label{fig:all_seq_lags}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_imp/GCSeq_TJTraces.png}
+        \caption{Kinetic traces of the fast spectroscopic response show varied degrees of shifting based on sequences.
+More info from Brennan on subtle differences here, discuss stretching coefficients?
+(could include this in main text as well)}
+        \label{fig:fast_stretching}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.PNG}
+        \caption{Full 5' diffusion maps including degenerate third diffusion map mode.}
         \label{fig:GC-end_dmaps_full}
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/GC-end_dmaps_3prime.PNG}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_3prime.PNG}
         \caption{Full 3' diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct "looping" region for intact GC bonds}
@@
 
+% add figure for slow and fast response fitting (just GC-core)
 \begin{figure}[ht!]
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_physical_cvs_labels.PNG}
-        \caption{All sequences and metastable macrostates plotted on frayed/shifted collective variable axes.}
-        \label{fig:all_seq_physical_cvs_labels}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.png}
+        \caption{Slow and Fast modes were determined by fitting over a distribution of 120 independent simulations.
+For each temperature in the series, a relaxation curve was fit to the distribution and the associated exponential coefficient was used to the calculate rate.
+Temperatures series between -5 - +15K relative to empirically determined sequence melting temperature were explored.
+This process was repeated for each sequence (only GC-core is shown above).}
+        \label{fig:fast_slow_fits}
 	\end{center}
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/all_seq_tica_cvs.PNG}
-        \caption{All sequences plotted on combined TICA axes. Might include in SI or not at all.}
-        \label{fig:all_seq_tica_cvs}
-	\end{center}
-\end{figure}
+%add anymore figures for the time analysis: As shown bleow:
+\begin{comment}
 
@@
     \item MSM construction
+
 \end{enumerate}
+\end{comment}
 
 \begin{comment}
-
-Other ways to investigate GC-mix:
-\begin{enumerate}
-
-	\item Track GC-mix trajectories along transition region
-	\item Discuss misaligned configurations here instead of in the GC-core section
-\end{enumerate}
-
-Compare frayed thermo to Paul's lattice model predictions.
 
@@
 - Open termini state shows new fraying mode of 5ps (I think shows free base motions)
-
 \end{comment}

### 2020-11-23 manuscript.tex
Updates from Overleaf

@@
 \usepackage{changepage}
+\usepackage{mathrsfs}
 
@@
 In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement.
-We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Models (MSMs) construction.
+We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model (MSM) construction.
 Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides.
 For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states.
-In contrast, G:C pairs near the center of the duplex induce more rapid fraying dynamics.
-In both cases, hybridization/dissociation mechanisms deviate from an “all-or-nothing” model.
-Our computational predictions are in excellent accord with experiment, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
+In contrast, G:C pairs near the center of the duplex induce more pronounced fraying dynamics through which hybridization and dissociation are facilitated.
+In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model.
+Our computational predictions show agreement with experiments, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
 
@@
 From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}.
-DNA dynamical behavior has become increasingly important in these technologies and a point of focus in fundamental biological processes such as transcription and gene regulation.
-\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}.
-Our understanding of DNA hybridization -- the formation of a duplex from two amenable single strands -- has been integral to these developments.
-The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for many mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
+In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}.
+A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
+The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
 Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}.
@@
 
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
-More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags altering strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
 Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
@@
 Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates.
-For example, metastable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, , Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}).
-The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequences dependent.
+For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}.
+The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent.
 Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
 Sanstead et al.
-highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}.
-Together, these dynamics represent the most substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
-
-In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
+highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides \citep{Sanstead2016}.
+Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
+
+In this work, we study the same four sequences explored by Sanstead et al.
+in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
 We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
 We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
-Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.
-MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies.
-\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}.
+Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system  -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
+MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
+\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent.
 We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
@@
 By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
-Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models.
-Additionally, we can compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs.
+Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible.
+Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs.
 Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
@@
 The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}.
-3spn2 has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
+3spn2 been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
 \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.
@@
 We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}.
-We specified a 240 mM implicit salt concentration and used a A Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
-We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
+We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
+We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat to account for implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
 Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states.
-We use a 20 fs time step and $13\times10^{9}$ steps in each simulation, saving frames every 100 ps.
-For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors for each simulations.
-Half of runs were initialized from the hybridized state and half from a dissociated state.
+We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps.
+For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation.
+Half of runs were initialized from the native hybridized state and half from a dissociated state.
 The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence.
-Depending on sequence, we observed between 40-100 hybridizing and dissociating transitions across all runs.
+Depending on sequence, we observed between 55-100 hybridizing and dissociating transitions across all runs.
 
@@
 Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together.
-This permutation reduction follows a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
+To account for this, we averaged permutable distances (45 pairs in total) together following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
 The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}.
-The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
+The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
+(Can include equations or additional info in SI?) We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordiantes (figure \ref{fig:allseq_features_vamp2}).
+We found that reciprocal coordinates performed better across the board, but we saw no substantial difference between the three featurization methods.
+This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.
+The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
+These features were normalized and passed into sequence-specific SRVs.
 
@@
 
-We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances and found small differences between the three that varied by sequences, but overall we did not observe a loss in kinetic information when using the symmetrized feature set (\ref{fig:allseq_features_vamp2}).
-We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal symmetrized coordinates to train the model.
-The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
-These features were normalized and passed into sequence-specific SRVs.
-
 \subsection{\label{sec:methods}SRVs}
@@
 as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}.
-The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis},and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}.
+The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}.
 The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}.
@@
 
-Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
-In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
-We used a batch size of 50000 and ran each model for a total of 20 training epochs.
-We used two hidden layers and set the size of each layer to 100.
-For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
-SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
-
-% might want to include these training params in the SI?
-
 \subsection{\label{sec:methods}SRV-MSMs}
@@
 MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}.
-The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time.
+Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
@@
 The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
-The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.
-Additional step-by-step details on SRV-MSM construction for each sequence are provided in the supplemental information and on Github (https://github.com/mrjoness/...).
+The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.
 
@@
 We used an Ewald summation to calculate long range Coulombic interaction between DNA and ions using a real space cutoff of 2 nm.
-
-% A Debye-Huckle screening potential was used to monitor account for ion concentration .
-
-\begin{comment} 
-% SI equations
-In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
-The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
-
-\begin{align*}
- 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
- 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
- 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
-	\\
- 	VAMP2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
- 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
-\end{align*}\label{CK1}
-
-\end{comment}
 
@@
 
-\subsection{All SRV-MSMs}
+\subsection{Interpreting PCCA+ Macrostates}
 
@@
 
-\subsubsection{SRV optimization and analysis}
+\subsection{SRV training}
+
+Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
+In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
+We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
+In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
+We used a batch size of 50000 and ran each model for a total of 20 training epochs.
+We used two hidden layers and set the size of each layer to 100.
+For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
+SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
+SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
+
+\subsection{SRV-MSM walkthrough and analysis}
 In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the simplest dynamics and displayed a clear spectral gap between modes.
@@
 \begin{comment} 
+% SI equations
+In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
+The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
+
+\begin{align*}
+ 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
+ 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
+ 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
+	\\
+ 	VAMP2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
+ 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
+\end{align*}\label{CK1}
+
+\end{comment}
+
+\begin{comment}
 

### 2020-11-23 manuscript.tex
Updates from Overleaf

@@
 Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}.
-We also show that PCCA state assignments are capturing free energy minima in tICA space as independent states.
+We also show that PCCA+ clustering assigns states to relative free energy minima in the tICA space.
 
@@
 Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
-It is informative to compare our MSM predictions against NN calculations to understand how inert tail or kinetic effects cause deviation from these predictions.
+It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
 For AT-all, NN calculations (figure \ref{fig:NN_table}) predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
@@
 For GC-end, we consider C:T and G:A mismatches in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends.
-Based on this treatment, NN calculations yield higher overall free energies (due to fewer native base pair contacts), but, contrary to PCCA+ results, predict that 3' out-of-register states to be more stable than 5' states.
+Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.
+Contrary to PCCA+ results, however, 3' out-of-register states to be more stable than 5' states.
 This leads us to believe that 5' vs.
@@
 
-Beyond state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
-Figure \ref{fig:allseq_transition}!!!change ref!!!
-shows the probability of moving from one state to another within the MSM lag time.
+In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
+Figure \ref{fig:allseq_transition} shows the probability of moving from one state to another within the MSM lag time.
 For AT-all, we observe approximately equal probability of transitioning from D to any other state, indicating that this is primarily a diffusion-driven process.
@@
 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
-Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
+Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
 Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
-In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than 3S2 $\rightarrow$ 3S1.
-Indeed for GC-end, we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
+In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path to native hybridization.
+Indeed, for GC-end we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
 The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D.
@@
 Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
-A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of H $\rightarrow$ D pathways pass through some out-of-register state.
+A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways pass through some out-of-register state.
 While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states.
@@
 For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale.
-We expect that fraying in out-of-register conformations would occur at different rates than intact fraying.
-Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contribute to the ensemble.
-Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, sliding mechanisms during the dissociation process, or some combination of the two.
+We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
+Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble.
+Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two.
 In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.
 We interrogate these differences by examining high temperature simulations later in this manuscript.
+(Can discuss more details about fig \ref{fig:fast_stretching} here and potentially include this fig in the main text)?
 
@@
 
-\subsection{GC-core fraying} % not sure if I should make this its own section?
+\subsection{GC-core metastability} % not sure if I should make this its own section?
 
@@
 Previous studies suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
-Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived metastable state.
+Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
 This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
-We observe these events to occur with equal probability on either permutable side of the helix.
-The hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state.
-This is expected considering that this state is more accessible from an already bound helix.
-Moreover, once oligos are in F4, they are over 10x more likely to return to H than to D.
+We observe these events to occur with equal probability on either permutable end of the helix.
+The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D.
 Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed it.
 On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
-
-% try to split this up between an experimental section (all SRV-MSMs) and more detail on diffusion maps
-\subsection{GC-core Comparison to Experiment}
 
@@
 Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
-We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.
+We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
 Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
@@
 Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
-As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively.
+Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model.
+We did, however, we do observe substantial fraying of the two AT termini in the simulation data.
 Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
-While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation.
-On the other hand, the F4 state in GC-core had a substantial probability of fully dissociated rather reforming an intact helix.
+While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state.
 Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
@@
 For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
-During hybridization, we observe the formations of some key base pair contacts before the full duplex forms.
-We observe that first contacts tend to involve one of the G:C bonds, which is not surprising that these are more evenly spaced out than in GC-core and GC-end.
+During hybridization, we observed the formations of some key base pair contacts before the full duplex formed.
+Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T.
 This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}.
-We obverse dissociation events proceeded by two base pair fraying on one side of the duplex followed by more rapid dissociation of the central base pairs.
+For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
 There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs.
@@
         \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
-        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
+        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
+Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs.
+The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined.
+Fluctuation between states is common during the transition.}
         \label{fig:GC-mix_transitions}
@@
 % add new diffusion map section introduction
-Although our SRV-MSMs provide useful insight into slow processes, we are also interested in the ensemble of configurations within PCCA+ macrostates.
+Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates.
 These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population.
@@
 We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most popluated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps.PNG}
+        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
+Color maps show inverse distances between out-of-register ends and complementary ends.}
+        \label{fig:GC-end_dmaps}
+	\end{center}
+\end{figure}
 
@@
 We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
-This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands.
+This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands.
 In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}).
-Diffusion maps built from samples of the 3' shifted states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).
+Diffusion maps built from samples of the 3S1 states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).
 
 The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatch C:T-pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+This reveals that the mismatched C:T pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
 This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
@@
 Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
-These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
+These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations is confirmed by molecular renderings sampled from these regions.
+Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
 
 % need to work on this if adding to this section (might belong somewhere else or not at all?
-Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
+%Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
 In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps.PNG}
-        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
-Color maps show inverse distances between out-of-register ends and complementary ends.}
-        \label{fig:GC-end_dmaps}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
+        \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
+Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
+        \label{fig:GC-core_dmaps}
 	\end{center}
@@
 We applied a similar approach to investigate the structural composition of the F4 macrostate.
-We build diffusion maps using 10000 frames sampled from the macrostate \ref{fig:GC-core_dmaps}.
+We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
 This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
 Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
-We found that a larger distance between the third and fourth A:T pairs increased the PCCA+ probability of inclusion into macrostate.
+We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate.
 This distance also correlated closely with the second diffusion map mode.
 Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core.
-In these conformations, one of the strands maintains some helical character while the other twist out of place, resulting in WC bonds being obstructed by the oligo backbone.
+We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
+In these conformations, one of the strands maintains some helical character while the other twists out of place, resulting in WC bonds being obstructed by the oligo backbone.
 These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
-        \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
-Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the the overlap between adjacent AT basepairs and the GC core.}
-        \label{fig:GC-core_dmaps}
+\begin{figure} %[ht!]
+	\begin{center}
+        \includegraphics[width=120mm]{Figs/figs_imp/GC-core_tracking_modes.png}
+        \caption{Leading GC-core SRV coordinates correlated with identifiable physical coordinates during sample hybridization (a) and dissociation (b) events.
+(c) Pearson correlations between all 100 intermolecular base pair distances and each the three leading slow modes.
+SRV2 correlations were calculated in the H and F4 states; SRV correlations were calculated in the D state.}
+        \label{fig:GC-core_tracking_modes}
 	\end{center}
@@
 For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they shared behavior.
-Two representative trajectories are shown in figure: (\ref{fig:GC-core_tracking_modes}).
+Two representative trajectories are shown in figure (\ref{fig:GC-core_tracking_modes}).
 Complementary G:C pairs are the best indicators for a hybridization/dissociation event, and we see a sharp change in the first SRV mode (SRV1) as these bonds form or break.
@@
 There is a small signal for fraying at the outer base pairs, but the mode overwhelmingly learns about these neighboring A:T/G:C bonds.
-This behavior reflects movement in and out of the F4 state in the corresponding SRV-MSM.
+Moreover, this behavior reflects movement in and out of the F4 state in the corresponding SRV-MSM discussed above.
 SRV3 is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
@@
 
-Despite the strong trends we observe between physical coordinates and leading SRV modes, we only find high Pearson correlations between SRV1 and physical coordinates.
-For the next two SRV modes, the sign of their correlation switches depending on whether the oligos are in the hybridized or dissociated state.
-This shows that these modes are providing support on top of the first mode -- which serves as an indicator function for hybridized vs.
-dissociated state -- and thus can display very different behavior in either state.
+Despite the qualitative trends we observe between physical coordinates and leading SRV modes, we found it difficult to find correlations across full trajectories.
+With the exception of SRV1, Pearson correlations between physical coordinates and SRV modes were near zero and showed erroneous trends.
+However, when we performed this analysis along smaller sections of trajectories we noticed that the sign of SRV2 and SRV3 correlations switched depending on whether the oligos were in the hybridized or dissociated state.
+This shows that these modes are inherently nonlinear and provide support on top of the first mode -- which serves as an indicator function for hybridized vs.
+dissociated state.
 With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in the H and F4 states -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state.
@@
 
-\begin{figure} %[ht!]
-	\begin{center}
-        \includegraphics[width=120mm]{Figs/figs_imp/GC-core_tracking_modes.png}
-        \caption{Leading GC-core SRV coordinates correlated with identifiable physical coordinates during sample hybridization (a) and dissociation (b) events.
-(c) Pearson correlations between all 100 intermolecular base pair distances and each the three leading slow modes.
-SRV2 correlations were calculated in the H and F4 states; SRV correlations were calculated in the D state.}
-        \label{fig:GC-core_tracking_modes}
-	\end{center}
-\end{figure}
-
 \subsection{\label{sec:Results}Temperature-dependent timescale comparisons}
@@
 We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core.
-Additionally, we repeated the protocol in Sandstead et al.
-to generate temperature series data for AT-all and GC-mix sequences \citep{Sanstead2018DirectDehybridization}.
+Additionally, we repeated the protocol in \citet{Sanstead2018DirectDehybridization} to generate temperature series data for AT-all and GC-mix sequences.
 
@@
 Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
-Experimental comparisons at high temperatures were limited due to mixing between the fast and slow responses.
 
@@
 In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater thermodynamic barrier associated with G:C fraying.
-We observe similar trends in the experimental data, especially after applying a consistent acceleration factor as shown in figure \ref{fig:relaxation-comparison}.
+We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}).
 We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about two order of magnitude relative to experiments.
@@
 
-Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge on relevant faster processes such as duplex nucleation and zippering.
+Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge relevant faster processes such as duplex nucleation and zippering.
 These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.
@@
 We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
-In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the exact ion conditions specified in the \citet{Sanstead2016} using the 3spn2 explicit implementation \citep{Hinckley2015}.
+In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3spn2 explicit implementation \citep{Hinckley2015}.
 These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities.
@@
 
-(Should add something else here or re-arrange above...) Going forward, we believe that this work will contribute novel insights to DNA design strategies, and can be leveraged as a prediction mechanism for the broader sequences spaces.
+Going forward, we believe that this work can be extended to predict more general trends in sequence-dependent hybridization and strategies for experimental comparisons.
+These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.
 
@@
 %In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff.
+
+\subsection{VAMP-2 Scoring}
+Optimization of hyperparameters including feature choice, number of microstates, and number of slow modes was evaluated with via the VAMP-2 score.
+In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
+The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
+
+\begin{align*}
+ 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
+ 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
+ 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
+	\\
+ 	VAMP-2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
+ 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
+\end{align*}\label{CK1}
 
@@
 \begin{comment}
-% SI equations
-In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
-The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
-
-\begin{align*}
- 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
- 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
- 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
-	\\
- 	VAMP2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
- 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
-\end{align*}\label{CK1}
-
-\end{comment}
-
-\begin{comment}
 

### 2020-12-01 manuscript.tex
ALF tex updates, bib updates, manuscript_annotatedALF2.pdf
Updates from Overleaf

@@
 %include hyperlinks to figures
-\documentclass{article}
+%\documentclass{article}
 \usepackage{graphicx}
@@
 \usepackage{framed}
-\usepackage[hidelinks]{hyperref}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}
+%\usepackage[hidelinks]{hyperref}
+%\usepackage{algorithm}
+%\usepackage{algpseudocode}
+%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}
 
@@
 
-%\author{I. Ken Groupleader}
-%\altaffiliation{A shared footnote}
-%\email{i.k.groupleader@unknown.uu}
-%\phone{+123 (0)123 4445556}
-%\fax{+123 (0)123 4445557}
-%\affiliation[Unknown University]
-%{Department of Chemistry, Unknown University, Unknown Town}
-%\alsoaffiliation[Second University]
-%{Department of Chemistry, Second University, Nearby Town}
-
-%\author{Susanne K. Laborator}
-%\email{s.k.laborator@bigpharma.co}
-%\affiliation[BigPharma]
-%{Lead Discovery, BigPharma, Big Town, USA}
-
 %\author{Kay T. Finally}
@@
 Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
-Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
-
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
+
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
 More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
-Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
+Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
 Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
@@
 Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
-Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
-Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.
+Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
 
@@
 in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
-We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
-We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
-Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system  -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
+We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
 MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
 \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent.
-We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
-Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed.
-We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
-
-We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
-By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
+We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
+
+% this section should look more like a conclusion foreshadowing conclusions and main take aways 
+We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
+Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation.
+Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer folding dynamics populated by a diversity of long-lived metastable states.
+Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior.
+Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process.
+Analysis within metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs.
+This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+
+% example text from Andy's conclusions:
+
+% Correlation of the data-driven kinetic order parameters extracted by SRVs provides new understanding of the sequence-dependent kinetic mechanisms by which oligomers hybridize and reveal these to be richer than classical base-pair shifting and “all-or-nothing” behaviors.
+
+% original last paragraph
+%By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
 Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible.
@@
 
+% introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
+Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed.
+We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
+
 % include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior.
@@
 The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
-Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
 The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}.
-3spn2 been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
+The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
 \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.
 
-We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}.
+We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}.
 We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
-We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat to account for implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}.
-Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states.
+We ran our simulations in the NVT ensemble.
+Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.
+\citep{Schneider1978Molecular-dynamicsTransitions}.
+Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
 We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps.
-For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation.
+For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation.
 Half of runs were initialized from the native hybridized state and half from a dissociated state.
 The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence.
-Depending on sequence, we observed between 55-100 hybridizing and dissociating transitions across all runs.
+We observed between 55-100 hybridization/dissociation events for each of the four sequences.
 
@@
 
-All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
-Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
+A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
+Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library \citep{McGibbon2015MDTraj:Trajectories}.
+Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
+To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
 The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}.
-The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
-(Can include equations or additional info in SI?) We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordiantes (figure \ref{fig:allseq_features_vamp2}).
-We found that reciprocal coordinates performed better across the board, but we saw no substantial difference between the three featurization methods.
-This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.
-The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
-These features were normalized and passed into sequence-specific SRVs.
+Additional details on the VAMP-2 score can be found in the supplemental information.
+By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
+Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
 
@@
         scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
-        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set.
-The inverse distances showed improvement across sequences, and the 55-dimension symmetrized coordinate set performed about as well as larger features set.}
+        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates.
+Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods.
+This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
         \label{fig:allseq_features_vamp2}
@@
 
+The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed.
+While most pipelines employ tICA to generate a linear transformation of input features, we use SRVS as a more powerful modular replacement for tICA.
 SRVs were first developed by Chen et al.
@@
 This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
-The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
-For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.
+The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
 
@@
 
+% need some more detail here -- talk to Andy about this
 MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}.
-Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time.
+After Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
@@
 % Include analysis on overall hybridization rates here or at the end?
-% Show difference between between full hybridization and partial events (GC-end and AT-all)
-
-\subsection{SRV-MSMs provide better resolution than tICA-MSMs}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=400, scale=1]{Figs/figs_imp/AT-all_implied_timescales.png}
-        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
-        \label{fig:AT-all_dynamic}
-	\end{center}
-\end{figure}
-
-As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach.
-Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model \citep{Sidky}.
-Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
-Similar plots for the other sequences are included in the SI.
-We also note that the infrequency of transitions relative to the individual trajectory length leads to slower convergence of the leading mode.
-Because this mode corresponds the overall hybridization and dissociation process, we found that it displayed similar behavior across all sequences and that higher order modes were more informative for lag time selection.
-
-%addition to timescale comparison, we found that tICA-MSMs often produced inconsistent PCCA+ clusterings between runs and tended to be less physically meaningfully
-
+
+\subsection{Sequence-dependent coarse-grained kinetics agree with T-jump IR data}
+
+% rerun these at the actual final temps and just see what happens? Doesn't really matter where Tjump occurs and could make things easier as a whole...
+% talk about rerunning AT-all and GC-mix=
+
+As an initial validation of the 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) experiments.
+Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
+As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.
+Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}.
+We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
+To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
+We derived relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
+
+% need to discuss tmelt shift
+We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
+We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
+We noted an average temperature shift of about 4C between experiment and simulation exponentials.
+While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
+Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is again sensitive to the exact definition of melting temperature.
+After accounting for the melting temperature shift and acceleration factor, we saw agreement between the experimental data and simulated relaxation fits.
+GC-core showed the largest deviation experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
+
+% GC-core fast timescales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
+% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor.
+We maintained a consistent accegqrelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses.
+It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.
+
+% need to update this fig with labels and temp shift
+
+\begin{figure}  %[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
+        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
+The effective simulation temperature is shifted by 4 degrees celcius to account for systematic differences in simulation and experimental melting temperature.
+Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
+        \label{fig:relaxation-comparison}
+	\end{center}
+\end{figure}
+
+% A documented short-coming for the 3spn2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
+This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
+
+The fast response -- which Sanstead et.
+al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
+All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
+Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
+Instead, we measured the fast fraying response by counting frames until either duplex terminal end splits beyond a cutoff.
+This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
+This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events.
+Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
+
+For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
+In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater thermodynamic barrier associated with G:C fraying.
+We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}).
+We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about 100-200x relative to experiments.
+It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
+In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation.
+Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
+and that 3spn2 can capture some sequence-dependent fraying effects.
+% need a better transition here back into SRV-MSMs?
+
+\subsection{SRV-MSMs for each sequence}
+
+% change state names and scientific notation
+% include D here? can see what it would look like...
 \begin{figure}[ht!]
@@
         \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates.
-The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
+Shifted state abbreviations indicate the direction of the shifted overhang (5' vs.
+3') and the number of shifted motifs relative the native state (1 vs.
+2)The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
         \label{fig:allseq_table}
@@
 
+% potentially get rid of State representation, ask Brennan if they are meaningful?
 \begin{figure}[hb!]
@@
 
-\subsection{Interpreting PCCA+ Macrostates}
-
 Following the analysis pipeline described above, we generate SRV-MSMs for each sequence.
 We identified seven kinetically relevant states that are captured within the resolution of the model.
-These include the fully hybridized state (H) in which all native base pairings are intact, the fully dissociated state (D), four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, and a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound.
-Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs.
+Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities.
+By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
+Arrows between states show the conditional transition probability to another state within the MSM lag time.
+We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.
+
+%Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs.
 3') and the number of shifted motifs relative the native state (1 vs.
@@
 % rewrite this into one paragraph
-\subsection{Shifted state stability}
-
-The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}.
+\subsection{MSM stabilities agree with nearest neighbor thermodynamic models}
+
+%The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}.
 The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
@@
 Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
+
+Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}.
+The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}.
 It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
-For AT-all, NN calculations (figure \ref{fig:NN_table}) predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
-We find qualitative agreement to our PCCA+ free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone.
-For GC-end, we consider C:T and G:A mismatches in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends.
+In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN calculations (figure \ref{fig:NN_table}).
+For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
+We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone.
+For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends.
 Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.
-Contrary to PCCA+ results, however, 3' out-of-register states to be more stable than 5' states.
-This leads us to believe that 5' vs.
-3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.
+Contrary to macrostate results, however, the 3S1 state is more stable than 5S1 states.
+Moreover, we do not observe a stable 3S2 cluster for GC-end.
+This indicates that 5' vs.
+3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.
 
@@
 
-\subsection{Shifting dynamics}
+\subsection{Out-of-register dynamics facilitate hybridization and dissociation}
 %\subsubsection{Transition probabilities between states}
@@
 In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
-Figure \ref{fig:allseq_transition} shows the probability of moving from one state to another within the MSM lag time.
-For AT-all, we observe approximately equal probability of transitioning from D to any other state, indicating that this is primarily a diffusion-driven process.
+For AT-all, we observe approximately equal probability of transitioning from D to any other state.
 Once a transition has been made, however, the 5' vs.
 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
-Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation.
+Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
 Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
 In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path to native hybridization.
-Indeed, for GC-end we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
+Indeed, for GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
 The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D.
@@
 These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
-Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and subtle kinetic traces.
+Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and response similarity to other processes such as fraying.
 To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}.
@@
 
-\subsection{Shifting experimental comparisons}
-% how to discuss this without first introducing the fast response?
-
-In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses.
+\subsection{Stretched T-jump IR responses correspond to shifting sequences}
+
+% add intro sentence here before jumping into data -- keep working on this
+%AT-all responses are less clear given that there are no G:C base pairs for comparison.
+
+In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response.
+In particular, the stretched fast response for AT-all and GC-end indicate that there may be a combination of processes occurring on the fast timescale.
+Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response.
+On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T vibration that points to A:T fraying as the predominant dynamic source.
+Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
+In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and a more stretched fast response.
+
+%The origin of the AT-all and GC-end fast response are most difficult to assign because both shifting and fraying could occur, but the more stretched profile of the fast response for AT-all and GC-end suggests that there may be a combination of processes occurring on the fast timescale.
+Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble.
+Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two.
+In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.
+
+%In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses.
 For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale.
@@
 
-\subsection{GC-core metastability} % not sure if I should make this its own section?
+\subsection{Centralizing GC placement induces long-lived, kinetically relevant fraying} % not sure if I should make this its own section?
 
@@
 Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4).
-Previous studies suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
+Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
 Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
@@
 The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D.
-Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed it.
+Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed.
 On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
@@
 defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}.
-When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
-found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}.
-As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core.
+<Add Brennan example> When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+found that the GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}.
+As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
 While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
 Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
-We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
+We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
 Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
@@
 \subsection{GC-mix displays more canonical hybridization and dissociation}
+
+Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
 Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix.
@@
 Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model.
-We did, however, we do observe substantial fraying of the two AT termini in the simulation data.
+We did, however, observe substantial fraying of the two AT termini in the simulation data.
 Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
@@
 
-Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
-For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
+To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
 During hybridization, we observed the formations of some key base pair contacts before the full duplex formed.
@@
 %%%%%%% start of "interesting phenomena" section   %%%%% 
-\subsection{Diffusion maps show structural diversity within PCCA+ macrostates}
+\subsection{Macrostate distributions inform state stability and kinetics}
 
@@
 Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most popluated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.
+We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most populated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.
 
@@
 
-% use this to substantiate the shifted fast response
-\subsubsection{5s1 state analysis}
+% need to gear this more towards the kinetics
+\subsubsection{5s1 state distributions inform differences between AT-all and GC-end kinetics}
 While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
@@
 
-The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
+%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
 This reveals that the mismatched C:T pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.PNG}
         \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
@@
 
-\subsubsection{F4 state analysis}  % shift to include this in the overall diffusion map section
+\subsubsection{The F4 state interrupts GC-core hybridization and dissociation}  % shift to include this in the overall diffusion map section
 %\subsubsection{Diffusion maps show ensemble of frayed states}
 
-We applied a similar approach to investigate the structural composition of the F4 macrostate.
+We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
 We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
@@
 
+% note that similar analysis on AT-all and GC-end can be found in the SI (move both dynamics plots and descriptions there)
+
+% Additional notes on the temp-dependent secttion
+%Again, GC-core behavior deviates from simulations prediction with consistently lower than expected fast responses.
+We believe this could be due to slower contributions from internal A:T base pairs sequentially fraying.
+
+% Experiments fast responses are difficult to discern at higher temperatures due to mixing with slow responses. This is especially true for GC-end because to the signals are closer to begin with.
+% add new info here, not sure if shifting discussion is necessary
+% take away message is that these are good proxies for the responses
+% although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response
+
+%In addition to fraying, we notice that the rate of shifting increases at higher temperature.
+Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
+For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
+It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
+
+%The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
+Previous work has shows that kinetic association rates were accelerated by about one to two orders of magnitude relative to experiment \citep{Hinckley2013AnHybridization}.
+It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not extensively parameterized on dynamical properties properties \citep{Hinckley2013AnHybridization}.
+
+%We also did not find available data on other acceleration studies using explicit ions.
+
+\section{\label{sec:conc}Conclusion}
+
+We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
+We replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
+We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
+We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
+On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
+This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense
+
+Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge relevant faster processes such as duplex nucleation and zippering.
+These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.
+Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process.
+We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations.
+Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence, and were working near memory limits when training SRVs and building SRV-MSMs.
+
+In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
+For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
+In general, coarse-grained models produce a smoother free-energy surface which can results in must faster motions between states.
+This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment.
+Furthermore, it may be easier to cross between states -- e.g.
+a 5S1 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
+
+Going forward, we believe that this work can be extended to predict more general trends in sequence-dependent hybridization and strategies for experimental comparisons.
+These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.
+
+\newcommand{\sectionbreak}{\clearpage}
+\section{\label{sec:Results}Supplemental Information}
+
+% add secion on dmap and high temp methods?
+%\subsection{Additional methods}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_imp/allseq_event_count.png}
+        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in dissociated state.
+The fraction of hybridized duplexes averaged across these sets is shown over time.
+For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive Tmelt difficult to identify.
+The number of hybridization/dissocation combined across trajectories is shown.
+Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics.
+AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.}
+        \label{fig:allseq_event_count}
+	\end{center}
+\end{figure}
+
+\subsection{VAMP-2 Scoring}
+The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
+Hyperparameters, including feature choice, number of microstates, and number of slow modes were evaluated and selection with via the VAMP-2 score.
+In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
+The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
+
+\begin{align*}
+ 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
+ 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
+ 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
+	\\
+ 	VAMP-2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
+ 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
+\end{align*}\label{CK1}
+
+\subsection{SRV training}
+
+Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
+In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
+We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
+In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
+We used a batch size of 50000 and ran each model for a total of 20 training epochs.
+We used two hidden layers and set the size of each layer to 100.
+For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
+SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
+SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
+
+\subsection{SRV-MSMs provide better resolution than tICA-MSMs}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=0.9\textwidth]{Figs/figs_imp/AT-all_implied_timescales.png}
+        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
+        \label{fig:AT-all_dynamic}
+	\end{center}
+\end{figure}
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=80mm, 
+        scale=0.5]{Figs/figs_imp/all_seq_implied_times.png}
+        \caption{SRV-MSM vs.
+tICA-MSM implied timescales convergence for GC-end, GC-core, and GC-mix.
+Progressively larger spectral gaps are observed between the first mode and higher order modes.
+We observe convegenve of resolvable higher order modes at a shared lag time of 1.2 ns.
+It is difficult to converge the leading mode in this regime due to the infrequency of hybridization/dissociation events.}
+        \label{fig:all_seq_lags}
+	\end{center}
+\end{figure}
+
+\subsection{SRV-MSM walkthrough and analysis}
+In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the most discernable spectral gap between modes.
+For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step.
+Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
+We found that most implied timescales converge at an SRV lag time of 1.2 ns.
+Next we selected an optimal number of SRV components to include in our analysis.
+After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
+It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
+For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
+As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
+
+%\subsubsection{SRV-MSM construction and optimization}
+Although SRV coordinates alone provide some information, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
+Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
+We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
+Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
+This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous tICA basis.
+Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
+Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
+Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
+Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=120mm, 
+        scale=0.5]{Figs/figs_imp/allseq_crossval.png}
+        \caption{5-fold cross validation procedure to select number of SRV coordinates.
+We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
+        \label{fig:allseq_srv_crossval}
+	\end{center}
+\end{figure}
+
+Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
+The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most relevant to each process.
+For example, the first slow mode shows a positive correlation to each distance and the strongest correlation with native base pair distances (shown along the main diagonal).
+Given these correlations and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
+The next four SRV components all show a relatively high correlation along offset diagonals.
+These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
+Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
+
+We found GC-end had a similar implied timescales distribution to AT-all, with a distinct spectral gap after the fourth mode.
+For AT-all and GC-end, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
+For GC-core we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
+We found that four macrostate clustering was unstable -- likely because the third mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized (H), dissociated (D), and 4 4 base pair frayed (F4) states.
+The GC-mix sequence showed a similar implied timescale distribution to GC-core, however we no longer saw a converged slow mode corresponding to multi-base fraying behavior.
+Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
+These correlate closely with the first and third GC-core SRV modes.
+Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
+As such, we designated this transitions as effectively two-state within the resolution of our model.
+
+%For GC-core and GC-mix, we found this clustering convention to be unstable -- likely because the last converged mode was providing information about dissociation dynamics -- and performed PCCA+ with one less macrostate.
+
+To visualize macrostates, we project the data into the two leading tICA coordinates.
+Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
+Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
+After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment.
+Means and standard deviations were calculated from this ensemble.
+
+As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach.
+Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model \citep{Sidky}.
+Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
+Similar plots for the other sequences are included in the SI.
+We also note that the infrequency of transitions relative to the individual trajectory length leads to slower convergence of the leading mode.
+Because this mode corresponds the overall hybridization and dissociation process, we found that it displayed similar behavior across all sequences and that higher order modes were more informative for lag time selection.
+
+\subsection{SRV modes reveal hierarchical clustering structure}  % shift to include this in
+
 \begin{figure} %[ht!]
@@
 % not sure if this should directly follow the SRV-MSM GC-core description or directly follow the GC-core dmaps?
-\subsection{SRV correlations to physical coordinates}
+% try cutting this out entirely ?
+%\subsection{SRV correlations to physical coordinates}
 
@@
 We found these GC-core modes to be of particular interest as they reveal the hierarchical and nonlinear nature of the dynamical encoding.
-In particular, we examined a collection of "trimmed" trajectories centered on both hybridization or dissociation events.
+We examined a collection of "trimmed" trajectories centered on both hybridization or dissociation events.
 For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they shared behavior.
@@
 
-% note that similar analysis on AT-all and GC-end can be found in the SI (move both dynamics plots and descriptions there)
-
-\subsection{\label{sec:Results}Temperature-dependent timescale comparisons}
-
-Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone.
-To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence.
-We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core.
-Additionally, we repeated the protocol in \citet{Sanstead2018DirectDehybridization} to generate temperature series data for AT-all and GC-mix sequences.
-
-We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a cutoff.
-We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
-Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature which can vary between simulation experiments.
-After accounting for the acceleration factor, we saw agreement between the experimental data and simulated relaxation fits.
-GC-core showed the largest deviation experiment, with dissociation rate increasing much more quickly with temperature in simulations.
-
-% GC-core fast timescales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
-% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor.
-We maintained a consistent accerelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses.
-It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.
-
-\begin{figure}  %[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
-        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
-The effective simulation temperature is shifted by 4 degrees K to account for systematic differences in simulation and experimental melting temperature.
-Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
-        \label{fig:relaxation-comparison}
-	\end{center}
-\end{figure}
-
-% A documented short-coming for the 3spn2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
-This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
-
-The fast response -- which Sanstead et.
-al attributed to base pair fraying signatures -- was more difficult to compare against simulation observables.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
-All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
-Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
-Instead, we measured the fast fraying response by counting frames until either duplex terminal end to split beyond a cutoff.
-This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
-This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events.
-Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
-
-For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
-In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater thermodynamic barrier associated with G:C fraying.
-We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}).
-We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about two order of magnitude relative to experiments.
-It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
-In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation.
-Although we should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
-and that 3spn2 can capture some sequence-dependent fraying effects.
-
-%Again, GC-core behavior deviates from simulations prediction with consistently lower than expected fast responses.
-We believe this could be due to slower contributions from internal A:T base pairs sequentially fraying.
-
-% Experiments fast responses are difficult to discern at higher temperatures due to mixing with slow responses. This is especially true for GC-end because to the signals are closer to begin with.
-% add new info here, not sure if shifting discussion is necessary
-% take away message is that these are good proxies for the responses
-% although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response
-
-%In addition to fraying, we notice that the rate of shifting increases at higher temperature.
-Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
-For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
-It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
-
-%The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
-Previous work has shows that kinetic association rates were accelerated by about one to two orders of magnitude relative to experiment \citep{Hinckley2013AnHybridization}.
-It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not extensively parameterized on dynamical properties properties \citep{Hinckley2013AnHybridization}.
-
-%We also did not find available data on other acceleration studies using explicit ions.
-
-\section{\label{sec:conc}Conclusion}
-
-We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
-We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
-We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
-On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense.
-Furthermore, we replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
-
-\subsection{\label{sec:conc}Limitations}
-
-Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge relevant faster processes such as duplex nucleation and zippering.
-These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.
-Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process.
-We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations.
-Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence, and were working near memory limits when training SRVs and building SRV-MSMs.
-
-In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
-For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
-In general, coarse-grained models produce a smoother free-energy surface which can results in must faster motions between states.
-This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment.
-Furthermore, it may be easier to cross between states -- e.g.
-a 5S1 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
-
-\subsubsection{\label{sec:conc}Not sure whether to discuss explicit comparisons} 
+\subsection{Explicit model comparisons}
+
 We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
@@
 
-Going forward, we believe that this work can be extended to predict more general trends in sequence-dependent hybridization and strategies for experimental comparisons.
-These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.
-
-\section{\label{sec:Results}SI Methods notes}
-
-% add secion on dmap and high temp methods?
-%\subsection{Additional methods}
-
-%In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff.
-
-\subsection{VAMP-2 Scoring}
-Optimization of hyperparameters including feature choice, number of microstates, and number of slow modes was evaluated with via the VAMP-2 score.
-In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
-The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
-
-\begin{align*}
- 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
- 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
- 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
-	\\
- 	VAMP-2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
- 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
-\end{align*}\label{CK1}
-
-\subsection{SRV training}
-
-Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
-In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
-We used a batch size of 50000 and ran each model for a total of 20 training epochs.
-We used two hidden layers and set the size of each layer to 100.
-For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
-SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
-
-\subsection{SRV-MSM walkthrough and analysis}
-In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the simplest dynamics and displayed a clear spectral gap between modes.
-For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step.
-Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that most implied timescales converge at an SRV lag time of 1.2 ns.
-Next we selected an optimal number of SRV components to include in our analysis.
-After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
-It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
-For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
-
-%\subsubsection{SRV-MSM construction and optimization}
-Although SRV coordinates alone provide some information, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
-Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
-We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
-Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
-This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous tICA basis.
-Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
-Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
-Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
-Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
-
-Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
-The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most relevant to each process.
-For example, the first slow mode shows a positive correlation to each distance and the strongest correlation with native base pair distances (shown along the main diagonal).
-Given these correlations and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
-The next four SRV components all show a relatively high correlation along offset diagonals.
-These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
-Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
-
-We found GC-end had a similar implied timescales distribution to AT-all, with a distinct spectral gap after the fourth mode.
-For AT-all and GC-end, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
-For GC-core we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
-We found that four macrostate clustering was unstable -- likely because the third mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized (H), dissociated (D), and 4 4 base pair frayed (F4) states.
-The GC-mix sequence showed a similar implied timescale distribution to GC-core, however we no longer saw a converged slow mode corresponding to multi-base fraying behavior.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
-These correlate closely with the first and third GC-core SRV modes.
-Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as effectively two-state within the resolution of our model.
-
-%For GC-core and GC-mix, we found this clustering convention to be unstable -- likely because the last converged mode was providing information about dissociation dynamics -- and performed PCCA+ with one less macrostate.
-
-To visualize macrostates, we project the data into the two leading tICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
-Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
-After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment.
-Means and standard deviations were calculated from this ensemble.
-
-\subsection{\label{sec:Results}SI figs}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_imp/allseq_event_count.png}
-        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in dissociated state.
-The fraction of hybridized duplexes averaged across these sets is shown over time.
-For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive Tmelt difficult to identify.
-The number of hybridization/dissocation combined across trajectories is shown.
-Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics.
-AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.}
-        \label{fig:allseq_event_count}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_imp/allseq_crossval.png}
-        \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
-        \label{fig:allseq_srv_crossval}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=80mm, 
-        scale=0.5]{Figs/figs_imp/all_seq_implied_times.png}
-        \caption{SRV-MSM vs.
-tICA-MSM implied timescales convergence for GC-end, GC-core, and GC-mix.
-Progressively larger spectral gaps are observed between the first mode and higher order modes.
-We observe convegenve of resolvable higher order modes at a shared lag time of 1.2 ns.
-It is difficult to converge the leading mode in this regime due to the infrequency of hybridization/dissociation events.}
-        \label{fig:all_seq_lags}
-	\end{center}
-\end{figure}
-
 \begin{figure}[ht!]
@@
         scale=0.5]{Figs/figs_imp/GCSeq_TJTraces.png}
-        \caption{Kinetic traces of the fast spectroscopic response show varied degrees of shifting based on sequences.
-More info from Brennan on subtle differences here, discuss stretching coefficients?
-(could include this in main text as well)}
+        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations.
+The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions.
+Each T-jump was performed with a temperature change of ~15 C to a final temperature near the Tm of each respective sequence.}
         \label{fig:fast_stretching}
@@
 \begin{comment}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/fast-slow_all-mix.png}
-        \caption{Slow and Fast fits for GC-mix and AT-all, no acceleration values were calculated as there was not the same temperature-dependent data available for these sequences.
-GC-mix again shows barrierles fraying at the terminus, with a noticeable barrier for G:C bonds breaking at the second base pair in.
-AT-all shows a more exponential relationship for terminal fraying, likely because the simulations are run at lower relative temperatures.}
-        \label{fig:relaxation-comparison_mix_and_all}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_0804/relaxation_core_and_end_allcuts.PNG}
-        \caption{Fits across multiple cutoffs and basepairs for GC-core and GC-end fast response. (This is a placeholder until I reformat these to match the main text).}
-        \label{fig:relaxation_core_and_end_allcuts}
-	\end{center}
-\end{figure}
 

### 2020-12-01 manuscript_1125.tex
Updates from Overleaf

--- 2020-11-23 manuscript.tex
@@
 Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
-Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
-
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
+Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
+
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes back to the hybridized state \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
 More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
-Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
+Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
 Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
@@
 Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
-Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
+Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
 Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.
@@
 in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
-We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
-We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another.
-Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system  -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
+We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
+We synthesize the simulation trajctoeis using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
 MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
 \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent.
-We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
-Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed.
-We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
-
-We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior.
+We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
+
+% this section should look more like a conclusion foreshadowing conclusions and main take aways
+Our MSM analyses reveal G:C base pair placement in decamer oligonucleotides to have a substantial effect on dynamical behavior.
 By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
@@
 
-All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}.
-Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other.
+Some featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
+Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library \citep{McGibbon2015MDTraj:Trajectories}.
+Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
 To account for this, we averaged permutable distances (45 pairs in total) together following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
@@
 	\begin{center}
-        \includegraphics[width=400, scale=1]{Figs/figs_imp/AT-all_implied_timescales.png}
+        \includegraphics[width=0.9\textwidth]{Figs/figs_imp/AT-all_implied_timescales.png}
         \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}

### 2020-12-01 manuscript.tex
Updates from Overleaf

@@
 
-%%%%%%%% Intro comments %%%%%%%%%%%%%%%%%
-
-%Progress in DNA science and engineering has been bolstered by a well-established understanding of sequence-dependent thermodynamics.
-Thousands of DNA and DNA-complex crystal structures have been captured, and the equilibrium structure of every tetranucleotide motif were simulated in a distributed computational effort \citep{Manuscript2014The2012}.
-Hybridization thermodynamics can predicted with high fidelity by nearest-neighbor stacking, and these calculations have been experimentally verified to account for mismatched pairs, dangling ends, and other non-native motifs \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
-
-% Secondary DNA structures such as hairpins and quadruplexes have also been studied in depth and leverage for nanotechnology applications.
-
-%Furthermore several computational models have been paramaterized to capture these thermodynamic characteristics.
-
 %These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}.
-Certain dynamical models for hybridization -- such the nucleation-zipper mechanism and fraying-peeling mechanism for dissociation -- have been well-established for decades (\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition}).
-Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
-Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
 
@@
 Experimentally, however, it is difficult to pick up on dynamics that have similar spectroscopic signatures and to tease apart temporally close responses that might be combined into one stretched exponential signal.
-
-% introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}.
-Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed.
-We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).
 
@@
         \includegraphics[width=80mm,
-        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
+        scale=0.5]{Fig1.pdf}
         \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates.
@@
 While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
-Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is again sensitive to the exact definition of melting temperature.
-After accounting for the melting temperature shift and acceleration factor, we saw agreement between the experimental data and simulated relaxation fits.
-GC-core showed the largest deviation experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
+We used to this temperature shift to inform our equilibrium simulations in the next section.
+We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
+After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits.
+At lower temperatures, T-jump slow responses likely contain a mix of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone.
+GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
+        \includegraphics[width=\textwidth]{Fig2.pdf}
         \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
@@
 Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
-Instead, we measured the fast fraying response by counting frames until either duplex terminal end splits beyond a cutoff.
+Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff.
 This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
@@
         \includegraphics[width=160mm, 
-        scale=1]{Figs/figs_imp/allseq_states.png}
+        scale=1]{Fig3.pdf}
         \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates.
@@
 3') and the number of shifted motifs relative the native state (1 vs.
-2)The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
+2).}
         \label{fig:allseq_table}
@@
 % potentially get rid of State representation, ask Brennan if they are meaningful?
-\begin{figure}[hb!]
-	\begin{center}
-        \includegraphics[width=160mm, 
-        scale=1]{Figs/figs_imp/allseq_transitions.png}
-        \caption{Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions.
-Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each state.
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=120mm, 
+        scale=1]{Fig4.pdf}
+        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate.
 Arrows indicate the probability of transitioning out of the state within the lag time.
 Circle areas are proportional to state stability.}
+
+        %Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions. }
         \label{fig:allseq_transition}
@@
 Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities.
-By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 3S1, 5S2, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
 Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
@@
 It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
-In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN calculations (figure \ref{fig:NN_table}).
+In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}).
 For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
@@
 	\begin{center} 
-        \includegraphics[width=100mm, scale=1]{Figs/figs_imp/shifted_FE_table.png}
+        \includegraphics[width=120mm, scale=1]{Fig5.pdf}
         \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations.
@@
 Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
-In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path to native hybridization.
-Indeed, for GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
-The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D.
+In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path towards native hybridization.
+
+For GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
+The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
 This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.
@@
 Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
-A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways pass through some out-of-register state.
-While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states.
+A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
+The more shifted 3S2 and 5s2 states are more prone to dissociation and thus contribute to about 5\% of these pathways.
+While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
 These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
@@
 Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
-In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and a more stretched fast response.
-
-%The origin of the AT-all and GC-end fast response are most difficult to assign because both shifting and fraying could occur, but the more stretched profile of the fast response for AT-all and GC-end suggests that there may be a combination of processes occurring on the fast timescale.
-Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble.
-Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two.
-In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.
-
-%In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses.
-For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale.
-We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
-Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble.
-Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two.
-In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.
-We interrogate these differences by examining high temperature simulations later in this manuscript.
-(Can discuss more details about fig \ref{fig:fast_stretching} here and potentially include this fig in the main text)?
+Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response.
+(Can briefly mention further experimental work to isolate shifting responses here?)
+
+% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
 
 %It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
-%Although reproducing these stretching responses is not within the scope of this analysis, we take these
 
@@
 We observe these events to occur with equal probability on either permutable end of the helix.
+
 The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D.
@@
 On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
+We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently then in experiment.
+Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.
 
@@
 
-\subsection{GC-mix displays more canonical hybridization and dissociation}
+\subsection{Heterogeneous base pair distribution promotes canonical hybridization and dissociation}
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
+        \includegraphics[width=450, scale=1]{Fig6.pdf}
         \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps.PNG}
+        \includegraphics[width=450, scale=1]{Fig7.pdf}
         \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
@@
 % need to gear this more towards the kinetics
-\subsubsection{5s1 state distributions inform differences between AT-all and GC-end kinetics}
+\subsection{5s1 state distributions reveal differences between AT-all and GC-end kinetics}
+
 While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.PNG}
-        \caption{Plotting the first two non-trivial diffusion map eigenfunctions.
+        \includegraphics[width=450, scale=1]{Fig8.pdf}
+        \caption{Diffusion map representation configurations sampled from the F4 state.
+The x and y axes represent the the first two non-trivial diffusion map eigenfunctions.
 Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
@@
 
-\subsubsection{The F4 state interrupts GC-core hybridization and dissociation}  % shift to include this in the overall diffusion map section
+\subsection{Structural diversity in the F4 state informs fraying dynamics}  % shift to include this in the overall diffusion map section
 %\subsubsection{Diffusion maps show ensemble of frayed states}
 
-We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
+(might just want to move this to the SI) We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
 We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
@@
 
-\section{\label{sec:conc}Conclusion}
-
-We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
-We replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
-We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
-We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
-On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense
+\subsection{Limitations}
 
@@
 
-Going forward, we believe that this work can be extended to predict more general trends in sequence-dependent hybridization and strategies for experimental comparisons.
+\section{\label{sec:conc}Conclusion}
+
+We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
+We replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
+We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
+We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
+On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
+This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense.
+
+Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
 These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.

### 2020-12-03 manuscript.tex
updates dmap section

@@
 
-%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
-This reveals that the mismatched C:T pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
-The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all.
-Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole.
-These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations is confirmed by molecular renderings sampled from these regions.
-Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
+The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
+As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
+The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, highlights structural variances within the two sequences distributions.
+We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration.
+We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register.
+These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated.
+This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state lifetime being substantially shorter than the AT-all equivalent.
+
+%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
+
+%Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
 
@@
 
-\subsection{Structural diversity in the F4 state informs fraying dynamics}  % shift to include this in the overall diffusion map section
+\subsection{Structural diversity in the F4 state informs fraying dynamics} 
+(might just want to move this to the SI)
+
+% shift to include this in the overall diffusion map section
 %\subsubsection{Diffusion maps show ensemble of frayed states}
 
-(might just want to move this to the SI) We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
+We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
 We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
-This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
 Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
@@
 We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
-In these conformations, one of the strands maintains some helical character while the other twists out of place, resulting in WC bonds being obstructed by the oligo backbone.
-These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated.
-We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart.
-This might be another consequence of differential excluded volume radii in the 3spn2 force field.
+At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place.
+Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation.
+Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
+
+%This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
 

### 2020-12-04 manuscript.tex
added new direct hybridization dmap analysis

@@
 
-%%%%%%% start of "interesting phenomena" section   %%%%% 
 \subsection{Macrostate distributions inform state stability and kinetics}
@@
 
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=300, scale=1]{Fig9.pdf}
+        \caption{Sequence-dependent population distributions for trimmed hybridization events.
+$Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression.
+Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins.
+Heat maps show the average intermolecular base distances in nm at increasing values of $Psi_2$, where the main diagonal shows native base pair distances.}
+        \label{fig:Fig9}
+	\end{center}
+\end{figure}
+
+\subsection{Differential fraying behavior determines direct hybridization pathways}
+
+(I just completed most of this analysis on Friday.
+I think this shows a more direct comparison to the Tokmakoff kinetic model -- mainly all-or-nothing deviations caused by fraying are ordered GC-core > GC-mix > AT-all > GC-end.
+In addition we see a distinct state form for GC-core which serves to verify the F4 state and its kinetic relevance.
+Lastly it provides another means by which MSMs can inform analysis techniques for clearer comparison between sequences.
+I could repeat this procedure and analysis for direct dissociation as well if that is of interest)
+
+To more directly compare hybridization events across all sequences, we discarded transitions that passed through long-lived out-of-register states.
+To achieve higher resolutions, we reran simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) transitions with a 10x higher save rate.
+We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences where each event consisted of 1000 frames.
+We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance.
+Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time a -0.94 correlation with strand core distance with minimal variation by sequence.
+This shows that $\Psi_2$ is strongly correlated with the hybridization process for all sequences and represents a consistent collective variable for comparison between sequences.
+For each sequence, we compared population distributions along $\Psi_2$ and noted statistically meaningful differences in the range $ 0.001 <\Psi_2 < 0.0035$ (Figure \ref{fig:GC-core_dmaps}).
+In this region, some base pair contacts have formed but frayed structures dominate the distribution.
+As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs.
+GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state.
+This is consistent with a the downhill "nucleation-zippering" mechanism noted above.
+
 %This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
 
-% note that similar analysis on AT-all and GC-end can be found in the SI (move both dynamics plots and descriptions there)
-
-% Additional notes on the temp-dependent secttion
-%Again, GC-core behavior deviates from simulations prediction with consistently lower than expected fast responses.
-We believe this could be due to slower contributions from internal A:T base pairs sequentially fraying.
-
-% Experiments fast responses are difficult to discern at higher temperatures due to mixing with slow responses. This is especially true for GC-end because to the signals are closer to begin with.
 % add new info here, not sure if shifting discussion is necessary
@@
 % although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response
-
-%In addition to fraying, we notice that the rate of shifting increases at higher temperature.
-Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex.
-For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above.
-It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.
-
-%The temperature-dependent analysis can also inform out interpretation of the 3spn2 model.
-Previous work has shows that kinetic association rates were accelerated by about one to two orders of magnitude relative to experiment \citep{Hinckley2013AnHybridization}.
-It is not surprising that different dynamics might be accelerated at different rates, especially considering that 3spn2 was not extensively parameterized on dynamical properties properties \citep{Hinckley2013AnHybridization}.
-
-%We also did not find available data on other acceleration studies using explicit ions.
 

### 2020-12-07 manuscript.tex
Updates from Overleaf
move f4 to supplement, added more refs, fixed fig size

@@
 We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
-We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.
+We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}.
 We noted an average temperature shift of about 4C between experiment and simulation exponentials.
 While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
-We used to this temperature shift to inform our equilibrium simulations in the next section.
+We used to this temperature shift to inform our equilibrium simulations in the proceeding section.
 We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
@@
 For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
-In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater thermodynamic barrier associated with G:C fraying.
+In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying.
 We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}).
@@
 The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
-This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.
+This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.
+This phenomenon will be explored more in depth below.
+
+% This mechanism is explored more in depth later on?
 
@@
 
-Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
-A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
-The more shifted 3S2 and 5s2 states are more prone to dissociation and thus contribute to about 5\% of these pathways.
+Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
+A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
+The 3S2 and 5s2 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.
 While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
-These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
-Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and response similarity to other processes such as fraying.
+These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior.
+Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences.
+Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
+These processes can occur even when terminal mismatches are present in the non-native structure.
+
+Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying.
 To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}.
-Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}.
+Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}.
 Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
-It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs.
-This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
-
-\subsection{Stretched T-jump IR responses correspond to shifting sequences}
+It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-obref-register transitions compared to AT motifs.
+This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
+
+\subsection{Out-of-register dynamics may stretch T-jump IR responses}
 
@@
 % discuss computational comparisons needs work
-Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019}.
+Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA}.
 Araque et al.
 defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}.
-<Add Brennan example> When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
 found that the GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}.
@@
 %Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.
+
+\begin{figure} %[ht!]
+	\begin{center}
+        \includegraphics[width=450, scale=1]{Fig6.pdf}
+        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
+Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs.
+The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined.
+Fluctuation between states is common during the transition.}
+        \label{fig:GC-mix_transitions}
+	\end{center}
+\end{figure}
 
@@
 
-\begin{figure} %[ht!]
-	\begin{center}
-        \includegraphics[width=450, scale=1]{Fig6.pdf}
-        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
-Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs.
-The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined.
-Fluctuation between states is common during the transition.}
-        \label{fig:GC-mix_transitions}
-	\end{center}
-\end{figure}
-
-\subsection{Macrostate distributions inform state stability and kinetics}
-
-% add new diffusion map section introduction
-Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates.
-These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population.
-Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most populated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.
-
 \begin{figure}[ht!]
@@
 \subsection{5s1 state distributions reveal differences between AT-all and GC-end kinetics}
+
+Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods.
+These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population.
+Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
 
@@
 These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated.
-This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state lifetime being substantially shorter than the AT-all equivalent.
+This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.
 
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=450, scale=1]{Fig8.pdf}
-        \caption{Diffusion map representation configurations sampled from the F4 state.
-The x and y axes represent the the first two non-trivial diffusion map eigenfunctions.
-Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
-        \label{fig:GC-core_dmaps}
-	\end{center}
-\end{figure}
-
-\subsection{Structural diversity in the F4 state informs fraying dynamics} 
-(might just want to move this to the SI)
-
-% shift to include this in the overall diffusion map section
-%\subsubsection{Diffusion maps show ensemble of frayed states}
-
-We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
-We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
-Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
-We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate.
-This distance also correlated closely with the second diffusion map mode.
-Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core.
-We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
-At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place.
-Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation.
-Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=300, scale=1]{Fig9.pdf}
+\subsection{Sequence-dependent fraying determines direct hybridization pathways}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=300, scale=1]{Fig8.pdf}
         \caption{Sequence-dependent population distributions for trimmed hybridization events.
@@
 Heat maps show the average intermolecular base distances in nm at increasing values of $Psi_2$, where the main diagonal shows native base pair distances.}
-        \label{fig:Fig9}
-	\end{center}
-\end{figure}
-
-\subsection{Differential fraying behavior determines direct hybridization pathways}
+        \label{fig:Fig8}
+	\end{center}
+\end{figure}
 
 (I just completed most of this analysis on Friday.
-I think this shows a more direct comparison to the Tokmakoff kinetic model -- mainly all-or-nothing deviations caused by fraying are ordered GC-core > GC-mix > AT-all > GC-end.
+I think this shows a more direct comparison to Paul's 2016 kinetic model -- mainly all-or-nothing deviations caused by fraying are ordered GC-core > GC-mix > AT-all > GC-end.
 In addition we see a distinct state form for GC-core which serves to verify the F4 state and its kinetic relevance.
@@
 
-To more directly compare hybridization events across all sequences, we discarded transitions that passed through long-lived out-of-register states.
-To achieve higher resolutions, we reran simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) transitions with a 10x higher save rate.
-We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences where each event consisted of 1000 frames.
+To more directly compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
+To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event.
+We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences.
 We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance.
-Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time a -0.94 correlation with strand core distance with minimal variation by sequence.
-This shows that $\Psi_2$ is strongly correlated with the hybridization process for all sequences and represents a consistent collective variable for comparison between sequences.
-For each sequence, we compared population distributions along $\Psi_2$ and noted statistically meaningful differences in the range $ 0.001 <\Psi_2 < 0.0035$ (Figure \ref{fig:GC-core_dmaps}).
+Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence.
+This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.
+
+For each sequence, we compared population distributions along $\Psi_2$ and noted statistically meaningful differences in the range $ 0.001 <\Psi_2 < 0.0035$ (Figure \ref{fig:Fig8}).
 In this region, some base pair contacts have formed but frayed structures dominate the distribution.
@@
 GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state.
-This is consistent with a the downhill "nucleation-zippering" mechanism noted above.
+This is consistent with the downhill "nucleation-zippering" mechanism described above.
+Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et.
+al.
+regarding fdeviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}.
+This qualitative agreement shows that our MSM information can be adapted to identify dynamic behavior within the lag time and explore different definitions of two-state deviations.
 
@@
 For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
-In general, coarse-grained models produce a smoother free-energy surface which can results in must faster motions between states.
+In general, coarse-grained models produce a smoother free-energy surface which can results in much faster motions between states.
 This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment.
@@
 
-\subsection{SRV training}
+\subsection{SRV hyperparameters}
 
@@
 Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that most implied timescales converge at an SRV lag time of 1.2 ns.
+We found that an SRV lag time of 1.2 ns showed good convergence across the relevant timescales for all sequences.
 Next we selected an optimal number of SRV components to include in our analysis.
@@
 
+\subsection{Structural diversity in the F4 state informs fraying dynamics}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=450, scale=1]{Fig9.pdf}
+        \caption{Diffusion map representation configurations sampled from the F4 state.
+The x and y axes represent the the first two non-trivial diffusion map eigenfunctions.
+Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
+        \label{fig:GC-core_dmaps}
+	\end{center}
+\end{figure}
+
+% shift to include this in the overall diffusion map section
+%\subsubsection{Diffusion maps show ensemble of frayed states}
+
+We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
+We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
+Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
+We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate.
+This distance also correlated closely with the second diffusion map mode.
+Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core.
+We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
+At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place.
+Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation.
+Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
+
 \subsection{Explicit model comparisons}

### 2020-12-09 manuscript.tex
added Brenna fig, and reformatted

@@
 More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
-Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}.
+Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}.
 Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
@@
 
-We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}.
+We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}.
 We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
@@
 
-% add intro sentence here before jumping into data -- keep working on this
-%AT-all responses are less clear given that there are no G:C base pairs for comparison.
+\begin{figure}[ht!] % right place for NN table? could go to SI
+	\begin{center} 
+        \includegraphics[width=150mm, scale=1]{Fig6.pdf}
+        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations.
+The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions.
+Each T-jump was performed with a temperature change of 15 °C to a final temperature near the Tm of each respective sequence.
+Each time trace was fit to the sum of a stretched exponential and two exponentials (solid lines).
+The stretched exponential describes the component from 10 ns to 1 μs ($tau_{Fast}$), and the two exponentials describe the signal rise from 1 - 300 μs ($tau_{Slow}$) and thermal relaxation of the sample back to the initial temperature.
+The stretch factor (β) from each fit is listed.}
+
+        \label{fig:Fig6}
+	\end{center}
+\end{figure}
 
 In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response.
-In particular, the stretched fast response for AT-all and GC-end indicate that there may be a combination of processes occurring on the fast timescale.
+In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}).
 Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response.
-On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T vibration that points to A:T fraying as the predominant dynamic source.
-Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
+On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source.
+We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-ends and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
 Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response.
-(Can briefly mention further experimental work to isolate shifting responses here?)
 
@@
 	\begin{center}
-        \includegraphics[width=450, scale=1]{Fig6.pdf}
+	\includegraphics[width=150mm, scale=1]{Fig7.pdf}
         \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
@@
 
-\subsection{Heterogeneous base pair distribution promotes canonical hybridization and dissociation}
+\subsection{Heterogeneous base pair distribution promotes nucleation-zippering hybridization and dissociation}
 
@@
 
-To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}).
+To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (Figure \ref{fig:GC-mix_transitions}).
 During hybridization, we observed the formations of some key base pair contacts before the full duplex formed.
@@
 	\begin{center}
-        \includegraphics[width=450, scale=1]{Fig7.pdf}
+        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
         \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
-Color maps show inverse distances between out-of-register ends and complementary ends.}
+Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends.
+The lower right region of the diffusion map space shows a high concentration of "shifted loop" base pairs that may facilitate the out-of-register GC-end transition.}
         \label{fig:GC-end_dmaps}
@@
 	\begin{center}
-        \includegraphics[width=300, scale=1]{Fig8.pdf}
+        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
         \caption{Sequence-dependent population distributions for trimmed hybridization events.
-$Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression.
+$\Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression.
 Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins.
-Heat maps show the average intermolecular base distances in nm at increasing values of $Psi_2$, where the main diagonal shows native base pair distances.}
+Heat maps show the average intermolecular base distances in nm at increasing values of $\Psi_2$, where the main diagonal shows native base pair distances.}
         \label{fig:Fig8}
@@
 
-For each sequence, we compared population distributions along $\Psi_2$ and noted statistically meaningful differences in the range $ 0.001 <\Psi_2 < 0.0035$ (Figure \ref{fig:Fig8}).
-In this region, some base pair contacts have formed but frayed structures dominate the distribution.
+Once we determined a strong hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
+The $ <\Psi_2 < 0$ reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence once first WC contacts are made.
+In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $<\Psi_2$ increases.
 As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs.
 GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state.
-This is consistent with the downhill "nucleation-zippering" mechanism described above.
+This is consistent with the downhill nucleation-zippering mechanism described above.
 Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et.
 al.
-regarding fdeviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}.
-This qualitative agreement shows that our MSM information can be adapted to identify dynamic behavior within the lag time and explore different definitions of two-state deviations.
+regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}.
+This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
+Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.
 
@@
 
-We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization.
-We replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales.
-We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results.
-We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
+We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics.
+By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior.
+In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
 On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense.
-
+Our computational approach and results show strong comparisons T-jump experiments at ns and us timescales.
 Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
@@
 	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_imp/GCSeq_TJTraces.png}
-        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations.
-The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions.
-Each T-jump was performed with a temperature change of ~15 C to a final temperature near the Tm of each respective sequence.}
-        \label{fig:fast_stretching}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
         \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.PNG}
@@
 
-%\bibliography{references}
-\bibliography{refs_mendeley2}
+\bibliography{references}
+%\bibliography{refs_mendeley2}
 

### 2020-12-10 manuscript.tex
spellcheck

@@
 While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
-We used to this temperature shift to inform our equilibrium simulations in the proceeding section.
+We used this temperature shift to inform our equilibrium simulations in the proceeding section.
 We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
 After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits.
-At lower temperatures, T-jump slow responses likely contain a mix of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone.
+At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone.
 GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
@@
 
-Following the analysis pipeline described above, we generate SRV-MSMs for each sequence.
+Following the analysis pipeline described above, we generated SRV-MSMs for each sequence.
 We identified seven kinetically relevant states that are captured within the resolution of the model.
@@
 Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
-It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-obref-register transitions compared to AT motifs.
+It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs.
 This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
@@
 We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-ends and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
-Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response.
+Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
 
@@
 The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D.
-Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed.
-On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
+Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed.
+On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
 We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently then in experiment.
@@
 As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}.
+While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs \citep{Phys2019}.
 Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
 We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
-Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
+
+%Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
 
@@
 	\begin{center}
-	\includegraphics[width=150mm, scale=1]{Fig7.pdf}
+	\includegraphics[width=170mm, scale=1]{Fig7.pdf}
         \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
-Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs.
+Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until full dissociation occurs.
 The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined.
@@
 
-To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (Figure \ref{fig:GC-mix_transitions}).
+To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near MSM-identified hybridization and dissociation events.
+We present two representative events in Figure \ref{fig:GC-mix_transitions} along with all native base pair distances over time.
 During hybridization, we observed the formations of some key base pair contacts before the full duplex formed.
@@
 There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs.
-In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes.
+In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes.
 As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}.
@@
 
-Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods.
-These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population.
+Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
+These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population.
 Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+% add something here on previous dmap refs?
 
@@
 This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands.
-In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}).
-Diffusion maps built from samples of the 3S1 states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).
+In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI.
+Diffusion maps built from samples of the 3S1 states are also shown in the SI.
 
@@
 Once we determined a strong hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
-The $ <\Psi_2 < 0$ reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence once first WC contacts are made.
-In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $<\Psi_2$ increases.
+The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first WC contacts are made.
+In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases.
 As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs.
@@
 By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior.
-In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all.
+In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these dynamics are more relevant for AT-all.
 On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-Our computational approach and results show strong comparisons T-jump experiments at ns and us timescales.
+Our computational approach and results show strong comparisons to T-jump experiments at ns and us timescales.
 Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
@@
 The fraction of hybridized duplexes averaged across these sets is shown over time.
-For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive Tmelt difficult to identify.
+For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive $T_m$ difficult to identify.
 The number of hybridization/dissocation combined across trajectories is shown.

### 2020-12-18 manuscript.tex
Updates from Overleaf

@@
 Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides, Roman o2013DNADependence, Hinckley2014Coarse-grainedEffects}.
 Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
@@
 We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
-
 % this section should look more like a conclusion foreshadowing conclusions and main take aways 
@@
 Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation.
-Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer folding dynamics populated by a diversity of long-lived metastable states.
+Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer dynamics populated by a diversity of long-lived metastable states.
 Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior.
 Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process.
-Analysis within metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs.
+Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs.
 This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence.
@@
 
-We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}.
+We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016}.
 We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
@@
 
+\subsection{\label{sec:methods}Diffusion Maps}
+Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
+We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies \citep{Shmilovich, Wang2017MesoscaleLandscapes}.
+We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.
+
 %%%%%%%% Methods comments %%%%%%%%%%%%%%%%%
@@
 Shifted state abbreviations indicate the direction of the shifted overhang (5' vs.
-3') and the number of shifted motifs relative the native state (1 vs.
-2).}
+3') and the number of shifted base pairs relative the native state (2 vs.
+4).}
         \label{fig:allseq_table}
@@
 	\begin{center}
-        \includegraphics[width=120mm, 
+        \includegraphics[width=160mm, 
         scale=1]{Fig4.pdf}
@@
 Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities.
-By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 3S1, 5S2, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
 Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
@@
 In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}).
-For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2).
+For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4).
 We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone.
@@
 Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.
-Contrary to macrostate results, however, the 3S1 state is more stable than 5S1 states.
-Moreover, we do not observe a stable 3S2 cluster for GC-end.
+Contrary to macrostate results, however, the 3S2 state is more stable than 5S2 states.
+Moreover, we do not observe a stable 3S4 cluster for GC-end.
 This indicates that 5' vs.
@@
 % Although NN calculations predict 5' vs.
-3' stabilities for AT-all, we observe that GC-end 5S1 state remains more populated than the 3S1 state contrary to NN predictions.
-Indeed, we do not observe the formation of a stable 3S2 state for GC-end, further indicating that these states 3' states are unfavorable.
+3' stabilities for AT-all, we observe that GC-end 5S2 state remains more populated than the 3S2 state contrary to NN predictions.
+Indeed, we do not observe the formation of a stable 3S4 state for GC-end, further indicating that these states 3' states are unfavorable.
 This leads us to believe that 5' vs.
@@
 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
-Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
-In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path towards native hybridization.
-
-For GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H.
-The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
-This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.
+Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S2 $\rightarrow$ H is more favorable than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
+Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
+In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.
+
+For GC-end we observe no significant flux between the 5S4 state and the structurally similar 5S2 state, indicating that the former acts more as a kinetic trap than a pathway to H.
+The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
+This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed.
 This phenomenon will be explored more in depth below.
@@
 A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
-The 3S2 and 5s2 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.
+The 3S4 and 5S4 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.
 While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
@@
         \includegraphics[width=150mm, scale=1]{Fig8.pdf}
-        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end.
+        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end.
 Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends.
@@
 % need to gear this more towards the kinetics
-\subsection{5s1 state distributions reveal differences between AT-all and GC-end kinetics}
+\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}
 
 Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
-These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population.
-Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-% add something here on previous dmap refs?
-
-While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
+These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
 We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding.
-To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences.
+To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S2 state of both sequences.
 We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
@@
 In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI.
-Diffusion maps built from samples of the 3S1 states are also shown in the SI.
+Diffusion maps built from samples of the 3S2 states are also shown in the SI.
 
@@
 These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated.
-This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.
+This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.
 
@@
 
-(I just completed most of this analysis on Friday.
-I think this shows a more direct comparison to Paul's 2016 kinetic model -- mainly all-or-nothing deviations caused by fraying are ordered GC-core > GC-mix > AT-all > GC-end.
-In addition we see a distinct state form for GC-core which serves to verify the F4 state and its kinetic relevance.
-Lastly it provides another means by which MSMs can inform analysis techniques for clearer comparison between sequences.
-I could repeat this procedure and analysis for direct dissociation as well if that is of interest)
-
-To more directly compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
+To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
 To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event.
@@
 
-Once we determined a strong hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
+Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
 The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first WC contacts are made.
@@
 Furthermore, it may be easier to cross between states -- e.g.
-a 5S1 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
+a 5S2 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
 
@@
         scale=0.5]{Figs/figs_imp/allseq_event_count.png}
-        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in dissociated state.
-The fraction of hybridized duplexes averaged across these sets is shown over time.
+        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in the dissociated state.
+Here we show fraction of hybridized duplexes averaged across these sets over time.
 For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive $T_m$ difficult to identify.
@@
 Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics.
-AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.}
+AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.
+GC-core displays the most native hybridization events, perhaps due to stronger nucleation sites at the central G:C contact points.}
         \label{fig:allseq_event_count}
+	\end{center}
+\end{figure}
+
+% add figure for slow and fast response fitting (just GC-core)
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.png}
+        \caption{Slow and Fast modes were determined by fitting over a distribution of 120 independent simulations.
+For each temperature in the series, a relaxation curve was fit to the distribution and the associated exponential coefficient was used to the calculate rate.
+Temperatures series between -5 - +15K relative to empirically determined sequence melting temperature were explored.
+This process was repeated for each sequence (only GC-core is shown above).}
+        \label{fig:fast_slow_fits}
 	\end{center}
@@
         \includegraphics[width=80mm, 
-        scale=0.5]{Figs/figs_imp/all_seq_implied_times.png}
+        scale=0.5]{Figs/figs_imp/allseq_lags.png}
         \caption{SRV-MSM vs.
@@
 	\begin{center}
-        \includegraphics[width=450, scale=1]{Fig9.pdf}
+        \includegraphics[width=450, scale=1]{Figs/figs_imp/GC-core_tracking_modes.png}
         \caption{Diffusion map representation configurations sampled from the F4 state.
@@
 
-\subsection{Explicit model comparisons}
-
-We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
+%\subsection{Explicit model comparisons}
+
+%We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
 In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3spn2 explicit implementation \citep{Hinckley2015}.
@@
         \label{fig:GC-end_dmaps_3prime}
-	\end{center}
-\end{figure}
-
-% add figure for slow and fast response fitting (just GC-core)
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.png}
-        \caption{Slow and Fast modes were determined by fitting over a distribution of 120 independent simulations.
-For each temperature in the series, a relaxation curve was fit to the distribution and the associated exponential coefficient was used to the calculate rate.
-Temperatures series between -5 - +15K relative to empirically determined sequence melting temperature were explored.
-This process was repeated for each sequence (only GC-core is shown above).}
-        \label{fig:fast_slow_fits}
 	\end{center}

### 2020-12-18 manuscript.tex
Updates from Overleaf

@@
 Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides, Roman o2013DNADependence, Hinckley2014Coarse-grainedEffects}.
+Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects}.
 Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
@@
 We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
-MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
-\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent.
+MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
+\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
 We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
-% this section should look more like a conclusion foreshadowing conclusions and main take aways 
-We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
+
+We find that 3spn2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
 Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation.
-Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer dynamics populated by a diversity of long-lived metastable states.
-Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior.
+Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
+Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior.
 Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process.
@@
 This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
 
@@
 
-The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
+The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
 Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
@@
 The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed.
-While most pipelines employ tICA to generate a linear transformation of input features, we use SRVS as a more powerful modular replacement for tICA.
+While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA.
 SRVs were first developed by Chen et al.
@@
 MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}.
-After Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time.
+Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
 To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}.
-Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis.
+Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis.
 To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
@@
 The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
-The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.
+The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.
 
@@
 
-As an initial validation of the 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) experiments.
+As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments.
 Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
@@
 To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
-We derived relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
+We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
 
 % need to discuss tmelt shift
-We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff.
+We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
 We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}.
@@
 While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
-We used this temperature shift to inform our equilibrium simulations in the proceeding section.
 We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
-After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits.
-At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone.
+After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})).
+At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone.
 GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
@@
 
-Following the analysis pipeline described above, we generated SRV-MSMs for each sequence.
+We followed the SRV-MSM analysis pipeline described above to generate sequence-dependent kinetic models for each sequence.
 We identified seven kinetically relevant states that are captured within the resolution of the model.
@@
 Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
-Arrows between states show the conditional transition probability to another state within the MSM lag time.
+Arrows between states show the probability of a transition to another state within the MSM lag time.
 We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.
@@
 
-%The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}.
-The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails.
-Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
+\begin{figure}[ht!] % right place for NN table? could go to SI
+	\begin{center} 
+        \includegraphics[width=120mm, scale=1]{Fig5.pdf}
+        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations.
+All free energies are normalized such that that the native hybridized state is set to zero.
+Calculations included dangling end contributions but do not take inert tails effects.}
+
+        \label{fig:NN_table}
+	\end{center}
+\end{figure}
 
@@
 Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.
-Contrary to macrostate results, however, the 3S2 state is more stable than 5S2 states.
+Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state.
 Moreover, we do not observe a stable 3S4 cluster for GC-end.
@@
 
+%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
+Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails.
+Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
+
 % and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
@@
 
-% Although NN calculations predict 5' vs.
-3' stabilities for AT-all, we observe that GC-end 5S2 state remains more populated than the 3S2 state contrary to NN predictions.
-Indeed, we do not observe the formation of a stable 3S4 state for GC-end, further indicating that these states 3' states are unfavorable.
-This leads us to believe that 5' vs.
+% This leads us to believe that 5' vs.
 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.
-
-\begin{figure}[ht!] % right place for NN table? could go to SI
-	\begin{center} 
-        \includegraphics[width=120mm, scale=1]{Fig5.pdf}
-        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations.
-All free energies are normalized such that that the native hybridized state is set to zero.
-Calculations included dangling end contributions but do not take inert tails effects.}
-
-        \label{fig:NN_table}
-	\end{center}
-\end{figure}
-
-%if AT-all has highest frequency and GC-end has lowest then can describe role of metastable states as increasing rate for the former but decreasing rate for the later (for kinetic traps that do not terminate in native hybridization and delay the process as a whole).
 
@@
 On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
-We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently then in experiment.
+We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently than in experiment.
 Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.
@@
 When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
-found that the GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}.
+found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}.
 As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
@@
 We present two representative events in Figure \ref{fig:GC-mix_transitions} along with all native base pair distances over time.
-During hybridization, we observed the formations of some key base pair contacts before the full duplex formed.
+During hybridization, we observed the some key base pair contacts before the full duplex formed.
 Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T.
-This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}.
+This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization}.
 For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
@@
 Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
-These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
+These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural understanding of the configuration in these regions \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
 While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
@@
 We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register.
-These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated.
+These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an transition state through which hybridization is facilitated.
 This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.
@@
 
-To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
+To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
 To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event.
@@
 We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance.
-Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence.
+Indeed, $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence.
 This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.
@@
 
-Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge relevant faster processes such as duplex nucleation and zippering.
+Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering.
 These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.

### 2020-12-18 manuscript.tex
Updates from Overleaf

@@
 Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
-From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}.
-In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}.
-A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
-The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.
-Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}.
-Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}.
-Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.
-
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}.
-More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}.
-Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}.
-Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}.
-Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects}.
-Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}.
-Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}.
-Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
-
-Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}.
-Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates.
-For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}.
-The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent.
-Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}.
-Sanstead et al.
-highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides \citep{Sanstead2016}.
-Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
+From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
+The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}
+
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
+
+Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates.
+For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent.
+Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al.
+highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
 
 In this work, we study the same four sequences explored by Sanstead et al.
-in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}.
-We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.
-We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}.
-MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}.
-\citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
+in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
 We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
@@
 
-The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}.
-Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
-The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}.
-The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.
-\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.
-
-We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016}.
-We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
+The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}
+
+We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
 We ran our simulations in the NVT ensemble.
-Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.
-\citep{Schneider1978Molecular-dynamicsTransitions}.
-Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
+Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
 We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps.
@@
 A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
-Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library \citep{McGibbon2015MDTraj:Trajectories}.
-Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}.
-The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}.
-Additional details on the VAMP-2 score can be found in the supplemental information.
+Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
+To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
 By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
@@
 SRVs were first developed by Chen et al.
-as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}.
-The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}.
-The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}.
-The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}.
-SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
+as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
 This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
-The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.
+The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}
 
@@
 % need some more detail here -- talk to Andy about this
-MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}.
-Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
+MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
-To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}.
-Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis.
-To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}.
-In a similar approach to Sidky et al.
-we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}.
-The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
+To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible.\citep{Pande2010EverythingAsk} Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis.
+To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence.\citep{Scherer2015PyEMMAModels} In a similar approach to Sidky et al.
+we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
 The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.
@@
 \subsection{\label{sec:methods}Diffusion Maps}
-Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
-We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies \citep{Shmilovich, Wang2017MesoscaleLandscapes}.
-We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.
+Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
+We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies.\citep{Shmilovich, Wang2017MesoscaleLandscapes} We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.
 
@@
 As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.
-Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}.
-We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
+Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
 To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
@@
 We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
-We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}.
-We noted an average temperature shift of about 4C between experiment and simulation exponentials.
+We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4C between experiment and simulation exponentials.
 While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
@@
 al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}.
-All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}.
-Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
 Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff.
@@
 
-Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}.
-The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}.
-It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
+Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
 In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}).
@@
 This indicates that 5' vs.
-3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.
+3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}
 
@@
 Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences.
-Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
-These processes can occur even when terminal mismatches are present in the non-native structure.
+Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands.\citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel} These processes can occur even when terminal mismatches are present in the non-native structure.
 
 Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying.
-To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}.
-Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}.
-Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
+To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
 It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs.
@@
 Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4).
-Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}.
-Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
+Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
 This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
@@
 % discuss computational comparisons needs work
-Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA}.
-Araque et al.
-defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}.
-When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
-found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}.
-As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs \citep{Phys2019}.
-Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}.
-We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
+Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA} Araque et al.
+defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions).\citep{Sanstead2016} As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
+While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs.\citep{Phys2019} Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
 
@@
 We did, however, observe substantial fraying of the two AT termini in the simulation data.
-Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.
-While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state.
+Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state.
 Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
@@
 Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T.
-This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization}.
-For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
+This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
 There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs.
 In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes.
-As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}.
-We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
+As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
 
@@
 Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
-These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural understanding of the configuration in these regions \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
-While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
+These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural understanding of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
 We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding.
@@
 al.
-regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}.
-This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
+regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end.\citep{Sanstead2016} This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
 Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.
@@
 In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}.
-In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
+We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise.\citep{McGibbon2015VariationalKinetics} In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
 We used a batch size of 50000 and ran each model for a total of 20 training epochs.
@@
 SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.
+SRV training was implemented using Keras and Tensorflow.\citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}
 
@@
 For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step.
-Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}.
-We found that an SRV lag time of 1.2 ns showed good convergence across the relevant timescales for all sequences.
+Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest.\citep{Phys2018MarkovValidation} We found that an SRV lag time of 1.2 ns showed good convergence across the relevant timescales for all sequences.
 Next we selected an optimal number of SRV components to include in our analysis.
-After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}.
-It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}.
-For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
+After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics.\citep{McGibbon2015VariationalKinetics} It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant.\citep{Pande2010EverythingAsk} For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
 As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
@@
 Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
-Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}.
-Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
+Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc).\citep{Pinamonti2017PredictingModels, PinamontiTheModels} Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
 
@@
 These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
-Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}.
+Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying.\citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}
 
@@
 To visualize macrostates, we project the data into the two leading tICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}.
-Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
+Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments.\citep{Sidky} Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
 After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment.
@@
 As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach.
-Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model \citep{Sidky}.
-Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
+Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.\citep{Sidky} Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
 Similar plots for the other sequences are included in the SI.

### 2020-12-18 manuscript.tex
Updates from Overleaf

@@
 In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model.
-Our computational predictions show agreement with experiments, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
+Our computational predictions show agreement with experiments and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
 
@@
 
-Over the last couple decades, DNA has proved to be much more than a vessel for genetic information.
-From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
+Over the last of couple decades, DNA has proven to be much more than a vessel for genetic information.
+From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
 The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}
 
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
 
-Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates.
+Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates.
 For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent.
@@
 In this work, we study the same four sequences explored by Sanstead et al.
-in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
+in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
 We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
@@
 
-We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
-We ran our simulations in the NVT ensemble.
+We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff and ran our simulations in the NVT ensemble.
 Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
@@
 We identified seven kinetically relevant states that are captured within the resolution of the model.
-Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities.
+Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities.
 By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
@@
 Arrows between states show the probability of a transition to another state within the MSM lag time.
-We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.
+We now proceed to analyze the sequence-dependent thermodynamics and kinetic features of each MSM.
 
@@
 This indicates that 5' vs.
-3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}
+3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}
 
@@
 % This leads us to believe that 5' vs.
-3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.
+3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.
 
@@
 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
-Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S2 $\rightarrow$ H is more favorable than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
+Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g.
+5S2 $\rightarrow$ H is more likely than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
 Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
@@
 
-Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying.
+Experimentally, out-of-register states are suspected to occur but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying.
 To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
@@
 Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4).
-Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
+Previous studies suggest that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
 This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
@@
 Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state.
-Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
+Furthermore, one or two-base-pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
 
@@
 We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations.
-Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence, and were working near memory limits when training SRVs and building SRV-MSMs.
+Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence and were working near memory limits when training SRVs and building SRV-MSMs.
 

### 2021-01-27 manuscript.tex
Updates from Overleaf

@@
 In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement.
-We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model (MSM) construction.
+We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model construction.
 Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides.
@@
 
+\begin{comment}
+
 \begin{figure}[ht!]
@@
 \end{figure}
+
+\end{comment}
 

### 2021-03-18 manuscript.tex
added DNA-PAINT refs

@@
 
-Over the last of couple decades, DNA has proven to be much more than a vessel for genetic information.
-From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
-The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}
+Over the last couple of decades, DNA has proven to be much more than a vessel for genetic information.
+From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing.
+\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Indeed, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
+The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}
 
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
 
 Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates.
-For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent.
+For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally and appears highly sequence-dependent.
 Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al.
@@
 In this work, we study the same four sequences explored by Sanstead et al.
-in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
+in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3SPN2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
 We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
 
-We find that 3spn2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
+We find that 3SPN2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
 Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation.
-Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
+Consistent with previous studies, \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
 Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior.
@@
 Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+These can results inform a generalized model between DNA sequence and hybridization mechanism with application towards equilibrium binding technologies such as DNA-PAINT and barcoding.
 
@@
 
-The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
-The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}
+The 3SPN2 coarse-grained model was designed to accelerate DNA simulations relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3SPN2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}
 
-We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff and ran our simulations in the NVT ensemble.
-Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
+We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
+We ran our simulations in the NVT ensemble with a Langevin thermostat that modeled interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
 We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps.
@@
 Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
-By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
+To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
+By comparing scores across several candidate feature sets (Figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
 Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
@@
 as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
-This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence.
+This is a key attribute to our system as 10 million frames with 55 features per frame are used for each sequence.
 The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}
@@
 
-As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments.
+As an initial validation of 3SPN2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments.
 Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
@@
 Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
-To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
+To compare results with 3SPN2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
 We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
@@
 We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
-We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4C between experiment and simulation exponentials.
-While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
+We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4$^{\circ}$C between experiment and simulation fits.
+While 3SPN2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
 We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
-After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})).
+After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (Figure \ref{fig:relaxation-comparison})).
 At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone.
@@
 
-% A documented short-coming for the 3spn2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
+% A documented short-coming for the 3SPN2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
 This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
@@
 al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN2 base, we would not expect to resolve this diverse collection of states and dynamics.
 Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff.
@@
 In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying.
-We observe similar trends in the experimental data, although comparisons at high temperatures were limited due to mixing between the fast and slow responses (figure \ref{fig:relaxation-comparison}).
-We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3spn2 fraying dynamics are accelerated by about 100-200x relative to experiments.
+We observe similar trends in the experimental data, although comparisons at the highest temperatures were limited due to mixing between the fast and slow responses (Figure \ref{fig:relaxation-comparison}).
+We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN2 fraying dynamics are accelerated by about 100-200x relative to experiment.
 It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
@@
 Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
-and that 3spn2 can capture some sequence-dependent fraying effects.
+and that 3SPN2 can recapitulate some sequence-dependent fraying effects.
 % need a better transition here back into SRV-MSMs?
@@
         \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate.
-Arrows indicate the probability of transitioning out of the state within the lag time.
-Circle areas are proportional to state stability.}
+Arrows indicate the probability of transitioning between states within the lag time.
+Circle areas are proportional to the log of stationary probabilities to enhance visualization of less stable states}
 
@@
 
-We followed the SRV-MSM analysis pipeline described above to generate sequence-dependent kinetic models for each sequence.
-We identified seven kinetically relevant states that are captured within the resolution of the model.
+We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer.
+We identified seven kinetically relevant states that were captured within the resolution of the model.
 Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities.
-By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
 Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
@@
 % rewrite this into one paragraph
-\subsection{MSM stabilities agree with nearest neighbor thermodynamic models}
+\subsection{Inert tails produce deviations from nearest neighbor thermodynamic models}
 
@@
 Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
-In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}).
+In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (Figure \ref{fig:NN_table}).
 For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4).
@@
 This indicates that 5' vs.
-3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}
+3' inert tail differences may out-weigh NN stacking effects alone.
+These inert tail effects show qualitative agreement with previous experimental studies, which attribute the differential behavior to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure.
+\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization}.
 
@@
 To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
-It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs.
+It is unclear if these differences were a consequence of varied simulation conditions or if GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs.
 This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
@@
 We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-ends and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
-Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
+Although reproducing these stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
 
@@
 On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
-We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently than in experiment.
-Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.
+We previously noted that frayed dynamics are substantially accelerated in the 3SPN2 model, which may lead to the F4 state being sampled more frequently than in experiment.
+Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in Figure \ref{fig:relaxation-comparison}.
 
@@
 Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA} Araque et al.
-defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+defined a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
 found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions).\citep{Sanstead2016} As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
@@
 We did, however, observe substantial fraying of the two AT termini in the simulation data.
-Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state.
+Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and transient that very few progress to full dissociation compared to the F4 state.
 Furthermore, one or two-base-pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
@@
 This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
-There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs.
-In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes.
+Qualitatively, we observed some short-lived states composed of 2-4 base pair contacts immediately before full dissociation occurs.
+However, in contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes.
 As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
@@
 Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
-These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural understanding of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
+These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural details of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
 We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding.
@@
 We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register.
-These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an transition state through which hybridization is facilitated.
-This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.
+These "shifted-loop" bonds appear uniquely stable for GC-end conformations, and suggest a transition state through which hybridization is facilitated.
+This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the equivalent AT-all pathways.
 
-%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.
+%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3SPN2 \citep{Hinckley2013AnHybridization}.
 
@@
 %Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
-In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+In the context of the 3SPN2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
 The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
 
-%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3spn2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
+%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3SPN2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
 
@@
 To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
-To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event.
+To achieve higher resolutions, we reran selected simulation with a 10x higher save rate and analyzed a 1000 frame region (10 ns simulation time) for each event.
 We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences.
@@
 Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
-The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first WC contacts are made.
+The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first Watson-Crick contacts are made.
 In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases.
@@
 This is consistent with the downhill nucleation-zippering mechanism described above.
-Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et.
-al.
-regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end.\citep{Sanstead2016} This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
+Although differences between AT-all and GC-end are more subtle, overall deviations from two-state behavior follow predictions by Sanstead et.
+al.\citep{Sanstead2016}  where GC-core > GC-mix > AT-all > GC-end.
+This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
 Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.
@@
 Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
-These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.
+These insights should be leveraged for enhanced understanding of
+\emph{in vivo} hybridization processes, optimized sequence design in application such as DNA-PAINT and barcoding, and novel developments in the growing field of dynamic DNA nanotechnology.
 
@@
 We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
-We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}).
+We build diffusion maps using 10000 frames sampled from the macrostate (Figure \ref{fig:GC-core_dmaps}).
 Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
@@
 %We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
-In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3spn2 explicit implementation \citep{Hinckley2015}.
+In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3SPN2 explicit implementation \citep{Hinckley2015}.
 These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities.

### 2021-03-30 manuscript.tex
typos

@@
 Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
-These can results inform a generalized model between DNA sequence and hybridization mechanism with application towards equilibrium binding technologies such as DNA-PAINT and barcoding.
+These results can inform a generalized model between DNA sequence and hybridization mechanism with application towards equilibrium binding technologies such as DNA-PAINT and barcoding.
 

### 2021-05-01 SI.tex
Andy edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article]{achemso}

%\documentclass[english,aps,preprint,pre,floatfix,nofootinbib,showpacs,showkeys]{revtex4-1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later. Do NOT use any
%% packages which require e-TeX (for example etoolbox): the e-TeX
%% extensions are not currently available on the ACS conversion
%% servers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}

\SectionNumbersOn

%include hyperlinks to figures
%\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=false,
    pdfborder={0 0 0},
    linkcolor=blue
]{hyperref}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{CJK}
\usepackage{framed}
%\usepackage[hidelinks]{hyperref}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{mathrsfs}
\usepackage{comment}
\usepackage[normalem]{ulem}
\usepackage{algorithm}
\usepackage{subfig}
\usepackage{algpseudocode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\makeatletter
    \setlength\@fptop{0\p@}
\makeatother

\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\rood}[1]{#1}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
%\newcommand*{\blauw}[1]{#1}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}

\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
%\newcommand*{\noter}[1]{}					% notes off

%\usepackage[draft]{todonotes}   % notes shown
\usepackage[disable]{todonotes} % notes hidden

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael S. Jones}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Brennan Ashwood}
\affiliation{%
  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Andrei Tokmakoff}
\affiliation{%
  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Andrew L. Ferguson}
\email{andrewferguson@uchicago.edu}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

%\author{Kay T. Finally}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]{SUPPLEMENTAL INFORMATION \\ Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\abbreviations{IR -- infrared, NMR -- nuclear magnetic resonance, UV -- ultraviolet}
%\keywords{Takens}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.  The document should begin with an
%% abstract, if appropriate.  If one is given and should not be, the
%% contents will be gobbled.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\renewcommand{\thepage}{S\arabic{page}} 
\renewcommand{\thesection}{S\arabic{section}}  
\renewcommand{\thetable}{S\arabic{table}}  
\renewcommand{\thefigure}{S\arabic{figure}}

% Fig. S1
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=0.9\textwidth, 
        scale=0.5]{Figs/figs_imp/allseq_crossval.pdf}
        \caption{Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence. A knee in the VAMP-2 plot at approximately the fifth slow mode for each of the four sequences motivates the choice of a 5D projection. The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that the 5D model is not overfitted.}
        \label{fig:allseq_srv_crossval}
	\end{center}
\end{figure}

\clearpage
\newpage

% Fig. S2
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=0.7\textwidth, 
        scale=0.5]{Figs/figs_imp/allseq_lags.pdf}
        \caption{Convergence of the MSM implied time scales $t_i$ as a function of lag time $\tau$. Solid lines indicate maximum likelihood result while dashed lines show the Bayesian ensemble means. The implied time scales for all sequences converge at a lag time of $\tau$ = 1.2 ns (vertical line). The black solid curve marks equality of the implied time scale and lag time and delimits the shaded region wherein the implied time scales are shorter than the lag time and cannot be resolved. \noter{(i) Eliminate tICA data. (ii) Eliminate legend. (iii) Add Bayesian ensemble uncertainties as colored shading behind the solid line. PyEMMA will do this calculation for you.}}
        \label{fig:all_seq_lags}
	\end{center}
\end{figure}

\clearpage
\newpage

% Fig. S3
\begin{figure}[ht!]
	\begin{center}
        %\includegraphics[width=0.7\textwidth, scale=0.5]{Figs/figs_imp/allseq_lags.pdf}
        \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by an MSM constructed at the $\tau$ = 1.2 ns lag time (dashed blue line) versus those computed from an MSM constructed at the particular lag time (solid black line). The good agreement between these two results provides numerical validation of the Markovian nature of the $\tau$ = 1.2 ns lag time MSM.}
        \label{fig:all_seq_lags}
	\end{center}
\end{figure}

\clearpage
\newpage

% Fig. S4
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.pdf}
        \caption{\noter{Can this be (i) simplified -- just show the left panels, the right panels are in the main text, (ii) report the kd values in the legends for the fits and provide the equation in the caption, and (iii) show data for all four seqs. Rewrite the caption and axes labels to reflect the terminology used in the main text (e.g., $k_d^\mathrm{slow}$, report Tm by writing it as text within white space of each figure for each sequence).}}
        \label{fig:fast_slow_fits}
	\end{center}
\end{figure}

\clearpage
\newpage

% Fig. S5
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=0.92\textwidth]{Figs/SX-1.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=0.75\textwidth]{Figs/SX-2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
	\begin{center}
        \caption{\noter{Add caption mirroring that from the main text.}}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

\clearpage
\newpage

\subsection*{Nearest neighbor model of duplex thermodynamics}

\rood{TBA -- full accounting complete with mathematics and references for model and its parameters of the application of NN model to predict free energy of each macrostate for each sequence} \cite{SantaLucia1998AThermodynamics}

\clearpage
\newpage

%\section{Supplemental Information}

% add secion on dmap and high temp methods?
%\subsection{Additional methods}

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=0.9\textwidth, 
%        scale=0.5]{Figs/figs_imp/allseq_event_count.pdf}
%        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in the dissociated state. Here we show fraction of hybridized duplexes averaged across these sets over time. For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive $T_m$ difficult to identify. The number of hybridization/dissocation combined across trajectories is shown. Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics. AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state. GC-core displays the most native hybridization events, perhaps due to stronger nucleation sites at the central G:C contact points.}
%        \label{fig:allseq_event_count}
%	\end{center}
%\end{figure}

% add figure for slow and fast response fitting (just GC-core)

%\subsection{VAMP-2 Scoring}
%The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models. Hyperparameters, including feature choice, number of microstates, and number of slow modes were evaluated and selection with via the VAMP-2 score. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.

%\begin{align*}
% 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
% 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
% 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
%	\\
% 	VAMP-2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
% 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
%\end{align*}\label{CK1}

%\subsection{SRV hyperparameters}

%Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise.\citep{McGibbon2015VariationalKinetics} In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data. We used a batch size of 50000 and ran each model for a total of 20 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs. SRV training was implemented using Keras and Tensorflow.\citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}

%\subsection{SRV-MSMs provide better resolution than tICA-MSMs}

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=0.9\textwidth]{Figs/figs_imp/AT-all_implied_timescales.pdf}
%        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
%        \label{fig:AT-all_dynamic}
%	\end{center}
%\end{figure}

%\subsection{SRV-MSM walkthrough and analysis}

%In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the most discernable spectral gap between modes. For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest.\citep{Phys2018MarkovValidation} We found that an SRV lag time of 1.2 ns showed good convergence across the relevant timescales for all sequences. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics.\citep{McGibbon2015VariationalKinetics} It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant.\citep{Pande2010EverythingAsk} For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.

%\subsubsection{SRV-MSM construction and optimization}
%Although SRV coordinates alone provide some information, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM. Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates. We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score. Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process. This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous tICA basis. Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate. Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures. Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc).\citep{Pinamonti2017PredictingModels, PinamontiTheModels} Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.

%Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands. The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most relevant to each process. For example, the first slow mode shows a positive correlation to each distance and the strongest correlation with native base pair distances (shown along the main diagonal). Given these correlations and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process. The next four SRV components all show a relatively high correlation along offset diagonals. These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order ``shifting'' processes between sets of such base pairings. Previous ``inchworm'' and ``pseudoknot'' mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying.\citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}

%We found GC-end had a similar implied timescales distribution to AT-all, with a distinct spectral gap after the fourth mode. For AT-all and GC-end, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs. For GC-core we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above. We found that four macrostate clustering was unstable -- likely because the third mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized (H), dissociated (D), and 4 4 base pair frayed (F4) states.  The GC-mix sequence showed a similar implied timescale distribution to GC-core, however we no longer saw a converged slow mode corresponding to multi-base fraying behavior. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated. These correlate closely with the first and third GC-core SRV modes. Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate. As such, we designated this transitions as effectively two-state within the resolution of our model.

%For GC-core and GC-mix, we found this clustering convention to be unstable -- likely because the last converged mode was providing information about dissociation dynamics -- and performed PCCA+ with one less macrostate.

%To visualize macrostates, we project the data into the two leading tICA coordinates. Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments.\citep{Sidky} Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states. After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment. Means and standard deviations were calculated from this ensemble.

%As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach. Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.\citep{Sidky} Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales. Similar plots for the other sequences are included in the SI. We also note that the infrequency of transitions relative to the individual trajectory length leads to slower convergence of the leading mode. Because this mode corresponds the overall hybridization and dissociation process, we found that it displayed similar behavior across all sequences and that higher order modes were more informative for lag time selection.

%\subsection{SRV modes reveal hierarchical clustering structure}  % shift to include this in

%\begin{figure} %[ht!]
%	\begin{center}
%        \includegraphics[width=0.8\textwidth]{Figs/figs_imp/GC-core_tracking_modes.pdf}
%        \caption{Leading GC-core SRV coordinates correlated with identifiable physical coordinates during sample hybridization (a) and dissociation (b) events. (c) Pearson correlations between all 100 intermolecular base pair distances and each the three leading slow modes. SRV2 correlations were calculated in the H and F4 states; SRV correlations were calculated in the D state.}
%        \label{fig:GC-core_tracking_modes}
%	\end{center}
%\end{figure}

% not sure if this should directly follow the SRV-MSM GC-core description or directly follow the GC-core dmaps?
% try cutting this out entirely ?
%\subsection{SRV correlations to physical coordinates}

%Having constructed and interpreted SRV-MSMs, we revisited our original SRV basis to investigate how slow modes are constructed from input features. We found these GC-core modes to be of particular interest as they reveal the hierarchical and nonlinear nature of the dynamical encoding. We examined a collection of ``trimmed'' trajectories centered on both hybridization or dissociation events. For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they shared behavior. Two representative trajectories are shown in figure (\ref{fig:GC-core_tracking_modes}). Complementary G:C pairs are the best indicators for a hybridization/dissociation event, and we see a sharp change in the first SRV mode (SRV1) as these bonds form or break. The second slow mode (SRV2) is most active when G:C pairs are bound but the adjacent AT pairs are not. There is a small signal for fraying at the outer base pairs, but the mode overwhelmingly learns about these neighboring A:T/G:C bonds. Moreover, this behavior reflects movement in and out of the F4 state in the corresponding SRV-MSM discussed above. SRV3 is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs. We attribute this to the SRV learning about the diffusive motions of the two body system and evaluating the likelihood of an imminent hybridization event. The third mode also peaks when the oligos are close together but configured in such a way that is not amenable hybridization. These misaligned conformations include inverse contacts where 5$^\prime$/5$^\prime$ and 3$^\prime$/3$^\prime$ ends meet and looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts.

%Despite the qualitative trends we observe between physical coordinates and leading SRV modes, we found it difficult to find correlations across full trajectories. With the exception of SRV1, Pearson correlations between physical coordinates and SRV modes were near zero and showed erroneous trends. However, when we performed this analysis along smaller sections of trajectories we noticed that the sign of SRV2 and SRV3 correlations switched depending on whether the oligos were in the hybridized or dissociated state. This shows that these modes are inherently nonlinear and provide support on top of the first mode -- which serves as an indicator function for hybridized vs. dissociated state. With respect to our SRV-MSM macroscates, we found that SRV2 was ``turned on'' in the H and F4 states -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state. Accordingly, we calculated Pearson correlations between each SRV mode and all distances in states where the modes are active. Figure \label{fig:GC-core_tracking_modes} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5$^\prime$/5$^\prime$ and 3$^\prime$/3$^\prime$ pairs which tend to approach each when the duplex is in the F4 state. We also observed a highly symmetric correlation between SRV3 and central base pairs distances, which indicative of overall diffusive behavior. Taken together, these analyses reveal how the SRV learns and represents the dynamical space in an hierarchical manner, providing enhanced resolution to a linear method like tICA and producing the MSM results we observe above.

%\subsection{Structural diversity in the F4 state informs fraying dynamics}

\begin{comment}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Figs/figs_imp/GC-core_tracking_modes.pdf}
        \caption{Diffusion map representation configurations sampled from the F4 state. The x and y axes represent the the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

\end{comment}

% shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

%We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (Figure \ref{fig:GC-core_dmaps}). Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in ``overlap'' distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place. Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation. Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.

%\subsection{Explicit model comparisons}

%We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation. In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3SPN2 explicit implementation \citep{Hinckley2015}. These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities. Comparisons were limited by the higher computational demand required to simulate 5x the beads in the box, a shorter integration timestep, and reliance on Ewald summation to account for electrostatics. We caution over-interpretation of these results as the explicit model has not been as widely adopted or thoroughly validated compared to the implicit model.

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.pdf}
%        \caption{Full 5$^\prime$ diffusion maps including degenerate third diffusion map mode.}
%        \label{fig:GC-end_dmaps_full}
%	\end{center}
%\end{figure}

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_3prime.pdf}
%        \caption{Full 3$^\prime$ diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct ``looping'' region for intact GC bonds}
%        \label{fig:GC-end_dmaps_3prime}
%	\end{center}
%\end{figure}

%add anymore figures for the time analysis: As shown bleow:
\begin{comment}

\subsubsection{\label{sec:Results}Code to provide on GitHub} 
\begin{enumerate}
    \item Cross val data
    \item lammps run script and each sequence input files
    \item Featurization with mdtraj (reindexing and sparse saving)
    \item Dmaps for GC-core and GC-end
    \item timescales comparison
    \item SRV cross valss
    \item MSM construction

\end{enumerate}
\end{comment}

\begin{comment}

\section{\label{sec:Discussion}Discussion}

\citep{Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}
Long simulations shows that terminal fraying does not converge on the picosecond timescale

\citep{Nonin1995TerminalFraying}
Most simulations suppress fraying by placing GC pairs at termini.

\citep{Andreatta2006UltrafastHelix}
- Base opening process is slower than 40 ns but faster than 1ms
- Open termini state shows new fraying mode of 5ps (I think shows free base motions)
\end{comment}

\begin{comment}

%%%%%% Other MSM and timescales approaches  %%%%%%%%%%%

McGibbon, R. T., & Pande, V. S. (2015). Variational cross-validation of slow dynamical modes in molecular kinetics : Importance of not generating too many microstates and employting cross-validation to avoid artificially high scores or artificially slow dynamics. Both of these lead to models fitting statistical noise which is the root of boosted scores.

Pinamonti, G., Paul, F., Rodriguez, A., & Bussi, G. (n.d.). analyzed with core-set Markov state models, 43: Good ideas on core-based clustering. Assigning microstates based on physical characteristics is interesting but didn't seem to work for me in practice (could try doing this based on sim.log data with number of base pairs bound or some energy metric). Lot's of notes on this paper regarding comparisons to timescales. Good notes on higher resolution fraying interactions that we likely can't resolve.

Sua, E., Adelman, J. L., & Zuckerman, D. M. (2016). Accurate Estimation of Protein Folding and Unfolding Times : Beyond Markov State Models, 1. https://doi.org/10.1021/acs.jctc.6b00339:
More accurate ways to adapt MFPT into a rate

Refs for DNA fraying:
\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro). Use TPS to focus on kinetic pathway by which singl base pair bind/unbind. Focus on end base of 3-bp oligomer in explicit solvent. Finds order parameters tracking base pair binding as well as intra-strand stacking. (energy and distance for each).

\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,

6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding

\end{comment}

\clearpage
\newpage

\bibliography{references}
%\bibliography{refs_mendeley2}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\end{document}

### 2021-05-09 SI.tex
Deleted most old figures and large excess files

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[]{SUPPLEMENTAL INFORMATION \\ Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy}
+\title[]{SUPPORTING INFORMATION \\ Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy}
 
@@
 
+\setlength{\marginparwidth}{2cm}
+
 \begin{document}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\clearpage
 \newpage
@@
         \includegraphics[width=0.9\textwidth, 
-        scale=0.5]{Figs/figs_imp/allseq_crossval.pdf}
+        scale=0.5]{SIFig1.pdf}
         \caption{Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
-A knee in the VAMP-2 plot at approximately the fifth slow mode for each of the four sequences motivates the choice of a 5D projection.
-The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that the 5D model is not overfitted.}
-        \label{fig:allseq_srv_crossval}
+A knee in the VAMP-2 plot was identified at approximately the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively.
+An embedding of corresponding dimensionality was then used to cluster frames into discrete states.
+The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
+        \label{fig:SIFig1}
 	\end{center}
@@
 \begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.7\textwidth, 
-        scale=0.5]{Figs/figs_imp/allseq_lags.pdf}
+	\centering
+    \includegraphics[width=0.7\textwidth, scale=0.5]{SIFig2.pdf}
     \caption{Convergence of the MSM implied time scales $t_i$ as a function of lag time $\tau$.
@@
 The implied time scales for all sequences converge at a lag time of $\tau$ = 1.2 ns (vertical line).
-The black solid curve marks equality of the implied time scale and lag time and delimits the shaded region wherein the implied time scales are shorter than the lag time and cannot be resolved.
-\noter{(i) Eliminate tICA data.
-(ii) Eliminate legend.
-(iii) Add Bayesian ensemble uncertainties as colored shading behind the solid line.
-PyEMMA will do this calculation for you.}}
-        \label{fig:all_seq_lags}
-	\end{center}
+The black solid curve marks equality of the implied time scale and lag time and delimits the shaded region wherein the implied time scales are shorter than the lag time and cannot be resolved.}
+    \label{fig:SIFig2}
 \end{figure}
@@
 	\begin{center}
-        %\includegraphics[width=0.7\textwidth, scale=0.5]{Figs/figs_imp/allseq_lags.pdf}
+        \includegraphics[width=1.1\textwidth]{SIFig3.pdf}
         \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by an MSM constructed at the $\tau$ = 1.2 ns lag time (dashed blue line) versus those computed from an MSM constructed at the particular lag time (solid black line).
 The good agreement between these two results provides numerical validation of the Markovian nature of the $\tau$ = 1.2 ns lag time MSM.}
-        \label{fig:all_seq_lags}
+        \label{fig:SIFig3}
 	\end{center}
@@
 \begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.pdf}
-        \caption{\noter{Can this be (i) simplified -- just show the left panels, the right panels are in the main text, (ii) report the kd values in the legends for the fits and provide the equation in the caption, and (iii) show data for all four seqs.
-Rewrite the caption and axes labels to reflect the terminology used in the main text (e.g., $k_d^\mathrm{slow}$, report Tm by writing it as text within white space of each figure for each sequence).}}
-        \label{fig:fast_slow_fits}
-	\end{center}
+	\centering
+    \includegraphics[width=.8\textwidth]{SIFig4.pdf}
+    \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
+From 120 independent 1 $\mu$s simulations, we compiled the distribution of times at which terminal A:T and central base pairs first separate to a distance of of 1.3 nm.
+We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
+Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.}
+    \label{fig:SIFig4}
 \end{figure}
@@
 \begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.92\textwidth]{Figs/SX-1.pdf}
-	\end{center}
+	\centering
+    \includegraphics[width=0.8\textwidth]{SIFig5a.pdf}
 \end{figure}
 \begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.75\textwidth]{Figs/SX-2.pdf}
-	\end{center}
+	\centering
+    \includegraphics[width=0.8\textwidth]{SIFig5b.pdf}
 \end{figure}
+
 \begin{figure}[ht!]
-	\begin{center}
-        \caption{\noter{Add caption mirroring that from the main text.}}
-        \label{fig:allseq_table}
-	\end{center}
+	\centering
+    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
+Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional hybridization and dissociation events.
+Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
+Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C WC pair and adjacent A:T pair or central A:T pairs first form prior to rapid formation of the complete duplex.
+Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
+    \label{fig:SIFig5}
 \end{figure}
@@
 
-\rood{TBA -- full accounting complete with mathematics and references for model and its parameters of the application of NN model to predict free energy of each macrostate for each sequence} \cite{SantaLucia1998AThermodynamics}
+DNA duplex hybridization thermodynamics are most commonly predicted using nearest-neighbor (NN) models where the stability of a given base pair is assumed to solely depend on the identity and orientation of its adjacent base pairs.
+We apply the NN model reported by SantaLucia\citep{SantaLucia1998AThermodynamics}  to estimate the stability of fully intact, shifted, and frayed duplex configurations observed in our Markov State Model.
+
+The NN model from SantaLucia assumes hybridization proceeds in a two-state all-or-nothing manner.
+The Gibbs free energy difference between the hybridized and dissociated states can be determined from calculation of NN enthalpy and entropy contributions.
+\begin{equation}\label{SIe1}
+\Delta G_{NN}^{\circ}(T) = \Delta H_{NN}^{\circ} - T\Delta S_{NN}^{\circ}
+\end{equation}
+Here,  and  are assumed to be independent of temperature.   and   are determined as the sum of all NN terms in a given configuration. 
+\begin{equation}\label{SIe2}
+\Delta H_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta H_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta H_{DE, j}^{\circ} + \Delta H_{init}^{\circ}
+\end{equation}
+
+\begin{equation}\label{SIe3}
+\Delta S_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta S_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta S_{DE, j}^{\circ} + \Delta S_{init}^{\circ} + \Delta S_{sym}^{\circ}
+\end{equation}
+
+$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA.
+$\Delta H_{DE, j}^{\circ}$  corresponds to the enthalpic contribution from a given dangling end  base next to the duplex\citep{Santalucia2004TM}.
+$\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
+$\Delta S_{NN}^{\circ}$  is constructed from analogous NN, dangling end, and initiation terms in addition to a symmetry term, $\Delta S_{sym}^{\circ}$, which applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.
+The SantaLucia NN parameters are defined for a solution condition at neutral pH with a sodium concentration of 1M.
+We use the empirical salt correction developed by Owczarzy\citep{Owczarzy2008PredictingCations} to determine the NN parameters at the sodium concentration used for simulations in this work.
+The Owczarzy model assumes the affect of counterions on DNA hybridization thermodynamics is purely entropic.
+
+\begin{equation}\label{SIe3}
+\Delta S_{NN}^{\circ}(c_{Na^+}) = S_{NN}^{\circ}(1M) + \Delta H_{NN}^{\circ}[(4.29f_{GC}-3.95)\times 10^{-5} \ln{(c_{Na^+})}+ 9.4\times 10^{-6} (\ln{(c_{Na^+})})^2]
+\end{equation}
+
+$f_{GC}$ is the fractional G:C content of the duplex and $c_{Na^+}$ is the concentration of sodium counterions in M.
+
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=.85\textwidth]{SIFig6.pdf}
+    \caption{Schematic of nearest neighbor contributions for an intact hybridized and out-of-register duplex with dangling ends (DE) stabilization.}
+    \label{fig:SIFig4}
+\end{figure}
 
@@
 Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
-
 %\subsection{Explicit model comparisons}
@@
 
-\bibliography{references}
+\bibliography{references_210505}
 %\bibliography{refs_mendeley2}
@@
 
-\clearpage
+%\clearpage
 

### 2021-05-09 manuscript_210505.tex
Deleted most old figures and large excess files

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article]{achemso}

%\documentclass[english,aps,preprint,pre,floatfix,nofootinbib,showpacs,showkeys]{revtex4-1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later. Do NOT use any
%% packages which require e-TeX (for example etoolbox): the e-TeX
%% extensions are not currently available on the ACS conversion
%% servers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}

\SectionNumbersOn

%include hyperlinks to figures
%\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=true,
    pdfborder={0 0 0},
    linkcolor=blue,
    citecolor=black
]{hyperref}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{CJK}
\usepackage{framed}

%add refs to SI
%\usepackage{xr}
%\externaldocument{SI}

%\usepackage[hidelinks]{hyperref}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{mathrsfs}
\usepackage{comment}
\usepackage[normalem]{ulem}
\usepackage{algorithm}
\usepackage{subfig}
\usepackage{algpseudocode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\makeatletter
    \setlength\@fptop{0\p@}
\makeatother

\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off

%\usepackage[draft]{todonotes}   % notes shown
\usepackage[disable]{todonotes} % notes hidden

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael S. Jones}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Brennan Ashwood}
\affiliation{%
  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Andrei Tokmakoff}
\affiliation{%
  Department of Chemistry, Institute for Biophysical Dynamics, and James Franck Institute, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

\author{Andrew L. Ferguson}
\email{andrewferguson@uchicago.edu}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

%\author{Kay T. Finally}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]{Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\abbreviations{IR -- infrared, NMR -- nuclear magnetic resonance, UV -- ultraviolet}
%\keywords{Takens}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.  The document should begin with an
%% abstract, if appropriate.  If one is given and should not be, the
%% contents will be gobbled.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\begin{abstract} % 250 words max for JACS

\noindent A robust understanding of the sequence-dependent thermodynamics of DNA hybridization has enabled rapid advances in DNA nanotechnology. A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped. In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy. For a repetitive ATATATATAT sequence \noteb{not sure if we want to use 5$^\prime$-ATATATATAT-3$^\prime$ notation here as well?}, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations. Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways. Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical two-state model of hybridization/dehybridization over a smooth landscape with no intermediate metastable states. Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.

%\noindent \url{https://www.overleaf.com/project/5e9e5110c524b8000192c548}

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{\label{sec:intro}Introduction}

Over the last couple of decades, DNA has proven to be much more than a vessel for genetic information. From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing. \cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnological applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the \textit{dynamical}, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves. \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence. \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence, Shah2019}. A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved. For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the extent to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent. The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization. Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}. The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} \noteb{these references correspond to CG enhanced sampling studies, I can refer to enhanced sampling on AA here instead (\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations}) but I think the CG studies should be mentioned since not much has been done for AA hybridization} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways. These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacements \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami structure \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.

In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms. This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models. \citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new and existing experimental data and reinterpret our prior experimental observations in light of the new computational understanding. Consistent with previous studies, \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states. Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states. Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states that facilitate the hybridization/dehybridization process. Analysis of the configurational ensemble within each metastable state reveals the conformational heterogeneity and dynamical interconversions proceeding on time scales that lie below those resolvable within the long-time kinetic model. Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to sculpt and qualitatively alter the dynamical behaviors of the system. We anticipate that this new foundational understanding, and the extension of our approach to more extensive families of DNA sequences, can guide the rational design of DNA oligomers with tailored kinetic properties engineered for DNA nanotechnology applications such as DNA-PAINT. \citep{Shah2019, Strauss2020UpDNA-PAINT}

% example text from Andy's conclusions:

% Correlation of the data-driven kinetic order parameters extracted by SRVs provides new understanding of the sequence-dependent kinetic mechanisms by which oligomers hybridize and reveal these to be richer than classical base-pair shifting and “all-or-nothing” behaviors.

% original last paragraph
%By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible. Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.

%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}.

%Still, rich dynamics have been observed short nucleotides \citep{Sanstead2018DirectDehybridization} and the sequence-dependent effects on these dynamics are not well understood.

% include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior. The slow response was attributed to complete strand dissociation, while the fast response described base fraying. We perform longer equilibrium sampling near Tm to get a sense of sequence equilibrium behavior, as well shorter simulations at a range of higher temperatures to replicate T-jump data. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.

\section{\label{sec:methods}Methods}

\subsection{Computational Methods}

\subsubsection{Molecular dynamics simulations}

We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC3-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix). We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing. \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation. \citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers.} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature. The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was set, according to 3SPN.2 documentation, to the coordinates of Arnott et al.\ ~\citep{Arnott1976CRCBiology} The dissociated state was generated by shifting one oligomer from the hybridized state by 1 nm in the x, y, and z directions. The magnitude of this shift did not affect the results so long as all native WC bonds were sufficiently separated. \noteb{No energy minimization (no 3spn2 precedent on this). The lammps default emin techniques may not work for this type of CG forcefield. Given the sampling it seems that  1 $\mu$s of equilibration should be sufficient?} Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the system temperature. \addref{Maxwell-Boltzmann --- \noteb{not sure what to use here, usually not cited?}} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

\subsubsection{Markov state model construction}

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionV1.0} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances. We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the $(10 \times 10)$ 100-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair. This featurization naturally accounts for intermolecular configurations via the pairwise distances between bases on different strands. \noteb{previous text had suggested we used the 190-dim features, but we actually used 100-45 = 55-dim features since the intramolecular features contributed no additional kinetic variance} One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances. Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2. \citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands. VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization. \citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top 5 modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using 5-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA GK210GL GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively \noteb{not sure the best way to explain this succinctly}. (Figure 1 in the Supporting Information). This motivated the construction of a n-dimensional embedding corresponding to n leading SRV modes retained for each sequence.

\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering. The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.

\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM. The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2. We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns. At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively. The identification of $(i-1)$ implied time scales \noteb{(may be 1-2 additional converging, but these correspond to diffusive processes that are irrelevant or unhelpful in determining the hybridization mechanisms. Still, this might come up when looking at the the SI Fig 2} implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates. We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates. These macrostate MSMs constitute our terminal kinetic models. We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}. This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$. We present in SI Figure 3 the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four kinetic models.

%\subsection{Featurization}

%A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a ``symmetrized'' feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\cite{ Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (Figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=0.55\textwidth]{Fig1.pdf}
%        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates. Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
%        \label{fig:allseq_features_vamp2}
%	\end{center}
%\end{figure}

%\subsection{\label{sec:methods}SRVs}

%The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and time scales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features per frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}

%\subsection{\label{sec:methods}SRV-MSMs}

% need some more detail here -- talk to Andy about this
%MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible.\citep{Pande2010EverythingAsk} Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence.\citep{Scherer2015PyEMMAModels} In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied time scales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.

%\subsection{\label{sec:methods}Diffusion Maps}
%Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time. We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies.\citep{Shmilovich, Wang2017MesoscaleLandscapes} We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.

%%%%%%%% Methods comments %%%%%%%%%%%%%%%%%

%We found close agreement with explicit ion simulations performed with 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} (supplemental) but 7x slower simulation speed made adequate sampling more difficult to achieve under these conditions.

%We initialized explicit ions such that 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. We used a periodic box size of 7.774 nm and an effective oligo concentration of 7 mM. We used an Ewald summation to calculate long range Coulombic interaction between DNA and ions using a real space cutoff of 2 nm.

\subsection{Experimental Methods}

\subsubsection{Sample Preparation}

Each DNA oligonucleotide was purchased from Integrated DNA Technologies (IDT) at desalt grade purity. Oligonucleotides were purified with 3 kD cutoff centrifugal filters (Amicon). All labile protons were exchanged in deuterium oxide (\ce{D2O}, Cambridge Isotopes, 99.9\%). Oligonucleotides were prepared at a total strand concentration of 2 mM in 50 mM pD 7.2 sodium phosphate buffer with 240 mM NaCl and 18 mM \ce{MgCl2}. Prior to each measurement, DNA solutions were placed in a water bath at 90$^{\circ}$ C for 3 minutes and then cooled to room temperature under ambient conditions.

\subsubsection{T-Jump IR Spectroscopy}\label{sec:Tjump}

The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}. The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO). Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump. T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs. t-HDVE spectra were acquired with a parallel pulse polarization scheme. Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum. t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.

The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a home-built brass jacket. The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket. The T-jump temperature change ($\Delta$T) was set to $\sim$15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam. The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.

\subsubsection{Determination of Fast and Slow Dissociation Rates}\label{sec:TjumpAnalysis}

To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates were taken as the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response ( $k_d^\mathrm{fast}$) is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant ($k_d^\mathrm{slow}$) was extracted from the observed rate of the slow response ($\lambda^\mathrm{slow}$) using a two-state model for self-complementary oligomers.~\citep{Bernasconi2012RelaxationKinetics}
	\begin{equation}\label{e1}
	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a
	\end{equation}
$[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a$ is the association rate constant. In practice, Equation~\ref{e1} is recast in terms of the dissociation equilibrium constant ($K_d$) to solve for $k_d^\mathrm{slow}$ and $k_a$.
	\begin{equation}\label{e2a}
	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d}{K_d+4[S]_{T_f}}
	\end{equation}
	\begin{equation}\label{e2b}
	k_a = \frac{k_d}{K_d}
	\end{equation}
$k_d^\mathrm{slow}$ and $[S]_{T_f}$ come from melting curves measured with FTIR spectroscopy that are fit to a two-state all-or-nothing dissociation model.~\citep{Marky1987CalculatingCurves} We used the FTIR melting curves previously reported in Ref.~\citep{Sanstead2016}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{\label{sec:Results}Results}

\subsection{Sequence-dependent coarse-grained kinetics recapitulate T-jump IR measurements} \label{sec:time}

% rerun these at the actual final temps and just see what happens? Doesn't really matter where Tjump occurs and could make things easier as a whole...
% talk about rerunning AT-all and GC-mix=

We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments. We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' (10-30 $\mu$s) and ``fast'' (70-100 ns) observed rates in the T-jump response (Section \ref{sec:Tjump}). As discussed in Section \ref{sec:TjumpAnalysis}, the fast ($k_d^\mathrm{fast}$) experimental rate is an observed rate whereas we have derived a dissociation rate constant ($k_d^\mathrm{slow}$) for the slow process.\noteb{is this stipulation clear enough?} The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response is considered to correspond to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds,~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} and all-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (SI Figure 4). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which the central base pairs first separate to a distance of 1.3 nm. This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state. We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the Watson-Crick (WC) interaction. We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.

We present in Fig.~\ref{fig:relaxation-comparison} a comparison of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ estimated by computation and experiment. Although the 3SPN.2 model was not directly fitted against kinetic data,~\cite{Hinckley2013AnHybridization} its predictions of sequence dependent T-jump relaxation rates are, within a systematic scaling factor in time and systematic shift in temperature, in good agreement with observed experimental trends. It is well-known that the smoothing of the underlying free energy landscape induced by coarse-graining artificially accelerates the kinetics of coarse-grained molecular simulations and that different degrees of freedom may be accelerated by different factors.~\citep{Marrink2007TheSimulations, Fritz2011MultiscaleDynamics, Marrink2013PerspectiveModel} We find that the simulated slow responses corresponding to center-of-mass translation of the strands during dissociation of the duplex is $\sim$10$\times$ accelerated relative to experiment, whereas the fast responses corresponding to fraying of the terminal bases is $\sim$120$\times$ accelerated. We apply these sequence-independent scaling factors to our reported computational values. Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal +4 K shift corrective calibration to our computational results. We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison. The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Fig1.pdf}
        \caption{Experimental measurements and computational predictions of slow and fast at T-jump IR responses. Experimental results are shown in terms of the final T-jump temperature (a) The simulated and T-jump dissociation rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales. (b) The Simulated fraying rate constant and fast observed rate $k_d^\mathrm{fast}$ \noteb{again trying to make the distinction between observed rate and dissociation rate here} corresponding to terminal base-pair fraying on short time scales. The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics. The simulated temperature in all cases is also corrected by +4 K to correct for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

In Fig.~\ref{fig:relaxation-comparison}a we report an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves. The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone. Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general trends exponential trend is preserved. This may be a result of its high propensity to fray as we will discuss further in Section \ref{sec:GC-core} \noteb{Is this sufficient discussion here since we bring it back up later?} In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end. This can be understood via the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences. Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements. The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.

%Coarse-grained time scales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation time scales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded ``fast'' and ``slow'' amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3SPN.2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
%We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4$^{\circ}$C between experiment and simulation fits. While 3SPN.2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (Figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

% GC-core fast time scales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor. We maintained a consistent accegqrelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses. It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.

% need to update this fig with labels and temp shift

% A documented short-coming for the 3SPN.2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}. This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.

%The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with time scales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN.2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted time scales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying time scales in order to extract rate approximations for each sequence at a series of temperatures.

%For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process. In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying. We observe similar trends in the experimental data, although comparisons at the highest temperatures were limited due to mixing between the fast and slow responses (Figure \ref{fig:relaxation-comparison}). We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN.2 fraying dynamics are accelerated by about 100-200x relative to experiment. It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom. In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation. Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al. and that 3SPN.2 can recapitulate some sequence-dependent fraying effects.
% need a better transition here back into SRV-MSMs?

\subsection{SRV-MSMs for each sequence}

We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent long-time kinetic models. MSMs are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates. We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences. Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium. By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F). In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates. The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model. The identity and fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways. Immediately we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior. In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM. These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates. The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization. Since each system is simulated at its respective melting temperature it is not surprising that the leading time scale is approximately equal for all four systems. The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates. The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes. The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization. We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.

%We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer. We identified seven kinetically relevant states that were captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities.  Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the probability of a transition to another state within the MSM lag time.

%Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5$^\prime$ vs. 3$^\prime$) and the number of shifted motifs relative the native state (1 vs. 2). GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states. Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}. We also show that PCCA+ clustering assigns states to relative free energy minima in the tICA space.

% rewrite this into one paragraph

% change state names and scientific notation
% include D here? can see what it would look like...

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=170mm, scale=1]{Fig2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
        \caption{ Summary of thermodynamics and kinetics quantities extracted from sequence-specific MSMs. (a) Seven metastable macrostates sampled by at least one of the four sequences. Cartoons represent 10bp sequences in the 5$^\prime$ to 3$^\prime$ direction, where red-to-blue contacts indicate WC base pairing and black indicates an unbound bases. Molecular rendering show 3D examples of each configuration. (b) Number of frames (out of $10^7$ total) clustered into each of the seven macrostates. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimator and are very small compared to total populations. (c) Coarse-grained MSMs and molecular renderings for each sequence. Arrow thickness corresponds to the flux between between states (using of minimum flux cutoff of $3\times10^{-6}$), and the overlaid number represents the probability of making a transition within the 1.2 ns lag time. The radii of state circles correspond to the log of state populations. (d) Distribution of leading timescales for each sequence. The leading timescales for each sequence corresponds to hybridization/dehybridization process. AT-all and GC-end display a spectrum of converged slow modes corresponding to shifting processes. GC-core has one faster timescale corresponding to fraying processes.
        \noteb{Previous (more compact) version Fig2 alternate}
    }
    \label{fig:allseq_table}
\end{figure}

\subsection{Comparison of MSM thermodynamic predictions with NN model} \label{sec:NN}

We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association. The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends. We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences. The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies. Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare the free energies predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model. A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the Supporting Information.

As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized (H) state by as much as 14.7 kJ/mol. (The free energy of the hybridized state is set to zero by construction for both the MSM and NN models.) The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement. Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the 5S2, 3S2, 5S4, and 3S4 states? First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data. There are approximations inherent in the 3SPN.2 model and the NN model was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions. Second, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.\citep{Michele2014EHybridization} Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability. \noteb{not sure the best way to phrase this, but they could technically still be considered inert tails}. In this case the NN and MSM free energy predictions are in good agreement. Two of our sequences, AT-all and GC-end, occupy macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}). \noteb{made more changes here} For both sequences, our simulations show that 5S2 is both the most stable of shifted states and has the smallest offset (4.0 kJ/mol for AT-all and 10.0 kJ/mole for GC-end) compared to NN predictions \ref{fig:NN_table}b. The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol for AT-all and 14.3-14.7 kJ/mole for GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction. It is known from differential scanning calorimetry studies~\citep{Dickman2012ThermodynamicDNAs} that 3$^\prime$ inert tails are more destabilizing than 5$^\prime$ tails, with the differential behavior attributed to a combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure. \citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization}. Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$). For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all. This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all. Given that terminal mismatches tend to be slightly stabilizing \citep{Santalucia2004TM}, we might expect a lower discrepancy than the  10.1-14.7 kJ/mol offset that we observe between NN and MSM results \noteb{May also want to note that 3spn.2 may not be fully capturing the stabilizing effects of dangling ends (there are four of these for each shifted state): Although our NN calculations include four dangling end contributions for all shifted states, 3SPN.2 may not fully capture the stabilizing effects of dangling ends. Indeed, cross-stacking and base stacking interactions should provide stabilization from dangling ends, but more granular interactions at the duplex termni may not be present. }. This offset may arise in part from the excluded volume interactions used to account for mismatched bases in 3SPN.2, however we attribute it mainly to inert tail effects. \noteb{could add something here about GC-end shifted states having a higher population (and therefore greater contribution to the kinetic landscape) if terminal mismatches were better accounted for in the model. But this would be speculative...} The GC-end NN predictions run contrary to the expectation that the 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the MSM predictions are consistent with this trend. Indeed, the MSM model for GC-end does not even register the 3S4 macrostate as a stable metastable conformation for the duplex.

In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails. In these instances the NN model overestimates macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol. The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero. With the exception of the frayed state (F4), the NN model predicts lower free energies than the MSM and therefore tends to overestimate macrostate stability relative to the hybridized state. (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction. (b) Free energy difference between the NN prediction and simulated value for each macrostate. \noteb{Should the y-axis be more descriptive here?}}
        \label{fig:NN_table}
	\end{center}
\end{figure}

%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of ``dangling ends'' -- unpaired bases adjacent to the paired duplex -- and ``inert tails'' -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3$^\prime$ inert tails have stronger effects than 5$^\prime$ tails. Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.

% and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths. It has been shown that 5$^\prime$ dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3$^\prime$ ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}. These effects might be attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure \citep{Doktycz1990ThermodynamicATGC}.

% This leads us to believe that 5$^\prime$ vs. 3$^\prime$ differences -- attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.

\subsection{Out-of-register states facilitate hybridization and dissociation dynamics (AT-all, GC-end)}

In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms. We now proceed to analyze these predictions for each of the four sequences and illuminate the relationship between sequence and dynamics. Two of our sequences, AT-all and GC-end, support out-of-register metastable states, and we commence our analysis with the role of these shifted states.

AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c). Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization. Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H. Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. Once in one of the four shifted states, the 5$^\prime$ vs. 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate. Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states. The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D. Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D. The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state. A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the union of direct and indirect pathways) proceed through one or more out-of-register shifted states. Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time. Mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s. The longer D $\dashrightarrow$ H time scale is expected given the number of out-of-register states that can form and re-dissociated before reaching H. These values are in close agreement with the leading implied time scale shown in  Fig.~\ref{fig:allseq_table}.  \noteb{can make note of the fact that Meta $\rightarrow$ (all other states) MFPTs align with the higher order modes as well? Not sure if this would require a more in depth discussion}

%Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g. 5S2 $\rightarrow$ H is more likely than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.

%Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state. The 3S4 and 5S4 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.

GC-end comprises the next most complex MSM with the introduction of the terminal G:C pairs maximally preserving the repetitive tract of A:T base pairings. The GC-end MSM possesses all of the same macrostates in its dynamical landscape with the exception of the 3S4 state (Fig.~\ref{fig:allseq_table}c). As discussed in Section \ref{sec:NN}, the 3S4 state is rendered unstable within the lag time of our MSM due to the presence of the destabilizing 3$^\prime$ inert tail and only four WC base pairings compared to six in the case of AT-all. Analysis of the transition probabilities reveal significant differences compared to those in the AT-all kinetic network. Commencing from the dissociated state D, we observe a similar transition probability to the 5S4 state as for AT-all, but once in the 5S4 state there are no significant transition probabilities to any other state except back to D. As such, the 5S4 state acts as a kinetic trap rather than as an intermediate to hybridization. The D $\rightarrow$ H and H $\rightarrow$ D transition probabilities are commensurate with those for AT-all. However, the D $\rightarrow$ 5S2 and D $\rightarrow$ 3S2 transition probabilities are half or less of those in AT-all, and the reverse transitions are an order of magnitude larger. This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that contain only six WC pairs compared to eight in AT-all (Fig.~\ref{fig:NN_table}). The 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H transition probabilities are more than an order of magnitude larger than in AT-all, which may again be attributed to the lower thermodynamic stability of the two shifted states relative to the hybridized state H. Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately 6.8 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. A transition path theory analysis of the MSM reveals that only 7\% of hybridization events D $\dashrightarrow$ H and dehybridization events H $\dashrightarrow$ D proceed through one or more out-of-register shifted states. The significantly reduced role for out-of-register shifted states in mediating the hybridization and dissociation pathways for GC-end relative to AT-all is consistent with the reduced thermodynamic stability of these states due to the elimination of possible out-of-register WC base pairing for the terminal G:C pairs and therefore a reduced accessibility of these states in the GC-end kinetic network. A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are both notably lower than AT-all values. This may be attributed in part to shorter meta-stable lifetimes, or to the higher temperature at which this sequences was simulated. Again, the average of these MFPTs align well with the leading implied time scale values in Fig.~\ref{fig:allseq_table}.

%The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all. This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed.

%While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states. These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior. Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences.

The out-of-register kinetic landscape that define AT-all and GC-end hybridization have been informed by numerous previous computational studies. Simulations have identified internal displacement mechanisms capable of correcting base pair alignment in 3SPN.2 \citep{Hinckley2014Coarse-grainedEffects} as well as in the coarse-grained oxDNA \citep{Romano2013DNADependence} and BioModi \citep{Markegard2015} models \noteb{not sure what you think of naming these specifically}. In all cases, these mechanisms were shown to be crucial components of the hybridization pathway for homogeneous and repetitive sequences. Xiao et al.\ performed an all-atom energy landscape-based analysis of 5$^\prime$-GGGGGG-3$^\prime$ and 5$^\prime$-GCGCGC-3$^\prime$ hexamers.~\citep{Xiao2019} Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism. (In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.) This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events. It is conceivable that the stronger hydrogen bonding in G:C WC pairs relative to A:T pairs may render out-of-register shifted states less favorable to hybridization by suppressing fluctuation-driven rearrangements,~\citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} but additional studies would be required to reconcile these observations.

\subsection{Central GC placement induces long-lived frayed states (GC-core)} \label{sec:GC-core}

The GC-core MSM represents a departure from the relatively rich and complex kinetic networks dominated by out-of-register shifted states to a much simpler one dominated by fraying (Fig.~\ref{fig:allseq_table}c). The MSM contains only three states -- hybridized H, dehybridized D, and frayed F4. The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs (it may be less due to rapid 1-2 terminal base pair fraying) on one side or other of the core; the other run of four A:T pairs remains free. (As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4. Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns. The stability of the state is attributable to the enthalpic stabilization offered by the six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.

Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25-times more likely than dissociation F4 $\rightarrow$ D. Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation. The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure how meaningful this is considering they are coming from different source states?}. We have noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation. Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a). Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s. \noteb{these are longer than both AT-all and GC-end, (cite repetitive sequences have slower hybridization time scales). The F4 is unlikely to contribute to the difference between states, rather we attribute this to a slight deviation in the total state populations -- the D equilibrium probability is about 9\% higher than H and F4 combined. Despite extensive sampling, these asymmetries arise from DNA's high temperature sensitivity and create the impression of a shorter H $\dashrightarrow$ D MFPT relative to D $\dashrightarrow$ H}.

Lattice models have previously identified frayed states as putative intermediates in DNA hybridization/dehybridization.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2016} Araque et al.\ studied a 5$^\prime$-ATGCGCAT-3$^\prime$ octomer using a lattice model and identified a symmetrically A:T frayed state as a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} We previously studied the four sequences that are the subject of the present work using T-jump IR and 2D IR spectroscopy and identified GC-core as possessing the highest deviation from two-state behavior during dissociation when neglecting out-of-register contributions.\citep{Sanstead2016} This result was interpreted to arise from loss of A:T contacts and fraying around the central G:C core, and this hypothesis was supported by lattice model calculations that predicted the GC-core conformational ensemble to possess substantially more frayed configurations than the other three sequences.~\citep{Phys2019} Follow-up T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} These prior results are consistent with the present findings that expose the GC-core sequence to be the only sequence that occupies the F4 frayed state and therefore the only one possessing a metastable frayed state on time scales exceeding the 1.2 ns lag time of our MSMs.

%When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al. found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions).\citep{Sanstead2016} As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core. While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs.\citep{Phys2019} Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.

%Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar time scale.

%Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.

\subsection{Disruption of repetitive AT tracts promotes two-state ``all-or-nothing'' kinetics (GC-mix)}

The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable on the 1.2 ns time scale of our MSM. The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states. We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM. This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the configurational entropy of the frayed two-base AT tails relative to the frayed four-base AT-tails, and the more limited disruption of the helical duplex by the shorter tails that offers less kinetic inhibition to its reformation from the transiently frayed structures. Simulations of GC-mix were conducted at the melting temperature of the sequence, so the simple two-state MSM should have, by construction, equal populations of, and equal transition probabilities between, the H and D states. As we observed for GC-core, a higher dissociated state population produces a larger effective MFPT for D $\rightarrow$ H (2.9 $\mu$s) than H $\rightarrow$ D is (2.3 $\mu$s). Again, the average of these time scales match the dominant leading implied time scale for this sequence (Fig.~\ref{fig:allseq_table}c).

%Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix. Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively. Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model. We did, however, observe substantial fraying of the two AT termini in the simulation data. Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and transient that very few progress to full dissociation compared to the F4 state. Furthermore, one or two-base-pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.

Given the very simple two-state ``all-or-nothing'' behavior of GC-mix and the absence of any intermediate metastable states, we sought to interrogate our simulation trajectory data to elucidate the hybridization and dehybridization mechanisms. To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events. We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more of each in SI Figure 3. In hybridization, we observe a global decrease in all 10 distances as the strands approach one another, but uniformly observe the formation of some key native WC base pair contacts immediately prior to duplex formation. Specifically, formation of one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs \noteb{this is a more general observation which is in line with all events} This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs. Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes. These dissociation dynamics are similar to previously reported ``fraying-peeling'' mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}

\begin{figure}[ht!]
	\begin{center}
	\includegraphics[width=180mm, scale=1]{Fig4.pdf}
        \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling. Tracking of the 10 intermolecular distances between native WC base pairs over the course of an (a) hybridization event and (b) dehybridization event. Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot. Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange). Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C WC pair and adjacent A:T pair or central A:T pairs first form prior to rapid formation of the complete duplex. Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex. Four additional hybridization events and four additional dehybridization events are presented in SI Figure 5.
        }
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

\subsection{T-jump IR responses are correlated with the complexity of the dynamical landscape}

We hypothesized that the complexity of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models. As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively. Although there are no distinct ``middle'' responses at the expected shifting timescales, the response profile can inform the distribution of dynamics contributing to the signal \citep{Hamm2006StretchedFolding, Johnston2006StretchedDecays}. In particular, the AT-all fast response exhibits more stretched exponential character ($\beta$ = 0.3) than any of the other three sequences (Figure \ref{fig:T-jump_stretching}). We can interpret the stretch as fraying from many configurations, each having a slightly different rate. The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM. For the GC-end response, however, the stretching factor is not appreciably different from the other sequences and comparison of 1660 and 1600 cm$^{-1}$ wavelengths indicate that the fast response is much stronger in the G/T rather than A/T region. This further suggests that it corresponds to terminal fraying rather than shifting. Overall, there are several reasons why the T-jump response may not be directly sensitive to shifting. In the case of GC-end, the low population of shifted states alone may diminish contributions below a detectable limit. For AT-all, the particularly slow timescales associated with the most common shifting process (H $\leftrightarrow$ 5S2) can blend in with the slow dissociation response. In both cases, it is unclear how populations of shifted states change with temperature, making these processes difficult to identify with T-jump techniques.

These comparisons also highlight the importance of considering and defining multiple types of deviations from two-state behavior. In this work, we define this deviation by the complexity of the MSM network and number of accessible metastable states. This definition suggests the following order of kinetic complexity: AT-all $>$ GC-end $>$ GC-core $>$ GC-mix. As we have shown, this trend is closely related with the degree of repetitive AT motifs in each sequences. In our previous work, we did not consider out-of-register states as part of the experimental or lattice model analysis, and therefore determined our ordering (GC-core $>$ GC-mix $>$ AT-all $>$ GC-end) in terms of frayed state populations prior to duplex dissociation\citep{Sanstead2016, Sanstead2018DirectDehybridization}. Depending on the biological system or nanotechnology application of interest, these forms of deviation may hold varied importance. For example,  \noteb{could something here, but don't want to dwell on this too much}

%In addition to identifying relaxation time scales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response. In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}). Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response. On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source. We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-end and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two. Although reproducing these stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.

% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.

%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.

%We emphasize that this does necessarily reflect shifting relaxation time scales themselves but more likely changes in the fast responses -- e.g. fraying out-of-register -- within the collection of metastable states.

\begin{figure}[ht!] % right place for NN table? could go to SI
	\begin{center} 
        \includegraphics[width=175mm, scale=1]{Fig5.pdf}
        \caption{T-Jump IR responses reflect sequence-dependent conformational heterogeneity. (a) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s. Normalized time traces for each sequence are shown at (b) 1600 cm$^{-1}$ and (c) 1660 cm$^{-1}$. The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations. Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$. The stretched exponential describes the process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature. The stretch factor $\beta$ from each fit is listed.}

        \label{fig:T-jump_stretching}
	\end{center}
\end{figure}

%\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}

%Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states. These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural details of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S2 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical ``top'' and ``bottom'' of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI. Diffusion maps built from samples of the 3S2 states are also shown in the SI.

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
%        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end. Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends. The lower right region of the diffusion map space shows a high concentration of ``shifted loop'' base pairs that may facilitate the out-of-register GC-end transition.}
%        \label{fig:GC-end_dmaps}
%	\end{center}
%\end{figure}

%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3$^\prime$ end and its shifted complementary pair. As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. The second diffusion mode, which correlates highly with the average distance between 3$^\prime$ and 5$^\prime$ ends, highlights structural variances within the two sequences distributions. We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration. We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register. These ``shifted-loop'' bonds appear uniquely stable for GC-end conformations, and suggest a transition state through which hybridization is facilitated. This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the equivalent AT-all pathways.

%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3SPN.2 \citep{Hinckley2013AnHybridization}.

%Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.

% need to work on this if adding to this section (might belong somewhere else or not at all?
%Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}. In the context of the 3SPN.2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}. The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.

%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3SPN.2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%

%\subsection{Sequence-dependent fraying determines direct hybridization pathways}

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
%        \caption{Sequence-dependent population distributions for trimmed hybridization events. $\Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression. Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins. Heat maps show the average intermolecular base distances in nm at increasing values of $\Psi_2$, where the main diagonal shows native base pair distances.}
%        \label{fig:Fig8}
%	\end{center}
%\end{figure}

%To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states. To achieve higher resolutions, we reran selected simulation with a 10x higher save rate and analyzed a 1000 frame region (10 ns simulation time) for each event. We built diffusion maps on a set of 48 such ``trimmed'' events evenly sampled across sequences. We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance. Indeed, $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence. This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.

%Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}). The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first Watson-Crick contacts are made. In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases. As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs. GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state. This is consistent with the downhill nucleation-zippering mechanism described above. Although differences between AT-all and GC-end are more subtle, overall deviations from two-state behavior follow predictions by Sanstead et. al.\citep{Sanstead2016}  where GC-core > GC-mix > AT-all > GC-end. This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time. Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.

%This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.

% add new info here, not sure if shifting discussion is necessary
% take away message is that these are good proxies for the responses
% although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response

%\subsection{Limitations}

%Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering. These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other time scales of interest. Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process. We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations. Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence and were working near memory limits when training SRVs and building SRV-MSMs.

%In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations. For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends. In general, coarse-grained models produce a smoother free-energy surface which can results in much faster motions between states. This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment. Furthermore, it may be easier to cross between states -- e.g. a 5S2 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.

\section{\label{sec:conc}Conclusions}

We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix) \noteb{again, do we want keep the 5$^\prime$-ATATATATAT-3$^\prime$ format here or not bother}. We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics. T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the predictions of the Markov state models. Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex. In particular, we found AT-all to possess the richest and most complex kinetic landscape of the four sequences that is dominated by out-of-register shifted states that participate in 33\% of hybridization and dehybridization events. These results are in accord with observations from HDVE T-jump IR spectroscopy that a spectrum of fast fraying responses likely originating from an ensemble of out-of-register states. Introduction of the G:C pairs at the end of the strand maintains a eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all. Similar to AT-all, \rood{XX}\% and \rood{XX}\% \noteb{not sure if we want to change the structure here given that these values appear to be the same (and are each only 7\%)} of hybridization and dehybridization events, respectively, in GC-end proceed through one or more shifted states. Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core. In this case, no metastable out-of-register shifted states are observed in our models with the hybridization and dehybridization pathways all passing through a strongly metastable frayed state in which one or other of the four-base AT-tracts is unbound resulting in two free AT-tails. Finally, placing the G:C bases between the center and end of the strand to maximally disrupt the repetitive AT tracts results in no metastable out-of-register or frayed states and results in simple two-state ``all-or-nothing'' hybridization/dehybridization behavior. Analysis of the specific pathways reveals hybridization to largely proceed by a nucleation-zippering mechanism and dehybridization to proceed by a fraying-peeling mechanism. The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix \noteb{not sure exactly what to say here}

In sum, our results demonstrate the profound effect of sequence upon the kinetic landscapes, metastable states, and hybridization/dehybridization mechanisms of short DNA oligomers. Our analysis of this small family of sequences expose preliminary design principles for the (meta)stability of out-of-register and frayed states but we anticipate much greater richness in the landscapes will emerge with studies of longer and more diverse sequences. Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons. We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for applications such as DNA-PAINT,~\citep{Strauss2020UpDNA-PAINT} DNA barcoding,~\citep{Shah2019} and novel developments in dynamic DNA nanotechnology.

%We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics. By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior. In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these dynamics are more relevant for AT-all. On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted. Our computational approach and results show strong comparisons to T-jump experiments at ns and us time scales. Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons. These insights should be leveraged for enhanced understanding of \emph{in vivo} hybridization processes, optimized sequence design in application such as DNA-PAINT and barcoding, and novel developments in the growing field of dynamic DNA nanotechnology.

\clearpage
\newpage

%\bibliographystyle{unsrt}
\bibliography{references_210505}

%\bibliography{refs_mendeley2}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
\newpage

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC.pdf}
        \caption*{TOC image \noteb{could consider adding an experimental aspect to this?}}
        \label{fig:TOC}
	\end{center}
\end{figure}

\end{document}

### 2021-05-10 manuscript.tex
added thermo section without dangling ends

@@
 \usepackage[%  
-    colorlinks=false,
+    colorlinks=true,
     pdfborder={0 0 0},
-    linkcolor=blue
+    linkcolor=blue,
+    citecolor=black
 ]{hyperref}
@@
 \usepackage{framed}
+
+%add refs to SI
+%\usepackage{xr}
+%\externaldocument{SI}
+
 %\usepackage[hidelinks]{hyperref}
@@
 \usepackage{mathrsfs}
+\usepackage{comment}
+\usepackage[normalem]{ulem}
+\usepackage{algorithm}
+\usepackage{subfig}
+\usepackage{algpseudocode}
 
@@
 \newcommand{\state}[1]{$\mathcal{S}_{#1}$}
-
-\newcommand*{\rood}[1]{#1}
-\newcommand*{\blauw}[1]{#1}
-\newcommand*{\groen}[1]{#1}
-\newcommand*{\blauwr}[1]{#1}
 \newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
@@
 
-%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\bood}[1]{{\color{blue}{#1}}}
+\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
+\newcommand*{\groen}[1]{\textcolor{green}{#1}}
 
-\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
+\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
 %\newcommand*{\noter}[1]{}					% notes off
@@
 \usepackage[disable]{todonotes} % notes hidden
-\usepackage{comment}
-
-\usepackage[normalem]{ulem}
-\usepackage{algorithm}
-\usepackage{subfig}
-\usepackage{algpseudocode}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[]{Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides}
+\title[]{Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy}
 
@@
 
-\begin{abstract}
+\begin{abstract} % 250 words max for JACS
 
-Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood.
-In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement.
-We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model construction.
-Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides.
-For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states.
-In contrast, G:C pairs near the center of the duplex induce more pronounced fraying dynamics through which hybridization and dissociation are facilitated.
-In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model.
-Our computational predictions show agreement with experiments and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.
+\noindent A robust understanding of the sequence-dependent thermodynamics of DNA hybridization has enabled rapid advances in DNA nanotechnology.
+A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped.
+In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
+For a repetitive ATATATATAT sequence \noteb{not sure if we want to use 5$^\prime$-ATATATATAT-3$^\prime$ notation here as well?}, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations.
+Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways.
+Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical two-state model of hybridization/dehybridization over a smooth landscape with no intermediate metastable states.
+Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.
 
-% revise "excellent accord" above?
-
-\noindent \url{https://www.overleaf.com/project/5e9e5110c524b8000192c548}
+%\noindent \url{https://www.overleaf.com/project/5e9e5110c524b8000192c548}
 
 \end{abstract}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing.
-\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Indeed, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields.
-The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}
+\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnological applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the \textit{dynamical}, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves.
+\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence.
+\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence, Shah2019}.
+A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved.
+For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the extent to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
+The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.
 
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.
+Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization.
+Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} \noteb{these references correspond to CG enhanced sampling studies, I can refer to enhanced sampling on AA here instead (\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations}) but I think the CG studies should be mentioned since not much has been done for AA hybridization} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacements \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami structure \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
 
-Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates.
-For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally and appears highly sequence-dependent.
-Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al.
-highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.
-
-In this work, we study the same four sequences explored by Sanstead et al.
-in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3SPN2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent.
-We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
-
-We find that 3SPN2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends.
-Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation.
+In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
+This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
+\citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new and existing experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
 Consistent with previous studies, \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
-Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior.
-Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process.
-Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs.
-This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
-These results can inform a generalized model between DNA sequence and hybridization mechanism with application towards equilibrium binding technologies such as DNA-PAINT and barcoding.
+Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states.
+Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states that facilitate the hybridization/dehybridization process.
+Analysis of the configurational ensemble within each metastable state reveals the conformational heterogeneity and dynamical interconversions proceeding on time scales that lie below those resolvable within the long-time kinetic model.
+Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to sculpt and qualitatively alter the dynamical behaviors of the system.
+We anticipate that this new foundational understanding, and the extension of our approach to more extensive families of DNA sequences, can guide the rational design of DNA oligomers with tailored kinetic properties engineered for DNA nanotechnology applications such as DNA-PAINT.
+\citep{Shah2019, Strauss2020UpDNA-PAINT}
 
@@
 Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.
 
@@
 
-%Progress has also been made in modeling kinetics for certain dynamic DNA systems, such as toehold exchanges and optical barcodes \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting}.
-Sequence effects on association and dissociation rates were predicted using combined experimental and machine learning approaches, however higher order dynamics are less well understood \citep{Zhang2018PredictingSequence}.
-Hairpin \citep{Woodside2006DirectAcid} and in duplex hybridization \citep{yer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization} studies have revealed sequence-dependent effects and varied differential deviations from "all or nothing" behavior.
-In temperature-jump spectroscopy, G:C placement alters the slow kinetic response attributed to dissociation and a fast response likely corresponding to terminal A:T fraying.
-Experimentally, however, it is difficult to pick up on dynamics that have similar spectroscopic signatures and to tease apart temporally close responses that might be combined into one stretched exponential signal.
-
 % include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior.
@@
 We perform longer equilibrium sampling near Tm to get a sense of sequence equilibrium behavior, as well shorter simulations at a range of higher temperatures to replicate T-jump data.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.
+Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.
 
@@
 
-\subsection{\label{sec:methods}Simulation set up}
+\subsection{Computational Methods}
 
-The 3SPN2 coarse-grained model was designed to accelerate DNA simulations relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
-The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3SPN2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}
+\subsubsection{Molecular dynamics simulations}
 
-We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff.
-We ran our simulations in the NVT ensemble with a Langevin thermostat that modeled interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states.
-We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps.
-For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation.
-Half of runs were initialized from the native hybridized state and half from a dissociated state.
-The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence.
-We observed between 55-100 hybridization/dissociation events for each of the four sequences.
+We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC3-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix).
+We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+\citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation.
+\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.
 
-\subsection{\label{sec:methods}Featurization}
+All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model.
+\citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
+Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers.} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s.
+\citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius.
+\citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat.
+\citep{Schneider1978Molecular-dynamicsTransitions}.
+Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature.
+The temperature at which the hybridized and dissociated populations were closest was selected.
+The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step.
+We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state.
+The initial hybridized state was set, according to 3SPN.2 documentation, to the coordinates of Arnott et al.\ ~\citep{Arnott1976CRCBiology} The dissociated state was generated by shifting one oligomer from the hybridized state by 1 nm in the x, y, and z directions.
+The magnitude of this shift did not affect the results so long as all native WC bonds were sufficiently separated.
+\noteb{No energy minimization (no 3spn2 precedent on this).
+The lammps default emin techniques may not work for this type of CG forcefield.
+Given the sampling it seems that  1 $\mu$s of equilibration should be sufficient?} Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the system temperature.
+\addref{Maxwell-Boltzmann -- \noteb{not sure what to use here, usually not cited?}} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
+The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.
 
-A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
+\subsubsection{Markov state model construction}
+
+Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionV1.0} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
+MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
+Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+Calculations were performed using the PyEMMA software package.
+\citep{Scherer2015PyEMMAModels}
+
+\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances.
+We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the $(10 \times 10)$ 100-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair.
+This featurization naturally accounts for intermolecular configurations via the pairwise distances between bases on different strands.
+\noteb{previous text had suggested we used the 190-dim features, but we actually used $100-45 = 55$-dim features since the intramolecular features contributed no additional kinetic variance (I've updated accordingly)} One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances.
+Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2.
+\citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands.
+VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization.
+\citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top 5 modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.
+
+\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering.
+The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA.
+\citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence.
+SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000.
+We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics.
+\citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
+\citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs.
+We guarded against overfitting using 5-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
+We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA GK210GL GPU card.
+A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively \noteb{not sure the best way to explain this succinctly}.
+(Figure 1 in the Supporting Information).
+This motivated the construction of a n-dimensional embedding corresponding to n leading SRV modes retained for each sequence.
+
+\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering.
+The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.
+
+\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM.
+The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2.
+We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
+At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively.
+The identification of $(i-1)$ implied time scales \noteb{(may be 1-2 additional converging, but these correspond to diffusive processes that are irrelevant or unhelpful in determining the hybridization mechanisms.
+Still, this might come up when looking at the the SI Fig 2} implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates.
+We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates.
+These macrostate MSMs constitute our terminal kinetic models.
+We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}.
+This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$.
+We present in SI Figure 3 the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four kinetic models.
+
+%\subsection{Featurization}
+
+%A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
 Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
+To account for this, we averaged permutable distances (45 pairs in total) together into a ``symmetrized'' feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\cite{ Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
 By comparing scores across several candidate feature sets (Figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
@@
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=80mm,
-        scale=0.5]{Fig1.pdf}
-        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates.
+%\begin{figure}[ht!]
+%	\begin{center}
+%        \includegraphics[width=0.55\textwidth]{Fig1.pdf}
+%        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates.
 Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods.
 This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
-        \label{fig:allseq_features_vamp2}
-	\end{center}
-\end{figure}
+%        \label{fig:allseq_features_vamp2}
+%	\end{center}
+%\end{figure}
 
-\subsection{\label{sec:methods}SRVs}
+%\subsection{\label{sec:methods}SRVs}
 
-The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed.
+%The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed.
 While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA.
 SRVs were first developed by Chen et al.
-as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
+as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and time scales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
 This is a key attribute to our system as 10 million frames with 55 features per frame are used for each sequence.
@@
 
-\subsection{\label{sec:methods}SRV-MSMs}
+%\subsection{\label{sec:methods}SRV-MSMs}
 
 % need some more detail here -- talk to Andy about this
-MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
+%MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
@@
 To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence.\citep{Scherer2015PyEMMAModels} In a similar approach to Sidky et al.
-we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.
+we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied time scales convergence.
 The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.
 
-\subsection{\label{sec:methods}Diffusion Maps}
-Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
+%\subsection{\label{sec:methods}Diffusion Maps}
+%Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
 We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies.\citep{Shmilovich, Wang2017MesoscaleLandscapes} We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.
@@
 
+\subsection{Experimental Methods}
+
+\subsubsection{Sample Preparation}
+
+Each DNA oligonucleotide was purchased from Integrated DNA Technologies (IDT) at desalt grade purity.
+Oligonucleotides were purified with 3 kD cutoff centrifugal filters (Amicon).
+All labile protons were exchanged in deuterium oxide (\ce{D2O}, Cambridge Isotopes, 99.9\%).
+Oligonucleotides were prepared at a total strand concentration of 2 mM in 50 mM pD 7.2 sodium phosphate buffer with 240 mM NaCl and 18 mM \ce{MgCl2}.
+Prior to each measurement, DNA solutions were placed in a water bath at 90$^{\circ}$ C for 3 minutes and then cooled to room temperature under ambient conditions.
+
+\subsubsection{T-Jump IR Spectroscopy}\label{sec:Tjump}
+
+The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}.
+The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO).
+Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump.
+T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs.
+t-HDVE spectra were acquired with a parallel pulse polarization scheme.
+Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum.
+t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.
+
+The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a home-built brass jacket.
+The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket.
+The T-jump temperature change ($\Delta$T) was set to $\sim$15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam.
+The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.
+
+\subsubsection{Determination of Fast and Slow Dissociation Rates}\label{sec:TjumpAnalysis}
+
+To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates were taken as the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response ( $k_d^\mathrm{fast}$) is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant ($k_d^\mathrm{slow}$) was extracted from the observed rate of the slow response ($\lambda^\mathrm{slow}$) using a two-state model for self-complementary oligomers.~\citep{Bernasconi2012RelaxationKinetics}
+	\begin{equation}\label{e1}
+	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a
+	\end{equation}
+$[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a$ is the association rate constant.
+In practice, Equation~\ref{e1} is recast in terms of the dissociation equilibrium constant ($K_d$) to solve for $k_d^\mathrm{slow}$ and $k_a$.
+	\begin{equation}\label{e2a}
+	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d}{K_d+4[S]_{T_f}}
+	\end{equation}
+	\begin{equation}\label{e2b}
+	k_a = \frac{k_d}{K_d}
+	\end{equation}
+$k_d^\mathrm{slow}$ and $[S]_{T_f}$ come from melting curves measured with FTIR spectroscopy that are fit to a two-state all-or-nothing dissociation model.~\citep{Marky1987CalculatingCurves} We used the FTIR melting curves previously reported in Ref.~\citep{Sanstead2016}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% Include analysis on overall hybridization rates here or at the end?
-
-\subsection{Sequence-dependent coarse-grained kinetics agree with T-jump IR data}
+\subsection{Sequence-dependent coarse-grained kinetics recapitulate T-jump IR measurements} \label{sec:time}
 
@@
 
-As an initial validation of 3SPN2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments.
-Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
+We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments.
+We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' (10-30 $\mu$s) and ``fast'' (70-100 ns) observed rates in the T-jump response (Section \ref{sec:Tjump}).
+As discussed in Section \ref{sec:TjumpAnalysis}, the fast ($k_d^\mathrm{fast}$) experimental rate is an observed rate whereas we have derived a dissociation rate constant ($k_d^\mathrm{slow}$) for the slow process.\noteb{is this stipulation clear enough?} The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response is has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds,~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} and all-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
+
+To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation.
+We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (SI Figure 4).
+First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which the central base pairs first separate to a distance of 1.3 nm.
+This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
+We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the Watson-Crick (WC) interaction.
+We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
+
+We present in Fig.~\ref{fig:relaxation-comparison} a comparison of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ estimated by computation and experiment.
+Although the 3SPN.2 model was not directly fitted against kinetic data,~\cite{Hinckley2013AnHybridization} its predictions of sequence dependent T-jump relaxation rates are, within a systematic scaling factor in time and systematic shift in temperature, in good agreement with observed experimental trends.
+It is well-known that the smoothing of the underlying free energy landscape induced by coarse-graining artificially accelerates the kinetics of coarse-grained molecular simulations and that different degrees of freedom may be accelerated by different factors.~\citep{Marrink2007TheSimulations, Fritz2011MultiscaleDynamics, Marrink2013PerspectiveModel} We find that the simulated slow responses corresponding to center-of-mass translation of the strands during dissociation of the duplex is $\sim$10$\times$ accelerated relative to experiment, whereas the fast responses corresponding to fraying of the terminal bases is $\sim$120$\times$ accelerated.
+We apply these sequence-independent scaling factors to our reported computational values.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal +4 K shift corrective calibration to our computational results.
+We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison.
+The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{Fig1.pdf}
+        \caption{Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
+Experimental results are shown in terms of the final T-jump temperature (a) The simulated and T-jump dissociation rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
+(b) The Simulated fraying rate constant and fast observed rate $k_d^\mathrm{fast}$ \noteb{again trying to make the distinction between observed rate and dissociation rate here} corresponding to terminal base-pair fraying on short time scales.
+The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
+The simulated temperature in all cases is also corrected by +4 K to correct for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
+        \label{fig:relaxation-comparison}
+	\end{center}
+\end{figure}
+
+In Fig.~\ref{fig:relaxation-comparison}a we report an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves.
+The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone.
+Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general trends exponential trend is preserved.
+This may be a result of its high propensity to fray as we will discuss further in Section \ref{sec:GC-core} \noteb{Is this sufficient discussion here since we bring it back up later?} In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end.
+This can be understood via the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences.
+Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements.
+The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.
+
+%Coarse-grained time scales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
 As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.
-Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
-To compare results with 3SPN2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
+Experimental temperature-dependent relaxation time scales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded ``fast'' and ``slow'' amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
+To compare results with 3SPN.2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
 We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
@@
 % need to discuss tmelt shift
-We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
+%We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
 We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4$^{\circ}$C between experiment and simulation fits.
-While 3SPN2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
+While 3SPN.2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
 We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
@@
 
-% GC-core fast timescales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
+% GC-core fast time scales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
 % in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor.
@@
 
-\begin{figure}  %[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Fig2.pdf}
-        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments.
-The effective simulation temperature is shifted by 4 degrees celcius to account for systematic differences in simulation and experimental melting temperature.
-Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
-        \label{fig:relaxation-comparison}
-	\end{center}
-\end{figure}
-
-% A documented short-coming for the 3SPN2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
+% A documented short-coming for the 3SPN.2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
 This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
 
-The fast response -- which Sanstead et.
+%The fast response -- which Sanstead et.
 al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN2 base, we would not expect to resolve this diverse collection of states and dynamics.
+Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with time scales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN.2 base, we would not expect to resolve this diverse collection of states and dynamics.
 Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff.
 This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
-This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events.
-Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
+This is a reasonable assumption given that the amplitude-weighted time scales should consist largely of terminal fraying events.
+Again we fit relaxation curves to the ensemble of fraying time scales in order to extract rate approximations for each sequence at a series of temperatures.
 
-For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
+%For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
 In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying.
 We observe similar trends in the experimental data, although comparisons at the highest temperatures were limited due to mixing between the fast and slow responses (Figure \ref{fig:relaxation-comparison}).
-We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN2 fraying dynamics are accelerated by about 100-200x relative to experiment.
+We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN.2 fraying dynamics are accelerated by about 100-200x relative to experiment.
 It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
@@
 Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
-and that 3SPN2 can recapitulate some sequence-dependent fraying effects.
+and that 3SPN.2 can recapitulate some sequence-dependent fraying effects.
 % need a better transition here back into SRV-MSMs?
@@
 
-% change state names and scientific notation
-% include D here? can see what it would look like...
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=160mm, 
-        scale=1]{Fig3.pdf}
-        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates.
-Shifted state abbreviations indicate the direction of the shifted overhang (5' vs.
-3') and the number of shifted base pairs relative the native state (2 vs.
-4).}
-        \label{fig:allseq_table}
-	\end{center}
-\end{figure}
+We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent long-time kinetic models.
+MSMs are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates.
+We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences.
+Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
+In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
+By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F).
+In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates.
+The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model.
+The identity and fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways.
+Immediately we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior.
+In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM.
+These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates.
+The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization.
+Since each system is simulated at its respective melting temperature it is not surprising that the leading time scale is approximately equal for all four systems.
+The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates.
+The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes.
+The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization.
+We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.
 
-% potentially get rid of State representation, ask Brennan if they are meaningful?
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=160mm, 
-        scale=1]{Fig4.pdf}
-        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate.
-Arrows indicate the probability of transitioning between states within the lag time.
-Circle areas are proportional to the log of stationary probabilities to enhance visualization of less stable states}
-
-        %Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions. }
-        \label{fig:allseq_transition}
-	\end{center}
-\end{figure}
-
-We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer.
+%We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer.
 We identified seven kinetically relevant states that were captured within the resolution of the model.
 Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities.
-By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
 Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
 Arrows between states show the probability of a transition to another state within the MSM lag time.
-We now proceed to analyze the sequence-dependent thermodynamics and kinetic features of each MSM.
 
-%Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs.
-3') and the number of shifted motifs relative the native state (1 vs.
+%Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5$^\prime$ vs.
+3$^\prime$) and the number of shifted motifs relative the native state (1 vs.
 2).
@@
 % rewrite this into one paragraph
-\subsection{Inert tails produce deviations from nearest neighbor thermodynamic models}
 
-\begin{figure}[ht!] % right place for NN table? could go to SI
+% change state names and scientific notation
+% include D here? can see what it would look like...
+
+\begin{figure}[ht!]
 	\begin{center}
-        \includegraphics[width=120mm, scale=1]{Fig5.pdf}
-        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations.
-All free energies are normalized such that that the native hybridized state is set to zero.
-Calculations included dangling end contributions but do not take inert tails effects.}
+        \includegraphics[width=170mm, scale=1]{Fig2.pdf}
+	\end{center}
+\end{figure}
+\begin{figure}[ht!]
+        \caption{ Summary of thermodynamics and kinetics quantities extracted from sequence-specific MSMs.
+(a) Seven metastable macrostates sampled by at least one of the four sequences.
+Cartoons represent 10bp sequences in the 5$^\prime$ to 3$^\prime$ direction, where red-to-blue contacts indicate WC base pairing and black indicates an unbound bases.
+Molecular rendering show 3D examples of each configuration.
+(b) Number of frames (out of $10^7$ total) clustered into each of the seven macrostates.
+Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimator and are very small compared to total populations.
+(c) Coarse-grained MSMs and molecular renderings for each sequence.
+Arrow thickness corresponds to the flux between between states (using of minimum flux cutoff of $3\times10^{-6}$), and the overlaid number represents the probability of making a transition within the 1.2 ns lag time.
+The radii of state circles correspond to the log of state populations.
+(d) Distribution of leading timescales for each sequence.
+The leading timescales for each sequence corresponds to hybridization/dehybridization process.
+AT-all and GC-end display a spectrum of converged slow modes corresponding to shifting processes.
+GC-core has one faster timescale corresponding to fraying processes.
+        \noteb{Previous (more compact) version Fig2 alternate}
+    }
+    \label{fig:allseq_table}
+\end{figure}
 
+\subsection{Comparison of MSM thermodynamic predictions with NN model} \label{sec:NN}
+
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association.
+The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
+We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences.
+The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
+Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare the free energies predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model.
+A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the Supporting Information.
+
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized (H) state by as much as 14.7 kJ/mol.
+(The free energy of the hybridized state is set to zero by construction for both the MSM and NN models.) The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
+Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the 5S2, 3S2, 5S4, and 3S4 states?
+First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
+There are approximations inherent in the 3SPN.2 model and the NN model was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions.
+Second, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.\citep{Michele2014EHybridization} Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability.
+\noteb{not sure the best way to phrase this, but they could technically still be considered inert tails}.
+In this case the NN and MSM free energy predictions are in good agreement.
+Two of our sequences, AT-all and GC-end, occupy macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}).
+For both sequences, our simulations show that 5S2 is both the most stable of shifted states and has the smallest offset (4.0 kJ/mol for AT-all and 10.0 kJ/mole for GC-end) compared to NN predictions \ref{fig:NN_table}b.
+The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol for AT-all and 14.3-14.7 kJ/mole for GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
+It is known from differential scanning calorimetry studies~\citep{Dickman2012ThermodynamicDNAs} that 3$^\prime$ inert tails are more destabilizing than 5$^\prime$ tails, with the differential behavior attributed to a combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure.
+\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization}.
+Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$).
+For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
+This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all.
+Given that terminal mismatches tend to be slightly stabilizing \citep{Santalucia2004TM}, we might expect a lower discrepancy than the  10.1-14.7 kJ/mol offset that we observe between NN and MSM results.
+This offset may arise in part from the excluded volume interactions used to account for mismatched bases in 3SPN.2, however we attribute it mainly to inert tail effects.
+The GC-end NN predictions run contrary to the expectation that the 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the MSM predictions are consistent with this trend.
+Indeed, the MSM model for GC-end does not even register the 3S4 macrostate as a stable metastable conformation for the duplex.
+
+In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails.
+In these instances the NN model overestimates macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol.
+The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.
+
+\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}
+
+\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well.
+It is important to note that we do not include dangling end contributions when calculating NN free energies.
+Although 3SPN.2 can capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the CG model was not parameterized to fully capture the enthalpic contributions of free bases interaction with terminal base pairs.
+For each macrostate, dangling ends would be expected to have stabilizing effects \noteb{Can show NN with dangling end contributions (original figure) in SI)} and therefore increase state populations compared to what we observe in simulation.}
+
+\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.
+We find that inert tail effects may be particularly impact in predicting out-of-register state stability.}
+
+\bood{For the AT-all sequence, we see close agreement between the macrostate and NN prediction for the 5S2 state.
+However, we observe deviations of 3.5 and 2.3 kJ/mol the 3S2 and 5S4 states, respectively.
+This is consistent with UV absorbance spectroscopy and differential scanning calorimetry experiments~\citep{Dickman2012ThermodynamicDNAs, Michele2014EHybridization} which show that (i) 3$^\prime$ tails are more destabilizing than 5$^\prime$ tails and (ii) longer tails are more destabilizing than shorter tails.
+These effects are compounded for the 3S4 state, where we observe an MSM free energy 5.1 kJ/mol higher than the NN model prediction.}
+
+\bood{For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
+This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all.
+The deviations between MSM and NN models follow a similar trend based on inert tail effects, but these are also consistently higher -- ranging from 3.2 to 7.8 kJ/mol.
+Indeed, the MSM model for GC-end does not even register the 3S4 macrostate as a stable metastable conformation for the duplex.
+We attribute this additional destabilization to excluded volume interactions between mismatched terminal pairs in the model, as complex and slightly stabilizing~\citep{Santalucia2004TM} terminal G:A and C:T interactions are not included in 3SPN.2.}
+
+\bood{Although the NN model tends to under-predict the MSM free energy for AT-all and GC-end, we see that the GC-core F4 state is about 2.5 kJ/mol more stable than predicted by NN.
+This can be attributed to an added probability of an ensemble of ``double-frayed'' state in which 1-2 terminal A:T base pairs are frayed and the configuration is clustered into F4.}
+
+\noteb{Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
+This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.}
+
+\begin{figure}[ht!]
+	\begin{center} 
+        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
+With the exception of the frayed state (F4), the NN model predicts lower free energies than the MSM and therefore tends to overestimate macrostate stability relative to the hybridized state.
+(a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
+(b) Free energy difference between the NN prediction and simulated value for each macrostate.}
         \label{fig:NN_table}
@@
 
-Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions.
-In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (Figure \ref{fig:NN_table}).
-For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4).
-We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone.
-For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends.
-Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.
-Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state.
-Moreover, we do not observe a stable 3S4 cluster for GC-end.
-This indicates that 5' vs.
-3' inert tail differences may out-weigh NN stacking effects alone.
-These inert tail effects show qualitative agreement with previous experimental studies, which attribute the differential behavior to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure.
-\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization}.
+\begin{figure}[ht!]
+	\begin{center} 
+        \includegraphics[width=160mm, scale=1]{Fig3_no_DE.pdf}
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
+These NN estimates do not include additional thermodynamic stabilization from dangling ends \noteb{can include this in SI?} (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
+(b) Free energy difference between the NN prediction and simulated value for each macrostate.
+\noteb{Should the y-axis be more descriptive here?}}
+        \label{fig:NN_table_no_de}
+	\end{center}
+\end{figure}
 
-%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails.
+%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of ``dangling ends'' -- unpaired bases adjacent to the paired duplex -- and ``inert tails'' -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
+Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3$^\prime$ inert tails have stronger effects than 5$^\prime$ tails.
 Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
@@
 % and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
-It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
-These effects might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure \citep{Doktycz1990ThermodynamicATGC}.
+It has been shown that 5$^\prime$ dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3$^\prime$ ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
+These effects might be attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure \citep{Doktycz1990ThermodynamicATGC}.
 
-% This leads us to believe that 5' vs.
-3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.
+% This leads us to believe that 5$^\prime$ vs.
+3$^\prime$ differences -- attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.
 
-\subsection{Out-of-register dynamics facilitate hybridization and dissociation}
-%\subsubsection{Transition probabilities between states}
+\subsection{Out-of-register states facilitate hybridization and dissociation dynamics (AT-all, GC-end)}
 
-In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states.
-For AT-all, we observe approximately equal probability of transitioning from D to any other state.
-Once a transition has been made, however, the 5' vs.
-3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate.
-Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g.
+In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms.
+We now proceed to analyze these predictions for each of the four sequences and illuminate the relationship between sequence and dynamics.
+Two of our sequences, AT-all and GC-end, support out-of-register metastable states, and we commence our analysis with the role of these shifted states.
+
+AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c).
+Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization.
+Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H.
+Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
+Once in one of the four shifted states, the 5$^\prime$ vs.
+3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate.
+Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states.
+The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D.
+Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D.
+The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state.
+A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the union of direct and indirect pathways) proceed through one or more out-of-register shifted states.
+Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time.
+Mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
+The longer D $\dashrightarrow$ H time scale is expected given the number of out-of-register states that can form and re-dissociated before reaching H.
+These values are in close agreement with the leading implied time scale shown in  Fig.~\ref{fig:allseq_table}.
+\noteb{can make note of the fact that Meta $\rightarrow$ (all other states) MFPTs align with the higher order modes as well?
+Not sure if this would require a more in depth discussion}
+
+%Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g.
 5S2 $\rightarrow$ H is more likely than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
@@
 
-For GC-end we observe no significant flux between the 5S4 state and the structurally similar 5S2 state, indicating that the former acts more as a kinetic trap than a pathway to H.
-The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
-This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed.
-This phenomenon will be explored more in depth below.
-
-% This mechanism is explored more in depth later on?
-
-%not sure to mention "inchworm" and "pseudoknot" since these were not directly observed/not in the scope of this work
-
-Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
+%Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
 A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
 The 3S4 and 5S4 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.
-While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
+
+GC-end comprises the next most complex MSM with the introduction of the terminal G:C pairs maximally preserving the repetitive tract of A:T base pairings.
+The GC-end MSM possesses all of the same macrostates in its dynamical landscape with the exception of the 3S4 state (Fig.~\ref{fig:allseq_table}c).
+As discussed in Section \ref{sec:NN}, the 3S4 state is rendered unstable within the lag time of our MSM due to the presence of the destabilizing 3$^\prime$ inert tail and only four WC base pairings compared to six in the case of AT-all.
+Analysis of the transition probabilities reveal significant differences compared to those in the AT-all kinetic network.
+Commencing from the dissociated state D, we observe a similar transition probability to the 5S4 state as for AT-all, but once in the 5S4 state there are no significant transition probabilities to any other state except back to D.
+As such, the 5S4 state acts as a kinetic trap rather than as an intermediate to hybridization.
+The D $\rightarrow$ H and H $\rightarrow$ D transition probabilities are commensurate with those for AT-all.
+However, the D $\rightarrow$ 5S2 and D $\rightarrow$ 3S2 transition probabilities are half or less of those in AT-all, and the reverse transitions are an order of magnitude larger.
+This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that contain only six WC pairs compared to eight in AT-all (Fig.~\ref{fig:NN_table}).
+The 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H transition probabilities are more than an order of magnitude larger than in AT-all, which may again be attributed to the lower thermodynamic stability of the two shifted states relative to the hybridized state H.
+Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately 6.8 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
+A transition path theory analysis of the MSM reveals that only 7\% of hybridization events D $\dashrightarrow$ H and dehybridization events H $\dashrightarrow$ D proceed through one or more out-of-register shifted states.
+The significantly reduced role for out-of-register shifted states in mediating the hybridization and dissociation pathways for GC-end relative to AT-all is consistent with the reduced thermodynamic stability of these states due to the elimination of possible out-of-register WC base pairing for the terminal G:C pairs and therefore a reduced accessibility of these states in the GC-end kinetic network.
+A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are both notably lower than AT-all values.
+This may be attributed in part to shorter meta-stable lifetimes, or to the higher temperature at which this sequences was simulated.
+Again, the average of these MFPTs align well with the leading implied time scale values in Fig.~\ref{fig:allseq_table}.
+
+%The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
+This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed.
+
+%While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
 These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior.
 Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences.
-Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands.\citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel} These processes can occur even when terminal mismatches are present in the non-native structure.
 
-Experimentally, out-of-register states are suspected to occur but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying.
-To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand.
-It is unclear if these differences were a consequence of varied simulation conditions or if GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs.
-This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.
+The out-of-register kinetic landscape that define AT-all and GC-end hybridization have been informed by numerous previous computational studies.
+Simulations have identified internal displacement mechanisms capable of correcting base pair alignment in 3SPN.2 \citep{Hinckley2014Coarse-grainedEffects} as well as in the coarse-grained oxDNA \citep{Romano2013DNADependence} and BioModi \citep{Markegard2015} models \noteb{not sure what you think of naming these specifically}.
+In all cases, these mechanisms were shown to be crucial components of the hybridization pathway for homogeneous and repetitive sequences.
+Xiao et al.\ performed an all-atom energy landscape-based analysis of 5$^\prime$-GGGGGG-3$^\prime$ and 5$^\prime$-GCGCGC-3$^\prime$ hexamers.~\citep{Xiao2019} Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
+(In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.) This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events.
+It is conceivable that the stronger hydrogen bonding in G:C WC pairs relative to A:T pairs may render out-of-register shifted states less favorable to hybridization by suppressing fluctuation-driven rearrangements,~\citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} but additional studies would be required to reconcile these observations.
 
-\subsection{Out-of-register dynamics may stretch T-jump IR responses}
+\subsection{Central GC placement induces long-lived frayed states (GC-core)} \label{sec:GC-core}
 
-\begin{figure}[ht!] % right place for NN table? could go to SI
-	\begin{center} 
-        \includegraphics[width=150mm, scale=1]{Fig6.pdf}
-        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations.
-The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions.
-Each T-jump was performed with a temperature change of 15 °C to a final temperature near the Tm of each respective sequence.
-Each time trace was fit to the sum of a stretched exponential and two exponentials (solid lines).
-The stretched exponential describes the component from 10 ns to 1 μs ($tau_{Fast}$), and the two exponentials describe the signal rise from 1 - 300 μs ($tau_{Slow}$) and thermal relaxation of the sample back to the initial temperature.
-The stretch factor (β) from each fit is listed.}
+The GC-core MSM represents a departure from the relatively rich and complex kinetic networks dominated by out-of-register shifted states to a much simpler one dominated by fraying (Fig.~\ref{fig:allseq_table}c).
+The MSM contains only three states -- hybridized H, dehybridized D, and frayed F4.
+The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs (it may be less due to rapid 1-2 terminal base pair fraying) on one side or other of the core; the other run of four A:T pairs remains free.
+(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
+Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns.
+The stability of the state is attributable to the enthalpic stabilization offered by the six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.
 
-        \label{fig:Fig6}
-	\end{center}
-\end{figure}
+Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25-times more likely than dissociation F4 $\rightarrow$ D.
+Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
+The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure how meaningful this is considering they are coming from different source states?}.
+We have noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
+Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a).
+Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s.
+The F4 state is unlikely to contribute significantly to the difference between MFPTs, rather we attribute this to a slight deviation in the total state populations -- the D equilibrium probability is about 9\% higher than H and F4 combined.
+Despite extensive sampling, these asymmetries arise from DNA's high temperature sensitivity and create the impression of a shorter H $\dashrightarrow$ D MFPT relative to D $\dashrightarrow$ H  \noteb{this should the origin of the discrepency (same for GC-mix) but I'm not sure if this is best way to explain it}.
 
-In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response.
-In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}).
-Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response.
-On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source.
-We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-ends and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
-Although reproducing these stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
+Lattice models have previously identified frayed states as putative intermediates in DNA hybridization/dehybridization.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2016} Araque et al.\ studied a 5$^\prime$-ATGCGCAT-3$^\prime$ octomer using a lattice model and identified a symmetrically A:T frayed state as a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} We previously studied the four sequences that are the subject of the present work using T-jump IR and 2D IR spectroscopy and identified GC-core as possessing the highest deviation from two-state behavior during dissociation when neglecting out-of-register contributions.\citep{Sanstead2016} This result was interpreted to arise from loss of A:T contacts and fraying around the central G:C core, and this hypothesis was supported by lattice model calculations that predicted the GC-core conformational ensemble to possess substantially more frayed configurations than the other three sequences.~\citep{Phys2019} Follow-up T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} These prior results are consistent with the present findings that expose the GC-core sequence to be the only sequence that occupies the F4 frayed state and therefore the only one possessing a metastable frayed state on time scales exceeding the 1.2 ns lag time of our MSMs.
 
-% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
-
-%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
-
-%We emphasize that this does necessarily reflect shifting relaxation timescales themselves but more likely changes in the fast responses -- e.g.
-fraying out-of-register -- within the collection of metastable states.
-
-\subsection{Centralizing GC placement induces long-lived, kinetically relevant fraying} % not sure if I should make this its own section?
-
-The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end.
-Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4).
-Previous studies suggest that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state.
-This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs.
-We observe these events to occur with equal probability on either permutable end of the helix.
-
-The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D.
-Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed.
-On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.
-We previously noted that frayed dynamics are substantially accelerated in the 3SPN2 model, which may lead to the F4 state being sampled more frequently than in experiment.
-Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in Figure \ref{fig:relaxation-comparison}.
-
-% discuss computational comparisons needs work
-Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA} Araque et al.
-defined a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
+%When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
 found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions).\citep{Sanstead2016} As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
@@
 
-%Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.
+%Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar time scale.
 
@@
 
-\begin{figure} %[ht!]
-	\begin{center}
-	\includegraphics[width=170mm, scale=1]{Fig7.pdf}
-        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.
-Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until full dissociation occurs.
-The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined.
-Fluctuation between states is common during the transition.}
-        \label{fig:GC-mix_transitions}
-	\end{center}
-\end{figure}
+\subsection{Disruption of repetitive AT tracts promotes two-state ``all-or-nothing'' kinetics (GC-mix)}
 
-\subsection{Heterogeneous base pair distribution promotes nucleation-zippering hybridization and dissociation}
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable on the 1.2 ns time scale of our MSM.
+The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
+We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
+This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the configurational entropy of the frayed two-base AT tails relative to the frayed four-base AT-tails, and the more limited disruption of the helical duplex by the shorter tails that offers less kinetic inhibition to its reformation from the transiently frayed structures.
+Simulations of GC-mix were conducted at the melting temperature of the sequence, so the simple two-state MSM should have, by construction, equal populations of, and equal transition probabilities between, the H and D states.
+As we observed for GC-core, a higher dissociated state population produces a larger effective MFPT for D $\rightarrow$ H (2.9 $\mu$s) than H $\rightarrow$ D is (2.3 $\mu$s).
+Again, the average of these time scales match the dominant leading implied time scale for this sequence (Fig.~\ref{fig:allseq_table}c).
 
-Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
+%Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
 Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix.
@@
 
-To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near MSM-identified hybridization and dissociation events.
-We present two representative events in Figure \ref{fig:GC-mix_transitions} along with all native base pair distances over time.
-During hybridization, we observed the some key base pair contacts before the full duplex formed.
-Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T.
-This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs.
-Qualitatively, we observed some short-lived states composed of 2-4 base pair contacts immediately before full dissociation occurs.
-However, in contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes.
-As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.
+Given the very simple two-state ``all-or-nothing'' behavior of GC-mix and the absence of any intermediate metastable states, we sought to interrogate our simulation trajectory data to elucidate the hybridization and dehybridization mechanisms.
+To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events.
+We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more of each in SI Figure 3.
+In hybridization, we observe a global decrease in all 10 distances as the strands approach one another, but uniformly observe the formation of some key native WC base pair contacts immediately prior to duplex formation.
+Specifically, formation of one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs \noteb{this is a more general observation which is in line with all events} This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
+Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
+These dissociation dynamics are similar to previously reported ``fraying-peeling'' mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}
 
@@
 	\begin{center}
-        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
-        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end.
-Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends.
-The lower right region of the diffusion map space shows a high concentration of "shifted loop" base pairs that may facilitate the out-of-register GC-end transition.}
-        \label{fig:GC-end_dmaps}
+	\includegraphics[width=180mm, scale=1]{Fig4.pdf}
+        \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
+Tracking of the 10 intermolecular distances between native WC base pairs over the course of an (a) hybridization event and (b) dehybridization event.
+Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
+Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C WC pair and adjacent A:T pair or central A:T pairs first form prior to rapid formation of the complete duplex.
+Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.
+Four additional hybridization events and four additional dehybridization events are presented in SI Figure 5.
+        }
+        \label{fig:GC-mix_transitions}
 	\end{center}
@@
 
-% need to gear this more towards the kinetics
-\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}
+\subsection{\label{sec:stretch_comparison}T-jump IR responses are correlated with the complexity of the dynamical landscape}
 
-Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
+We hypothesized that the complexity of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models.
+As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively.
+Although there are no distinct ``middle'' responses at the expected shifting timescales, the response profile can inform the distribution of dynamics contributing to the signal \citep{Hamm2006StretchedFolding, Johnston2006StretchedDecays}.
+In particular, the AT-all fast response exhibits more stretched exponential character ($\beta$ = 0.3) than any of the other three sequences (Figure \ref{fig:T-jump_stretching}).
+We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
+The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM.
+For the GC-end response, however, the stretching factor is not appreciably different from the other sequences and comparison of 1660 and 1600 cm$^{-1}$ wavelengths indicate that the fast response is much stronger in the G/T rather than A/T region.
+This further suggests that it corresponds to terminal fraying rather than shifting.
+Overall, there are several reasons why the T-jump response may not be directly sensitive to shifting.
+In the case of GC-end, the low population of shifted states alone may diminish contributions below a detectable limit.
+For AT-all, the particularly slow timescales associated with the most common shifting process (H $\leftrightarrow$ 5S2) can blend in with the slow dissociation response.
+In both cases, it is unclear how populations of shifted states change with temperature, making these processes difficult to identify with T-jump techniques.
+
+These comparisons also highlight the importance of considering and defining multiple types of deviations from two-state behavior.
+In this work, we define this deviation by the complexity of the MSM network and number of accessible metastable states.
+This definition suggests the following order of kinetic complexity: AT-all $>$ GC-end $>$ GC-core $>$ GC-mix.
+As we have shown, this trend is closely related to sequence repetitiveness, specifically the number of consecutive AT motifs.
+In our previous work, we did not consider out-of-register states as part of the experimental or lattice model analysis, and therefore determined our ordering (GC-core $>$ GC-mix $>$ AT-all $>$ GC-end) in terms of frayed state populations prior to duplex dissociation\citep{Sanstead2016, Sanstead2018DirectDehybridization}.
+Depending on the biological system or nanotechnology application of interest, these forms of deviation may hold varied importance \noteb{could add example here, but don't want to dwell on this too much?}.
+
+%In addition to identifying relaxation time scales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response.
+In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}).
+Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response.
+On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source.
+We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-end and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
+Although reproducing these stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
+
+% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
+
+%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
+
+%We emphasize that this does necessarily reflect shifting relaxation time scales themselves but more likely changes in the fast responses -- e.g.
+fraying out-of-register -- within the collection of metastable states.
+
+\begin{figure}[ht!] % right place for NN table? could go to SI
+	\begin{center} 
+        \includegraphics[width=175mm, scale=1]{Fig5.pdf}
+        \caption{T-Jump IR responses reflect sequence-dependent conformational heterogeneity.
+(a) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
+Normalized time traces for each sequence are shown at (b) 1600 cm$^{-1}$ and (c) 1660 cm$^{-1}$.
+The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
+Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+The stretched exponential describes the process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
+The stretch factor $\beta$ from each fit is listed.}
+
+        \label{fig:T-jump_stretching}
+	\end{center}
+\end{figure}
+
+%\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}
+
+%Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
 These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural details of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
@@
 We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
-This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands.
+This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical ``top'' and ``bottom'' of the strands.
 In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI.
@@
 
-The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair.
+%\begin{figure}[ht!]
+%	\begin{center}
+%        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
+%        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end.
+Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends.
+The lower right region of the diffusion map space shows a high concentration of ``shifted loop'' base pairs that may facilitate the out-of-register GC-end transition.}
+%        \label{fig:GC-end_dmaps}
+%	\end{center}
+%\end{figure}
+
+%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3$^\prime$ end and its shifted complementary pair.
 As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, highlights structural variances within the two sequences distributions.
+The second diffusion mode, which correlates highly with the average distance between 3$^\prime$ and 5$^\prime$ ends, highlights structural variances within the two sequences distributions.
 We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration.
 We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register.
-These "shifted-loop" bonds appear uniquely stable for GC-end conformations, and suggest a transition state through which hybridization is facilitated.
+These ``shifted-loop'' bonds appear uniquely stable for GC-end conformations, and suggest a transition state through which hybridization is facilitated.
 This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the equivalent AT-all pathways.
 
-%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3SPN2 \citep{Hinckley2013AnHybridization}.
+%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3SPN.2 \citep{Hinckley2013AnHybridization}.
 
@@
 %Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
-In the context of the 3SPN2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
+In the context of the 3SPN.2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
 The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
 
-%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3SPN2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
+%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3SPN.2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
 
-\subsection{Sequence-dependent fraying determines direct hybridization pathways}
+%\subsection{Sequence-dependent fraying determines direct hybridization pathways}
 
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
-        \caption{Sequence-dependent population distributions for trimmed hybridization events.
+%\begin{figure}[ht!]
+%	\begin{center}
+%        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
+%        \caption{Sequence-dependent population distributions for trimmed hybridization events.
 $\Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression.
@@
 Heat maps show the average intermolecular base distances in nm at increasing values of $\Psi_2$, where the main diagonal shows native base pair distances.}
-        \label{fig:Fig8}
-	\end{center}
-\end{figure}
+%        \label{fig:Fig8}
+%	\end{center}
+%\end{figure}
 
-To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
+%To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
 To achieve higher resolutions, we reran selected simulation with a 10x higher save rate and analyzed a 1000 frame region (10 ns simulation time) for each event.
-We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences.
+We built diffusion maps on a set of 48 such ``trimmed'' events evenly sampled across sequences.
 We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance.
@@
 
-Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
+%Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
 The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first Watson-Crick contacts are made.
@@
 
-\subsection{Limitations}
+%\subsection{Limitations}
 
-Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering.
-These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest.
+%Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering.
+These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other time scales of interest.
 Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process.
@@
 
-In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
+%In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
 For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
@@
 
-\section{\label{sec:conc}Conclusion}
+\section{\label{sec:conc}Conclusions}
 
-We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics.
+We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix) \noteb{again, do we want keep the 5$^\prime$-ATATATATAT-3$^\prime$ format here or not bother}.
+We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the predictions of the Markov state models.
+Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
+In particular, we found AT-all to possess the richest and most complex kinetic landscape of the four sequences that is dominated by out-of-register shifted states that participate in 33\% of hybridization and dehybridization events.
+These results are in accord with observations from HDVE T-jump IR spectroscopy that a spectrum of fast fraying responses likely originating from an ensemble of out-of-register states.
+Introduction of the G:C pairs at the end of the strand maintains a eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all.
+Similar to AT-all, \rood{XX}\% and \rood{XX}\% \noteb{not sure if we want to change the structure here given that these values appear to be the same (and are each only 7\%)} of hybridization and dehybridization events, respectively, in GC-end proceed through one or more shifted states.
+Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core.
+In this case, no metastable out-of-register shifted states are observed in our models with the hybridization and dehybridization pathways all passing through a strongly metastable frayed state in which one or other of the four-base AT-tracts is unbound resulting in two free AT-tails.
+Finally, placing the G:C bases between the center and end of the strand to maximally disrupt the repetitive AT tracts results in no metastable out-of-register or frayed states and results in simple two-state ``all-or-nothing'' hybridization/dehybridization behavior.
+Analysis of the specific pathways reveals hybridization to largely proceed by a nucleation-zippering mechanism and dehybridization to proceed by a fraying-peeling mechanism.
+The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix \noteb{not sure exactly what to say here given that Section \ref{sec:stretch_comparison} is different now}
+
+In sum, our results demonstrate the profound effect of sequence upon the kinetic landscapes, metastable states, and hybridization/dehybridization mechanisms of short DNA oligomers.
+Our analysis of this small family of sequences expose preliminary design principles for the (meta)stability of out-of-register and frayed states but we anticipate much greater richness in the landscapes will emerge with studies of longer and more diverse sequences.
+Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
+We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for applications such as DNA-PAINT,~\citep{Strauss2020UpDNA-PAINT} DNA barcoding,~\citep{Shah2019} and novel developments in dynamic DNA nanotechnology.
+
+%We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics.
 By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior.
@@
 On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-Our computational approach and results show strong comparisons to T-jump experiments at ns and us timescales.
+Our computational approach and results show strong comparisons to T-jump experiments at ns and us time scales.
 Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
-These insights should be leveraged for enhanced understanding of
-\emph{in vivo} hybridization processes, optimized sequence design in application such as DNA-PAINT and barcoding, and novel developments in the growing field of dynamic DNA nanotechnology.
-
-\newcommand{\sectionbreak}{\clearpage}
-\section{\label{sec:Results}Supplemental Information}
-
-% add secion on dmap and high temp methods?
-%\subsection{Additional methods}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_imp/allseq_event_count.png}
-        \caption{All sequences consist of 20 trajectories initialized in the hybridized states and 20 in the dissociated state.
-Here we show fraction of hybridized duplexes averaged across these sets over time.
-For all sequences these curves converge near 0.5, but stochasticity of rare-event makes a definitive $T_m$ difficult to identify.
-The number of hybridization/dissocation combined across trajectories is shown.
-Not every trajectory contained a transition event, but on average more than one full event occurred (H$\rightarrow$D$\rightarrow$H) per trajectory, providing adequate sampling of dynamics.
-AT-all undergoes substantially more transition, but many of these do not reach a native (in-register) hybridized state.
-GC-core displays the most native hybridization events, perhaps due to stronger nucleation sites at the central G:C contact points.}
-        \label{fig:allseq_event_count}
-	\end{center}
-\end{figure}
-
-% add figure for slow and fast response fitting (just GC-core)
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_fast_slow_fits.png}
-        \caption{Slow and Fast modes were determined by fitting over a distribution of 120 independent simulations.
-For each temperature in the series, a relaxation curve was fit to the distribution and the associated exponential coefficient was used to the calculate rate.
-Temperatures series between -5 - +15K relative to empirically determined sequence melting temperature were explored.
-This process was repeated for each sequence (only GC-core is shown above).}
-        \label{fig:fast_slow_fits}
-	\end{center}
-\end{figure}
-
-\subsection{VAMP-2 Scoring}
-The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.
-Hyperparameters, including feature choice, number of microstates, and number of slow modes were evaluated and selection with via the VAMP-2 score.
-In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$.
-The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
-
-\begin{align*}
- 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
- 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
- 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
-	\\
- 	VAMP-2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
- 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
-\end{align*}\label{CK1}
-
-\subsection{SRV hyperparameters}
-
-Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set.
-In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes.
-We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise.\citep{McGibbon2015VariationalKinetics} In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data.
-We used a batch size of 50000 and ran each model for a total of 20 training epochs.
-We used two hidden layers and set the size of each layer to 100.
-For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training.
-SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs.
-SRV training was implemented using Keras and Tensorflow.\citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}
-
-\subsection{SRV-MSMs provide better resolution than tICA-MSMs}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.9\textwidth]{Figs/figs_imp/AT-all_implied_timescales.png}
-        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
-        \label{fig:AT-all_dynamic}
-	\end{center}
-\end{figure}
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=80mm, 
-        scale=0.5]{Figs/figs_imp/allseq_lags.png}
-        \caption{SRV-MSM vs.
-tICA-MSM implied timescales convergence for GC-end, GC-core, and GC-mix.
-Progressively larger spectral gaps are observed between the first mode and higher order modes.
-We observe convegenve of resolvable higher order modes at a shared lag time of 1.2 ns.
-It is difficult to converge the leading mode in this regime due to the infrequency of hybridization/dissociation events.}
-        \label{fig:all_seq_lags}
-	\end{center}
-\end{figure}
-
-\subsection{SRV-MSM walkthrough and analysis}
-In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the most discernable spectral gap between modes.
-For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step.
-Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest.\citep{Phys2018MarkovValidation} We found that an SRV lag time of 1.2 ns showed good convergence across the relevant timescales for all sequences.
-Next we selected an optimal number of SRV components to include in our analysis.
-After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics.\citep{McGibbon2015VariationalKinetics} It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant.\citep{Pande2010EverythingAsk} For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest.
-As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
-
-%\subsubsection{SRV-MSM construction and optimization}
-Although SRV coordinates alone provide some information, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM.
-Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates.
-We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score.
-Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process.
-This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous tICA basis.
-Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate.
-Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures.
-Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc).\citep{Pinamonti2017PredictingModels, PinamontiTheModels} Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=120mm, 
-        scale=0.5]{Figs/figs_imp/allseq_crossval.png}
-        \caption{5-fold cross validation procedure to select number of SRV coordinates.
-We look for the converge of the VAMP-2 scores, inconsistent scores between folds, and deviation between training and test data as indicators that the model has begun fitting on statistical noise.}
-        \label{fig:allseq_srv_crossval}
-	\end{center}
-\end{figure}
-
-Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands.
-The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most relevant to each process.
-For example, the first slow mode shows a positive correlation to each distance and the strongest correlation with native base pair distances (shown along the main diagonal).
-Given these correlations and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process.
-The next four SRV components all show a relatively high correlation along offset diagonals.
-These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings.
-Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying.\citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}
-
-We found GC-end had a similar implied timescales distribution to AT-all, with a distinct spectral gap after the fourth mode.
-For AT-all and GC-end, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs.
-For GC-core we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above.
-We found that four macrostate clustering was unstable -- likely because the third mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized (H), dissociated (D), and 4 4 base pair frayed (F4) states.
-The GC-mix sequence showed a similar implied timescale distribution to GC-core, however we no longer saw a converged slow mode corresponding to multi-base fraying behavior.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated.
-These correlate closely with the first and third GC-core SRV modes.
-Although we built our SRV-MSM using these two coordinates, we again were unable to form a stable third state along the second coordinate.
-As such, we designated this transitions as effectively two-state within the resolution of our model.
-
-%For GC-core and GC-mix, we found this clustering convention to be unstable -- likely because the last converged mode was providing information about dissociation dynamics -- and performed PCCA+ with one less macrostate.
-
-To visualize macrostates, we project the data into the two leading tICA coordinates.
-Although SRVs outperform these coordinates for the purpose of MSM construction, tICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments.\citep{Sidky} Furthermore, we found that all macrostate were discernable on the first two tICA coordinates, whereas multiple SRV dimensions would be necessary to visualize states.
-After assigning these macrostates we calculated their stationary and transition probabilities averaged across 100 Gibbs sampled MSM and PCCA+ assignment.
-Means and standard deviations were calculated from this ensemble.
-
-As a preliminary check on SRV-MSM performance, we compared implied timescales with a more conventional tICA-MSM approach.
-Previous work showed that SRV-MSM implied timescales converged faster than tICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model.\citep{Sidky} Here we observed the same trend across all sequences, and we highlight these results for AT-all in figure \ref{fig:AT-all_dynamic}, emphasizing that faster convergence was observed across all leading timescales.
-Similar plots for the other sequences are included in the SI.
-We also note that the infrequency of transitions relative to the individual trajectory length leads to slower convergence of the leading mode.
-Because this mode corresponds the overall hybridization and dissociation process, we found that it displayed similar behavior across all sequences and that higher order modes were more informative for lag time selection.
-
-\subsection{SRV modes reveal hierarchical clustering structure}  % shift to include this in
-
-\begin{figure} %[ht!]
-	\begin{center}
-        \includegraphics[width=120mm]{Figs/figs_imp/GC-core_tracking_modes.png}
-        \caption{Leading GC-core SRV coordinates correlated with identifiable physical coordinates during sample hybridization (a) and dissociation (b) events.
-(c) Pearson correlations between all 100 intermolecular base pair distances and each the three leading slow modes.
-SRV2 correlations were calculated in the H and F4 states; SRV correlations were calculated in the D state.}
-        \label{fig:GC-core_tracking_modes}
-	\end{center}
-\end{figure}
-
-% not sure if this should directly follow the SRV-MSM GC-core description or directly follow the GC-core dmaps?
-% try cutting this out entirely ?
-%\subsection{SRV correlations to physical coordinates}
-
-Having constructed and interpreted SRV-MSMs, we revisited our original SRV basis to investigate how slow modes are constructed from input features.
-We found these GC-core modes to be of particular interest as they reveal the hierarchical and nonlinear nature of the dynamical encoding.
-We examined a collection of "trimmed" trajectories centered on both hybridization or dissociation events.
-For each trajectory, we compared the first three SRV coordinates with a corresponding collective variables with which they shared behavior.
-Two representative trajectories are shown in figure (\ref{fig:GC-core_tracking_modes}).
-Complementary G:C pairs are the best indicators for a hybridization/dissociation event, and we see a sharp change in the first SRV mode (SRV1) as these bonds form or break.
-The second slow mode (SRV2) is most active when G:C pairs are bound but the adjacent AT pairs are not.
-There is a small signal for fraying at the outer base pairs, but the mode overwhelmingly learns about these neighboring A:T/G:C bonds.
-Moreover, this behavior reflects movement in and out of the F4 state in the corresponding SRV-MSM discussed above.
-SRV3 is most active during dissociation, and seems to track closely with the average distance between all complementary base pairs.
-We attribute this to the SRV learning about the diffusive motions of the two body system and evaluating the likelihood of an imminent hybridization event.
-The third mode also peaks when the oligos are close together but configured in such a way that is not amenable hybridization.
-These misaligned conformations include inverse contacts where 5'/5' and 3'/3' ends meet and looped conformations where one strand is folded in on itself and preventing satisfactory WC contacts.
-
-Despite the qualitative trends we observe between physical coordinates and leading SRV modes, we found it difficult to find correlations across full trajectories.
-With the exception of SRV1, Pearson correlations between physical coordinates and SRV modes were near zero and showed erroneous trends.
-However, when we performed this analysis along smaller sections of trajectories we noticed that the sign of SRV2 and SRV3 correlations switched depending on whether the oligos were in the hybridized or dissociated state.
-This shows that these modes are inherently nonlinear and provide support on top of the first mode -- which serves as an indicator function for hybridized vs.
-dissociated state.
-With respect to our SRV-MSM macroscates, we found that SRV2 was "turned on" in the H and F4 states -- corresponding to the intact helix and frayed state, respectively -- and SRV3 was turned on in the dissociated state.
-Accordingly, we calculated Pearson correlations between each SRV mode and all distances in states where the modes are active.
-Figure \label{fig:GC-core_tracking_modes} shows the highest correlation between SRV2 and inner A:T pairs, weak correlation with outer A:T pairs, and an inverse correlation with 5'/5' and 3'/3' pairs which tend to approach each when the duplex is in the F4 state.
-We also observed a highly symmetric correlation between SRV3 and central base pairs distances, which indicative of overall diffusive behavior.
-Taken together, these analyses reveal how the SRV learns and represents the dynamical space in an hierarchical manner, providing enhanced resolution to a linear method like tICA and producing the MSM results we observe above.
-
-\subsection{Structural diversity in the F4 state informs fraying dynamics}
-
-\begin{comment}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=450, scale=1]{Figs/figs_imp/GC-core_tracking_modes.png}
-        \caption{Diffusion map representation configurations sampled from the F4 state.
-The x and y axes represent the the first two non-trivial diffusion map eigenfunctions.
-Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
-        \label{fig:GC-core_dmaps}
-	\end{center}
-\end{figure}
-
-\end{comment}
-
-% shift to include this in the overall diffusion map section
-%\subsubsection{Diffusion maps show ensemble of frayed states}
-
-We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
-We build diffusion maps using 10000 frames sampled from the macrostate (Figure \ref{fig:GC-core_dmaps}).
-Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
-We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate.
-This distance also correlated closely with the second diffusion map mode.
-Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core.
-We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
-At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place.
-Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation.
-Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
-
-%\subsection{Explicit model comparisons}
-
-%We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
-In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3SPN2 explicit implementation \citep{Hinckley2015}.
-These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities.
-Comparisons were limited by the higher computational demand required to simulate 5x the beads in the box, a shorter integration timestep, and reliance on Ewald summation to account for electrostatics.
-We caution over-interpretation of these results as the explicit model has not been as widely adopted or thoroughly validated compared to the implicit model.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.PNG}
-        \caption{Full 5' diffusion maps including degenerate third diffusion map mode.}
-        \label{fig:GC-end_dmaps_full}
-	\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_3prime.PNG}
-        \caption{Full 3' diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct "looping" region for intact GC bonds}
-        \label{fig:GC-end_dmaps_3prime}
-	\end{center}
-\end{figure}
-
-%add anymore figures for the time analysis: As shown bleow:
-\begin{comment}
-
-\subsubsection{\label{sec:Results}Code to provide on GitHub} 
-\begin{enumerate}
-    \item Cross val data
-    \item lammps run script and each sequence input files
-    \item Featurization with mdtraj (reindexing and sparse saving)
-    \item Dmaps for GC-core and GC-end
-    \item timescales comparison
-    \item SRV cross valss
-    \item MSM construction
-
-\end{enumerate}
-\end{comment}
-
-\begin{comment}
-
-\section{\label{sec:Discussion}Discussion}
-
-\citep{Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}
-Long simulations shows that terminal fraying does not converge on the picosecond timescale
-
-\citep{Nonin1995TerminalFraying}
-Most simulations suppress fraying by placing GC pairs at termini.
-
-\citep{Andreatta2006UltrafastHelix}
-- Base opening process is slower than 40 ns but faster than 1ms
-- Open termini state shows new fraying mode of 5ps (I think shows free base motions)
-\end{comment}
-
-\begin{comment}
-
-%%%%%% Other MSM and timescales approaches  %%%%%%%%%%%
-
-McGibbon, R.
-T., & Pande, V.
-S.
-(2015).
-Variational cross-validation of slow dynamical modes in molecular kinetics : Importance of not generating too many microstates and employting cross-validation to avoid artificially high scores or artificially slow dynamics.
-Both of these lead to models fitting statistical noise which is the root of boosted scores.
-
-Pinamonti, G., Paul, F., Rodriguez, A., & Bussi, G.
-(n.d.).
-analyzed with core-set Markov state models, 43: Good ideas on core-based clustering.
-Assigning microstates based on physical characteristics is interesting but didn't seem to work for me in practice (could try doing this based on sim.log data with number of base pairs bound or some energy metric).
-Lot's of notes on this paper regarding comparisons to timescales.
-Good notes on higher resolution fraying interactions that we likely can't resolve.
-
-Sua, E., Adelman, J. L., & Zuckerman, D. M. (2016). Accurate Estimation of Protein Folding and Unfolding Times : Beyond Markov State Models, 1. https://doi.org/10.1021/acs.jctc.6b00339:
-More accurate ways to adapt MFPT into a rate
-
-Refs for DNA fraying:
-\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro).
-Use TPS to focus on kinetic pathway by which singl base pair bind/unbind.
-Focus on end base of 3-bp oligomer in explicit solvent.
-Finds order parameters tracking base pair binding as well as intra-strand stacking.
-(energy and distance for each).
-
-\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,
-
-6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
-37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
-38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding
-
-\end{comment}
+These insights should be leveraged for enhanced understanding of \emph{in vivo} hybridization processes, optimized sequence design in application such as DNA-PAINT and barcoding, and novel developments in the growing field of dynamic DNA nanotechnology.
 
@@
 
-\bibliography{references}
+%\bibliographystyle{unsrt}
+\bibliography{references_210505}
+
 %\bibliography{refs_mendeley2}
@@
 \clearpage
+\newpage
+
+\begin{figure}[ht!]
+	\begin{center} 
+        \includegraphics[width=150mm, scale=1]{TOC.pdf}
+        \caption*{TOC image \noteb{could consider adding an experimental aspect to this?}}
+        \label{fig:TOC}
+	\end{center}
+\end{figure}
 

### 2021-05-10 manuscript.tex
update fig3 resolution

@@
 \textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM.
-The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2.
+The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2.
 We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.

### 2021-05-12 SI.tex
Andy edits

@@
 	\begin{center}
-        \includegraphics[width=0.9\textwidth, 
-        scale=0.5]{SIFig1.pdf}
+        \includegraphics[width=1.0\textwidth]{FigS1.pdf}
         \caption{Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
-A knee in the VAMP-2 plot was identified at approximately the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively.
+A knee in the VAMP-2 plot was identified at the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively.
 An embedding of corresponding dimensionality was then used to cluster frames into discrete states.
-The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
+The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.
+\noter{Please re-order panels into standard order with panel labels as per the caption.}}
         \label{fig:SIFig1}
@@
 	\centering
-    \includegraphics[width=0.7\textwidth, scale=0.5]{SIFig2.pdf}
+    \includegraphics[width=0.8\textwidth]{FigS2.pdf}
     \caption{Convergence of the MSM implied time scales $t_i$ as a function of lag time $\tau$.
@@
 	\begin{center}
-        \includegraphics[width=1.1\textwidth]{SIFig3.pdf}
+        \includegraphics[width=1.0\textwidth]{FigS3.pdf}
         \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by an MSM constructed at the $\tau$ = 1.2 ns lag time (dashed blue line) versus those computed from an MSM constructed at the particular lag time (solid black line).
@@
 	\centering
-    \includegraphics[width=.8\textwidth]{SIFig4.pdf}
+    \includegraphics[width=.75\textwidth]{FigS4.pdf}
     \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
-From 120 independent 1 $\mu$s simulations, we compiled the distribution of times at which terminal A:T and central base pairs first separate to a distance of of 1.3 nm.
+From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with \rood{both OR either} \noter{Pick the right word.} central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with \rood{both OR either} \noter{Pick the right word.} terminal Watson-Crick base pairs intact as a function of time.
+\noter{Do I have this correct?} We define a Watson-Crick base pair to be intact if the two comlementary bases lie within a linear distance of 1.3 nm.
 We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
@@
 	\centering
-    \includegraphics[width=0.8\textwidth]{SIFig5a.pdf}
+    \includegraphics[width=0.8\textwidth]{FigS5a.pdf}
 \end{figure}
@@
 	\centering
-    \includegraphics[width=0.8\textwidth]{SIFig5b.pdf}
+    \includegraphics[width=0.8\textwidth]{FigS5b.pdf}
 \end{figure}
@@
 
-The NN model from SantaLucia assumes hybridization proceeds in a two-state all-or-nothing manner.
-The Gibbs free energy difference between the hybridized and dissociated states can be determined from calculation of NN enthalpy and entropy contributions.
+The NN model assumes hybridization proceeds in a two-state all-or-nothing manner.
+The Gibbs free energy difference between the hybridized and dissociated states can be determined from calculation of NN enthalpy and entropy contributions,
 \begin{equation}\label{SIe1}
-\Delta G_{NN}^{\circ}(T) = \Delta H_{NN}^{\circ} - T\Delta S_{NN}^{\circ}
+\Delta G_{NN}^{\circ}(T) = \Delta H_{NN}^{\circ} - T\Delta S_{NN}^{\circ}.
 \end{equation}
-Here,  and  are assumed to be independent of temperature.   and   are determined as the sum of all NN terms in a given configuration. 
+$\Delta H_{NN}^{\circ}$ and $\Delta S_{NN}^{\circ}$ are assumed to be independent of temperature $T$, and are computed from the sum over all NN terms in a given configuration,
 \begin{equation}\label{SIe2}
-\Delta H_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta H_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta H_{DE, j}^{\circ} + \Delta H_{init}^{\circ}
+\Delta H_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta H_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta H_{DE, j}^{\circ} + \Delta H_{init}^{\circ},
 \end{equation}
+\begin{equation}\label{SIe3}
+\Delta S_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta S_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta S_{DE, j}^{\circ} + \Delta S_{init}^{\circ} + \Delta S_{sym}^{\circ}.
+\end{equation}                          
+$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TM}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
+A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig4}.
+$\Delta S_{NN}^{\circ}$ is constructed in an analogous manner from NN, dangling end, and initiation terms, plus a symmetry term, $\Delta S_{sym}^{\circ}$, that applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.
 
-\begin{equation}\label{SIe3}
-\Delta S_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta S_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta S_{DE, j}^{\circ} + \Delta S_{init}^{\circ} + \Delta S_{sym}^{\circ}
-\end{equation}
-
-$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA.
-$\Delta H_{DE, j}^{\circ}$  corresponds to the enthalpic contribution from a given dangling end  base next to the duplex\citep{Santalucia2004TM}.
-$\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
-$\Delta S_{NN}^{\circ}$  is constructed from analogous NN, dangling end, and initiation terms in addition to a symmetry term, $\Delta S_{sym}^{\circ}$, which applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.
 The SantaLucia NN parameters are defined for a solution condition at neutral pH with a sodium concentration of 1M.
 We use the empirical salt correction developed by Owczarzy\citep{Owczarzy2008PredictingCations} to determine the NN parameters at the sodium concentration used for simulations in this work.
-The Owczarzy model assumes the affect of counterions on DNA hybridization thermodynamics is purely entropic.
+The Owczarzy model assumes the effect of counter ions on DNA hybridization thermodynamics to be purely entropic,
+\begin{equation}\label{SIe3}
+\Delta S_{NN}^{\circ}(c_{Na^+}) = S_{NN}^{\circ}(1M) + \Delta H_{NN}^{\circ}[(4.29f_{GC}-3.95)\times 10^{-5} \ln{(c_{Na^+})}+ 9.4\times 10^{-6} (\ln{(c_{Na^+})})^2],
+\end{equation} 
+where $f_{GC}$ is the fractional G:C content of the duplex and $c_{Na^+}$ is the concentration of sodium counter ions measured in M.
 
-\begin{equation}\label{SIe3}
-\Delta S_{NN}^{\circ}(c_{Na^+}) = S_{NN}^{\circ}(1M) + \Delta H_{NN}^{\circ}[(4.29f_{GC}-3.95)\times 10^{-5} \ln{(c_{Na^+})}+ 9.4\times 10^{-6} (\ln{(c_{Na^+})})^2]
-\end{equation}
-
-$f_{GC}$ is the fractional G:C content of the duplex and $c_{Na^+}$ is the concentration of sodium counterions in M.
+\vspace{+0.5in}
 
@@
 	\centering
-    \includegraphics[width=.85\textwidth]{SIFig6.pdf}
-    \caption{Schematic of nearest neighbor contributions for an intact hybridized and out-of-register duplex with dangling ends (DE) stabilization.}
+    \includegraphics[width=.85\textwidth]{FigS6.pdf}
+    \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
     \label{fig:SIFig4}

### 2021-05-12 cover_letter/letterhead.tex
Andy edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 1.0 (28/12/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
% with extensive modifications by Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

%\documentclass[11pt,a4paper,draft]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\usepackage{letterbib}
\usepackage{graphicx} % Required for including pictures
\usepackage{microtype} % Improves typography
\usepackage{fancyhdr}
\usepackage[T1]{fontenc} % Required for accented characters
\usepackage[sort&compress,numbers,super]{natbib}
\usepackage{caption}
\usepackage{newfloat}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{soul}
\usepackage[svgnames]{xcolor}
\usepackage{framed}
\usepackage{calligra}
\usepackage{fp}
\FPset\marginWidth{0.7}	% margin width (all sides) in inches
\usepackage[margin=\marginWidth in]{geometry}
\usepackage{marginfix}
\usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter

\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%more new commands and definitions
\newcommand*{\rood}[1]{\textcolor{red}{#1}}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}
\newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
\newcommand{\scripty}[1]{\ensuremath{\mathcalligra{#1}}}

\newfloat{figure}{h}{lot}
\floatname{figure}{Figure}
\newfloat{sfigure}{h}{lot}
\floatname{sfigure}{Figure}
\newfloat{lfigure}{h}{lot}
\floatname{lfigure}{Figure}

\definecolor{shadecolor}{named}{LightGray}

\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\DeclareFontShape{T1}{calligra}{m}{n}{<->s*[2.2]callig15}{}

\renewcommand{\thesfigure}{S\arabic{sfigure}}
\renewcommand{\thelfigure}{\Alph{lfigure}}

%-----------------------------------------------------------------------------
% HEADER/FOOTER INFO
%-----------------------------------------------------------------------------
%\pagestyle{fancy}
\fancyhf{}
%\cfoot{\textit{telephone} 217-333-1441 $\bullet$ \textit{fax} 217-333-2736 $\bullet$ \textit{email} matse@illinois.edu}
\cfoot{\thepage}

%----------------------------------------------------------------------------------------
%	SENDER INFORMATION
%----------------------------------------------------------------------------------------

\def\Who{Andrew L. Ferguson} % Your name
\def\What{, PhD} % Your title
\def\Where{Pritzker School of Molecular Engineering\\University of Chicago} % Your department/institution
\def\department{Pritzker School of Molecular Engineering}
\def\institution{University of Chicago}
\def\Address{5640 South Ellis Avenue} % Your address
\def\CityZip{Chicago, IL 60637} % Your city, zip code, country, etc
\def\Email{e:~\url{andrewferguson@uchicago.edu}} % Your email address
\def\TEL{p:~(773)~702-5950} % Your phone number
\def\URL{w:~\url{http://andrewferguson.uchicago.edu}} % Your URL
\def\titleone{Associate Professor}

%----------------------------------------------------------------------------------------
%	HEADER AND FROM ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing institution logo indent based on margin width
\FPsub\logoIndent{7.55}{\tmp}

\address{
\includegraphics[width=3.5in]{UChicago_PME_logo.png} % Include the logo of your institution
%\hspace{\logoIndent in} % Position of the institution logo, increase to move left, decrease to move right
\hspace{2.8in} % Position of the institution logo, increase to move left, decrease to move right
\vskip -0.82in~\\ % Position of the text in relation to the institution logo, increase to move down, decrease to move up
\Large\hspace{0.75in} \hfill ~\\[0.05in] % First line of institution name, adjust hspace if your logo is wide
\hspace{0.75in} \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
\makebox[0ex][r]{\bf \Who \What }\hspace{0.65in} % Print your name and title with a little whitespace to the right
~\\[-0.11in] % Reduce the whitespace above the horizontal rule
\hspace{0in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
\hspace{\fill}\parbox[t]{4in}{ % Create a box for your details underneath the horizontal rule on the right
\footnotesize % Use a smaller font size for the details
%\Who \\  % Your name, all text after this will be italicized
%\\
\titleone
%\titletwo\\
%\titlethree\\
%\titlefour\\
\vskip 0.05in
\department\\ % Your department
\institution\\ % Your institution
\Address\\ % Your address
\CityZip
\vskip 0.05in
\TEL\\ % Your phone number
\Email\\ % Your email address
\URL\\ % Your URL
}
\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
\vspace{-1in} % Move the letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	TO ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing date indent based on margin width
\FPsub\dateIndent{4.15}{\tmp}

\def\opening#1{\thispagestyle{empty}
{\centering\fromaddress \vspace{1.1in} \\ % Print the header and from address here, add whitespace to move date down
%
\hspace*{4.15 in}\today\hspace*{\fill}\par
} % Print today's date, remove \today to not display it
{\raggedright \toname \\ \toaddress \par} % Print the to name and address
\vspace{0.05in} % White space after the to address
\noindent #1 % Print the opening line
% Uncomment the 4 lines below to print a footnote with custom text
%\def\thefootnote{}
%\def\footnoterule{\hrule}
%\footnotetext{\hspace*{0.18\textwidth}{Telephone 217-333-1441 $\bullet$ Fax 217-333-2736 $\bullet$ email matse@illinois.edu\footnotesize \hspace{3cm} \thepage}}
%\def\thefootnote{\arabic{footnote}}
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who \What} % The signature is a combination of your name and title

\long\def\closing#1{
\vspace{0.1in} % Some whitespace after the letter content and before the signature
\noindent % Stop paragraph indentation
%\hspace*{\longindentation} % Move the signature right
%\parbox{\indentedwidth}{\raggedright
#1 % Print the signature text
\vskip 0.0in
\includegraphics[width=0.25\textwidth]{signature_720}
%\vskip 0.65in % Whitespace between the signature text and your name
\\
\noindent
\fromsig
\vskip -0.05in
\titleone\\
\department\\
\institution\\
\vskip -0.1in
\noindent cc: M.~Zhao, K.~Lachowski, S. Alamdari, J.~Sampath, P.~Mu, L.D.~Pozzo, C.-L.~Chen, C.J.~Mundy, J.~Pfaendtner 
}

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}
{
Prof.\ Erick Carreira \\
Editor-in-Chief, \textit{The Journal of the American Chemical Society}
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Prof.\ Carreira}

We are pleased to submit a manuscript for consideration for publication in the \textit{The Journal of the American Chemical Society} an original research \textbf{Article} entitled \textbf{``Hierarchical Self-Assembly Pathways of Polypeptoid Helices and Sheets''} by Mingfei Zhao (U.~Chicago), Kacper Lachowski(U.W.~Seattle), Sarah Alamdari (U.W.~Seattle), Janani Sampath (PNNL),  Peng Mu (PNNL), Chun-Long Chen (PNNL, U.W.~Seattle), Lilo D.\ Pozzo (U.W.~Seattle), Christopher J.\ Mundy (PNNL, U.W.~Seattle), Jim Pfaendtner (U.W.~Seattle, PNNL), and Andrew L.\ Ferguson (U.~Chicago). This article has not been published previously and is not under consideration in any other journal. All authors have seen and approved the submission of the manuscript.

% background and why it is important to do this.
\textbf{Background.} Polypeptoids (poly-N-substituted glycines) are a class of highly tailorable synthetic peptidomic polymers with applications as drugs, antimicrobials, and catalysts that can be engineered to assemble into nanoaggregates including spheres, helices, tubes, and sheets. Amphiphilic diblock polypeptoids have been engineered to assemble high-aspect ratio 2D crystalline lattices with applications in catalysis, molecular separations, and photovoltaics. Assembly is induced by dissolving the peptoids in an organic solvent/water mixture and evaporating the organic phase to promote hydrophobic association, but the mechanistic pathways and thermodynamic/kinetic driving forces mediating assembly remain uncharacterized.

% what we did in this paper
\textbf{Key Results.}  In this work, we integrate all-atom molecular dynamics (MD) simulation, small angle x-ray scattering (SAXS), circular dichroism (CD), atomic force microscopy (AFM), and x-ray diffraction (XRD) to present an integrated computational/experimental study of the self-assembly of Nbrpe6Nc6 -- an amphiphilic diblock polypeptoid comprising an \ce{NH2} capped block of six hydrophobic N-((4-bromophenyl)ethyl)glycine residues conjugated to a hydrophilic 6-aminohexanal (\ce{NH3(CH2)5CO}) tail -- in organic solvent/water mixtures. Our MD simulations predict a hierarchical assembly pathway by which monomers first assemble into disordered aggregates that self-order into 1D chiral helical rods that subsequently assemble 2D achiral crystalline nanosheets. Experimentally, we observe stable, free-floating 1D rods in mixed solvent using SAXS and CD, and 2D crystalline sheets in pure water using XRD and AFM. MD predictions of the relative thermodynamic stability of disordered aggregates, helical rods, and sheets as a function of the organic solvent concentration support a thermodynamically-controlled assembly process.

% impact and importance of this paper
\textbf{Significance.} Our work integrates simulation and experiment to establish a previously unknown hierarchical polypeptoid assembly pathway wherein 0D monomers first assemble into disordered aggregates that ripen into 1D helical rods that subsequently form 2D crystalline nanosheets. This new understanding of the hierarchical assembly mechanisms and stable intermediaries presents new principles and guidance for the rational design of peptoid-based nanomaterials with potential biochemical, biomedical, and bioengineering applications including as drugs, antimicrobials, catalysts, and molecular sieves.

% fit for the journal
\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of polypeptoid self-assembly. This new understanding establishes new mechanistic understanding and a thermodynamic basis for existing experimental evaporation-induced assembly protocols and identifies previously unknown solvent conditions for the stable fabrication of 1D helical rods. These principles offer new strategies for the supramolecular assembly of diverse peptoid sequences into 1D and 2D supramolecular aggregates.

% reviewers
\textbf{Reviewers.} We propose the following scientists as well qualified to review this work.

\textbf{\textit{Simulation.}}

Stephen Whitelam \hfill Molecular Foundry, Lawrence Berkeley National Lab \\
(510) 495-2769 \hfill \url{swhitelam@lbl.gov}

$\circ$ \textit{Expert in biomolecular simulation and molecular modeling of peptoid assembly.}

Monica Olvera de la Cruz \hfill Materials Science and Engineering and Chemistry, Northwestern University \\
(847) 491-7801 \hfill \url{m-olvera@northwestern.edu}

$\circ$ \textit{Expert in molecular modeling and simulation of peptides, polymers, and materials assembly.}

Arthi Jayaraman \hfill Chemical and Biomolecular Engineering, University of Delaware \\
(302) 831-8682 \hfill \url{arthij@udel.edu}

$\circ$ \textit{Expert in theory and simulation of biomolecular assembly and molecular materials design.}

\textbf{\textit{Experiment.}}

Darrin Pochan \hfill Materials Science and Engineering, University of Delaware \\
(302) 831-3569 \hfill \url{pochan@udel.edu}

$\circ$ \textit{Expert in biomolecular and polymeric assembly and characterization.}

Rachel Segalman \hfill Chemical Engineering, UC Santa Barbara \\
(805) 893-3709 \hfill \url{segalman@ucsb.edu}

$\circ$ \textit{Expert in biological and polymeric materials assembly.}

Ron Zuckermann \hfill The Molecular Foundry, Lawrence Berkeley National Laboratory \\
(510) 387-9364 \hfill \url{rnzuckermann@lbl.gov}

$\circ$ \textit{Pioneer of peptoid science. Expert in peptoid synthesis and characterization.}

We look forward to hearing from you.

\closing{Yours Sincerely}

\end{letter}

%----------------------------------------------------------------------------------------

%\clearpage
%\newpage

%\begin{shaded}
%\textbf{Response to Reviewer \# 1}
%\end{shaded}

%\bibliographystyle{achemso}
%\bibliography{bibliography.bib}

\end{document}

### 2021-05-12 manuscript.tex
Andy edits

@@
 In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
-For a repetitive ATATATATAT sequence \noteb{not sure if we want to use 5$^\prime$-ATATATATAT-3$^\prime$ notation here as well?}, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations.
+For a repetitive ATATATATAT sequence, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations.
 Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways.
@@
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} \noteb{these references correspond to CG enhanced sampling studies, I can refer to enhanced sampling on AA here instead (\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations}) but I think the CG studies should be mentioned since not much has been done for AA hybridization} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
-These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacements \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami structure \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
 
@@
 \citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
-Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers.} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s.
+Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s.
 \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius.
@@
 We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state.
-The initial hybridized state was set, according to 3SPN.2 documentation, to the coordinates of Arnott et al.\ ~\citep{Arnott1976CRCBiology} The dissociated state was generated by shifting one oligomer from the hybridized state by 1 nm in the x, y, and z directions.
-The magnitude of this shift did not affect the results so long as all native WC bonds were sufficiently separated.
-\noteb{No energy minimization (no 3spn2 precedent on this).
-The lammps default emin techniques may not work for this type of CG forcefield.
-Given the sampling it seems that  1 $\mu$s of equilibration should be sufficient?} Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the system temperature.
-\addref{Maxwell-Boltzmann -- \noteb{not sure what to use here, usually not cited?}} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
+The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions.
+The magnitude of this displacement did not affect the results so long as all native WC bonds were completely broken.
+Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest.
+Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
 The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.
@@
 \textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances.
-We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the $(10 \times 10)$ 100-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair.
-This featurization naturally accounts for intermolecular configurations via the pairwise distances between bases on different strands.
-\noteb{previous text had suggested we used the 190-dim features, but we actually used $100-45 = 55$-dim features since the intramolecular features contributed no additional kinetic variance (I've updated accordingly)} One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances.
+We adopt intermolecular pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the 10$\times$10=100-element vector of pairwise distances between all bases within the complementary pair of 10-base strands.
+One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances.
 Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2.
-\citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands.
+\citep{Sengupta2019AutomatedSelf-assembly} Finally, we took the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands.
 VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization.
-\citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top 5 modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.
+\citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top five modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.
+Somewhat surprisingly, we observed that augmenting our set of intermolecular distances with intramolecular distances led to no improvement of the VAMP-2 score, indicating that the kinetically-relevant conformational state of the two strands are adequately represented via the intermolecular distances and leading us to employ only intermolecular distances within our featurization.
 
@@
 \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs.
-We guarded against overfitting using 5-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
-We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA GK210GL GPU card.
-A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively \noteb{not sure the best way to explain this succinctly}.
-(Figure 1 in the Supporting Information).
-This motivated the construction of a n-dimensional embedding corresponding to n leading SRV modes retained for each sequence.
+We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
+We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
+A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 3D embeddings, respectively.
 
 \textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering.
-The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.
+The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.
 
 \textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM.
-The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2.
-We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
-At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively.
-The identification of $(i-1)$ implied time scales \noteb{(may be 1-2 additional converging, but these correspond to diffusive processes that are irrelevant or unhelpful in determining the hybridization mechanisms.
-Still, this might come up when looking at the the SI Fig 2} implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates.
+The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
+We observe rapid convergence of the implied time scales $t_i$ with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
+For this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively.
+The identification of $(i-1)$ implied time scales implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates.
 We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates.
@@
 We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}.
-This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$.
-We present in SI Figure 3 the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four kinetic models.
+This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$ should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$.
+We present in \blauw{Fig.~S3} the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models perform very well in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four models.
 
@@
 Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump.
-T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs.
+T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between (-10) and 10 fs.
 t-HDVE spectra were acquired with a parallel pulse polarization scheme.
@@
 The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket.
-The T-jump temperature change ($\Delta$T) was set to $\sim$15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam.
+The T-jump temperature change ($\Delta$T) was set to 15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam.
 The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.
@@
 
-To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates were taken as the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response ( $k_d^\mathrm{fast}$) is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant ($k_d^\mathrm{slow}$) was extracted from the observed rate of the slow response ($\lambda^\mathrm{slow}$) using a two-state model for self-complementary oligomers.~\citep{Bernasconi2012RelaxationKinetics}
+To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
 	\begin{equation}\label{e1}
-	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a
+	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow},
 	\end{equation}
-$[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a$ is the association rate constant.
-In practice, Equation~\ref{e1} is recast in terms of the dissociation equilibrium constant ($K_d$) to solve for $k_d^\mathrm{slow}$ and $k_a$.
+where $[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a^\mathrm{slow}$ is the association rate constant.
+\noter{I renamed $k_a$ as $k_a^\mathrm{slow}$ for consistency.
+Is this OK?} In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d = k_d^\mathrm{slow} / k_a^\mathrm{slow}$,
 	\begin{equation}\label{e2a}
-	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d}{K_d+4[S]_{T_f}}
+	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d}{K_d+4[S]_{T_f}}.
 	\end{equation}
-	\begin{equation}\label{e2b}
-	k_a = \frac{k_d}{K_d}
-	\end{equation}
-$k_d^\mathrm{slow}$ and $[S]_{T_f}$ come from melting curves measured with FTIR spectroscopy that are fit to a two-state all-or-nothing dissociation model.~\citep{Marky1987CalculatingCurves} We used the FTIR melting curves previously reported in Ref.~\citep{Sanstead2016}
+We use this expression to estimate $k_d^\mathrm{slow}$, where we estimate $K_d$ by \rood{...}, $\lambda^\mathrm{slow}$ by \rood{...}, and $[S]_{T_f}$ by \rood{...} \noter{What was written seems incomplete -- it seems like we need to say how we estimate (i) $\lambda^\mathrm{slow}$, (ii) $K_d$, and (iii) $[S]_{T_f}$, not $k_d^\mathrm{slow}$ and $[S]_{T_f}$?
+Am I understanding Eqn.~2 correctly?} \rood{$k_d^\mathrm{slow}$ and $[S]_{T_f}$ come from melting curves measured with FTIR spectroscopy that are fit to a two-state all-or-nothing dissociation model.~\citep{Marky1987CalculatingCurves} We used the FTIR melting curves previously reported in Ref.~\citep{Sanstead2016}}
 
@@
 We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments.
-We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' (10-30 $\mu$s) and ``fast'' (70-100 ns) observed rates in the T-jump response (Section \ref{sec:Tjump}).
-As discussed in Section \ref{sec:TjumpAnalysis}, the fast ($k_d^\mathrm{fast}$) experimental rate is an observed rate whereas we have derived a dissociation rate constant ($k_d^\mathrm{slow}$) for the slow process.\noteb{is this stipulation clear enough?} The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response is has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds,~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} and all-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
+We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
+The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response is has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
 
 To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation.
-We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (SI Figure 4).
-First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which the central base pairs first separate to a distance of 1.3 nm.
+We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
+First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which \rood{both OR either} \noter{Pick the right word.} of the central base pairs first separate to a distance of 1.3 nm.
 This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
@@
 We apply these sequence-independent scaling factors to our reported computational values.
-Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal +4 K shift corrective calibration to our computational results.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal (+4) K corrective calibration to our computational results.
 We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison.
@@
         \caption{Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
-Experimental results are shown in terms of the final T-jump temperature (a) The simulated and T-jump dissociation rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-(b) The Simulated fraying rate constant and fast observed rate $k_d^\mathrm{fast}$ \noteb{again trying to make the distinction between observed rate and dissociation rate here} corresponding to terminal base-pair fraying on short time scales.
+Results are reported in terms of the final T-jump temperature.
+(a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
+(b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
 The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
-The simulated temperature in all cases is also corrected by +4 K to correct for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
+The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
         \label{fig:relaxation-comparison}
@@
 Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general trends exponential trend is preserved.
-This may be a result of its high propensity to fray as we will discuss further in Section \ref{sec:GC-core} \noteb{Is this sufficient discussion here since we bring it back up later?} In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end.
+This may be a result of its high propensity to fray (cf.\ Section \ref{sec:GC-core}).
+In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end.
 This can be understood via the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences.
@@
 
-We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent long-time kinetic models.
+We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models.
 MSMs are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates.
@@
 The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model.
-The identity and fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways.
+The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways.
 Immediately we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior.
@@
 The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization.
-Since each system is simulated at its respective melting temperature it is not surprising that the leading time scale is approximately equal for all four systems.
-The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates.
+Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization.
+The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways.
 The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes.
@@
 We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.
+
+\noter{In the caption to Figure 2: (i) Change panel labels to lower case.
+(ii) Improve resolution of images.
+(iii) Label the MSMs in panel c as you did for the bar charts in panel b; AT-all, GC-end, etc.
+(iv) You nicely label panel a ``Macrostates''.
+Can you also label the other panels following the labels I have suggested in the caption?}
 
@@
 	\begin{center}
-        \includegraphics[width=170mm, scale=1]{Fig2.pdf}
+        \includegraphics[width=1.0\textwidth]{Fig2.pdf}
 	\end{center}
@@
 \begin{figure}[ht!]
-        \caption{ Summary of thermodynamics and kinetics quantities extracted from sequence-specific MSMs.
-(a) Seven metastable macrostates sampled by at least one of the four sequences.
-Cartoons represent 10bp sequences in the 5$^\prime$ to 3$^\prime$ direction, where red-to-blue contacts indicate WC base pairing and black indicates an unbound bases.
-Molecular rendering show 3D examples of each configuration.
-(b) Number of frames (out of $10^7$ total) clustered into each of the seven macrostates.
-Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimator and are very small compared to total populations.
-(c) Coarse-grained MSMs and molecular renderings for each sequence.
-Arrow thickness corresponds to the flux between between states (using of minimum flux cutoff of $3\times10^{-6}$), and the overlaid number represents the probability of making a transition within the 1.2 ns lag time.
-The radii of state circles correspond to the log of state populations.
-(d) Distribution of leading timescales for each sequence.
-The leading timescales for each sequence corresponds to hybridization/dehybridization process.
-AT-all and GC-end display a spectrum of converged slow modes corresponding to shifting processes.
-GC-core has one faster timescale corresponding to fraying processes.
-        \noteb{Previous (more compact) version Fig2 alternate}
+        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mid.
+(a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D).
+The line drawings represent the 10-base self-complementary sequences, where red-to-blue contacts indicates (possible) WC base pairing and black indicates an unbound bases.
+Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate.
+(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the \rood{1 ms} \noter{Right?} simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
+Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
+(c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
+The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a.
+The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b.
+Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate.
+The grey arrows between macrostates indicate the presence of a kinetic flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux.
+(To avoid congesting the diagram, arrows are not reported for fluxes lower than $3\times10^{-6}$ \noter{Units?}.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM.
+Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions.
+(d) \textbf{Kinetic time scales.} Distribution of MSM implied time scales for each sequence.
+The leading implied time scale corresponds to the characteristic time scale for hybridization/dehybridization and is approximately equal for all systems since simulations were conducted at the same concentration and at the respective melting temperatures.
+The higher order implied time scales correspond to a spectrum of kinetic relaxations between the constituent macrostates in the MSM corresponding to shifted and/or frayed states.
     }
@@
 Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare the free energies predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model.
-A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the Supporting Information.
+A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the \blauw{Supporting Information}.
 
@@
 There are approximations inherent in the 3SPN.2 model and the NN model was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions.
-Second, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.\citep{Michele2014EHybridization} Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability.
+Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model.
+\noter{Agree with this sentence?} Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.\citep{Michele2014EHybridization}
+
+%Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability.
 \noteb{not sure the best way to phrase this, but they could technically still be considered inert tails}.
 In this case the NN and MSM free energy predictions are in good agreement.
-Two of our sequences, AT-all and GC-end, occupy macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}).
-For both sequences, our simulations show that 5S2 is both the most stable of shifted states and has the smallest offset (4.0 kJ/mol for AT-all and 10.0 kJ/mole for GC-end) compared to NN predictions \ref{fig:NN_table}b.
-The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol for AT-all and 14.3-14.7 kJ/mole for GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
+
+We can further explore the role of inert tails upon macrostate stability by analyzing the AT-all and GC-end sequences that occupy out-of-register shifted macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}).
+For both sequences, our simulations show that 5S2 is both the most stable of shifted states and has the smallest discrepancy (4.0 kJ/mol AT-all, 10.0 kJ/mol GC-end) compared to NN predictions (Fig.~\ref{fig:NN_table}b).
+The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol AT-all, 14.3-14.7 kJ/mol GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
 It is known from differential scanning calorimetry studies~\citep{Dickman2012ThermodynamicDNAs} that 3$^\prime$ inert tails are more destabilizing than 5$^\prime$ tails, with the differential behavior attributed to a combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure.
@@
 Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$).
-For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
-This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all.
-Given that terminal mismatches tend to be slightly stabilizing \citep{Santalucia2004TM}, we might expect a lower discrepancy than the  10.1-14.7 kJ/mol offset that we observe between NN and MSM results.
+For GC-end, the MSM and NN models both predict the four shifted states to be less stable than the corresponding states in AT-all.
+This is in line with expectations since the terminal G:C pairs in GC-end decrease by two the number of available WC pairings in out-of-register shifted states compared to AT-all.
+The GC-end NN predictions run contrary to the expectation that the 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the MSM predictions are consistent with this trend.
+Indeed, the MSM model for GC-end does not identify the 3S4 macrostate as a stable metastable conformation for the duplex.
+
+%Given that terminal mismatches tend to be slightly stabilizing \citep{Santalucia2004TM}, we might expect a lower discrepancy than the 10.1-14.7 kJ/mol offset that we observe between NN and MSM results.
 This offset may arise in part from the excluded volume interactions used to account for mismatched bases in 3SPN.2, however we attribute it mainly to inert tail effects.
-The GC-end NN predictions run contrary to the expectation that the 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the MSM predictions are consistent with this trend.
-Indeed, the MSM model for GC-end does not even register the 3S4 macrostate as a stable metastable conformation for the duplex.
 
@@
 
-\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}
+%\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}
 
-\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well.
+%\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well.
 It is important to note that we do not include dangling end contributions when calculating NN free energies.
@@
 
-\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.
+%\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.
 We find that inert tail effects may be particularly impact in predicting out-of-register state stability.}
 
-\bood{For the AT-all sequence, we see close agreement between the macrostate and NN prediction for the 5S2 state.
+%\bood{For the AT-all sequence, we see close agreement between the macrostate and NN prediction for the 5S2 state.
 However, we observe deviations of 3.5 and 2.3 kJ/mol the 3S2 and 5S4 states, respectively.
@@
 
-\bood{For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
+%\bood{For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
 This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all.
@@
 
-\bood{Although the NN model tends to under-predict the MSM free energy for AT-all and GC-end, we see that the GC-core F4 state is about 2.5 kJ/mol more stable than predicted by NN.
+%\bood{Although the NN model tends to under-predict the MSM free energy for AT-all and GC-end, we see that the GC-core F4 state is about 2.5 kJ/mol more stable than predicted by NN.
 This can be attributed to an added probability of an ensemble of ``double-frayed'' state in which 1-2 terminal A:T base pairs are frayed and the configuration is clustered into F4.}
 
-\noteb{Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
+%\noteb{Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
 This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.}
@@
         \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
-With the exception of the frayed state (F4), the NN model predicts lower free energies than the MSM and therefore tends to overestimate macrostate stability relative to the hybridized state.
-(a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
-(b) Free energy difference between the NN prediction and simulated value for each macrostate.}
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero.
+(b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model.
+With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.}
         \label{fig:NN_table}
@@
 
-\begin{figure}[ht!]
-	\begin{center} 
-        \includegraphics[width=160mm, scale=1]{Fig3_no_DE.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
+%\begin{figure}[ht!]
+%	\begin{center} 
+%        \includegraphics[width=160mm, scale=1]{Fig3_no_DE.pdf}
+%        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
 These NN estimates do not include additional thermodynamic stabilization from dangling ends \noteb{can include this in SI?} (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
@@
 \noteb{Should the y-axis be more descriptive here?}}
-        \label{fig:NN_table_no_de}
-	\end{center}
-\end{figure}
+%        \label{fig:NN_table_no_de}
+%	\end{center}
+%\end{figure}
 
@@
 The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state.
-A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the union of direct and indirect pathways) proceed through one or more out-of-register shifted states.
+A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states.
 Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time.
-Mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
-The longer D $\dashrightarrow$ H time scale is expected given the number of out-of-register states that can form and re-dissociated before reaching H.
-These values are in close agreement with the leading implied time scale shown in  Fig.~\ref{fig:allseq_table}.
-\noteb{can make note of the fact that Meta $\rightarrow$ (all other states) MFPTs align with the higher order modes as well?
-Not sure if this would require a more in depth discussion}
+A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
+These values are in close agreement with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d and, as expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, approximately equal to one another.
+
+%\noteb{can make note of the fact that Meta $\rightarrow$ (all other states) MFPTs align with the higher order modes as well? Not sure if this would require a more in depth discussion}
 
@@
 However, the D $\rightarrow$ 5S2 and D $\rightarrow$ 3S2 transition probabilities are half or less of those in AT-all, and the reverse transitions are an order of magnitude larger.
-This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that contain only six WC pairs compared to eight in AT-all (Fig.~\ref{fig:NN_table}).
+This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that comprise only six WC pairs compared to eight in AT-all (Fig.~\ref{fig:NN_table}).
 The 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H transition probabilities are more than an order of magnitude larger than in AT-all, which may again be attributed to the lower thermodynamic stability of the two shifted states relative to the hybridized state H.
-Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately 6.8 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
+Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately seven times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
 A transition path theory analysis of the MSM reveals that only 7\% of hybridization events D $\dashrightarrow$ H and dehybridization events H $\dashrightarrow$ D proceed through one or more out-of-register shifted states.
 The significantly reduced role for out-of-register shifted states in mediating the hybridization and dissociation pathways for GC-end relative to AT-all is consistent with the reduced thermodynamic stability of these states due to the elimination of possible out-of-register WC base pairing for the terminal G:C pairs and therefore a reduced accessibility of these states in the GC-end kinetic network.
-A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are both notably lower than AT-all values.
-This may be attributed in part to shorter meta-stable lifetimes, or to the higher temperature at which this sequences was simulated.
-Again, the average of these MFPTs align well with the leading implied time scale values in Fig.~\ref{fig:allseq_table}.
+We compute a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are again approximately equal and in good agreement with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
 
@@
 
-The out-of-register kinetic landscape that define AT-all and GC-end hybridization have been informed by numerous previous computational studies.
-Simulations have identified internal displacement mechanisms capable of correcting base pair alignment in 3SPN.2 \citep{Hinckley2014Coarse-grainedEffects} as well as in the coarse-grained oxDNA \citep{Romano2013DNADependence} and BioModi \citep{Markegard2015} models \noteb{not sure what you think of naming these specifically}.
+The out-of-register kinetic landscape that define AT-all and GC-end hybridization have been explored by a number of previous computational studies.
+Simulations have identified internal displacement mechanisms capable of correcting base pair alignment in 3SPN.2 \citep{Hinckley2014Coarse-grainedEffects} as well as in the coarse-grained oxDNA \citep{Romano2013DNADependence} and BioModi \citep{Markegard2015} models.
 In all cases, these mechanisms were shown to be crucial components of the hybridization pathway for homogeneous and repetitive sequences.
@@
 The MSM contains only three states -- hybridized H, dehybridized D, and frayed F4.
-The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs (it may be less due to rapid 1-2 terminal base pair fraying) on one side or other of the core; the other run of four A:T pairs remains free.
+The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free.
 (As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
 Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns.
-The stability of the state is attributable to the enthalpic stabilization offered by the six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.
+The stability of the state is attributable to the enthalpic stabilization offered by the up to six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.
 
@@
 Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
-The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure how meaningful this is considering they are coming from different source states?}.
-We have noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
+The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability.
+We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
 Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a).
 Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s.
-The F4 state is unlikely to contribute significantly to the difference between MFPTs, rather we attribute this to a slight deviation in the total state populations -- the D equilibrium probability is about 9\% higher than H and F4 combined.
-Despite extensive sampling, these asymmetries arise from DNA's high temperature sensitivity and create the impression of a shorter H $\dashrightarrow$ D MFPT relative to D $\dashrightarrow$ H  \noteb{this should the origin of the discrepency (same for GC-mix) but I'm not sure if this is best way to explain it}.
+Again these values are approximately equal and commensurate with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
 
-Lattice models have previously identified frayed states as putative intermediates in DNA hybridization/dehybridization.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2016} Araque et al.\ studied a 5$^\prime$-ATGCGCAT-3$^\prime$ octomer using a lattice model and identified a symmetrically A:T frayed state as a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} We previously studied the four sequences that are the subject of the present work using T-jump IR and 2D IR spectroscopy and identified GC-core as possessing the highest deviation from two-state behavior during dissociation when neglecting out-of-register contributions.\citep{Sanstead2016} This result was interpreted to arise from loss of A:T contacts and fraying around the central G:C core, and this hypothesis was supported by lattice model calculations that predicted the GC-core conformational ensemble to possess substantially more frayed configurations than the other three sequences.~\citep{Phys2019} Follow-up T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} These prior results are consistent with the present findings that expose the GC-core sequence to be the only sequence that occupies the F4 frayed state and therefore the only one possessing a metastable frayed state on time scales exceeding the 1.2 ns lag time of our MSMs.
+Lattice models have previously identified frayed states as putative intermediates in DNA hybridization/dehybridization.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2016} Araque et al.\ studied a 5$^\prime$-ATGCGCAT-3$^\prime$ octomer using a lattice model and identified a symmetrically A:T frayed state as a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} We previously studied the four sequences that are the subject of the present work using T-jump IR and 2D IR spectroscopy and identified GC-core as possessing the highest deviation from two-state behavior during dissociation when neglecting out-of-register contributions.\citep{Sanstead2016} This result was interpreted to arise from loss of A:T contacts and fraying around the central G:C core, and this hypothesis was supported by lattice model calculations that predicted the GC-core conformational ensemble to possess substantially more frayed configurations than the other three sequences.~\citep{Phys2019} Follow-up T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} These prior results are consistent with the present findings that expose the GC-core sequence to be the only sequence that occupies the F4 frayed state and therefore the only one possessing a metastable frayed state on time scales exceeding the $\tau$=1.2 ns lag time of our MSMs.
 
@@
 
-The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable on the 1.2 ns time scale of our MSM.
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
 The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
 We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
-This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the configurational entropy of the frayed two-base AT tails relative to the frayed four-base AT-tails, and the more limited disruption of the helical duplex by the shorter tails that offers less kinetic inhibition to its reformation from the transiently frayed structures.
-Simulations of GC-mix were conducted at the melting temperature of the sequence, so the simple two-state MSM should have, by construction, equal populations of, and equal transition probabilities between, the H and D states.
-As we observed for GC-core, a higher dissociated state population produces a larger effective MFPT for D $\rightarrow$ H (2.9 $\mu$s) than H $\rightarrow$ D is (2.3 $\mu$s).
-Again, the average of these time scales match the dominant leading implied time scale for this sequence (Fig.~\ref{fig:allseq_table}c).
+This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the frayed two-base AT tails relative to the frayed four-base AT-tails.
+Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s.
+Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
 
@@
 To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events.
-We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more of each in SI Figure 3.
-In hybridization, we observe a global decrease in all 10 distances as the strands approach one another, but uniformly observe the formation of some key native WC base pair contacts immediately prior to duplex formation.
-Specifically, formation of one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs \noteb{this is a more general observation which is in line with all events} This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
+We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more of each in \blauw{Fig.~S5}.
+During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another, but uniformly observe the formation of some key native WC base pair contacts immediately prior to duplex formation.
+Specifically, formation of one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
+This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
 Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
-These dissociation dynamics are similar to previously reported ``fraying-peeling'' mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}
+These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}
 
@@
 Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
-Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C WC pair and adjacent A:T pair or central A:T pairs first form prior to rapid formation of the complete duplex.
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the complete duplex.
 Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.
-Four additional hybridization events and four additional dehybridization events are presented in SI Figure 5.
+Four additional hybridization events and four additional dehybridization events are presented in \blauw{Fig.~S5}.
         }
@@
 
-\subsection{\label{sec:stretch_comparison}T-jump IR responses are correlated with the complexity of the dynamical landscape}
+\subsection{\label{sec:stretch_comparison}Long-lived metastable states predicted by the MSM are resolved by T-jump IR}
 
-We hypothesized that the complexity of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models.
+\noter{Rough cut largely for a suggested structure -- please clean up and rework as you see fit based on your discussions with Brennan.
+I probably also screwed up some of the experimental jargon.} We hypothesized that we should be able to validate particular kinetic predictions of our sequence-dependent MSMs using T-jump IR measurements.
+T-jump IR measurements commencing from low temperatures at which the hybridized state dominates the equilibrium ensemble apply a rapid step jump in temperature and probe the relaxation of the system to the dehybridized state.
+We hypothesized that sequences possessing long-lived intermediate states between the hybridized H and dehybridized D states may be detectable in a stretching of the time resolved T-jump IR t-HDVE signal.
+We report in Fig.~\ref{fig:T-jump_stretching}a the predicted life times of each macrostate for each of the four DNA sequences at their respective melting temperatures.
+Only the AT-all sequence contains long-lived partially hybridized states with life times in excess of \rood{XX} $\mu$s and commensurate with those of the hybridized H and dehybridized D states.
+The longest lived of these is the \rood{XX} state with a life time of \rood{XX} $\mu$s.
+GC-core also contains a number of out-of-register shifted states, but their life times are predicted to be \rood{XX}-fold shorter than that of the corresponding states in AT-all.
+This is likely attributable to their reduced stability due to the two fewer available WC pairings for any given out-of-register GC-end state relative to AT-all.
+The longest lived GC-end intermediate state is only \rood{XX} $\mu$s.
+The sole intermediate F4 frayed state occupied by GC-core has a life time of only \rood{XX} ns, and the GC-mid sequence possesses no metastable states at all.
+Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence possessing long-lived intermediate states with life times of up to \rood{XX} $\mu$s should be distinguishable from the GC-end, GC-core, and GC-mid sequences predicted to have no metastable states or only very short-lived states with life times less than \rood{XX} ns.
+We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mid due to the very short-lived nature of these intermediate states.
+
+We present in Fig.~\ref{fig:T-jump_stretching}b-d \noter{Can we use some other color than yellow for AT-all in the time traces -- it's quite hard to read.} our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$.
+The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
+Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to \noter{What physical process.} A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mid.
+This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = \rood{XX} compared to $\beta_\mathrm{GC-end}$ = \rood{XX}, $\beta_\mathrm{GC-core}$ = \rood{XX}, and $\beta_\mathrm{GC-mid}$ = \rood{XX}, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = \rood{XX} compared to $\beta_\mathrm{GC-end}$ = \rood{XX}, $\beta_\mathrm{GC-core}$ = \rood{XX}, and $\beta_\mathrm{GC-mid}$ = \rood{XX}.
+This result provides an experimental validation of the kinetic predictictions of the MSM models.
+
+%observe signatures  of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models.
 As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively.
@@
 
-These comparisons also highlight the importance of considering and defining multiple types of deviations from two-state behavior.
+%These comparisons also highlight the importance of considering and defining multiple types of deviations from two-state behavior.
 In this work, we define this deviation by the complexity of the MSM network and number of accessible metastable states.
@@
 
-\begin{figure}[ht!] % right place for NN table? could go to SI
+\begin{figure}[ht!]
 	\begin{center} 
-        \includegraphics[width=175mm, scale=1]{Fig5.pdf}
+        \includegraphics[width=1.0\textwidth]{Fig5.pdf}
         \caption{T-Jump IR responses reflect sequence-dependent conformational heterogeneity.
-(a) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
-Normalized time traces for each sequence are shown at (b) 1600 cm$^{-1}$ and (c) 1660 cm$^{-1}$.
+(a) MSM predictions of equilibrium macrostate life times at the sequence melting temperature.
+(b) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
+Normalized time traces for each sequence are shown at (c) 1600 cm$^{-1}$ and (d) 1660 cm$^{-1}$.
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
@@
 The stretched exponential describes the process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The stretch factor $\beta$ from each fit is listed.}
-
+The stretch factor $\beta$ for the fits at 1600 cm$^{-1}$ and 1660 cm$^{-1}$ are reported directly on the plots in panels c and d.
+        }
         \label{fig:T-jump_stretching}
@@
 
-We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix) \noteb{again, do we want keep the 5$^\prime$-ATATATATAT-3$^\prime$ format here or not bother}.
+We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix).
 We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
-T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the predictions of the Markov state models.
+T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the kinetic prediction of the Markov state models that the AT-all sequence should possess long-lived out-of-register shifted states that are detectable within T-jump IR t-HDVE time traces.
 Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
 In particular, we found AT-all to possess the richest and most complex kinetic landscape of the four sequences that is dominated by out-of-register shifted states that participate in 33\% of hybridization and dehybridization events.
-These results are in accord with observations from HDVE T-jump IR spectroscopy that a spectrum of fast fraying responses likely originating from an ensemble of out-of-register states.
-Introduction of the G:C pairs at the end of the strand maintains a eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all.
-Similar to AT-all, \rood{XX}\% and \rood{XX}\% \noteb{not sure if we want to change the structure here given that these values appear to be the same (and are each only 7\%)} of hybridization and dehybridization events, respectively, in GC-end proceed through one or more shifted states.
+Introduction of the G:C pairs at the end of the strand maintains an eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all.
+Destabilization of the GC-end shifted states relative to AT-all, however, results in a far more limited participation of these states with only 7\% of GC-end hybridization and dehybridization events passing through one or more shifted states.
 Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core.
@@
 Analysis of the specific pathways reveals hybridization to largely proceed by a nucleation-zippering mechanism and dehybridization to proceed by a fraying-peeling mechanism.
-The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix \noteb{not sure exactly what to say here given that Section \ref{sec:stretch_comparison} is different now}
+The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix -- with this trend largely dictates by sequence repetitiveness, specifically the number of consecutive AT motifs.
+We note that this ordering differs from our previously reported ordering in terms of deviation from two-state behavior of GC-core $>$ GC-mix $>$ AT-all $>$ GC-end.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} We can understand these two apparently discrepant orderings by understanding that the latter was deduced based on experimental analyses and lattice models that did not account for out-of-register states and focussed largely on fraying behaviors.
+Indeed, under the assumption that fraying is the dominant kinetic process relative to out-of-register shifting, we  can harmonize the predictions of the present work with our prior work by eliminating all out-of-register shifted states in our fitted MSMs (Fig.~\ref{fig:allseq_table}c), in which case we find GC-core to contain an F4 frayed intermediate and the remaining sequences to all have simple two-state dynamics such that the predicted ordering is GC-core $>$ GC-mix $\approx$ AT-all $\approx$ GC-end.
 
@@
 Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for applications such as DNA-PAINT,~\citep{Strauss2020UpDNA-PAINT} DNA barcoding,~\citep{Shah2019} and novel developments in dynamic DNA nanotechnology.
+We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
 
@@
 
+\section*{TOC Image}
+
 \begin{figure}[ht!]
@@
         \includegraphics[width=150mm, scale=1]{TOC.pdf}
-        \caption*{TOC image \noteb{could consider adding an experimental aspect to this?}}
+        %\caption*{TOC image}
         \label{fig:TOC}

### 2021-05-13 SI.tex
Update figures 2 and 5 and section 3.7

@@
 An embedding of corresponding dimensionality was then used to cluster frames into discrete states.
-The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.
-\noter{Please re-order panels into standard order with panel labels as per the caption.}}
+The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
         \label{fig:SIFig1}
@@
     \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
-From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with \rood{both OR either} \noter{Pick the right word.} central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with \rood{both OR either} \noter{Pick the right word.} terminal Watson-Crick base pairs intact as a function of time.
-\noter{Do I have this correct?} We define a Watson-Crick base pair to be intact if the two comlementary bases lie within a linear distance of 1.3 nm.
+From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time.
+We define a Watson-Crick base pair to be intact if the two comlementary bases lie within a linear distance of 1.3 nm.
 We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
@@
 The Owczarzy model assumes the effect of counter ions on DNA hybridization thermodynamics to be purely entropic,
-\begin{equation}\label{SIe3}
+\begin{equation}\label{SIe4}
 \Delta S_{NN}^{\circ}(c_{Na^+}) = S_{NN}^{\circ}(1M) + \Delta H_{NN}^{\circ}[(4.29f_{GC}-3.95)\times 10^{-5} \ln{(c_{Na^+})}+ 9.4\times 10^{-6} (\ln{(c_{Na^+})})^2],
@@
     \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
-    \label{fig:SIFig4}
+    \label{fig:SIFig6}
 \end{figure}

### 2021-05-13 manuscript.tex
Update figures 2 and 5 and section 3.7

@@
 
-We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC3-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix).
-We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
+We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix).
+We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) model that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
 \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation.
@@
 SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000.
-We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics.
+We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics.
 \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
@@
 We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
-A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 3D embeddings, respectively.
+A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
 
@@
 We observe rapid convergence of the implied time scales $t_i$ with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
-For this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively.
+For this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mix systems, respectively.
 The identification of $(i-1)$ implied time scales implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates.
-We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates.
+We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mix systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates.
 These macrostate MSMs constitute our terminal kinetic models.
@@
 
-To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
+To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
 	\begin{equation}\label{e1}
-	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow},
+	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow}
 	\end{equation}
 where $[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a^\mathrm{slow}$ is the association rate constant.
-\noter{I renamed $k_a$ as $k_a^\mathrm{slow}$ for consistency.
-Is this OK?} In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d = k_d^\mathrm{slow} / k_a^\mathrm{slow}$,
+In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d$ to solve for $k_d$ and $k_a$.
 	\begin{equation}\label{e2a}
-	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d}{K_d+4[S]_{T_f}}.
+	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d(T_f)}{K_d(T_f)+4[S]_{T_f}}
 	\end{equation}
-We use this expression to estimate $k_d^\mathrm{slow}$, where we estimate $K_d$ by \rood{...}, $\lambda^\mathrm{slow}$ by \rood{...}, and $[S]_{T_f}$ by \rood{...} \noter{What was written seems incomplete -- it seems like we need to say how we estimate (i) $\lambda^\mathrm{slow}$, (ii) $K_d$, and (iii) $[S]_{T_f}$, not $k_d^\mathrm{slow}$ and $[S]_{T_f}$?
-Am I understanding Eqn.~2 correctly?} \rood{$k_d^\mathrm{slow}$ and $[S]_{T_f}$ come from melting curves measured with FTIR spectroscopy that are fit to a two-state all-or-nothing dissociation model.~\citep{Marky1987CalculatingCurves} We used the FTIR melting curves previously reported in Ref.~\citep{Sanstead2016}}
+	\begin{equation}\label{e2b}
+	k_a^\mathrm{slow} = \frac{k_d^\mathrm{slow}}{K_d(T_f)}
+	\end{equation}
+
+FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement.
+	\begin{equation}\label{e3a}
+	[S]_{T_f} = [1 - \theta(T_f)]c_{tot}
+	\end{equation}
+	\begin{equation}\label{e3b}
+	K_d(T_f) = \frac{2c_{tot}[1 - \theta(T_f)]^2}{\theta(T_f)}
+	\end{equation}
+$c_{tot}$ is the total oligonucleotide concentration, which was 2 mM for all measurements.
 
@@
 We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
-The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response is has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
+The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
 
@@
 We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
-First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which \rood{both OR either} \noter{Pick the right word.} of the central base pairs first separate to a distance of 1.3 nm.
+First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm.
 This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
@@
 
-\noter{In the caption to Figure 2: (i) Change panel labels to lower case.
-(ii) Improve resolution of images.
-(iii) Label the MSMs in panel c as you did for the bar charts in panel b; AT-all, GC-end, etc.
-(iv) You nicely label panel a ``Macrostates''.
-Can you also label the other panels following the labels I have suggested in the caption?}
-
 %We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer.
@@
 \begin{figure}[ht!]
-        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mid.
+        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix.
 (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D).
@@
 Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate.
-(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the \rood{1 ms} \noter{Right?} simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
+(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
 Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
@@
 The grey arrows between macrostates indicate the presence of a kinetic flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux.
-(To avoid congesting the diagram, arrows are not reported for fluxes lower than $3\times10^{-6}$ \noter{Units?}.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM.
+(To avoid congesting the diagram, arrows are not reported for fluxes lower than $3\times10^{-6}$ \noteb{flux is in terms of $p_i*p_{ij}$ so I think it should still be unitless?}.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM.
 Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions.
@@
 Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model.
-\noter{Agree with this sentence?} Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.\citep{Michele2014EHybridization}
+Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{Michele2014EHybridization}
 
 %Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability.
-\noteb{not sure the best way to phrase this, but they could technically still be considered inert tails}.
 In this case the NN and MSM free energy predictions are in good agreement.
@@
 
-%\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}
-
 %\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well.
@@
 Although 3SPN.2 can capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the CG model was not parameterized to fully capture the enthalpic contributions of free bases interaction with terminal base pairs.
-For each macrostate, dangling ends would be expected to have stabilizing effects \noteb{Can show NN with dangling end contributions (original figure) in SI)} and therefore increase state populations compared to what we observe in simulation.}
+For each macrostate, dangling ends would be expected to have stabilizing effects (Can show NN with dangling end contributions in SI and therefore increase state populations compared to what we observe in simulation.)
 
@@
 
-%\noteb{Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
-This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.}
+%Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
+This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.
 
@@
 %        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
-These NN estimates do not include additional thermodynamic stabilization from dangling ends \noteb{can include this in SI?} (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
+These NN estimates do not include additional thermodynamic stabilization from dangling ends  (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
 (b) Free energy difference between the NN prediction and simulated value for each macrostate.
-\noteb{Should the y-axis be more descriptive here?}}
 %        \label{fig:NN_table_no_de}
@@
 A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
-These values are in close agreement with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d and, as expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, approximately equal to one another.
-
-%\noteb{can make note of the fact that Meta $\rightarrow$ (all other states) MFPTs align with the higher order modes as well? Not sure if this would require a more in depth discussion}
+\noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A.
+These MFPTs reflect the frequency of partial hybridization events and are in close agreement with the leading implied time scale of $t_1$ = 1.5 $\mu$s reported in Fig.~\ref{fig:allseq_table}d" This works for AT-all, but for all sequences the leading timescale tends to be lower by up to 2x than H $\dashrightarrow$ D MFTPs.
+I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)} As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.
 
@@
 The significantly reduced role for out-of-register shifted states in mediating the hybridization and dissociation pathways for GC-end relative to AT-all is consistent with the reduced thermodynamic stability of these states due to the elimination of possible out-of-register WC base pairing for the terminal G:C pairs and therefore a reduced accessibility of these states in the GC-end kinetic network.
-We compute a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are again approximately equal and in good agreement with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
+We compute a MFPT for D $\dashrightarrow$ H of 1.6 $\mu$s and for H $\dashrightarrow$ D of 2.1 $\mu$s, which are again approximately equal.
+
+\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}
 
@@
 Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a).
-Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s.
-Again these values are approximately equal and commensurate with the leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
+Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s, which are again approximately equal.
+
+\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}
 
@@
 This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the frayed two-base AT tails relative to the frayed four-base AT-tails.
-Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s.
-Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = \rood{XX} $\mu$s reported in Fig.~\ref{fig:allseq_table}d.
+Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s, which are again nearly equal.
+
+\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }
 
@@
 
-\noter{Rough cut largely for a suggested structure -- please clean up and rework as you see fit based on your discussions with Brennan.
-I probably also screwed up some of the experimental jargon.} We hypothesized that we should be able to validate particular kinetic predictions of our sequence-dependent MSMs using T-jump IR measurements.
-T-jump IR measurements commencing from low temperatures at which the hybridized state dominates the equilibrium ensemble apply a rapid step jump in temperature and probe the relaxation of the system to the dehybridized state.
-We hypothesized that sequences possessing long-lived intermediate states between the hybridized H and dehybridized D states may be detectable in a stretching of the time resolved T-jump IR t-HDVE signal.
-We report in Fig.~\ref{fig:T-jump_stretching}a the predicted life times of each macrostate for each of the four DNA sequences at their respective melting temperatures.
-Only the AT-all sequence contains long-lived partially hybridized states with life times in excess of \rood{XX} $\mu$s and commensurate with those of the hybridized H and dehybridized D states.
-The longest lived of these is the \rood{XX} state with a life time of \rood{XX} $\mu$s.
-GC-core also contains a number of out-of-register shifted states, but their life times are predicted to be \rood{XX}-fold shorter than that of the corresponding states in AT-all.
-This is likely attributable to their reduced stability due to the two fewer available WC pairings for any given out-of-register GC-end state relative to AT-all.
-The longest lived GC-end intermediate state is only \rood{XX} $\mu$s.
-The sole intermediate F4 frayed state occupied by GC-core has a life time of only \rood{XX} ns, and the GC-mid sequence possesses no metastable states at all.
-Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence possessing long-lived intermediate states with life times of up to \rood{XX} $\mu$s should be distinguishable from the GC-end, GC-core, and GC-mid sequences predicted to have no metastable states or only very short-lived states with life times less than \rood{XX} ns.
-We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mid due to the very short-lived nature of these intermediate states.
-
-We present in Fig.~\ref{fig:T-jump_stretching}b-d \noter{Can we use some other color than yellow for AT-all in the time traces -- it's quite hard to read.} our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$.
+\noteb{After running more low-temp simulations and looking at NN predictions for shifted states at low temps, it seems like the shifted equilibrium does not change too much (a relatively higher loss for GC-end than AT-all).
+I think the state lifetime argument becomes confusing since we (i) we are presenting simulation and experimental time scales on the same figure (ii) do not have an acceleration factor to readily convert these timescales (iii) do not have any evidence of mid-range relaxation that you might expect given this timescale spectrum.
+No need to mention any of this in the paper, but I think there is really a catch-22 with trying to detect these states with T-jump.
+The only process you can really resolve is going from shifted --> D since H--> shifted will be as slow as H-->D.
+If a state is unstable this process should happen at an intermediate timescale, but it contain such a small (less than 1\%) of the total population that it won't be picked up.
+If the metastable state is more stable then the population will be larger, but the process will occur near the same timescale as full dissociation and look the exact same.
+So all of that is to say that a stretched exponential can very reasonably be attributed to shifted-fraying response, but it becomes more of a thermodynamic argument than kinetic.}
+
+We hypothesized that we should be able to validate particular kinetic predictions of our sequence-dependent MSMs using T-jump IR spectroscopy.
+T-jump IR measurements commencing from low temperatures (at which the hybridized state dominates the equilibrium ensemble) apply a rapid step jump in temperature and probe the relaxation of the system to the dehybridized state.
+As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively.
+Shifting dynamics are more difficult to distinguish given that shifting events (e.g.
+H $\rightarrow$ 5S2) are shown by our MSMs to occur on similar time scales as H $\rightarrow$ D and would therefore be blended into the slow response.
+However, we do expect some portion of low temperature duplexes to initialize in shifted states and alter fast fraying dynamics.
+Furthermore, we expect this effect to be proportional to the population of shifted states relative to intact hybridized states.
+At our MSM equilibrium distribution, we find that AT-all shifted states make up 24\% of the hybridized ensemble (all states excluding D) compared to only 0.3\% for GC-end shifted states.
+GC-core and GC-mix would not be expected to have any shifted states in the starting ensemble.
+Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence, possessing significantly more heterogeneous state distribution after the T-jump, should be distinguishable from the GC-end, GC-core, and GC-mix sequences.
+We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mix due to the similarity of the starting ensemble.
+
+We present in Fig.~\ref{fig:T-jump_stretching} our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$.
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
@@
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to \noter{What physical process.} A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mid.
-This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = \rood{XX} compared to $\beta_\mathrm{GC-end}$ = \rood{XX}, $\beta_\mathrm{GC-core}$ = \rood{XX}, and $\beta_\mathrm{GC-mid}$ = \rood{XX}, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = \rood{XX} compared to $\beta_\mathrm{GC-end}$ = \rood{XX}, $\beta_\mathrm{GC-core}$ = \rood{XX}, and $\beta_\mathrm{GC-mid}$ = \rood{XX}.
-This result provides an experimental validation of the kinetic predictictions of the MSM models.
-
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and we attribute the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale.
+A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
+This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
+We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
+The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM.
+
+%%% possible additions to qualify what's above
+%Overall, there are several reasons why the T-jump response may not be directly sensitive to shifting.
+In the case of GC-end, the low population of shifted states alone may diminish contributions below a detectable limit.
+For AT-all, the particularly slow timescales associated with the most common shifting process (H $\leftrightarrow$ 5S2) can blend in with the slow dissociation response.
+In both cases, it is unclear how populations of shifted states change with temperature, making these processes difficult to identify with T-jump techniques.
+
+%Although there are no distinct ``middle'' responses at the expected shifting timescales, the response profile can inform the distribution of dynamics contributing to the signal \citep{Hamm2006StretchedFolding, Johnston2006StretchedDecays}.
+Shifted state are either (i) too low in population to detect a discernable Shifted $\rightarrow$ D response, or (ii) have lifetimes too long, (especially for the AT-all 5S2 state) to be discerned from the slow response.
+Most shifting behavior will not occur until similar timescales as full dissociation, given that H-->D transitions are approximately as likely as H-->D.
+Therefore, the only discernable traces would come from shifted state in the initial low temperature ensemble.
+For example, the AT-all 5S2 state, which contains 80\% of the shifted population under equilibrium conditions, is the only metastable state with a predicted lifetime within an order of magnitude of the hybridized state.
+Therefore it is likely that this relaxation response is blending with the slow response, while shorter-lived relaxations are too low in magnitude to detect.
+
+%%% Timescale approach %%%
+%We report in Fig.~\ref{fig:T-jump_stretching}a the predicted life times of each macrostate for each of the four DNA sequences at their respective melting temperatures.
+Only the AT-all sequence contains long-lived partially hybridized states with life times in excess of 50 ns and.
+The longest lived of these is the 5S2 state with a life time of 470 ns .
+GC-end also contains a number of out-of-register shifted states, but their life times are predicted to be an order of magnitude shorter than that of the corresponding states in AT-all.
+This is likely attributable to their reduced stability due to the two fewer available WC pairings for any given out-of-register GC-end state relative to AT-all.
+The sole intermediate F4 frayed state occupied by GC-core has a life time of only 2 ns, and the GC-mix sequence possesses no metastable states at all.
+Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence possessing long-lived intermediate states with life times of up to 470 ns should be distinguishable from the GC-end, GC-core, and GC-mix sequences predicted to have no metastable states or only very short-lived states with life times less than 30 ns.
+We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mix due to the very short-lived nature of these intermediate states.
+
+%We present in Fig.~\ref{fig:T-jump_stretching}b-d our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$.
+The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
+Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to \noter{What physical process.}  A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
+This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
+This result provides an experimental validation of the kinetic predictions of the MSM models.
+
+%%% orginal %%%
 %observe signatures  of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models.
@@
 In our previous work, we did not consider out-of-register states as part of the experimental or lattice model analysis, and therefore determined our ordering (GC-core $>$ GC-mix $>$ AT-all $>$ GC-end) in terms of frayed state populations prior to duplex dissociation\citep{Sanstead2016, Sanstead2018DirectDehybridization}.
-Depending on the biological system or nanotechnology application of interest, these forms of deviation may hold varied importance \noteb{could add example here, but don't want to dwell on this too much?}.
+Depending on the biological system or nanotechnology application of interest, these forms of deviation may hold varied importance...
 
@@
 
+\noteb{check out the options for TOC}
 \section*{TOC Image}
@@
 	\begin{center} 
-        \includegraphics[width=150mm, scale=1]{TOC.pdf}
+        \includegraphics[width=150mm, scale=1]{TOC_alternate2.pdf}
         %\caption*{TOC image}

### 2021-05-13 manuscript.tex
Andy edits

@@
     pdfborder={0 0 0},
-    linkcolor=blue,
-    citecolor=black
+    linkcolor=black,
+    citecolor=black,
+    urlcolor=black
 ]{hyperref}
@@
 Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate.
-The grey arrows between macrostates indicate the presence of a kinetic flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux.
-(To avoid congesting the diagram, arrows are not reported for fluxes lower than $3\times10^{-6}$ \noteb{flux is in terms of $p_i*p_{ij}$ so I think it should still be unitless?}.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM.
+The grey arrows between macrostates indicate the presence of a probability flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux.
+(To avoid congesting the diagram, arrows are not reported for probability fluxes lower than $3\times10^{-6}$.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM.
 Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions.
@@
 A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
-\noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A.
+As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.
+
+%\noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A.
 These MFPTs reflect the frequency of partial hybridization events and are in close agreement with the leading implied time scale of $t_1$ = 1.5 $\mu$s reported in Fig.~\ref{fig:allseq_table}d" This works for AT-all, but for all sequences the leading timescale tends to be lower by up to 2x than H $\dashrightarrow$ D MFTPs.
-I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)} As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.
+I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)}
 
@@
 
-\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}
+%\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}
 
@@
 
-\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}
+%\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}
 
@@
 
-\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }
+%\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }
 
@@
 
-\subsection{\label{sec:stretch_comparison}Long-lived metastable states predicted by the MSM are resolved by T-jump IR}
-
-\noteb{After running more low-temp simulations and looking at NN predictions for shifted states at low temps, it seems like the shifted equilibrium does not change too much (a relatively higher loss for GC-end than AT-all).
-I think the state lifetime argument becomes confusing since we (i) we are presenting simulation and experimental time scales on the same figure (ii) do not have an acceleration factor to readily convert these timescales (iii) do not have any evidence of mid-range relaxation that you might expect given this timescale spectrum.
-No need to mention any of this in the paper, but I think there is really a catch-22 with trying to detect these states with T-jump.
-The only process you can really resolve is going from shifted --> D since H--> shifted will be as slow as H-->D.
-If a state is unstable this process should happen at an intermediate timescale, but it contain such a small (less than 1\%) of the total population that it won't be picked up.
-If the metastable state is more stable then the population will be larger, but the process will occur near the same timescale as full dissociation and look the exact same.
-So all of that is to say that a stretched exponential can very reasonably be attributed to shifted-fraying response, but it becomes more of a thermodynamic argument than kinetic.}
-
-We hypothesized that we should be able to validate particular kinetic predictions of our sequence-dependent MSMs using T-jump IR spectroscopy.
-T-jump IR measurements commencing from low temperatures (at which the hybridized state dominates the equilibrium ensemble) apply a rapid step jump in temperature and probe the relaxation of the system to the dehybridized state.
-As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively.
-Shifting dynamics are more difficult to distinguish given that shifting events (e.g.
-H $\rightarrow$ 5S2) are shown by our MSMs to occur on similar time scales as H $\rightarrow$ D and would therefore be blended into the slow response.
-However, we do expect some portion of low temperature duplexes to initialize in shifted states and alter fast fraying dynamics.
-Furthermore, we expect this effect to be proportional to the population of shifted states relative to intact hybridized states.
-At our MSM equilibrium distribution, we find that AT-all shifted states make up 24\% of the hybridized ensemble (all states excluding D) compared to only 0.3\% for GC-end shifted states.
-GC-core and GC-mix would not be expected to have any shifted states in the starting ensemble.
-Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence, possessing significantly more heterogeneous state distribution after the T-jump, should be distinguishable from the GC-end, GC-core, and GC-mix sequences.
-We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mix due to the similarity of the starting ensemble.
+\subsection{\label{sec:stretch_comparison}Long-lived metastable shifted states predicted by the MSM are resolved by T-jump IR}
+
+Finally, we sought to validate the predictions of our sequence-dependent MSMs against experimental T-jump IR spectroscopy.
+T-jump IR measurements commence from a low temperatures, applies a step jump in temperature, and tracks the relaxation of the system to the dehybridized state.
+We hypothesized that the influence of the out-of-register shifted states present in the AT-all and GC-end sequences upon the system relaxation kinetics should be manifest in the slow and/or fast responses measured by T-jump IR.
+As discussed in Section \ref{sec:time}, the slow IR response corresponds to dissociation and the fast to terminal base fraying.
+With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on \rood{microsecond} time scales, which are commensurate with the $\sim$\rood{XX} $\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D).
+As such, we anticipate that the dynamical relaxations associated with out-of-register shifted states proceed on similar time scales to, and may not be distinguishable from, the relaxation to the dehybridized state.
+Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a \rood{stronger / slower / longer-lived} fast fraying response corresponding to a \rood{more stretched relaxation over \rood{XX}-\rood{XX} ns time scales}.
+Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
+In is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an \rood{elongation of the fast relaxation response associated with terminal base fraying}.
+% We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
 
@@
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and we attribute the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the \rood{heterogeneity of fraying dynamics at the fast timescale} \noter{Whatever you write here should be consistent with the text in the hypothesis paragraph above.}.
 A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
 This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
-We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
-The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM.
+This result supports our hypothesis and validates a testable dynamical prediction of our sequence-dependent MSMs.
 
@@
 
-\noteb{check out the options for TOC}
 \section*{TOC Image}
@@
 	\begin{center} 
-        \includegraphics[width=150mm, scale=1]{TOC_alternate2.pdf}
+        \includegraphics[width=150mm, scale=1]{TOC.pdf}
         %\caption*{TOC image}

### 2021-05-14 cover_letter/letterhead.tex
cover letter and citations

@@
 \vskip -0.1in
-\noindent cc: M.~Zhao, K.~Lachowski, S. Alamdari, J.~Sampath, P.~Mu, L.D.~Pozzo, C.-L.~Chen, C.J.~Mundy, J.~Pfaendtner 
+\noindent cc: M.~Jones, B.~Ashwood, A.~Tokmakoff
 }
@@
 %----------------------------------------------------------------------------------------
+\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
 
@@
 
-We are pleased to submit a manuscript for consideration for publication in the \textit{The Journal of the American Chemical Society} an original research \textbf{Article} entitled \textbf{``Hierarchical Self-Assembly Pathways of Polypeptoid Helices and Sheets''} by Mingfei Zhao (U.~Chicago), Kacper Lachowski(U.W.~Seattle), Sarah Alamdari (U.W.~Seattle), Janani Sampath (PNNL),  Peng Mu (PNNL), Chun-Long Chen (PNNL, U.W.~Seattle), Lilo D.\ Pozzo (U.W.~Seattle), Christopher J.\ Mundy (PNNL, U.W.~Seattle), Jim Pfaendtner (U.W.~Seattle, PNNL), and Andrew L.\ Ferguson (U.~Chicago).
+We are pleased to submit a manuscript for consideration for publication in the \textit{The Journal of the American Chemical Society} an original research \textbf{Article} entitled \textbf{``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy''} by Michael S.
+Jones (U.~Chicago), Brennan Ashwood(U.~Chicago), Andrei Tokmakoff(U.~Chicago), and Andrew L.\ Ferguson (U.~Chicago).
 This article has not been published previously and is not under consideration in any other journal.
@@
 % background and why it is important to do this.
-\textbf{Background.} Polypeptoids (poly-N-substituted glycines) are a class of highly tailorable synthetic peptidomic polymers with applications as drugs, antimicrobials, and catalysts that can be engineered to assemble into nanoaggregates including spheres, helices, tubes, and sheets.
-Amphiphilic diblock polypeptoids have been engineered to assemble high-aspect ratio 2D crystalline lattices with applications in catalysis, molecular separations, and photovoltaics.
-Assembly is induced by dissolving the peptoids in an organic solvent/water mixture and evaporating the organic phase to promote hydrophobic association, but the mechanistic pathways and thermodynamic/kinetic driving forces mediating assembly remain uncharacterized.
+\textbf{Background.}  A robust understanding of the sequence-dependent thermodynamics of DNA hybridization has enabled rapid advances in DNA nanotechnology.
+A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped.
+Although previous experimental and computational studies have identified deviations from an "all-or-nothing" hybridization/dehybridization model, few studies have integrated these approaches in order to evaluate and quantify sequence-dependent dynamics.
 
 % what we did in this paper
-\textbf{Key Results.}  In this work, we integrate all-atom molecular dynamics (MD) simulation, small angle x-ray scattering (SAXS), circular dichroism (CD), atomic force microscopy (AFM), and x-ray diffraction (XRD) to present an integrated computational/experimental study of the self-assembly of Nbrpe6Nc6 -- an amphiphilic diblock polypeptoid comprising an \ce{NH2} capped block of six hydrophobic N-((4-bromophenyl)ethyl)glycine residues conjugated to a hydrophilic 6-aminohexanal (\ce{NH3(CH2)5CO}) tail -- in organic solvent/water mixtures.
-Our MD simulations predict a hierarchical assembly pathway by which monomers first assemble into disordered aggregates that self-order into 1D chiral helical rods that subsequently assemble 2D achiral crystalline nanosheets.
-Experimentally, we observe stable, free-floating 1D rods in mixed solvent using SAXS and CD, and 2D crystalline sheets in pure water using XRD and AFM.
-MD predictions of the relative thermodynamic stability of disordered aggregates, helical rods, and sheets as a function of the organic solvent concentration support a thermodynamically-controlled assembly process.
+\textbf{Key Results.}  In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
+We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
 
 % impact and importance of this paper
-\textbf{Significance.} Our work integrates simulation and experiment to establish a previously unknown hierarchical polypeptoid assembly pathway wherein 0D monomers first assemble into disordered aggregates that ripen into 1D helical rods that subsequently form 2D crystalline nanosheets.
-This new understanding of the hierarchical assembly mechanisms and stable intermediaries presents new principles and guidance for the rational design of peptoid-based nanomaterials with potential biochemical, biomedical, and bioengineering applications including as drugs, antimicrobials, catalysts, and molecular sieves.
+\textbf{Significance.} Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.
 
 % fit for the journal
-\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of polypeptoid self-assembly.
-This new understanding establishes new mechanistic understanding and a thermodynamic basis for existing experimental evaporation-induced assembly protocols and identifies previously unknown solvent conditions for the stable fabrication of 1D helical rods.
-These principles offer new strategies for the supramolecular assembly of diverse peptoid sequences into 1D and 2D supramolecular aggregates.
+\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of DNA hybridization as a function of sequence.
+These sequence-dependent mechanisms are of increasing importance to both fundamental biophysical and biochemical processes and rapidly developing nanotechnology applications.
 
@@
 
-\textbf{\textit{Simulation.}}
+\textbf{\textit{Simulation.}} \noteb{Notes are just to show connection}
 
-Stephen Whitelam \hfill Molecular Foundry, Lawrence Berkeley National Lab \\
-(510) 495-2769 \hfill \url{swhitelam@lbl.gov}
+Joshua Lequieu \hfill Chemical and Biological Engineering, Drexel University \\
+(215) 895-6624 \hfill \url{lequieu@drexel.edu}
 
-$\circ$ \textit{Expert in biomolecular simulation and molecular modeling of peptoid assembly.}
+$\circ$ \textit{Expert in multi-scale biomolecular simulation with a focus in materials science and biology. \noteb{3spn2}}
 
-Monica Olvera de la Cruz \hfill Materials Science and Engineering and Chemistry, Northwestern University \\
-(847) 491-7801 \hfill \url{m-olvera@northwestern.edu}
+Jonathon K. Whitmer \hfill Chemical and Biomolecular Engineering, Notre Dame \\
+(574) 631-1417 \hfill \url{jwhitme1@nd.edu}
 
-$\circ$ \textit{Expert in molecular modeling and simulation of peptides, polymers, and materials assembly.}
+$\circ$ \textit{Expert in simulations of polyelectrolytes and self-assembling systems.  \noteb{3spn2}}
 
-Arthi Jayaraman \hfill Chemical and Biomolecular Engineering, University of Delaware \\
-(302) 831-8682 \hfill \url{arthij@udel.edu}
+Thomas Knotts \hfill Chemical Engineering, Brigham Young University \\
+(801) 422-9158 \hfill \url{thomas.knotts@byu.edu}
 
-$\circ$ \textit{Expert in theory and simulation of biomolecular assembly and molecular materials design.}
+$\circ$ \textit{Expert in theory and simulation of multi-molecular systems.  \noteb{3spn.0 and .1}}
+
+Shoji Takada \hfill Biophysics, Graduate School of Science, Kyoto University \\
++81-(0)75-753-4070 \hfill \url{takada@ biophys.kyoto-u.ac.jp}
+
+$\circ$ \textit{Expert in simulations of DNA-protein systems \noteb{3spn2 and  CG protein modeling}}
+
+Thomas Ouldridge \hfill  Bioengineering, Imperial College London \\
++44-(0)20-7589-5111 \hfill \url{t.ouldridge@implerial.ac.uk}
+
+$\circ$ \textit{Expert in simulations of biochemical systems for nanotechnology applications \noteb{oxDNA}}
+
+Modesto Orozco \hfill  Institute for Research in Biomedicine, University of Barcelona \\
++34-93-40-37156 \hfill \url{modesto.orozco@irbbarcelona.org}
+
+$\circ$ \textit{Expert in molecular dynamics simulations of nucleic acids \noteb{all-atom DNA model}}
 
@@
 
-Darrin Pochan \hfill Materials Science and Engineering, University of Delaware \\
-(302) 831-3569 \hfill \url{pochan@udel.edu}
+Anjum Ansari \hfill Physics, University of Illinois at Chicago\\
+(312) 996-8735 \hfill \url{ansari@uic.edu}
 
-$\circ$ \textit{Expert in biomolecular and polymeric assembly and characterization.}
+$\circ$ \textit{Expert in kinetics of biomolecular relaxation from non-equilibrium ensembles.}
 
-Rachel Segalman \hfill Chemical Engineering, UC Santa Barbara \\
-(805) 893-3709 \hfill \url{segalman@ucsb.edu}
+Neil Hunt \hfill Chemistry, University of York \\
++44-(0)19-0432-2511 \hfill \url{neil.hunt@york.ac.uk}
 
-$\circ$ \textit{Expert in biological and polymeric materials assembly.}
+$\circ$ \textit{Expert in ultrafast spectroscopy of complex fluids and biological systems.}
 
-Ron Zuckermann \hfill The Molecular Foundry, Lawrence Berkeley National Laboratory \\
-(510) 387-9364 \hfill \url{rnzuckermann@lbl.gov}
+Xinsheng Zhao \hfill Chemistry and Molecular Engineering, Peking University \\
++86-10-6275-1727 \hfill \url{zhaoxs@pku.edu.cn}
 
-$\circ$ \textit{Pioneer of peptoid science. Expert in peptoid synthesis and characterization.}
+$\circ$ \textit{Expert in dynamic mechanisms of biomolecular reactions and nucleic acid biophysics.}
 

### 2021-05-14 manuscript.tex
cover letter and citations

@@
 \usepackage{framed}
+\usepackage{achemso}
 
@@
 From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing.
-\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnological applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the \textit{dynamical}, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves.
+\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnological applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the \textit{dynamical}, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves.
 \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence.
@@
 A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved.
-For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the extent to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
+For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the extent to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
 The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.
@@
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
@@
 \citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new and existing experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
-Consistent with previous studies, \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
+Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
 Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states.
@@
 \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation.
-\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.
+\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSimulations, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.
 
 All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model.
-\citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
+\citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
 Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s.
@@
 
-Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionV1.0} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
+Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
@@
 \textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM.
-The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
+The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionSoftware} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
 We observe rapid convergence of the implied time scales $t_i$ with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
@@
 These macrostate MSMs constitute our terminal kinetic models.
-We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}.
+We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Prinz2011OptimalDynamics, Wehmeyer2019IntroductionSoftware}.
 This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$ should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$.
@@
 % need some more detail here -- talk to Andy about this
-%MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
+%MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Prinz2011OptimalDynamics, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
 The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
@@
 We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
-The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{PinamontiTheModels, Zgarbova2014BaseRNA}
+The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
 
@@
 We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association.
-The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
+The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
 We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences.
@@
 Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model.
-Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{Michele2014EHybridization}
+Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{DiMichele2014EffectHybridization}
 
@@
 It is known from differential scanning calorimetry studies~\citep{Dickman2012ThermodynamicDNAs} that 3$^\prime$ inert tails are more destabilizing than 5$^\prime$ tails, with the differential behavior attributed to a combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure.
-\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization}.
+\citep{Doktycz1990ThermodynamicC, Dickman2012ThermodynamicDNAs, DiMichele2014EffectHybridization}.
 Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$).
@@
         \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero.
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero.
 (b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model.
@@
 The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D.
-Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D.
+Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicC, Dickman2012ThermodynamicDNAs, DiMichele2014EffectHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D.
 The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state.
@@
 The 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H transition probabilities are more than an order of magnitude larger than in AT-all, which may again be attributed to the lower thermodynamic stability of the two shifted states relative to the hybridized state H.
-Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicATGC, Dickman2012ThermodynamicDNAs, Michele2014EHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately seven times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
+Again, the transition probabilities out of the 3S2 state to D or H are comparatively higher than those out of the 5S2 state, consistent with the increased destabilizing effect of 3$^\prime$ inert tails.~\citep{Doktycz1990ThermodynamicC, Dickman2012ThermodynamicDNAs, DiMichele2014EffectHybridization} Commencing from the hybridized state H, a direct transition to the dissociated state is approximately seven times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
 A transition path theory analysis of the MSM reveals that only 7\% of hybridization events D $\dashrightarrow$ H and dehybridization events H $\dashrightarrow$ D proceed through one or more out-of-register shifted states.
@@
 As discussed in Section \ref{sec:time}, the slow IR response corresponds to dissociation and the fast to terminal base fraying.
-With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on \rood{microsecond} time scales, which are commensurate with the $\sim$\rood{XX} $\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D).
+With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on microsecond time scales, which are commensurate with the 1.4 - 2.9$\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D).
 As such, we anticipate that the dynamical relaxations associated with out-of-register shifted states proceed on similar time scales to, and may not be distinguishable from, the relaxation to the dehybridized state.
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a \rood{stronger / slower / longer-lived} fast fraying response corresponding to a \rood{more stretched relaxation over \rood{XX}-\rood{XX} ns time scales}.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated and slightly faster fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a more stretched relaxation over 70-100 ns (experimental) time scales.
 Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
-In is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an \rood{elongation of the fast relaxation response associated with terminal base fraying}.
+It is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
 % We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
@@
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the \rood{heterogeneity of fraying dynamics at the fast timescale} \noter{Whatever you write here should be consistent with the text in the hypothesis paragraph above.}.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale.
 A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
@@
 %\bibliographystyle{unsrt}
-\bibliography{references_210505}
-
-%\bibliography{refs_mendeley2}
+%\bibliographystyle{achemso} % should be pre-defined in usepackage
+\bibliography{references_210514}
 

### 2021-05-14 SI.tex
cover letter and citations
Final edits before sending to Andrei

@@
         \includegraphics[width=1.0\textwidth]{FigS3.pdf}
-        \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by an MSM constructed at the $\tau$ = 1.2 ns lag time (dashed blue line) versus those computed from an MSM constructed at the particular lag time (solid black line).
+        \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by $k$ applications of an MSM constructed at the $\tau$ = 1.2 ns lag time $\mathbf{P}^k(\tau)$ (dashed blue line) versus those computed from an MSM constructed at that particular lag time $\mathbf{P}(k \tau)$ (solid black line).
 The good agreement between these two results provides numerical validation of the Markovian nature of the $\tau$ = 1.2 ns lag time MSM.}
@@
 $\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TM}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
-A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig4}.
+A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig6}.
 $\Delta S_{NN}^{\circ}$ is constructed in an analogous manner from NN, dangling end, and initiation terms, plus a symmetry term, $\Delta S_{sym}^{\circ}$, that applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.
@@
 \bibliography{references_210505}
-%\bibliography{refs_mendeley2}
 

### 2021-05-14 cover_letter/letterhead.tex
Final edits before sending to Andrei

@@
 \vskip -0.1in
-\noindent cc: M.~Jones, B.~Ashwood, A.~Tokmakoff
+\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
 }
@@
 We are pleased to submit a manuscript for consideration for publication in the \textit{The Journal of the American Chemical Society} an original research \textbf{Article} entitled \textbf{``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy''} by Michael S.
-Jones (U.~Chicago), Brennan Ashwood(U.~Chicago), Andrei Tokmakoff(U.~Chicago), and Andrew L.\ Ferguson (U.~Chicago).
+Jones, Brennan Ashwood, Andrei Tokmakoff, and Andrew L.\ Ferguson, all from the University of Chicago.
 This article has not been published previously and is not under consideration in any other journal.
@@
 A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped.
-Although previous experimental and computational studies have identified deviations from an "all-or-nothing" hybridization/dehybridization model, few studies have integrated these approaches in order to evaluate and quantify sequence-dependent dynamics.
+Although previous experimental and computational studies have identified deviations from an ``all-or-nothing'' hybridization/dehybridization model, few studies have integrated these approaches in order to evaluate and quantify sequence-dependent dynamics.
 
 % what we did in this paper
-\textbf{Key Results.}  In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
-We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+\textbf{Key Results.}  In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
+We conducted more than 1 ms of coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
 Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
@@
 % impact and importance of this paper
-\textbf{Significance.} Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.
+\textbf{Significance.} Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, offer quantitative models of their long-time kinetics, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.
 
 % fit for the journal
-\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of DNA hybridization as a function of sequence.
-These sequence-dependent mechanisms are of increasing importance to both fundamental biophysical and biochemical processes and rapidly developing nanotechnology applications.
+\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of sequence-dependent DNA hybridization/dehybridization pathways.
+These sequence-dependent mechanisms are of both fundamental interest in biophysical and biochemical processes and of increasing importance in burgeoning DNA nanotechnology applications reliant on quantitative understanding of the sequence-dependent kinetic pathways and rates.
 
@@
 
-\textbf{\textit{Simulation.}} \noteb{Notes are just to show connection}
+\textbf{\textit{Simulation.}}
 
@@
 
-$\circ$ \textit{Expert in multi-scale biomolecular simulation with a focus in materials science and biology. \noteb{3spn2}}
+$\circ$ \textit{Expert in multi-scale biomolecular simulation with a focus in materials science and biology.}
 
-Jonathon K. Whitmer \hfill Chemical and Biomolecular Engineering, Notre Dame \\
+Jon K. Whitmer \hfill Chemical and Biomolecular Engineering, Notre Dame \\
 (574) 631-1417 \hfill \url{jwhitme1@nd.edu}
 
-$\circ$ \textit{Expert in simulations of polyelectrolytes and self-assembling systems.  \noteb{3spn2}}
+$\circ$ \textit{Expert in simulations of biomolecular and self-assembling systems.}
 
@@
 
-$\circ$ \textit{Expert in theory and simulation of multi-molecular systems.  \noteb{3spn.0 and .1}}
+$\circ$ \textit{Expert in theory and simulation of coarse-grained biomolecular systems.}
 
@@
 
-$\circ$ \textit{Expert in simulations of DNA-protein systems \noteb{3spn2 and  CG protein modeling}}
+$\circ$ \textit{Expert in simulations of coarse-grained modeling of DNA-protein systems.}
 
@@
 
-$\circ$ \textit{Expert in simulations of biochemical systems for nanotechnology applications \noteb{oxDNA}}
+$\circ$ \textit{Expert in simulations of biochemical systems for nanotechnology applications.}
 
@@
 
-$\circ$ \textit{Expert in molecular dynamics simulations of nucleic acids \noteb{all-atom DNA model}}
+$\circ$ \textit{Expert in molecular dynamics simulations of nucleic acids.}
 

### 2021-05-14 manuscript.tex
Final edits before sending to Andrei

@@
 Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
-Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature.
-The temperature at which the hybridized and dissociated populations were closest was selected.
+Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The temperature at which the hybridized and dissociated populations were closest was selected.
 The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step.
@@
 The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions.
-The magnitude of this displacement did not affect the results so long as all native WC bonds were completely broken.
+The magnitude of this displacement did not affect the results so long as all native Watson-Crick (WC) bonds were completely broken.
 Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest.
@@
 
-Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
+Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
+In brief, a MSM extracts from simulation trajectories an ensemble of long-lived metastable macrostates, their equilibrium occupancy populations, and the equilibrium probability fluxes between them.
+In this manner, they provide an interpretable and predictive model of the system thermodynamics and long-time kinetics.
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
-Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 
-To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
+To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
 	\begin{equation}\label{e1}
-	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow}
+	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow},
 	\end{equation}
 where $[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a^\mathrm{slow}$ is the association rate constant.
-In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d$ to solve for $k_d$ and $k_a$.
+In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d$ to solve for $k_d^\mathrm{slow}$ and $k_a^\mathrm{slow}$ as a function of $K_d$ and $[S]_{T_f}$,
 	\begin{equation}\label{e2a}
-	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d(T_f)}{K_d(T_f)+4[S]_{T_f}}
+	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d(T_f)}{K_d(T_f)+4[S]_{T_f}},
 	\end{equation}
 	\begin{equation}\label{e2b}
-	k_a^\mathrm{slow} = \frac{k_d^\mathrm{slow}}{K_d(T_f)}
+	k_a^\mathrm{slow} = \frac{k_d^\mathrm{slow}}{K_d(T_f)}.
 	\end{equation}
 
-FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement.
+FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
 	\begin{equation}\label{e3a}
-	[S]_{T_f} = [1 - \theta(T_f)]c_{tot}
+	[S]_{T_f} = [1 - \theta(T_f)]c_{tot},
 	\end{equation}
 	\begin{equation}\label{e3b}
-	K_d(T_f) = \frac{2c_{tot}[1 - \theta(T_f)]^2}{\theta(T_f)}
+	K_d(T_f) = \frac{2c_{tot}[1 - \theta(T_f)]^2}{\theta(T_f)},
 	\end{equation}
-$c_{tot}$ is the total oligonucleotide concentration, which was 2 mM for all measurements.
+where $c_{tot}$ is the total oligonucleotide concentration, which was 2 mM for all measurements.
 
@@
 We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
-The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
+The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
 
@@
 We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the Watson-Crick (WC) interaction.
+Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction.
 We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
@@
 We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models.
-MSMs are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates.
+MSMs define the long-lived metastable macrostates of the system, their equilibrium occupancies, and the equilibrium transition probabilities between them.
+As such, they are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates.
 We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences.
@@
 The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways.
-Immediately we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior.
+Immediately, we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas at the other end of the spectrum GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior.
 In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM.
@@
 Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2.
-Once in one of the four shifted states, the 5$^\prime$ vs.
-3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate.
+Once in one of the four shifted states, the 5$^\prime$ vs.\ 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate.
 Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states.
@@
 
-Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25-times more likely than dissociation F4 $\rightarrow$ D.
+Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25 times more likely than dissociation F4 $\rightarrow$ D.
 Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
@@
 We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
-This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the frayed two-base AT tails relative to the frayed four-base AT-tails.
+This stands in contrast to the metastable F4 state populated by GC-core.
 Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s, which are again nearly equal.
@@
 Finally, we sought to validate the predictions of our sequence-dependent MSMs against experimental T-jump IR spectroscopy.
-T-jump IR measurements commence from a low temperatures, applies a step jump in temperature, and tracks the relaxation of the system to the dehybridized state.
+T-jump IR measurements commence from a low temperatures, apply a step jump in temperature, and track the relaxation of the system to the dehybridized state (Section \ref{sec:Tjump}).
 We hypothesized that the influence of the out-of-register shifted states present in the AT-all and GC-end sequences upon the system relaxation kinetics should be manifest in the slow and/or fast responses measured by T-jump IR.
 As discussed in Section \ref{sec:time}, the slow IR response corresponds to dissociation and the fast to terminal base fraying.
-With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on microsecond time scales, which are commensurate with the 1.4 - 2.9$\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D).
+With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on microsecond time scales, which are commensurate with the 1.4 - 2.9 $\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D) for each of the four sequences.
 As such, we anticipate that the dynamical relaxations associated with out-of-register shifted states proceed on similar time scales to, and may not be distinguishable from, the relaxation to the dehybridized state.
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated and slightly faster fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a more stretched relaxation over 70-100 ns (experimental) time scales.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated and slightly faster fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
 Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
@@
 We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix).
-We conducted more than 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+We conducted 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
 T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the kinetic prediction of the Markov state models that the AT-all sequence should possess long-lived out-of-register shifted states that are detectable within T-jump IR t-HDVE time traces.
@@
 Analysis of the specific pathways reveals hybridization to largely proceed by a nucleation-zippering mechanism and dehybridization to proceed by a fraying-peeling mechanism.
-The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix -- with this trend largely dictates by sequence repetitiveness, specifically the number of consecutive AT motifs.
+
+The ordering of the computationally predicted kinetic landscapes from most to least complex -- AT-all $>$ GC-end $>$ GC-core $>$ GC-mix -- is largely dictated by sequence repetitiveness, specifically the number of consecutive AT motifs.
 We note that this ordering differs from our previously reported ordering in terms of deviation from two-state behavior of GC-core $>$ GC-mix $>$ AT-all $>$ GC-end.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} We can understand these two apparently discrepant orderings by understanding that the latter was deduced based on experimental analyses and lattice models that did not account for out-of-register states and focussed largely on fraying behaviors.

### 2021-05-17 manuscript.tex
Brennan edits

@@
 The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket.
-The T-jump temperature change ($\Delta$T) was set to 15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam.
+The T-jump temperature change ($\Delta$T) was set to 14-16 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam.
 The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.
@@
 
-\section{\label{sec:Results}Results}
+\section{\label{sec:Results}Results and Discussion}
 
@@
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
-Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
@@
 
+\section*{Supporting Information Available}
+
+The following files are available free of charge.
+
+\begin{itemize}
+\item SI.pdf
+\end{itemize}
+
+\section*{Conflict of Interest Statement}
+
+A.L.F.\ is a co-founder and consultant of Evozyne, Inc.\ and a co-author of US Provisional Patents 62/853,919 and 62/900,420 and International Patent Applications PCT/US2020/035206 and PCT/US20/50466.
+
+\section*{Acknowledgements}
+
+This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805.
+This work was completed in part with resources provided by the University of Chicago Research Computing Center.
+We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No.
+DMR-1828629.
+
 \clearpage

### 2021-05-17 manuscript.tex
acs style, citation abbreviations, figure 3 label

@@
 
-%add refs to SI
-%\usepackage{xr}
-%\externaldocument{SI}
-
 %\usepackage[hidelinks]{hyperref}
-%\usepackage{algorithm}
 %\usepackage{algpseudocode}
@@
 
-\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
-
-\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
-\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
-\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
-
-\newcommand*{\rood}[1]{{\color{red}{#1}}}
-\newcommand*{\bood}[1]{{\color{blue}{#1}}}
+%\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+
+%\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
+%\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
+%\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+
+%\newcommand*{\rood}[1]{{\color{red}{#1}}}
+%\newcommand*{\bood}[1]{{\color{blue}{#1}}}
 \newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
-\newcommand*{\groen}[1]{\textcolor{green}{#1}}
-
-\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
-\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
-\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
+
+%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
 %\newcommand*{\noter}[1]{}					% notes off
@@
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA.
-\citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence.
+\citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky2019High-ResolutionVAMPnets} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence.
 SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000.
 We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics.
-\citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
+\citep{Phys2011MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
 \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs.
@@
 We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
-A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
+A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
 
@@
 This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805.
+B.A.
+acknowledges support from the NSF Graduate Research Fellowship Program.
 This work was completed in part with resources provided by the University of Chicago Research Computing Center.
@@
 %\bibliographystyle{achemso} % should be pre-defined in usepackage
-\bibliography{references_210514}
+\bibliography{references_210514_abv}
 
@@
         \includegraphics[width=150mm, scale=1]{TOC.pdf}
-        %\caption*{TOC image}
+        \caption*{For Table of Contents Only}
         \label{fig:TOC}

### 2021-05-18 manuscript.tex
Andrei edits

@@
 %\newcommand*{\rood}[1]{{\color{red}{#1}}}
-%\newcommand*{\bood}[1]{{\color{blue}{#1}}}
-\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
+%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
+\newcommand*{\blauw}[1]{{#1}}
 
 %\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
-%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+%\newcommand*{\noter}[1]{}					% notes off
 %\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
-%\newcommand*{\noter}[1]{}					% notes off
 
@@
   The University of Chicago, %
-  929 East 57th Street, Chicago, Illinois 60637, United States%
+  5640 South Ellis Avenue, Chicago, Illinois 60637, United States%
 }
@@
 
-We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix).
+We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix).
 We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) model that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing.
@@
 The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}.
-The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO).
-Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump.
+The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG laser sent through an optical parametric oscillator.
+Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump with a synchronized 1 kHz spectrometer.
 T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between (-10) and 10 fs.
-t-HDVE spectra were acquired with a parallel pulse polarization scheme.
-Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum.
-t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.
-
-The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a home-built brass jacket.
+t-HDVE spectra were acquired with a parallel pulse polarization scheme and presented as a dispersed pump-probe (t-DPP) spectrum.
+t-DPP data are reported as the difference spectra relative to the initial temperature.
+
+The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a brass jacket.
 The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket.
@@
 We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
-First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm.
+First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
 This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
@@
 Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal (+4) K corrective calibration to our computational results.
-We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison.
+We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but note that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
 The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.
@@
 Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
-(c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
+\noter{Add percentages to blue bars in white text.} (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
 The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a.
@@
 The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
-Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare the free energies predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model.
+Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare free energy differences predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model.
+For clarity of presentation, we specify the additive constant $C$ in both the MSM and NN models by arbitrarily assigning the free energy of the hybridized state H to define the zero of free energy.
 A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the \blauw{Supporting Information}.
 
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized (H) state by as much as 14.7 kJ/mol.
-(The free energy of the hybridized state is set to zero by construction for both the MSM and NN models.) The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
-Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the 5S2, 3S2, 5S4, and 3S4 states?
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized state by as much as 14.7 kJ/mol.
+The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
+Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the 5S2, 3S2, 5S4, and 3S4 states?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
-There are approximations inherent in the 3SPN.2 model and the NN model was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions.
+There are approximations inherent in the 3SPN.2 model, not least of which is the coarse-grained representation that integrates over atomic degrees of freedom, and in the NN model that was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions.
 Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model.
@@
 We can further explore the role of inert tails upon macrostate stability by analyzing the AT-all and GC-end sequences that occupy out-of-register shifted macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}).
-For both sequences, our simulations show that 5S2 is both the most stable of shifted states and has the smallest discrepancy (4.0 kJ/mol AT-all, 10.0 kJ/mol GC-end) compared to NN predictions (Fig.~\ref{fig:NN_table}b).
+For both sequences, our simulations show that 5S2 is both the most stable of shifted states relative to H (Fig.~\ref{fig:NN_table}a) and has the smallest discrepancy (4.0 kJ/mol AT-all, 10.0 kJ/mol GC-end) compared to NN predictions (Fig.~\ref{fig:NN_table}b).
 The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol AT-all, 14.3-14.7 kJ/mol GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
@@
 Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$).
-For GC-end, the MSM and NN models both predict the four shifted states to be less stable than the corresponding states in AT-all.
+For GC-end, the MSM and NN models both predict the four shifted states to be less stable relative to the hybridized state than the corresponding predictions for AT-all.
 This is in line with expectations since the terminal G:C pairs in GC-end decrease by two the number of available WC pairings in out-of-register shifted states compared to AT-all.
@@
 (b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model.
-With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.}
+With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.
+\noter{(i) y-axis label in a: ``free energy relative to hybridized state, F (kJ/mol)'' (ii) y-axis label in b: ``free energy discrepancy between MSM and NN model relative to hybridized state, $\Delta$F (kJ/mol)'' -- might have to go on two lines}}
         \label{fig:NN_table}
@@
 The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free.
-(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
+(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) \noter{Maybe add a sentence here on other frayed states containing 3, 2, 1 frayed bases but not metastable within our MSM.} Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
 Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns.
@@
 The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale.
-A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
+A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter (i.e., a more stretched response) relative to those for GC-end, GC-core, and GC-mix.
 This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
-This result supports our hypothesis and validates a testable dynamical prediction of our sequence-dependent MSMs.
+This result supports our hypothesis and validates a testable prediction of our sequence-dependent MSMs.
 
@@
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
-Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
 The stretched exponential describes the process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
@@
 
-\section*{Supporting Information Available}
-
-The following files are available free of charge.
-
-\begin{itemize}
-\item SI.pdf
-\end{itemize}
+\section*{Supporting Information}
+
+Details of nearest neighbor (NN) model calculations, supplementary figures illustrating selection of optimal number of SRV slow modes for each sequence, convergence of MSM implied time scales, MSM Chapman-Kolmogorov tests, ``computational T-jump'' calculations, and GC-mix hybridization and dehybridization trajectories.
 
@@
 This work was completed in part with resources provided by the University of Chicago Research Computing Center.
-We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No.
-DMR-1828629.
+We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No.\ DMR-1828629.
 

### 2021-05-19 manuscript.tex
Figure changes and acknowledgements

@@
 Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
-\noter{Add percentages to blue bars in white text.} (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
+(c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
 The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a.
@@
 We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association.
-The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
+The nearest neighbor (NN) model is a popular empirical model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
 We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences.
@@
 
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized state by as much as 14.7 kJ/mol.
-The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM \noter{and therefore predicts a higher metastable state populations.} The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
 Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the 5S2, 3S2, 5S4, and 3S4 states?
@@
 In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails.
-In these instances the NN model overestimates macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol.
+In these instances the NN model overestimates \noter{another mention of ``overestimate'' but perhaps we have qualified it more at this point in the text?} macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol.
 The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.
@@
 (b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model.
-With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.
-\noter{(i) y-axis label in a: ``free energy relative to hybridized state, F (kJ/mol)'' (ii) y-axis label in b: ``free energy discrepancy between MSM and NN model relative to hybridized state, $\Delta$F (kJ/mol)'' -- might have to go on two lines}}
+With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.}
         \label{fig:NN_table}
@@
 The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free.
-(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) \noter{Maybe add a sentence here on other frayed states containing 3, 2, 1 frayed bases but not metastable within our MSM.} Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
+(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to inter-convert faster than the lag time and are not metastable within our MSM.
+Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
 Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns.
@@
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to \noter{What physical process.}  A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to.
+A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
 This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
@@
 This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805.
+A.T.
+thanks the National Institute of General Medical Sciences of the National Institutes of Health (Award No.
+R01-GM118774) for support of this research.
 B.A.

### 2021-05-19 SI.tex
Andrei edits
Andy pre-submission edits

@@
 \usepackage[%  
-    colorlinks=false,
+    colorlinks=true,
     pdfborder={0 0 0},
-    linkcolor=blue
+    linkcolor=black,
+    citecolor=black,
+    urlcolor=black
 ]{hyperref}
@@
 
-\newcommand*{\rood}[1]{{\color{red}{#1}}}
+%\newcommand*{\rood}[1]{{\color{red}{#1}}}
 %\newcommand*{\rood}[1]{#1}
-\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
-%\newcommand*{\blauw}[1]{#1}
-\newcommand*{\groen}[1]{\textcolor{green}{#1}}
+%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
+\newcommand*{\blauw}[1]{#1}
+%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
 %\newcommand*{\groen}[1]{#1}
@@
 
-\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
 
-\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
+%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
 %\newcommand*{\noter}[1]{}					% notes off
@@
 %\usepackage[draft]{todonotes}   % notes shown
-\usepackage[disable]{todonotes} % notes hidden
+%\usepackage[disable]{todonotes} % notes hidden
 
@@
     \caption{Convergence of the MSM implied time scales $t_i$ as a function of lag time $\tau$.
-Solid lines indicate maximum likelihood result while dashed lines show the Bayesian ensemble means.
+Solid lines indicate maximum likelihood result while dashed lines show the Bayesian ensemble means superposed on shaded areas representing 95\% confidence intervals.
+The high simulation data volume used in our MSM construction means that the maximum likelihood and Bayesian ensemble means are in good agreement and confidence intervals are narrow.
 The implied time scales for all sequences converge at a lag time of $\tau$ = 1.2 ns (vertical line).
@@
         \caption{Chapman-Kolmogorov (CK) tests comparing the probabilities of remaining within each macrostates for each sequence as a function of lag time predicted by $k$ applications of an MSM constructed at the $\tau$ = 1.2 ns lag time $\mathbf{P}^k(\tau)$ (dashed blue line) versus those computed from an MSM constructed at that particular lag time $\mathbf{P}(k \tau)$ (solid black line).
+The blue shaded area indicates the estimated standard deviation in $\mathbf{P}^k(\tau)$.
 The good agreement between these two results provides numerical validation of the Markovian nature of the $\tau$ = 1.2 ns lag time MSM.}
@@
     \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
-Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional hybridization and dissociation events.
+Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events.
 Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
 Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
-Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C WC pair and adjacent A:T pair or central A:T pairs first form prior to rapid formation of the complete duplex.
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
 Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
@@
 
-DNA duplex hybridization thermodynamics are most commonly predicted using nearest-neighbor (NN) models where the stability of a given base pair is assumed to solely depend on the identity and orientation of its adjacent base pairs.
-We apply the NN model reported by SantaLucia\citep{SantaLucia1998AThermodynamics}  to estimate the stability of fully intact, shifted, and frayed duplex configurations observed in our Markov State Model.
+DNA duplex hybridization thermodynamics are commonly predicted using nearest-neighbor (NN) models wherein the stability of a given base pair is assumed to solely depend on the identity and orientation of its adjacent base pairs.
+We apply the popular NN model reported by SantaLucia\citep{SantaLucia1998AThermodynamics} to estimate the stability of fully intact, shifted, and frayed duplex configurations observed in our MSMs.
 
@@
 \end{equation}                          
-$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TM}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
+$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TheMotifs}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
 A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig6}.
@@
 
-\bibliography{references_210505}
+\bibliography{references_210514_abv}
 

### 2021-05-19 manuscript.tex
Andy pre-submission edits

@@
 %\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
-\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
 %\newcommand*{\noter}[1]{}					% notes off
@@
 A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped.
-In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
+In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical modes, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
 For a repetitive ATATATATAT sequence, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations.
 Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways.
-Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical two-state model of hybridization/dehybridization over a smooth landscape with no intermediate metastable states.
-Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.
+Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical ``all-or-nothing'' two-state model of hybridization/dehybridization with no intermediate metastable states.
+Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation for DNA nanotechnology applications.
 
@@
 From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing.
-\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnological applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the \textit{dynamical}, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves.
-\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence.
+\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnology applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the dynamical, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but these technologies have outpaced our fundamental understanding of the dynamics themselves.
+\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena over picosecond to millisecond time scales.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence.
 \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence, Shah2019}.
 A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved.
-For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the extent to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
+For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
 The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.
 
-Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization.
+Our understanding of hybridization dynamics is built upon decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated analyses under equilibrium conditions, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization.
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
 
-In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
+In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning of the slow dynamical modes, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
 This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
-\citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new and existing experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
-Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
+\citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
+Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of out-of-register shifted states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
 Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states.
-Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states that facilitate the hybridization/dehybridization process.
-Analysis of the configurational ensemble within each metastable state reveals the conformational heterogeneity and dynamical interconversions proceeding on time scales that lie below those resolvable within the long-time kinetic model.
+Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states.
 Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to sculpt and qualitatively alter the dynamical behaviors of the system.
@@
 \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation.
-\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSimulations, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.
+\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomena including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSimulations, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in good agreement with observed experimental trends.
 
@@
 \citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
-Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s.
+Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of 9.94 $\times$ 10$^{-11}$ m$^2$/s.
 \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius.
@@
 Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
-Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The temperature at which the hybridized and dissociated populations were closest was selected.
+Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
 The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step.
 We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state.
-The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions.
+The initial hybridized state was defined based on the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions.
 The magnitude of this displacement did not affect the results so long as all native Watson-Crick (WC) bonds were completely broken.
 Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest.
-Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
+Each simulation was conducted for 26 $\mu$s and frames saved to disc every 100 ps.
+Each simulation required $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
 The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.
@@
 
-Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
+Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ in this work to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation.
 In brief, a MSM extracts from simulation trajectories an ensemble of long-lived metastable macrostates, their equilibrium occupancy populations, and the equilibrium probability fluxes between them.
-In this manner, they provide an interpretable and predictive model of the system thermodynamics and long-time kinetics.
+In this manner, they provide an interpretable and predictive model of the system thermodynamics and kinetics.
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 \textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances.
-We adopt intermolecular pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the 10$\times$10=100-element vector of pairwise distances between all bases within the complementary pair of 10-base strands.
+We adopt intermolecular pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the 10$\times$10=100-element vector of inter-strand pairwise distances.
 One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances.
@@
 \citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top five modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.
-Somewhat surprisingly, we observed that augmenting our set of intermolecular distances with intramolecular distances led to no improvement of the VAMP-2 score, indicating that the kinetically-relevant conformational state of the two strands are adequately represented via the intermolecular distances and leading us to employ only intermolecular distances within our featurization.
+Somewhat surprisingly, we observed that augmenting our set of intermolecular distances between bases with intramolecular distances between bases on the same strand led to no improvement of the VAMP-2 score.
+This indicates that the kinetically-relevant conformational state of the two strands are adequately represented via the intermolecular distances and leading us to employ only intermolecular distances within our featurization.
 
@@
 The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA.
-\citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky2019High-ResolutionVAMPnets} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence.
+\citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA.\citep{Chen, Sidky2019High-ResolutionVAMPnets} SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence.
 SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000.
 We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics.
-\citep{Phys2011MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
-\citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs.
+\citep{Phys2011MarkovValidation} As observed by Husic and Pande, the lag time cannot be treated as a hyperparameter to be optimized via the VAMP-2 score, but must be selected as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest.
+\citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to high-resolution Markovian macrostate MSMs.
 We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
-We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
-A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
-
-\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering.
+We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
+A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
+
+\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms molecular simulation trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering.
 The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.
 
-\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM.
+\textbf{Macrostate clustering.} The 200 microstates comprising each system were coarsened into our terminal macrostate MSM.
 The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionSoftware} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
@@
 The identification of $(i-1)$ implied time scales implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates.
-We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mix systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates.
+We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mix systems, respectively, by projecting the molecular simulation trajectories into these discrete macrostates.
 These macrostate MSMs constitute our terminal kinetic models.
@@
 Oligonucleotides were prepared at a total strand concentration of 2 mM in 50 mM pD 7.2 sodium phosphate buffer with 240 mM NaCl and 18 mM \ce{MgCl2}.
-Prior to each measurement, DNA solutions were placed in a water bath at 90$^{\circ}$ C for 3 minutes and then cooled to room temperature under ambient conditions.
+Prior to each measurement, DNA solutions were placed in a water bath at 90$^{\circ}$ C for three minutes and then cooled to room temperature under ambient conditions.
 
@@
 
-To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
+To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates $\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$ were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
 	\begin{equation}\label{e1}
@@
 
-FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
+FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex $\theta$ as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
 	\begin{equation}\label{e3a}
@@
 We apply these sequence-independent scaling factors to our reported computational values.
-Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction in making the experimental comparison and so we apply a universal (+4) K corrective calibration to our computational results.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results.
 We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but note that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
@@
 
-In Fig.~\ref{fig:relaxation-comparison}a we report an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves.
+In Fig.~\ref{fig:relaxation-comparison}a we observe an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves.
 The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone.
-Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general trends exponential trend is preserved.
+Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general exponential trend is preserved.
 This may be a result of its high propensity to fray (cf.\ Section \ref{sec:GC-core}).
 In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end.
-This can be understood via the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences.
+These trends can be understood in light of the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences.
 Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements.
@@
 We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences.
-Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D).
+Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which four terminal A:T base pairs are unbound, and the fully dissociated state (D).
 We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
@@
 The nearest neighbor (NN) model is a popular empirical model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
-We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences.
+We apply the NN model to predict the free energy $F^\mathrm{NN}$ of each of the macrostates occupied by each of the four sequences.
+A full accounting of our application of the NN model is provided in the \blauw{Supporting Information}.
 The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
-Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare free energy differences predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model.
-For clarity of presentation, we specify the additive constant $C$ in both the MSM and NN models by arbitrarily assigning the free energy of the hybridized state H to define the zero of free energy.
-A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the \blauw{Supporting Information}.
-
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM \noter{and therefore predicts a higher metastable state populations.} The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement.
-Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the 5S2, 3S2, 5S4, and 3S4 states?
+We use this relationship to convert the equilibrium occupancy probabilities predicted by our MSM and reported in Fig.~\ref{fig:allseq_table}b into free energies $F^\mathrm{MSM}$.
+The unknown additive constants preclude comparisons of absolute free energies between the MSM and NN model, but it is legitimate to compare relative free energies between pairs of macrostates since the additive constant cancels in taking differences.
+As such, we arbitrarily set the additive constant $C$ in both the MSM and NN model such that the hybridized state H defines a zero free energy reference state and we report the stability of all macrostates relative to the hybridized state as $\Delta F^\mathrm{NN} = F^\mathrm{NN} - F_H^\mathrm{NN}$ for the NN model and $\Delta F^\mathrm{MSM} = F^\mathrm{MSM} - F_H^\mathrm{MSM}$ for the MSM.
+
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher relative free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 compared to the NN model, such that the MSM predicts lower equilibrium occupancies of these states.
+The NN and MSM predictions for the frayed state F4 are in good agreement.
+Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the shifted states?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
@@
 
-We can further explore the role of inert tails upon macrostate stability by analyzing the AT-all and GC-end sequences that occupy out-of-register shifted macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (Fig.~\ref{fig:allseq_table}).
-For both sequences, our simulations show that 5S2 is both the most stable of shifted states relative to H (Fig.~\ref{fig:NN_table}a) and has the smallest discrepancy (4.0 kJ/mol AT-all, 10.0 kJ/mol GC-end) compared to NN predictions (Fig.~\ref{fig:NN_table}b).
-The 3S2 and and 5S4 states have nearly the same deviation from NN predictions (7.0-7.2 kJ/mol AT-all, 14.3-14.7 kJ/mol GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
+\begin{figure}[ht!]
+	\begin{center} 
+        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
+(b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model.
+The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
+        \label{fig:NN_table}
+	\end{center}
+\end{figure}
+
+We can further explore the role of inert tails upon macrostate stability by analyzing the AT-all and GC-end sequences that occupy out-of-register shifted macrostates 5S2 and 3S2 comprising a one-base inert tail and 5S4 and 3S4 comprising a three-base inert tail (cf.\ Fig.~\ref{fig:allseq_table}a).
+For both sequences, our simulations show that 5S2 is both the most stable of shifted states relative to H (Fig.~\ref{fig:NN_table}a) and has the smallest discrepancy ($\sim$4 kJ/mol AT-all, $\sim$10 kJ/mol GC-end) compared to NN predictions (Fig.~\ref{fig:NN_table}b).
+The 3S2 and and 5S4 states have nearly the same deviation from NN predictions ($\sim$7 kJ/mol AT-all, $\sim$15 kJ/mol GC-end), indicating that longer tails in the 5$^\prime$ direction are as destabilizing as shorter tails in the 3$^\prime$ direction.
 It is known from differential scanning calorimetry studies~\citep{Dickman2012ThermodynamicDNAs} that 3$^\prime$ inert tails are more destabilizing than 5$^\prime$ tails, with the differential behavior attributed to a combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure.
@@
 Both the NN and MSM predictions for AT-all are consistent with this trend (i.e., $F_\mathrm{3S2} > F_\mathrm{5S2}$ and $F_\mathrm{3S4} > F_\mathrm{5S4}$).
-For GC-end, the MSM and NN models both predict the four shifted states to be less stable relative to the hybridized state than the corresponding predictions for AT-all.
+For GC-end, the MSM and NN models both predict the out-of-register shifted states to be less stable relative to the hybridized state than the corresponding predictions for AT-all.
 This is in line with expectations since the terminal G:C pairs in GC-end decrease by two the number of available WC pairings in out-of-register shifted states compared to AT-all.
@@
 In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails.
-In these instances the NN model overestimates \noter{another mention of ``overestimate'' but perhaps we have qualified it more at this point in the text?} macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol.
+In these instances the MSM predicts these macrostates to be less stable relative to the hybridized state compared to the NN model predictions by 4.0-14.5 kJ/mol.
 The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.
@@
 This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.
-
-\begin{figure}[ht!]
-	\begin{center} 
-        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero.
-(b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model.
-With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.}
-        \label{fig:NN_table}
-	\end{center}
-\end{figure}
 
@@
 A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
-As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.
+As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal.
 
@@
 
-GC-end comprises the next most complex MSM with the introduction of the terminal G:C pairs maximally preserving the repetitive tract of A:T base pairings.
-The GC-end MSM possesses all of the same macrostates in its dynamical landscape with the exception of the 3S4 state (Fig.~\ref{fig:allseq_table}c).
+GC-end comprises the next most complex MSM.
+The introduction of the G:C pairs at the termini of the strands maximally preserves the repetitive tract of A:T base pairings such that the GC-end MSM possesses all of the same macrostates in its dynamical landscape with the exception of the 3S4 state (Fig.~\ref{fig:allseq_table}c).
 As discussed in Section \ref{sec:NN}, the 3S4 state is rendered unstable within the lag time of our MSM due to the presence of the destabilizing 3$^\prime$ inert tail and only four WC base pairings compared to six in the case of AT-all.
@@
 However, the D $\rightarrow$ 5S2 and D $\rightarrow$ 3S2 transition probabilities are half or less of those in AT-all, and the reverse transitions are an order of magnitude larger.
-This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that comprise only six WC pairs compared to eight in AT-all (Fig.~\ref{fig:NN_table}).
+This may be attributed to the reduced thermodynamic stability of the 5S2 and 3S2 states in GC-end that comprise only six WC pairs compared to eight in AT-all (cf.\ Fig.~\ref{fig:NN_table}).
 The 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H transition probabilities are more than an order of magnitude larger than in AT-all, which may again be attributed to the lower thermodynamic stability of the two shifted states relative to the hybridized state H.
@@
 
-The out-of-register kinetic landscape that define AT-all and GC-end hybridization have been explored by a number of previous computational studies.
+The out-of-register kinetic landscape that defines AT-all and GC-end hybridization have been explored by a number of previous computational studies.
 Simulations have identified internal displacement mechanisms capable of correcting base pair alignment in 3SPN.2 \citep{Hinckley2014Coarse-grainedEffects} as well as in the coarse-grained oxDNA \citep{Romano2013DNADependence} and BioModi \citep{Markegard2015} models.
@@
 Xiao et al.\ performed an all-atom energy landscape-based analysis of 5$^\prime$-GGGGGG-3$^\prime$ and 5$^\prime$-GCGCGC-3$^\prime$ hexamers.~\citep{Xiao2019} Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
-(In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.) This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events.
+In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.
+This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events.
 It is conceivable that the stronger hydrogen bonding in G:C WC pairs relative to A:T pairs may render out-of-register shifted states less favorable to hybridization by suppressing fluctuation-driven rearrangements,~\citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} but additional studies would be required to reconcile these observations.
@@
 The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free.
-(As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to inter-convert faster than the lag time and are not metastable within our MSM.
+(As expected by symmetry, the particular AT run that is free occurs with equal probability on either side of the core.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to inter-convert faster than the lag time and are not registered as metastable within our MSM.
 Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
@@
 Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a).
-Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s, which are again approximately equal.
+Our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s, which are again approximately equal.
 
@@
 To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events.
-We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more of each in \blauw{Fig.~S5}.
-During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another, but uniformly observe the formation of some key native WC base pair contacts immediately prior to duplex formation.
-Specifically, formation of one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
+We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more in \blauw{Fig.~S5}.
+During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another and the formation of key native WC contacts immediately prior to duplex formation: specifically, one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
 This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
@@
 Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
-Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the complete duplex.
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
 Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.
@@
 Finally, we sought to validate the predictions of our sequence-dependent MSMs against experimental T-jump IR spectroscopy.
-T-jump IR measurements commence from a low temperatures, apply a step jump in temperature, and track the relaxation of the system to the dehybridized state (Section \ref{sec:Tjump}).
+T-jump IR measurements commence from a low temperatures, apply a step jump in temperature, and track the relaxation of the system to the dehybridized state.
 We hypothesized that the influence of the out-of-register shifted states present in the AT-all and GC-end sequences upon the system relaxation kinetics should be manifest in the slow and/or fast responses measured by T-jump IR.
-As discussed in Section \ref{sec:time}, the slow IR response corresponds to dissociation and the fast to terminal base fraying.
+As discussed in Section \ref{sec:time}, the slow IR response is largely attributed to dissociation events and the fast to terminal base fraying.
 With regards to the slow response, our MFPT analyses of our MSMs predict out-of-register shifting events (i.e., H $\rightarrow$ 3S2, 5S2, 3S4, 5S4) to proceed on microsecond time scales, which are commensurate with the 1.4 - 2.9 $\mu$s time scales for dehybridization (i.e., H $\rightarrow$ D) for each of the four sequences.
@@
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated and slightly faster fraying response over the course of the temperature relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
 Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
@@
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
-Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta_\mathrm{fast}})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
 The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ to the heterogeneity of fraying dynamics at the fast timescale.
-A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter (i.e., a more stretched response) relative to those for GC-end, GC-core, and GC-mix.
-This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
+The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta_\mathrm{fast}$ to the heterogeneity of fraying dynamics at the fast timescale.
+A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta_\mathrm{fast}$ parameter (i.e., a more stretched response) relative to those for GC-end, GC-core, and GC-mix.
+This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{fast}^\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{fast}^\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{fast}^\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{fast}^\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{fast}^\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{fast}^\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{fast}^\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{fast}^\mathrm{GC-mix}$ = 0.6.
 This result supports our hypothesis and validates a testable prediction of our sequence-dependent MSMs.
@@
 The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
-Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
+Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta_\mathrm{fast}})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
 The stretched exponential describes the process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The stretch factor $\beta$ for the fits at 1600 cm$^{-1}$ and 1660 cm$^{-1}$ are reported directly on the plots in panels c and d.
+The stretch factor $\beta_\mathrm{fast}$ for the fits at 1600 cm$^{-1}$ and 1660 cm$^{-1}$ are reported directly on the plots in panels c and d.
         }
@@
 Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core.
-In this case, no metastable out-of-register shifted states are observed in our models with the hybridization and dehybridization pathways all passing through a strongly metastable frayed state in which one or other of the four-base AT-tracts is unbound resulting in two free AT-tails.
-Finally, placing the G:C bases between the center and end of the strand to maximally disrupt the repetitive AT tracts results in no metastable out-of-register or frayed states and results in simple two-state ``all-or-nothing'' hybridization/dehybridization behavior.
+In this case, no metastable out-of-register shifted states are registered by our model with the hybridization and dehybridization pathways all passing through a strongly metastable frayed state in which one or other of the four-base AT-tracts is unbound to produce two free AT-tails.
+Finally, placing the G:C bases between the center and end of the strand to maximally disrupt the repetitive AT tracts results in no metastable out-of-register or frayed states for GC-mix and results in simple two-state ``all-or-nothing'' hybridization/dehybridization behavior.
 Analysis of the specific pathways reveals hybridization to largely proceed by a nucleation-zippering mechanism and dehybridization to proceed by a fraying-peeling mechanism.
@@
 A.T.
-thanks the National Institute of General Medical Sciences of the National Institutes of Health (Award No.
-R01-GM118774) for support of this research.
+thanks the National Institute of General Medical Sciences of the National Institutes of Health (Award No.\ R01-GM118774) for support of this research.
 B.A.

### 2021-05-20 cover_letter/letterhead.tex
Andy pre-submission edits
final submission edits

@@
 \usepackage[sort&compress,numbers,super]{natbib}
-\usepackage{caption}
+%\usepackage{caption}
 \usepackage{newfloat}
@@
             citecolor=blue]{hyperref}
-\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
+
 % Create a new command for the horizontal rule in the document which allows thickness specification
@@
 %more new commands and definitions
-\newcommand*{\rood}[1]{\textcolor{red}{#1}}
-\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
-\newcommand*{\groen}[1]{\textcolor{green}{#1}}
+\newcommand*{\rood}[1]{\color{red}{#1}}
+\newcommand*{\blauw}[1]{\color{blue}{#1}}
+\newcommand*{\groen}[1]{\color{green}{#1}}
+%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
+%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
+%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
 \newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
 \newcommand{\scripty}[1]{\ensuremath{\mathcalligra{#1}}}
+
+% code for custom comments
+\usepackage{tikz}
+\usepackage{xcolor}
+% markers (black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, olive, orange, pink, purple, red, teal, violet, white, yellow, more: https://en.wikibooks.org/wiki/LaTeX/Colors)
+\newcommand{\marker}[3]{
+  \tikz[baseline=(X.base)]{
+    \node [fill=#1!40,rounded corners] (X) {#2:};
+  }
+  {\color{#1!80!black}#3}
+}
+\newcommand{\khb}[1]{\marker{teal}{KHB}{#1}}  % Kathy Breen
+\newcommand{\replyhead}{\textbf{Author Reply: }}
 
@@
 \def\TEL{p:~(773)~702-5950} % Your phone number
-\def\URL{w:~\url{http://andrewferguson.uchicago.edu}} % Your URL
-\def\titleone{Associate Professor}
+\def\URL{w:~\url{www.ferglab.com}} % Your URL
+\def\titleone{Associate Professor\\ Deputy Dean for Equity, Diversity, and Inclusion}
 
@@
 \vskip 0.0in
-\includegraphics[width=0.25\textwidth]{signature_720}
+\includegraphics[width=0.2\textwidth]{signature_720}
 %\vskip 0.65in % Whitespace between the signature text and your name
@@
 \institution\\
-\vskip -0.1in
-\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
+%\vskip -0.1in
+%\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
 }
@@
 \textbf{Key Results.}  In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy.
-We conducted more than 1 ms of coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+We conducted more than 1 ms of coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence-dependent dynamics.
 Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
@@
 % fit for the journal
-\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of sequence-dependent DNA hybridization/dehybridization pathways.
+\textbf{Scope.}  This work fits within the scope of \textit{The Journal of the American Chemical Society} in that it reports new fundamental physical chemical understanding of the molecular pathways and mechanisms of sequence-dependent DNA hybridization/dehybridization.
 These sequence-dependent mechanisms are of both fundamental interest in biophysical and biochemical processes and of increasing importance in burgeoning DNA nanotechnology applications reliant on quantitative understanding of the sequence-dependent kinetic pathways and rates.
@@
 
-We look forward to hearing from you.
-
 \closing{Yours Sincerely}

### 2021-05-20 manuscript.tex
final submission edits

@@
 %\newcommand*{\noter}[1]{}					% notes off
-%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
+\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
 
@@
 In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning of the slow dynamical modes, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
-This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
+This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair \noteb{``of two G:C base pairs?''}within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
 \citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
@@
 \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation.
-\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomena including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSimulations, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in good agreement with observed experimental trends.
+\citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomena including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSimulations, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxations are, within a corrective scaling factor, in good agreement with observed experimental trends.
 
@@
 In this manner, they provide an interpretable and predictive model of the system thermodynamics and kinetics.
-MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation.
+MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
 A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
@@
 It is well-known that the smoothing of the underlying free energy landscape induced by coarse-graining artificially accelerates the kinetics of coarse-grained molecular simulations and that different degrees of freedom may be accelerated by different factors.~\citep{Marrink2007TheSimulations, Fritz2011MultiscaleDynamics, Marrink2013PerspectiveModel} We find that the simulated slow responses corresponding to center-of-mass translation of the strands during dissociation of the duplex is $\sim$10$\times$ accelerated relative to experiment, whereas the fast responses corresponding to fraying of the terminal bases is $\sim$120$\times$ accelerated.
-We apply these sequence-independent scaling factors to our reported computational values.
+We apply these sequence-independent scaling factors to our reported computational values \noteb{I know we clarify this at the end of the paragraph, but maybe add ``our reported computational values in Fig.~\ref{fig:relaxation-comparison}'' to be clear?}.
 Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results.
-We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but note that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
+We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
 The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.
@@
 In Fig.~\ref{fig:relaxation-comparison}a we observe an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves.
-The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone.
+The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone \noteb{This effect is now minimized by the fact that we are using an experimental dissociation rates instead of the pure slow response (previous data showed worse agreement at low temp).
+Still, there could be some hybridization contribution so we could leave this sentence in}.
 Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general exponential trend is preserved.
@@
 The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model.
-The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways.
+The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hybridization/dehybridization pathways.
 Immediately, we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas at the other end of the spectrum GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior.
@@
 The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a.
-The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b.
+The radii of the circles is proportional to the logarithm of the equilibrium occupancy populations reported in panel b.
 Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate.
@@
 
-We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association.
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models \noteb{does ``model'' imply we're looking at more than NN?} for the thermodynamics of DNA association.
 The nearest neighbor (NN) model is a popular empirical model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
@@
 The NN and MSM predictions for the frayed state F4 are in good agreement.
-Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the shifted states?
+Although the \noteb{shifted state} trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the shifted states?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
@@
 Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
-The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability.
+The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure this is correct?
+F4 $\rightarrow$ H is 50\% whereas the previous two are much lower.
+Maybe remove this sentence unless it is needed to support what comes next?}.
 We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
@@
 
-The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
 The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
-We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
+We note that we do observe substantial transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
 This stands in contrast to the metastable F4 state populated by GC-core.
@@
 Finally, we sought to validate the predictions of our sequence-dependent MSMs against experimental T-jump IR spectroscopy.
-T-jump IR measurements commence from a low temperatures, apply a step jump in temperature, and track the relaxation of the system to the dehybridized state.
+T-jump IR measurements commence from a low temperature, apply a step jump in temperature, and track the relaxation of the system to the dehybridized state.
 We hypothesized that the influence of the out-of-register shifted states present in the AT-all and GC-end sequences upon the system relaxation kinetics should be manifest in the slow and/or fast responses measured by T-jump IR.
@@
 Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
+Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium \noteb{this is 24\% of any partially hybridized state.
+Can just say ``hybridized ensemble''} ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
 It is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
@@
 Introduction of the G:C pairs at the end of the strand maintains an eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all.
-Destabilization of the GC-end shifted states relative to AT-all, however, results in a far more limited participation of these states with only 7\% of GC-end hybridization and dehybridization events passing through one or more shifted states.
+Destabilization of the GC-end shifted states relative to AT-all, however, results in a far more limited participation of these states with only 7\% of GC-end \noteb{a more general note, but I think a lot of people would consider any kind of stable duplex as hybridized, perhaps it would be worth stating near the beginning of our ananlysis that we are referring by default to fully in-register duplexes?} hybridization and dehybridization events passing through one or more shifted states.
 Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core.

### 2021-05-20 SI.tex
final submission edits
Andy final edits prior to submission

@@
 	\centering
-    \includegraphics[width=0.8\textwidth]{FigS2.pdf}
+    \includegraphics[width=0.75\textwidth]{FigS2.pdf}
     \caption{Convergence of the MSM implied time scales $t_i$ as a function of lag time $\tau$.
@@
 	\centering
-    \includegraphics[width=.75\textwidth]{FigS4.pdf}
+    \includegraphics[width=.70\textwidth]{FigS4.pdf}
     \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
 From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time.
-We define a Watson-Crick base pair to be intact if the two comlementary bases lie within a linear distance of 1.3 nm.
+We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm.
 We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
@@
 We use the empirical salt correction developed by Owczarzy\citep{Owczarzy2008PredictingCations} to determine the NN parameters at the sodium concentration used for simulations in this work.
-The Owczarzy model assumes the effect of counter ions on DNA hybridization thermodynamics to be purely entropic,
+The Owczarzy model assumes the effect of counter-ions on DNA hybridization thermodynamics to be purely entropic,
 \begin{equation}\label{SIe4}
@@
 \end{equation} 
-where $f_{GC}$ is the fractional G:C content of the duplex and $c_{Na^+}$ is the concentration of sodium counter ions measured in M.
+where $f_{GC}$ is the fractional G:C content of the duplex and $c_{Na^+}$ is the concentration of sodium counter-ions measured in M.
 

### 2021-05-20 manuscript.tex
Andy final edits prior to submission

@@
 %\newcommand*{\noter}[1]{}					% notes off
-\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
+%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
 
@@
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
@@
 In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning of the slow dynamical modes, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
-This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair \noteb{``of two G:C base pairs?''}within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
+This family -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of two G:C base pairs within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
 \citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
@@
 Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states.
-Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to sculpt and qualitatively alter the dynamical behaviors of the system.
+Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to control the dynamical behaviors of the system.
 We anticipate that this new foundational understanding, and the extension of our approach to more extensive families of DNA sequences, can guide the rational design of DNA oligomers with tailored kinetic properties engineered for DNA nanotechnology applications such as DNA-PAINT.
 \citep{Shah2019, Strauss2020UpDNA-PAINT}
-
-% example text from Andy's conclusions:
-
-% Correlation of the data-driven kinetic order parameters extracted by SRVs provides new understanding of the sequence-dependent kinetic mechanisms by which oligomers hybridize and reveal these to be richer than classical base-pair shifting and “all-or-nothing” behaviors.
-
-% original last paragraph
-%By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process.
-Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible.
-Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs.
-Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes.
-Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.
-
-%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}.
-
-%Still, rich dynamics have been observed short nucleotides \citep{Sanstead2018DirectDehybridization} and the sequence-dependent effects on these dynamics are not well understood.
-
-% include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior.
-The slow response was attributed to complete strand dissociation, while the fast response described base fraying.
-We perform longer equilibrium sampling near Tm to get a sense of sequence equilibrium behavior, as well shorter simulations at a range of higher temperatures to replicate T-jump data.
-Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative time scales.
 
@@
 
-%\subsection{Featurization}
-
-%A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries.
-Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other.
-To account for this, we averaged permutable distances (45 pairs in total) together into a ``symmetrized'' feature set following a similar procedure used in tICAgg coordinate construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\cite{ Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information.
-By comparing scores across several candidate feature sets (Figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure.
-Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.
-
-%\begin{figure}[ht!]
-%	\begin{center}
-%        \includegraphics[width=0.55\textwidth]{Fig1.pdf}
-%        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates.
-Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods.
-This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
-%        \label{fig:allseq_features_vamp2}
-%	\end{center}
-%\end{figure}
-
-%\subsection{\label{sec:methods}SRVs}
-
-%The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed.
-While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA.
-SRVs were first developed by Chen et al.
-as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and time scales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches.
-This is a key attribute to our system as 10 million frames with 55 features per frame are used for each sequence.
-The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}
-
-%\subsection{\label{sec:methods}SRV-MSMs}
-
-% need some more detail here -- talk to Andy about this
-%MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Prinz2011OptimalDynamics, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time.
-The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted.
-To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible.\citep{Pande2010EverythingAsk} Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis.
-To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence.\citep{Scherer2015PyEMMAModels} In a similar approach to Sidky et al.
-we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied time scales convergence.
-The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.
-
-%\subsection{\label{sec:methods}Diffusion Maps}
-%Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time.
-We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies.\citep{Shmilovich, Wang2017MesoscaleLandscapes} We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.
-
-%%%%%%%% Methods comments %%%%%%%%%%%%%%%%%
-
-%We found close agreement with explicit ion simulations performed with 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} (supplemental) but 7x slower simulation speed made adequate sampling more difficult to achieve under these conditions.
-
-%We initialized explicit ions such that 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}.
-We used a periodic box size of 7.774 nm and an effective oligo concentration of 7 mM.
-We used an Ewald summation to calculate long range Coulombic interaction between DNA and ions using a real space cutoff of 2 nm.
-
 \subsection{Experimental Methods}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{\label{sec:Results}Results and Discussion}
@@
 \subsection{Sequence-dependent coarse-grained kinetics recapitulate T-jump IR measurements} \label{sec:time}
-
-% rerun these at the actual final temps and just see what happens? Doesn't really matter where Tjump occurs and could make things easier as a whole...
-% talk about rerunning AT-all and GC-mix=
 
@@
 To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation.
-We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
+We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
 First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
@@
 It is well-known that the smoothing of the underlying free energy landscape induced by coarse-graining artificially accelerates the kinetics of coarse-grained molecular simulations and that different degrees of freedom may be accelerated by different factors.~\citep{Marrink2007TheSimulations, Fritz2011MultiscaleDynamics, Marrink2013PerspectiveModel} We find that the simulated slow responses corresponding to center-of-mass translation of the strands during dissociation of the duplex is $\sim$10$\times$ accelerated relative to experiment, whereas the fast responses corresponding to fraying of the terminal bases is $\sim$120$\times$ accelerated.
-We apply these sequence-independent scaling factors to our reported computational values \noteb{I know we clarify this at the end of the paragraph, but maybe add ``our reported computational values in Fig.~\ref{fig:relaxation-comparison}'' to be clear?}.
+We apply these sequence-independent scaling factors to our reported computational values in Fig.~\ref{fig:relaxation-comparison}.
 Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results.
@@
 In Fig.~\ref{fig:relaxation-comparison}a we observe an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves.
-The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone \noteb{This effect is now minimized by the fact that we are using an experimental dissociation rates instead of the pure slow response (previous data showed worse agreement at low temp).
-Still, there could be some hybridization contribution so we could leave this sentence in}.
 Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general exponential trend is preserved.
@@
 The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.
-
-%Coarse-grained time scales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom.
-As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.
-Experimental temperature-dependent relaxation time scales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded ``fast'' and ``slow'' amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence.
-To compare results with 3SPN.2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence.
-We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.
-
-% need to discuss tmelt shift
-%We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff.
-We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4$^{\circ}$C between experiment and simulation fits.
-While 3SPN.2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects.
-We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature.
-After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (Figure \ref{fig:relaxation-comparison})).
-At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone.
-GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.
-
-% GC-core fast time scales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
-% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor.
-We maintained a consistent accegqrelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses.
-It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.
-
-% need to update this fig with labels and temp shift
-
-% A documented short-coming for the 3SPN.2 model is a high proclivity for termini fraying relative to what has been observed in vitro \citep{2013}.
-This supports why we observe significantly more acceleration for the fast response, but relative sequence and temperature effects still compare well with T-jump results.
-
-%The fast response -- which Sanstead et.
-al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables.
-Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with time scales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3SPN.2 base, we would not expect to resolve this diverse collection of states and dynamics.
-Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff.
-This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal.
-This is a reasonable assumption given that the amplitude-weighted time scales should consist largely of terminal fraying events.
-Again we fit relaxation curves to the ensemble of fraying time scales in order to extract rate approximations for each sequence at a series of temperatures.
-
-%For A:T terminal sequences, the simulated fast response appears linear with temperature, indicating a barrierless and diffusion-driven process.
-In contrast, the GC-end responses are distinctly slower and increase exponentially with temperature, likely due to a greater enthalpic barrier associated with G:C fraying.
-We observe similar trends in the experimental data, although comparisons at the highest temperatures were limited due to mixing between the fast and slow responses (Figure \ref{fig:relaxation-comparison}).
-We found the optimal acceleration factor to be dependent on the choice of the cutoff parameter, but we can approximate that 3SPN.2 fraying dynamics are accelerated by about 100-200x relative to experiment.
-It is not surprising that we see different rates of acceleration for the dissociation and fraying processes given that coarse-grained effects can vary across different degrees of freedom.
-In particular, the simplified treatment of the fraying process may smoothen the free energy landscape and speed up dynamics relative to a more global processes like dissociation.
-Although one should rely on higher resolution models to study in depth mechanisms of fraying, our results indicate that terminal fraying is a reasonable assignment for the spectroscopic fast response in Sanstead et al.
-and that 3SPN.2 can recapitulate some sequence-dependent fraying effects.
-% need a better transition here back into SRV-MSMs?
 
@@
 We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.
-
-%We followed the SRV-MSM analysis pipeline described in the Methods section to generate sequence-dependent kinetic models for each oligomer.
-We identified seven kinetically relevant states that were captured within the resolution of the model.
-Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities.
-Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models.
-Arrows between states show the probability of a transition to another state within the MSM lag time.
-
-%Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5$^\prime$ vs.
-3$^\prime$) and the number of shifted motifs relative the native state (1 vs.
-2).
-GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states.
-Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}.
-We also show that PCCA+ clustering assigns states to relative free energy minima in the tICA space.
-
-% rewrite this into one paragraph
-
-% change state names and scientific notation
-% include D here? can see what it would look like...
 
@@
 
-We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models \noteb{does ``model'' imply we're looking at more than NN?} for the thermodynamics of DNA association.
-The nearest neighbor (NN) model is a popular empirical model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
+The NN model predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
 We apply the NN model to predict the free energy $F^\mathrm{NN}$ of each of the macrostates occupied by each of the four sequences.
@@
 
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher relative free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 compared to the NN model, such that the MSM predicts lower equilibrium occupancies of these states.
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 relative to the H state compared to the NN model.
 The NN and MSM predictions for the frayed state F4 are in good agreement.
-Although the \noteb{shifted state} trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the shifted states?
+Although the shifted state trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
@@
 Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{DiMichele2014EffectHybridization}
-
-%Visualizations show that, although free A:T ends in the F4 state could be considered inert tails, they tend to extend away from the intact section of the duplex and therefore may have less of an impact on duplex stability.
-In this case the NN and MSM free energy predictions are in good agreement.
 
@@
 
-%Given that terminal mismatches tend to be slightly stabilizing \citep{Santalucia2004TM}, we might expect a lower discrepancy than the 10.1-14.7 kJ/mol offset that we observe between NN and MSM results.
-This offset may arise in part from the excluded volume interactions used to account for mismatched bases in 3SPN.2, however we attribute it mainly to inert tail effects.
-
 In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails.
@@
 The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.
-
-%\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well.
-It is important to note that we do not include dangling end contributions when calculating NN free energies.
-Although 3SPN.2 can capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the CG model was not parameterized to fully capture the enthalpic contributions of free bases interaction with terminal base pairs.
-For each macrostate, dangling ends would be expected to have stabilizing effects (Can show NN with dangling end contributions in SI and therefore increase state populations compared to what we observe in simulation.)
-
-%\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex.
-We find that inert tail effects may be particularly impact in predicting out-of-register state stability.}
-
-%\bood{For the AT-all sequence, we see close agreement between the macrostate and NN prediction for the 5S2 state.
-However, we observe deviations of 3.5 and 2.3 kJ/mol the 3S2 and 5S4 states, respectively.
-This is consistent with UV absorbance spectroscopy and differential scanning calorimetry experiments~\citep{Dickman2012ThermodynamicDNAs, Michele2014EHybridization} which show that (i) 3$^\prime$ tails are more destabilizing than 5$^\prime$ tails and (ii) longer tails are more destabilizing than shorter tails.
-These effects are compounded for the 3S4 state, where we observe an MSM free energy 5.1 kJ/mol higher than the NN model prediction.}
-
-%\bood{For GC-end, the MSM and NN models both predict the four shifted states to be less stable than for AT-all.
-This is in line with expectations since the terminal G:C pairs in GC-end decrease the number of available WC pairings by two compared to AT-all.
-The deviations between MSM and NN models follow a similar trend based on inert tail effects, but these are also consistently higher -- ranging from 3.2 to 7.8 kJ/mol.
-Indeed, the MSM model for GC-end does not even register the 3S4 macrostate as a stable metastable conformation for the duplex.
-We attribute this additional destabilization to excluded volume interactions between mismatched terminal pairs in the model, as complex and slightly stabilizing~\citep{Santalucia2004TM} terminal G:A and C:T interactions are not included in 3SPN.2.}
-
-%\bood{Although the NN model tends to under-predict the MSM free energy for AT-all and GC-end, we see that the GC-core F4 state is about 2.5 kJ/mol more stable than predicted by NN.
-This can be attributed to an added probability of an ensemble of ``double-frayed'' state in which 1-2 terminal A:T base pairs are frayed and the configuration is clustered into F4.}
-
-%Can add something here (or in kinetic section) about how, given the MSMs seem not to account for dangling end stability, there would likely be a higher population of each state than we observe in experiment.
-This could (i) increase out-of-register role in the hybridization process my increasing membership in intermediate states or (ii) decrease the likelihood of shifting behavior due to higher barriers between those states.
-
-%\begin{figure}[ht!]
-%	\begin{center} 
-%        \includegraphics[width=160mm, scale=1]{Fig3_no_DE.pdf}
-%        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} The arbitrary additive constant to the free energy is specified such that the free energy of the hybridized macrostate is zero.
-These NN estimates do not include additional thermodynamic stabilization from dangling ends  (a) Free energy estimates relative to the hybridized state for both the simulated macrostates and corresponding NN prediction.
-(b) Free energy difference between the NN prediction and simulated value for each macrostate.
-%        \label{fig:NN_table_no_de}
-%	\end{center}
-%\end{figure}
-
-%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of ``dangling ends'' -- unpaired bases adjacent to the paired duplex -- and ``inert tails'' -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}.
-Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3$^\prime$ inert tails have stronger effects than 5$^\prime$ tails.
-Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.
-
-% and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths.
-It has been shown that 5$^\prime$ dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3$^\prime$ ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}.
-These effects might be attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure \citep{Doktycz1990ThermodynamicATGC}.
-
-% This leads us to believe that 5$^\prime$ vs.
-3$^\prime$ differences -- attributed to some combination of 5$^\prime$ tails preferentially stacking on the core duplex and 3$^\prime$ tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.
 
@@
 The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state.
-A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states.
-Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time.
+A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) and dehybridization trajectories H $\dashrightarrow$ D proceed through one or more out-of-register shifted states.
+Among these out-of-register hybridization pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time.
 A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s.
-As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal.
-
-%\noteb{Can add additional analysis "When considering the mean MFPT between D and any state A (A = 5S2 $\cup$ 3S2 $\cup$ 5S4 $\cup$ 3S4 $\cup$ H), we observe the shorter time scales 1.3 $\mu$s for A $\rightarrow$ D 1.3 $\mu$s for D $\rightarrow$ A.
-These MFPTs reflect the frequency of partial hybridization events and are in close agreement with the leading implied time scale of $t_1$ = 1.5 $\mu$s reported in Fig.~\ref{fig:allseq_table}d" This works for AT-all, but for all sequences the leading timescale tends to be lower by up to 2x than H $\dashrightarrow$ D MFTPs.
-I suspect this is because these leading timescale takes longer to converge, but I'm not sure if we want to draw attention to this and/or report the leading timescale for each sequence since its exact meaning is not always clear)}
-
-%Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g.
-5S2 $\rightarrow$ H is more likely than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation.
-Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation.
-In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.
-
-%Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways.
-A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state.
-The 3S4 and 5S4 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways.
+As expected by the fact that the calculations are performed at the melting temperature, the MFPTs are approximately equal.
 
@@
 
-%\noteb{and in good agreement with the leading implied time scale of $t_1$ = 1.4 $\mu$s reported in Fig.~\ref{fig:allseq_table}d.}
-
-%The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all.
-This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed.
-
-%While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states.
-These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior.
-Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences.
-
 The out-of-register kinetic landscape that defines AT-all and GC-end hybridization have been explored by a number of previous computational studies.
@@
 Xiao et al.\ performed an all-atom energy landscape-based analysis of 5$^\prime$-GGGGGG-3$^\prime$ and 5$^\prime$-GCGCGC-3$^\prime$ hexamers.~\citep{Xiao2019} Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
-In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.
-This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events.
+(In contrast, slithering through out-of-register shifted states and zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.) This stands in contrast to our results for our AT-all (5$^\prime$-ATATATATAT-3$^\prime$) sequence, in which out-of-register states participated in 33\% of productive hybridization events.
 It is conceivable that the stronger hydrogen bonding in G:C WC pairs relative to A:T pairs may render out-of-register shifted states less favorable to hybridization by suppressing fluctuation-driven rearrangements,~\citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} but additional studies would be required to reconcile these observations.
@@
 The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free.
-(As expected by symmetry, the particular AT run that is free occurs with equal probability on either side of the core.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to inter-convert faster than the lag time and are not registered as metastable within our MSM.
+(As expected by symmetry, the particular AT run that is free occurs with equal probability on either side of the core.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to interconvert faster than the lag time and are not registered as metastable within our MSM.
 Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4.
@@
 Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25 times more likely than dissociation F4 $\rightarrow$ D.
-Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
-The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure this is correct?
-F4 $\rightarrow$ H is 50\% whereas the previous two are much lower.
-Maybe remove this sentence unless it is needed to support what comes next?}.
+Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are less likely to proceed to complete dissociation D.
 We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
@@
 
-%\noteb{This MFPT shows the largest deviation from the leading timescale $t_1$ = 1.6 $\mu$s}
-
 Lattice models have previously identified frayed states as putative intermediates in DNA hybridization/dehybridization.\citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2016} Araque et al.\ studied a 5$^\prime$-ATGCGCAT-3$^\prime$ octomer using a lattice model and identified a symmetrically A:T frayed state as a crucial part of the duplex transition path.\citep{Araque2016LatticeCooperativity} We previously studied the four sequences that are the subject of the present work using T-jump IR and 2D IR spectroscopy and identified GC-core as possessing the highest deviation from two-state behavior during dissociation when neglecting out-of-register contributions.\citep{Sanstead2016} This result was interpreted to arise from loss of A:T contacts and fraying around the central G:C core, and this hypothesis was supported by lattice model calculations that predicted the GC-core conformational ensemble to possess substantially more frayed configurations than the other three sequences.~\citep{Phys2019} Follow-up T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} These prior results are consistent with the present findings that expose the GC-core sequence to be the only sequence that occupies the F4 frayed state and therefore the only one possessing a metastable frayed state on time scales exceeding the $\tau$=1.2 ns lag time of our MSMs.
-
-%When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al.
-found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions).\citep{Sanstead2016} As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core.
-While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs.\citep{Phys2019} Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states.\citep{Sanstead2018DirectDehybridization} We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.
-
-%Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar time scale.
-
-%Our diffusion map analysis further shows how the ensemble of 4-bp frayed configurations impede helix formation.
 
@@
 Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s, which are again nearly equal.
-
-%\noteb{Again, these are nearly equal and commensurate with the dominant leading implied time scale of $t_1$ = 2.0 $\mu$s reported in Fig.~\ref{fig:allseq_table}d. }
-
-%Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix.
-Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix.
-Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively.
-Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model.
-We did, however, observe substantial fraying of the two AT termini in the simulation data.
-Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and transient that very few progress to full dissociation compared to the F4 state.
-Furthermore, one or two-base-pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.
 
@@
 Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium \noteb{this is 24\% of any partially hybridized state.
-Can just say ``hybridized ensemble''} ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
-It is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
-% We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
+Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
+It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
 
@@
 This result supports our hypothesis and validates a testable prediction of our sequence-dependent MSMs.
-
-%%% possible additions to qualify what's above
-%Overall, there are several reasons why the T-jump response may not be directly sensitive to shifting.
-In the case of GC-end, the low population of shifted states alone may diminish contributions below a detectable limit.
-For AT-all, the particularly slow timescales associated with the most common shifting process (H $\leftrightarrow$ 5S2) can blend in with the slow dissociation response.
-In both cases, it is unclear how populations of shifted states change with temperature, making these processes difficult to identify with T-jump techniques.
-
-%Although there are no distinct ``middle'' responses at the expected shifting timescales, the response profile can inform the distribution of dynamics contributing to the signal \citep{Hamm2006StretchedFolding, Johnston2006StretchedDecays}.
-Shifted state are either (i) too low in population to detect a discernable Shifted $\rightarrow$ D response, or (ii) have lifetimes too long, (especially for the AT-all 5S2 state) to be discerned from the slow response.
-Most shifting behavior will not occur until similar timescales as full dissociation, given that H-->D transitions are approximately as likely as H-->D.
-Therefore, the only discernable traces would come from shifted state in the initial low temperature ensemble.
-For example, the AT-all 5S2 state, which contains 80\% of the shifted population under equilibrium conditions, is the only metastable state with a predicted lifetime within an order of magnitude of the hybridized state.
-Therefore it is likely that this relaxation response is blending with the slow response, while shorter-lived relaxations are too low in magnitude to detect.
-
-%%% Timescale approach %%%
-%We report in Fig.~\ref{fig:T-jump_stretching}a the predicted life times of each macrostate for each of the four DNA sequences at their respective melting temperatures.
-Only the AT-all sequence contains long-lived partially hybridized states with life times in excess of 50 ns and.
-The longest lived of these is the 5S2 state with a life time of 470 ns .
-GC-end also contains a number of out-of-register shifted states, but their life times are predicted to be an order of magnitude shorter than that of the corresponding states in AT-all.
-This is likely attributable to their reduced stability due to the two fewer available WC pairings for any given out-of-register GC-end state relative to AT-all.
-The sole intermediate F4 frayed state occupied by GC-core has a life time of only 2 ns, and the GC-mix sequence possesses no metastable states at all.
-Based on these predictions from our equilibrium MSMs, we hypothesized that the AT-all sequence possessing long-lived intermediate states with life times of up to 470 ns should be distinguishable from the GC-end, GC-core, and GC-mix sequences predicted to have no metastable states or only very short-lived states with life times less than 30 ns.
-We do not expect to be able to resolve the small differences between GC-end, GC-core, and GC-mix due to the very short-lived nature of these intermediate states.
-
-%We present in Fig.~\ref{fig:T-jump_stretching}b-d our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$.
-The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
-Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.
-The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature.
-The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to.
-A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix.
-This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6.
-This result provides an experimental validation of the kinetic predictions of the MSM models.
-
-%%% orginal %%%
-%observe signatures  of the MSM kinetic landscapes would be reflected in T-jump IR measurements and sought experimental validations for the predictions of our molecularly-detailed kinetic models.
-As discussed in Section \ref{sec:time}, the fast and slow IR responses correspond well to terminal base fraying and dissociation, respectively.
-Although there are no distinct ``middle'' responses at the expected shifting timescales, the response profile can inform the distribution of dynamics contributing to the signal \citep{Hamm2006StretchedFolding, Johnston2006StretchedDecays}.
-In particular, the AT-all fast response exhibits more stretched exponential character ($\beta$ = 0.3) than any of the other three sequences (Figure \ref{fig:T-jump_stretching}).
-We can interpret the stretch as fraying from many configurations, each having a slightly different rate.
-The greater fraying heterogeneity is in agreement with the diverse kinetic landscape of the AT-all MSM.
-For the GC-end response, however, the stretching factor is not appreciably different from the other sequences and comparison of 1660 and 1600 cm$^{-1}$ wavelengths indicate that the fast response is much stronger in the G/T rather than A/T region.
-This further suggests that it corresponds to terminal fraying rather than shifting.
-Overall, there are several reasons why the T-jump response may not be directly sensitive to shifting.
-In the case of GC-end, the low population of shifted states alone may diminish contributions below a detectable limit.
-For AT-all, the particularly slow timescales associated with the most common shifting process (H $\leftrightarrow$ 5S2) can blend in with the slow dissociation response.
-In both cases, it is unclear how populations of shifted states change with temperature, making these processes difficult to identify with T-jump techniques.
-
-%These comparisons also highlight the importance of considering and defining multiple types of deviations from two-state behavior.
-In this work, we define this deviation by the complexity of the MSM network and number of accessible metastable states.
-This definition suggests the following order of kinetic complexity: AT-all $>$ GC-end $>$ GC-core $>$ GC-mix.
-As we have shown, this trend is closely related to sequence repetitiveness, specifically the number of consecutive AT motifs.
-In our previous work, we did not consider out-of-register states as part of the experimental or lattice model analysis, and therefore determined our ordering (GC-core $>$ GC-mix $>$ AT-all $>$ GC-end) in terms of frayed state populations prior to duplex dissociation\citep{Sanstead2016, Sanstead2018DirectDehybridization}.
-Depending on the biological system or nanotechnology application of interest, these forms of deviation may hold varied importance...
-
-%In addition to identifying relaxation time scales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response.
-In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}).
-Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response.
-On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source.
-We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-end and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two.
-Although reproducing these stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all $>$ GC-end $>$ GC-core, GC-mix) and increased stretching in the fast response.
-
-% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.
-
-%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
-
-%We emphasize that this does necessarily reflect shifting relaxation time scales themselves but more likely changes in the fast responses -- e.g.
-fraying out-of-register -- within the collection of metastable states.
 
@@
 
-%\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}
-
-%Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states.
-These configurations interchange significantly faster than the SRV-MSM lag time, but we can use a diffusion map approach to glean a structural details of the configuration in these regions.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register.
-We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding.
-To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S2 state of both sequences.
-We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation.
-This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical ``top'' and ``bottom'' of the strands.
-In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI.
-Diffusion maps built from samples of the 3S2 states are also shown in the SI.
-
-%\begin{figure}[ht!]
-%	\begin{center}
-%        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
-%        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end.
-Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends.
-The lower right region of the diffusion map space shows a high concentration of ``shifted loop'' base pairs that may facilitate the out-of-register GC-end transition.}
-%        \label{fig:GC-end_dmaps}
-%	\end{center}
-%\end{figure}
-
-%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3$^\prime$ end and its shifted complementary pair.
-As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region.
-The second diffusion mode, which correlates highly with the average distance between 3$^\prime$ and 5$^\prime$ ends, highlights structural variances within the two sequences distributions.
-We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration.
-We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register.
-These ``shifted-loop'' bonds appear uniquely stable for GC-end conformations, and suggest a transition state through which hybridization is facilitated.
-This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the equivalent AT-all pathways.
-
-%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3SPN.2 \citep{Hinckley2013AnHybridization}.
-
-%Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.
-
-% need to work on this if adding to this section (might belong somewhere else or not at all?
-%Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}.
-In the context of the 3SPN.2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}.
-The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.
-
-%AT-all looping could be a consequence of multiple bps sharing the same wc partner (I believe this is possible in 3SPN.2, because types of bonds are only mutually exclusive between cross-stacking, bp binding, and excluded volume)%
-
-%\subsection{Sequence-dependent fraying determines direct hybridization pathways}
-
-%\begin{figure}[ht!]
-%	\begin{center}
-%        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
-%        \caption{Sequence-dependent population distributions for trimmed hybridization events.
-$\Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression.
-Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins.
-Heat maps show the average intermolecular base distances in nm at increasing values of $\Psi_2$, where the main diagonal shows native base pair distances.}
-%        \label{fig:Fig8}
-%	\end{center}
-%\end{figure}
-
-%To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states.
-To achieve higher resolutions, we reran selected simulation with a 10x higher save rate and analyzed a 1000 frame region (10 ns simulation time) for each event.
-We built diffusion maps on a set of 48 such ``trimmed'' events evenly sampled across sequences.
-We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance.
-Indeed, $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence.
-This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.
-
-%Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}).
-The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first Watson-Crick contacts are made.
-In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases.
-As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs.
-GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state.
-This is consistent with the downhill nucleation-zippering mechanism described above.
-Although differences between AT-all and GC-end are more subtle, overall deviations from two-state behavior follow predictions by Sanstead et.
-al.\citep{Sanstead2016}  where GC-core > GC-mix > AT-all > GC-end.
-This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time.
-Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.
-
-%This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
-
-% add new info here, not sure if shifting discussion is necessary
-% take away message is that these are good proxies for the responses
-% although there is not perfect agreement, we show that these are very reasonable coordinates assing fast and slow response
-
-%\subsection{Limitations}
-
-%Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering.
-These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other time scales of interest.
-Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process.
-We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations.
-Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence and were working near memory limits when training SRVs and building SRV-MSMs.
-
-%In any high-level model there are inevitable simulation artefacts produced by coarse-grained approximations.
-For example, the treatment of non-interacting base pairs as an excluded volume potential alone may not be representative of dynamics produced from mismatched dangling ends.
-In general, coarse-grained models produce a smoother free-energy surface which can results in much faster motions between states.
-This is illustrated by substantial accelerations in temperature-dependent responses when compared to experiment.
-Furthermore, it may be easier to cross between states -- e.g.
-a 5S2 $\rightarrow$ H transition -- when the usually rough free energy path becomes more easily traversed.
-
 \section{\label{sec:conc}Conclusions}
@@
 Our results reveal that the specific placement of interrupting G:C pairs within an otherwise repetitive AT sequence can have a profound impact on the kinetic pathways and mechanisms for association and dissociation of the DNA duplex.
-In particular, we found AT-all to possess the richest and most complex kinetic landscape of the four sequences that is dominated by out-of-register shifted states that participate in 33\% of hybridization and dehybridization events.
+In particular, we found AT-all to possess the richest and most complex kinetic landscape of the four sequences that is dominated by out-of-register shifted states that participate in 33\% of complete hybridization events -- pathways leading from the dissociated state to full duplex formation -- and dehybridization events -- pathways from the complete duplex to full dissociation.
 Introduction of the G:C pairs at the end of the strand maintains an eight-base pair repetitive AT tract and the GC-end kinetic landscape possess all but one of the same out-of-register shifted states as AT-all.
-Destabilization of the GC-end shifted states relative to AT-all, however, results in a far more limited participation of these states with only 7\% of GC-end \noteb{a more general note, but I think a lot of people would consider any kind of stable duplex as hybridized, perhaps it would be worth stating near the beginning of our ananlysis that we are referring by default to fully in-register duplexes?} hybridization and dehybridization events passing through one or more shifted states.
+Destabilization of the GC-end shifted states relative to AT-all, however, results in a far more limited participation of these states with only 7\% of GC-end hybridization and dehybridization events passing through one or more shifted states.
 Placing the G:C pairs in the center of the strand maintains two four-base AT tracts either side of the core and results in qualitatively different kinetic behaviors for GC-core.
@@
 We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
-
-%We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics.
-By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior.
-In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these dynamics are more relevant for AT-all.
-On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted.
-Our computational approach and results show strong comparisons to T-jump experiments at ns and us time scales.
-Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons.
-These insights should be leveraged for enhanced understanding of \emph{in vivo} hybridization processes, optimized sequence design in application such as DNA-PAINT and barcoding, and novel developments in the growing field of dynamic DNA nanotechnology.
 
@@
         \includegraphics[width=150mm, scale=1]{TOC.pdf}
-        \caption*{For Table of Contents Only}
+        %\caption*{For Table of Contents Only}
         \label{fig:TOC}

### 2021-05-20 manuscript_comments.tex
Andy final edits prior to submission

--- 2021-05-20 manuscript.tex
@@
 In this work, we study a family of self-complementary pairs of 10-mer DNA oligomers using coarse-grained molecular simulation, machine learning of the slow dynamical modes, and data-driven inference of long-time kinetic models to establish new understanding of the influence of sequence upon hybridization/dehybridization kinetics and mechanisms.
-This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of a G:C pair \noteb{``of two G:C base pairs?''}within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
+This family of oligomers -- 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and 5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix) -- was designed to probe the influence of the placement of two G:C base pairs within an otherwise repetitive A:T sequence and has been the subject of our prior experimental investigations using temperature-jump infrared spectroscopy and simple lattice models.
 \citep{Sanstead2016} We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
@@
 The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.
+
+%The computational rates do tend to systematically under-predict the experimental values at low temperatures, an effect that we ascribe to the fact that the experimental T-jump responses likely contain a mixture of dissociation and hybridization events -- the latter of which persist to lower temperatures than the former -- whereas the computational analysis considers dissociation alone.
 
@@
 The NN and MSM predictions for the frayed state F4 are in good agreement.
-Although the \noteb{shifted state} trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the shifted states?
+Although the shifted state trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
@@
 Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation.
-The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability \noteb{Not sure this is correct?
-F4 $\rightarrow$ H is 50\% whereas the previous two are much lower.
-Maybe remove this sentence unless it is needed to support what comes next?}.
 We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation.
@@
 Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 24\% of the equilibrium \noteb{this is 24\% of any partially hybridized state.
-Can just say ``hybridized ensemble''} ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.3\% for GC-end, and 0\% for GC-core and GC-mix.
+Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
 It is our conjecture that the substantial presence of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.

### 2021-05-20 cover_letter/letterhead.tex
Cover letter updates

@@
 \institution\\
-%\vskip -0.1in
-%\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
+\vskip -0.1in
+\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
 }
@@
 
-Shoji Takada \hfill Biophysics, Graduate School of Science, Kyoto University \\
-+81-(0)75-753-4070 \hfill \url{takada@ biophys.kyoto-u.ac.jp}
+%Shoji Takada \hfill Biophysics, Graduate School of Science, Kyoto University \\
+%+81-(0)75-753-4070 \hfill \url{takada@ biophys.kyoto-u.ac.jp}
 
-$\circ$ \textit{Expert in simulations of coarse-grained modeling of DNA-protein systems.}
+%$\circ$ \textit{Expert in simulations of coarse-grained modeling of DNA-protein systems.}
 
-Thomas Ouldridge \hfill  Bioengineering, Imperial College London \\
-+44-(0)20-7589-5111 \hfill \url{t.ouldridge@implerial.ac.uk}
+%Thomas Ouldridge \hfill  Bioengineering, Imperial College London \\
+%+44-(0)20-7589-5111 \hfill \url{t.ouldridge@implerial.ac.uk}
 
-$\circ$ \textit{Expert in simulations of biochemical systems for nanotechnology applications.}
+%$\circ$ \textit{Expert in simulations of biochemical systems for nanotechnology applications.}
 
-Modesto Orozco \hfill  Institute for Research in Biomedicine, University of Barcelona \\
-+34-93-40-37156 \hfill \url{modesto.orozco@irbbarcelona.org}
+%Modesto Orozco \hfill  Institute for Research in Biomedicine, University of Barcelona \\
+%+34-93-40-37156 \hfill \url{modesto.orozco@irbbarcelona.org}
 
-$\circ$ \textit{Expert in molecular dynamics simulations of nucleic acids.}
+%$\circ$ \textit{Expert in molecular dynamics simulations of nucleic acids.}
 

### 2021-07-14 cover_letter/response_letter.tex
"response letter figures"

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 1.0 (28/12/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
% with extensive modifications by Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% may not be compatible with rest of template
% \documentclass[journal=jpcbfk,manuscript=article]{achemso}

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

%\documentclass[11pt,a4paper,draft]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\usepackage{letterbib}
\usepackage{graphicx} % Required for including pictures
\usepackage{microtype} % Improves typography
\usepackage{fancyhdr}
\usepackage[T1]{fontenc} % Required for accented characters
\usepackage[sort&compress,numbers,super]{natbib}
\usepackage{caption}
\usepackage{newfloat}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{soul}
\usepackage[svgnames]{xcolor}
\usepackage{framed}
\usepackage{calligra}
\usepackage{fp}
\FPset\marginWidth{0.7}	% margin width (all sides) in inches
\usepackage[margin=\marginWidth in]{geometry}
\usepackage{marginfix}
\usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter

\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%more new commands and definitions
%\newcommand*{\rood}[1]{\color{red}{#1}}
\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\noteg}[1]{\textcolor{green}{[[#1]]}}		% notes

\newcommand*{\blauw}[1]{\color{blue}{#1}}
\newcommand*{\groen}[1]{\color{green}{#1}}
\newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
\newcommand{\scripty}[1]{\ensuremath{\mathcalligra{#1}}}

% code for custom comments
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{mathtools}

% markers (black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, olive, orange, pink, purple, red, teal, violet, white, yellow, more: https://en.wikibooks.org/wiki/LaTeX/Colors)
\newcommand{\marker}[3]{
  \tikz[baseline=(X.base)]{
    \node [fill=#1!40,rounded corners] (X) {#2:};
  }
  {\color{#1!80!black}#3}
}
\newcommand{\khb}[1]{\marker{teal}{KHB}{#1}}  % Kathy Breen
\newcommand{\replyhead}{\textbf{Author Reply: }}

\newfloat{figure}{h}{lot}
\floatname{figure}{Figure}
\newfloat{sfigure}{h}{lot}
\floatname{sfigure}{Figure}
\newfloat{lfigure}{h}{lot}
\floatname{lfigure}{Figure}

\definecolor{shadecolor}{named}{LightGray}

\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\DeclareFontShape{T1}{calligra}{m}{n}{<->s*[2.2]callig15}{}

\renewcommand{\thesfigure}{S\arabic{sfigure}}
\renewcommand{\thelfigure}{\Alph{lfigure}}

%-----------------------------------------------------------------------------
% HEADER/FOOTER INFO
%-----------------------------------------------------------------------------
%\pagestyle{fancy}
\fancyhf{}
%\cfoot{\textit{telephone} 217-333-1441 $\bullet$ \textit{fax} 217-333-2736 $\bullet$ \textit{email} matse@illinois.edu}
\cfoot{\thepage}

%----------------------------------------------------------------------------------------
%	SENDER INFORMATION
%----------------------------------------------------------------------------------------

\def\Who{Andrew L. Ferguson} % Your name
\def\What{, PhD} % Your title
\def\Where{Pritzker School of Molecular Engineering\\University of Chicago} % Your department/institution
\def\department{Pritzker School of Molecular Engineering}
\def\institution{University of Chicago}
\def\Address{5640 South Ellis Avenue} % Your address
\def\CityZip{Chicago, IL 60637} % Your city, zip code, country, etc
\def\Email{e:~\url{andrewferguson@uchicago.edu}} % Your email address
\def\TEL{p:~(773)~702-5950} % Your phone number
\def\URL{w:~\url{www.ferglab.com}} % Your URL
\def\titleone{Associate Professor\\ Deputy Dean for Equity, Diversity, and Inclusion}

%----------------------------------------------------------------------------------------
%	HEADER AND FROM ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing institution logo indent based on margin width
\FPsub\logoIndent{7.55}{\tmp}

\address{
\includegraphics[width=3.5in]{UChicago_PME_logo.png} % Include the logo of your institution
%\hspace{\logoIndent in} % Position of the institution logo, increase to move left, decrease to move right
\hspace{2.8in} % Position of the institution logo, increase to move left, decrease to move right
\vskip -0.82in~\\ % Position of the text in relation to the institution logo, increase to move down, decrease to move up
\Large\hspace{0.75in} \hfill ~\\[0.05in] % First line of institution name, adjust hspace if your logo is wide
\hspace{0.75in} \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
\makebox[0ex][r]{\bf \Who \What }\hspace{0.65in} % Print your name and title with a little whitespace to the right
~\\[-0.11in] % Reduce the whitespace above the horizontal rule
\hspace{0in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
\hspace{\fill}\parbox[t]{4in}{ % Create a box for your details underneath the horizontal rule on the right
\footnotesize % Use a smaller font size for the details
%\Who \\  % Your name, all text after this will be italicized
%\\
\titleone
%\titletwo\\
%\titlethree\\
%\titlefour\\
\vskip 0.05in
\department\\ % Your department
\institution\\ % Your institution
\Address\\ % Your address
\CityZip
\vskip 0.05in
\TEL\\ % Your phone number
\Email\\ % Your email address
\URL\\ % Your URL
}
\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
\vspace{-1in} % Move the letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	TO ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing date indent based on margin width
\FPsub\dateIndent{4.15}{\tmp}

\def\opening#1{\thispagestyle{empty}
{\centering\fromaddress \vspace{1.1in} \\ % Print the header and from address here, add whitespace to move date down
%
\hspace*{4.15 in}\today\hspace*{\fill}\par
} % Print today's date, remove \today to not display it
{\raggedright \toname \\ \toaddress \par} % Print the to name and address
\vspace{0.05in} % White space after the to address
\noindent #1 % Print the opening line
% Uncomment the 4 lines below to print a footnote with custom text
%\def\thefootnote{}
%\def\footnoterule{\hrule}
%\footnotetext{\hspace*{0.18\textwidth}{Telephone 217-333-1441 $\bullet$ Fax 217-333-2736 $\bullet$ email matse@illinois.edu\footnotesize \hspace{3cm} \thepage}}
%\def\thefootnote{\arabic{footnote}}
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who \What} % The signature is a combination of your name and title

\long\def\closing#1{
\vspace{0.1in} % Some whitespace after the letter content and before the signature
\noindent % Stop paragraph indentation
%\hspace*{\longindentation} % Move the signature right
%\parbox{\indentedwidth}{\raggedright
#1 % Print the signature text
\vskip 0.0in
\includegraphics[width=0.2\textwidth]{signature_720}
%\vskip 0.65in % Whitespace between the signature text and your name
\\
\noindent
\fromsig
\vskip -0.05in
\titleone\\
\department\\
\institution\\
\vskip -0.1in
\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
}

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}
{
Prof.\ Hanadi Sleiman \\
Associate Editor, \textit{The Journal of the American Chemical Society}
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Prof.\ Sleiman}

\noteg{update below}

Thank you for your 23-Jul-2020 message regarding our article “Molecular Latent Space Simulators” assigned ID SC-EDG-07-2020-003635. We thank you for handling our submission and the anonymous reviewers for their careful reading and critical assessment of our work.

We reproduce below the full text of the reviewer comments in blue along with our responses, reproduction of the changes made to our manuscript, and the locations of these edits. The modifications within our revised submission in response to the reviewer comments are indicated in red text.

In light of these changes, we hope that our revised submission is suitable for publication in Chemical Science.

\closing{Yours Sincerely}

\end{letter}

%----------------------------------------------------------------------------------------

\clearpage
\newpage

Brennan comments:
\begin{enumerate}
    \item Get rid of "elevated" fraying response
    \item Rev 1, either view to show
    \item Figure out if the conc correction makes sense -- how much does it actually improve things?
    \item Kd hybridization does not actually increase with temp
\end{enumerate}

\begin{shaded}
\textbf{Response to Reviewer \# 1}
\end{shaded}

\textcolor{blue}{In the manuscript “Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy”, Jones et al. provide a well-designed and compelling study of DNA hybridization kinetics and dynamics that succinctly encompasses both experiment and theory. They demonstrate that the coarse-grained model of DNA, 3SPN.2, can not only well-reproduce the relative kinetic rates of sequence-dependent hybridization but also can reveal critical insights into their physical underpinnings. These insights are driven by the use of state-of-the-art applications of machine learning to Markov state model construction and well-validated with temperature jump infrared spectroscopy. In particular, the demonstration of how the placement of two G:C base pairs within the sequence can dramatically affect the energy landscape is truly impactful for the design of DNA sequences with desired properties. We believe this work is well suited for JACS publication after the Authors address a few major and minor points.}

\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased they find it significant to JACS readers.

\textcolor{blue}{Major point.
1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D. While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ‘representative’ structures shown in Figures 2A, C). A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern. This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations. At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
}

\textbf{Author Reply:}   We acknowledge that this information in crucial to the core conclusions of the manuscript and that some information regarding state definitions and representative structures was unclear.

When defining macrostates, we used both visual inspection of configurations and plots along a variety of structural metrics. In order to demonstrate that configuration in a given macrostates are well-defined by the representations in figure 2, we show the distribution of several structural metrics for each macrostate and sequence combination. These metrics are a collection of mean intermolecular distance that we have found work well to differentiate between states.

During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate. Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5\prime and 3\prime directions. We have visualized these features below, and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation. Each color map corresponds to a different macrostate according to the legend, and lighter regions correspond to higher probability at equilibrium (lower free energy).

We agree with the reviewer that it is important readers to understand the physical definitions and distributions of each state, and we have added this figure to the supplemental information.

\noteg{state distribution figure}

With regard the representative states shown in figure 2A, these were chosen from microstates with the highest probability of being assigned to the macrostates they represent. Accordingly, they should correspond to structures  that are representative of that macrostate as whole. As the reviewer points out, it is unclear how these selected microstates were defined and we include the following addition to section 3.2 to make this point explicit:

``We present these seven macrostates in Fig. 2a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. \rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole''}

\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering. I'be tried to address both aspect above, but I think}

\textcolor{blue}{1)      It is unclear how five datasets were selected used for cross-validation in the SRVs (P. 10 L. 43). One would naively think a 50:50 split of a single trajectory would result in only one dataset. A bit more detail could be provided here.}

\textbf{Author Reply:}   Five-fold cross-validation was conducted by randomly partitioning 40 independent trajectories into two sets of 20. In this way, all data is included in each fold but with distinct test/train sets. This approach prevents rare events from being left out of small validation sets and was used in our previous study, where one long trajectory was broken up into 100 \citep{Sidky2019High-ResolutionVAMPnets} smaller trajectories for training. We have  clarified this in the text by adding the following sentence to section 2.1:

We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits \rood{such that random combinations of 20 trajectories were partitioned into each set}.\citep{Sidky2019High-ResolutionVAMPnets}

\textcolor{blue}{2)      For the validation of embedding dimensions (P. 11 L. 3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Figure S1. Could a few additional dimensions be provided?}

\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence. As suggested by the reviewer, we added two additional modes to the cross-validation procedure showing this kink after five modes and have updated the AT-all subplot of Figure S1 as follows in Figure \ref{fig:AT-all_crossval}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=100mm]{cover_letter/revision_figures/AT-all_added_modes.png}
        \caption{Cross-validation for the AT-all sequence with two additional modes to show the presence of a kink in the VAMP-2 score after the fifth mode.}
        \label{fig:AT-all_crossval}
	\end{center}
\end{figure}

\textcolor{blue}{3)      The values associated with the “slow” and “fast” rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}

\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14. We agree that this sentence in unclear in its current form and have made the following change:

``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively''}

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Figure 1 has been updated to include units in s$^{-1}$. In accordance with suggestions by reviewer \#3 we have also displayed the slow responses on a y-log axis, and used the finite size effect correction mentioned by reviewer \#2 to improve the fit (Figure \ref{fig:conc_correction}). \noteg{Should we include the figure here instead? Thought it made more sense in the R2 section since we are addressing a more substantial change}

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all Figure S4 plots to include goodness-of-fit values determined by the $R^2$ values of log fits (Figure \ref{goodness-of-fit}). We have also re-plotted these on the log axis for easier comparisons between the simulated data and fits. Although all fits are strong, it does appear that GC-core may be slightly worse than the others sequences. As the reviewer suggests, this may be a contributing factor to the weaker experimental agreement for GC-core as well, however it is difficult for us to make that claim definitely given the data.  \noteg{Not sure if we really want to go into this or update the main text?}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=100mm]{cover_letter/revision_figures/goodness_of_fit.png}
        \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. We have updated this supplemental figure to include goodness-of-fit values determined by the $R^2$ values of log fits and plotted both the slow and fast fits on a log scale.}
        \label{fig:goodness-of-fit}
	\end{center}
\end{figure}

\textcolor{blue}{6)       The authors chose to compute the temperature jump relaxation rates from fits to quantities obtained directly from the MD simulation. However, it is not clear why the authors did not also compute the relaxation of observables from estimated Markov state models following (P. N. A. S. USA. 2011;108(12):4822-4827). This approach has also been previously used to compare kinetics obtained from Markov state models with other spectroscopic temperature-jump based experiments in DNA (J. Phys. Chem. B. 2019;123(10):2291-2304.) and RNA (J. Chem. Theory Comput. 2017;108(2):926–934) oligomers which should be mentioned. If possible, it would be interesting to compare or discuss the differences between their approach and these previous approaches.}

\textbf{Author Reply:}   The reviewer raises an important point regarding previous methodologies that have been developed to directly recapitulate experimental relaxation rates from MD-derived MSMs.

We have since experimented with some of these techniques using built-in PyEMMA functions, and although the results are interesting, they do not provide the most informative analysis of these particular dynamics or a strong experimental connection. We believe this is the case for the following reasons: i) shifting dynamics occur on timescales approaching that of dissociation and ii) shifting populations are small in amplitude relative to fraying and full dissociation and iii) a shifting signal is difficult to interpret experimentally without a more advanced technique such a as isotope labeling. \noteg{check with Brennan here} The combined result of these effects is that shifting dynamics are thought to make small contributions blended with the slow dissociation response. We support this hypothesis with data shown in Figure \ref{fig:responses_all}) \noteg{not sure how much to go into here}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=100mm]{cover_letter/revision_figures/responses_all.png}
        \caption{a) Experimental rate representations for each sequence. Both the 1600 cm$-1$ corresponding to A,T dynamics and 1660$-1$ corresponding to G,T dynamics are shown. b) Simulated relaxation using MSM built from each sequence. An A:T and G:C response are calculated based on the distance between available WC pairs. c) Change in response when factoring in shifting motions. Because shifting is low in amplitude and occurs on timescales approaching full dissociation, the response is not expected to be clearly discernible experimentally.  \noteg{need more description here, but not sure how much of this to include} }
        \label{fig:responses_all}
	\end{center}
\end{figure}

% Can also describe fraying issue
% The appraoch in PNAS assumes that the dynamical observable can be captured by a markovian lag time
The second problem we encountered with the MSM relaxation approach is that it does not adequately describe fast-termini fraying. As discussed in section 3.1, the experimental fast responses is thought to corresponds to single base fraying of (mostly A:T) termini. This is a process that occurs even faster in simulation as shown by our derived acceleration factor. We found that although fraying does occur for each sequence, this process cannot be captured by a Markovian lag time. On the other hand, multi-base fraying observed in GC-core is slower and corresponds to the second leading slow mode for that sequence. As shown in Figure \ref{fig:responses_all})b, we only see a distinct fast response for GC-core despite the fact that termini fraying is occurring in each sequence (as demonstrated by Figure 1 in the main text).

In addition to the issues discussed above, the main reason that we extracted responses directly from MD simulations was that effects of coarse graining, concentration, salt, etc. were unknown and could not be accounted for at a single temperature. For the slow response in particular,  timescales have an exponential dependence on temperature which could not be sufficiently standardized between experiment and simulation. Furthermore, generating equilibrium simulation data over a range of temperatures for each sequence was computationally prohibitive.

The additional references cited by the reviewer highlight important related work and should be included \citep{Noe2011DynamicalExperiments, Pinamonti2017, Remington2019MolecularKinetics}. Although we discussed the connection to (J. Chem. Theory Comput. 2017;108(2):926–934)\citep{Pinamonti2017} in the introduction, we failed to reference it with respect to relaxation rate simulations. We do believe that our work is quite distinct from these previous studies, given that much smaller RNA oligonucleotide systems were investigated. Furthermore, the RNA studies were either lacking experimental data or compared against older data which suffered some experimental flaws. This highlights the difficulty in making strong connections uses MSM relaxation techniques alone.

We have added the following text to section 3.1 make some of these difficulties more explicit and qualify our approach:

\noteg{Not totally sure where do put this, maybe end of first paragraph of 3.1 or at the end of that section?}

\rood{"Previous studies have used a dynamical fingerprinting approach \citep{Noe2011DynamicalExperiments} to compute rates from MSMs. These techniques were used to study base stacking dynamics in 2-4 base RNA oligomers \citep{Pinamonti2017, Remington2019MolecularKinetics} and showed some reasonable agreement to experiment. We tried this techniques with our SRV-MSMs as well, but found that i) termini fraying timescales were too fast for a Markovian lag time and ii) coarse-graining effects could not be properly standardized between experiment and simulations at single temperature. This motivated us to directly analyze responses from simulation data over a range of temperatures."}

\begin{shaded}
\textbf{Response to Reviewer \# 2}
\end{shaded}

\textcolor{blue}{This paper reports simulation results on the hybridization and dehybridization mechanisms near to the melting temperature of a number of short DNA sequences, and accompanying temperature jump IR experiments. The sequences are chosen to have a repeating AT motif that potentially allows relatively stable out-of register duplex states, with a focus being on how these states are disrupted, by the presence of GC base pairs. The authors show that the introduction of just 2 appropriately-positioned base pairs can change the kinetics to a simple two-state behaviour from one in which multiple states contribute.}

\textbf{Author Reply:}   We thank the reviewer... \noteg{need to say anything here?}

\textcolor{blue}{I found this an interesting study that nicely illustrated some of the potential effects of sequence on (de)hybridization mechanisms. However, I found the results unsurprising, and these types of effects have been previously explored in the literature. Although I can't therefore recommend publication in JACS, I think it would be very appropriate for one of the more specialized ACS journals.}

\textbf{Author Reply:}  We are glad that the reviewer finds our results interesting, and we acknowledge that there is precedent in the literature for mechanisms that we discuss.  Indeed, although similar states have been identified previously, the transition probabilities between states and sequence-dependent role in hybridization had not been identified. As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states \noteg{how easy will this be to do...} reflects the main significance of this work rather than the discovery of the states themselves.

\rood{update intro to focus on transition probabilities and not just states}

\textcolor{blue}{One limitation of the study is its reliance on``brute-force'' simulations (rather than using advanced sampling/rare-event techniques) to probe the transition. This means that systems of very high concentration (7 M) and temperatures close to the melting point have to be used, which are probably not the conditions of most interest.}

\textbf{Author Reply:}    We appreciate the reviewer raising this point, and we believe that more discussion is required to motivate our choice of unbiased simulations in this work.

The principal reasons that we performed brute force simulations were that we observed strong experimental agreement and obtained adequate sampling on the ms scale without needing to bias the system.  Indeed methods just as transition path sampling and forward-flux sampling have been successful in elucidating hybridization kinetics \citep{Sambriski2009, Hinckley2014Coarse-grainedEffects, Jin2019}, however these methods do rely on some knowledge and manual specification of relevant states or collective variables \emph{a priori}. In contrast, our unbiased approached allowed us to use a fully \noteg{can we say this?} unsupervised machine learning pipeline to cluster configuration and build MSMs. To summarize these points, we have added the following to \noteg{section} of the manuscript:

\rood{``A couple sentences on why we chose unbiased simulation''}

We also note that the 7 M concentration should have been 7 mM (only 3.5x higher than the 2 mM experimental concentration), and we appreciate both reviewers \#2 and \#3 for pointing out the unphysically high reported value. We have updated this change in the manuscript in section 2.5:

``A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand \rood{concentration of 7 mMol/L.''}

\textcolor{blue}{Very small points:
+ p16 line 3: "Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results". This sentence confused me at first. I presume you are just shifting your simulated kd slow curves to better match experiment, rather than shifting the data so that the melting temperatures agree. You may want to be more explicit as I read it as the latter at first. (Note one shouldn't expect the melting temperatures to agree as very different concentrations are being used in simulation and experiment, and one also has to correct for finite-size effects when estimating melting points from single duplex simulations (https://doi.org/10.1088/0953-8984/22/10/104102)}

\textbf{Author Reply:}   We greatly appreciate this comment and suggestion, especially in the context of questions from other reviewers regarding temperature and concentration.

We have since incorporated the finite size correction for homodimers reported in  https://doi.org/10.1088/0953-8984/22/10/104102 \citep{Ouldridge2010ExtractingSimulations} when computing the slow dissociation fits. We find that the results show even better agreement with experiment, motivating us to update figure 1 with these data as shown below \ref{fig:conc_correction}. In addition, we have added the units and log scale changes proposed by reviewers \#1 and \#3, respectively.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_4K_shift.png}
        \caption{Updated figure 1 of the main text. We made three changes: i) adding units to both plots, ii) accounting for finite size effects when computing the slow response, and iii) displaying the slow response on a log axis. These were proposed by reviewers \#1, \#2, and \#3 respectively}
        \label{fig:conc_correction}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_4K_shift.png}
        \caption{Updated figure 1 of the main text. We added units to both plots and displayed the slow response on a log axis as proposed by reviewers \#1 and \#3, respectively}
        \label{fig:conc_correction}
	\end{center}
\end{figure}

The following was added to the methods section to describe the fluctuation correction:

\noteg{Add a small methods paragraph to describe the temp-dependent fits? I think this would also help address reviewer \#3 confusions about the various temps used }

\noteg{Everything below would only be if we choose to include the fluctuation corrections}

\rood{The following calculates the dimer fraction in bulk solution from simulations performed in a periodic box. We apply this to the dissociation fraction in order to better compare with the experimental slow response. We compute these values from the dimer fraction (DD) and monomer fraction (D, D).}

    \begin{equation}\label{}
    \phi = \frac{\mathrm{probability (DD))}}{\mathrm{probability (D, D)}}
    \end{equation}

    \begin{equation}\label{}
	f_{\infty} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
	\end{equation}

With regard to the longer simulations used to construct SRV-MSMs, we maintain the same empirical melting temperatures so as to maximize (de)hybridization events. Although selecting a slightly lower temperature for each sequence may better approximate the bulk melting temperature, we do not expect this small shift to have a substantial impact on the sequence-dependent kinetic landscapes we have reported. Furthermore, as the reviewer indicated we are already operating at a higher concentration than experiment and would not be able to recover the exact conditions given any temperature shift alone.

\textcolor{blue}{+ p15 line 20 "beyond with" -> "beyond which"}

\textbf{Author Reply:}    This typo has been corrected.

\begin{shaded}
\textbf{Response to Reviewer \# 3}
\end{shaded}

\textcolor{blue}{Jones et al present a combined experimental and computational paper that probes the dynamics of the DNA hybridization/de-hybridization transition. Through a Markov state model analysis of coarse-grained simulation, the authors identify states that contribute to the process, and present experimental evidence to support the role of these states. In particular, repetitive sequences are observed in simulations to have a number of misaligned states that are both appreciably stable and contribute to hybridization and de-hybridization pathways as intermediates. The role of these states is diminished by base pairs that disrupt the repetitive sequence. In the special case of a sequence with a strong core and weaker extremities, metastable "frayed" states are also observed. The presence of these states is assessed experimentally trough the analysis of relaxation timescales observed in T-jump IR spectroscopy; fast relaxation is attributed to fraying equilibria, and a broader fast relaxation to an increased heterogeniety due to the presence of mis-aligned intermediates.}

\textbf{Author Reply:}   We appreciate ...

\textcolor{blue}{The paper is generally well-written, and, I believe, represents a good attempt to tackle this important process. I personally have had a manuscript "desk-rejected" by JACS on the basis that it is "too focussed on DNA biotechnology", and "primarily of interest to RNA and DNA chemists". The same statements could be made about this paper. However, I disagree with the conclusion that such papers are not appropriate for JACS. I would therefore recommend publication, subject to satisfactory responses to the comments below.  }

\textbf{Author Reply:}   We are pleased that the reviewer finds the paper both well-written and of high significance to JACS readers.

\textcolor{blue}{1.   I was frequently confused as to the temperatures used in the experiments and analysis. At one point, it is noted that simulations are performed at melting temperatures. However, simulations are actually performed at a range of temperatures and it isn't always clear when looking at a figure which temperatures were used for simulation, NN model comparison or experiments (see eg. Figs 2 and 3 and 5). I would recommend noting the temperature used to obtain the data in each case.}

\textbf{Author Reply:}   Good comment, add some clarification to the text. Be more explicit here. The “main” data is all taken at Tm, but maybe we should make more not of the temperature range in the intro/methods?

\textcolor{blue}{P7 line 13 - surely these simulations were not really done at 7M? That sounds impossibly dense.}

\textbf{Author Reply:}2.    We appreciate the reviewer pointing out this error in our reported units. The value should be 7 mM, not 7 M as shown in the below calculation. This has been corrected in the manuscript.

    \begin{equation}\label{}
	\frac{2 \mathrm{\:oligos}}{\mathrm{box\:volume}} = \frac{2}{7.80\mathrm{\:nm}^3} = 
	\frac{3.32 \times 10^{-24} \mathrm{\:moles}}{4.75 \times 10^{-25} \mathrm{\:m}^3} = 
	6.99 \frac{\mathrm{moles}}{\mathrm{m}^3} \approx 7\mathrm{mM}
	\end{equation}

\textcolor{blue}{3.  The authors repeatedly refer to "slithering" mechanisms for rearrangement. I believe this is unfortunate. In the context of the 3SPN model, the "slithering" term was first introduced in Ref. 49 by Sambriski et al, and describes two helical single strands sliding along each other. As noted by Hinckley et al in Ref 18, that mechanism is largely an artifact of the form of the potential in 3SPN.1. The rearrangement mechanism in 3SPN.2 is more of a "defect diffusion", analogous to the "inchworm mechanism" previously identified in the oxDNA model (ref 52 - note, the author list for this reference is wrong). Given this context, it seems unnecessarily confusing to talk about "slithering".}

\textbf{Author Reply:}    We share the reviewers concerns with using ``slithering'' to describe a process that would be more appropriately described as ``defect diffusion'' or the ``inchworm mechanism.`` In the manuscript, we made the effort to only use this term in reference to studies that were focused on this processes and described it as such -- namely refs 17 and 49 \noteg{update these after adding ref}. In our revisions, we have removed the term from the introduction so that it now reads as follows:

``An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms.~\citep{Xiao2019}''}

The term is also referenced on page 27 but enclosed in quotation marks to show that we are using terminology from ref 17 and not implying that this processes defines the mechanisms we observe in this work.

\noteg{No change below yet. Could add quotes to the second instance of slithering or just remove the word entirely? Not sure if either is necessary}

\rood{``Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism. (In contrast, out-of-register slithering and in-register zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.)'' }

We also appreciate the reviewer pointing out the incorrect author list in ref 52, and we have fixed the issue.

\textcolor{blue}{4. P15: I am a bit surprised that a separation of 1.3nm between the central base pairs is enough to guarantee strand separation, given that frayed end can separate by 1.3nm when the core is intact. Can the authors verify that all their duplexes had truly separated at this point?}

\textbf{Author Reply:}   This a good point of clarification. Given the importance of the core contacts during duplex dissociation, we found that any cutoff value from 0.8 nm - 2.5 nm is indicative of full dissociation and yields very similar results when calculating the slow response. On the other hand, the fraying cutoff should measure complete termini dissociation without requiring multiple bases to dissociate and thus should be chosen more carefully.

To demonstrate the flexibility of this parameter, we show the slow response from Figure 1a with dissociation cutoffs at 1.3 nm to 2.0 nm. These plots, shown in Figure \ref{fig:compare_cutoffs}, also incorporate the finite size correction proposed by reviewer \#2 and log y-axis suggested in the next comment.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{slow_response_compare_cutoff.png}
        \caption{Comparing the experimental slow response when increasing the cutoff parameter from 1.3 nm to 2.0 nm}
        \label{fig:compare_cutoffs}
	\end{center}
\end{figure}

\textcolor{blue}{5. The authors refer to "exponential" increase in Fig. 1. However, it's hard to see that an increase is exponential by eye unless the data is plotted on a log scale.}

\textbf{Author Reply:}   We agree with the reviewer that a log plot is an optimal representation of the temperature-dependent slow responses. We have made this change to to Figure 1 which has been also been shown above \ref{fig:conc_correction}.

\textcolor{blue}{6. Do the authors have any data on the scaling of k\_hybridization with T? In experiment, rates appear to increase with temperature (albeit with a much weaker dependence than de-hybridization rates). Simple "nucleation and zipper" models would predict the opposite, as does the oxDNA model (see ref 52).}

\textbf{Author Reply:}   We are interested in temperature and sequence affects on  k\_hybridization and exploring a some future work in this vein. Indeed, it is possible to use similar methods to investigate the hybridization process, however these calculation require much greater computational expense in order to gather adequate statistics. As this section served more as a validation for the model’s ability recapitulate experimental responses, we do not think this work is in scope of this work. Furthermore, the k\_hybridization values gathered by T-jump are significantly less reliable than k\_dissociation as the system cools back down and timescales approach the millisecond regime. \noteg{check with Brennan on this}

\noteg{can cite some experimental work here, but the trend they describe here is not really correct}

\textcolor{blue}{7. The authors could explain a bit more carefully (perhaps in the SI) how they handle the multiple nearest-neighbour states that correspond to their microstates (eg. F4 clearly involves more than one NN configuration).}

\textbf{Author Reply:}   \noteg{ask Brennan about this...} NN calculations we only explicitly considered configurations when we have four unbound on either side, and the rest are fully bound. We could also add configurations with partial fraying on one side but this is unlikely to change the values substantially

\textcolor{blue}{8. I don't understand how the population of F4 could be >50\% that of H for the GC-core system, when the free energy of F4 is reported as > 5kJ/mol. My back-of-the-envelope maths predicts a ratio of 8:1 for a 5kJ/mol gap. Incidentally, 8:1 is roughly the ratio I got from running the configurations through NUPACK.}

\textbf{Author Reply:}   We appreciate this comment, and we feel that it illustrates some confusion regarding the log population plot shown in Figure 2. Indeed, it may appear that the F4 state is >50\% that of H for the GC-core, however it is only about 1\% of the of H. We have decided to keep the log plot in order to capture small populations of shifted and frayed states, however we had added the following lines to the main text figure 2 caption:

``(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. \rood{Values are reported on a log y-axis to properly visualize less low-population states}''

\textcolor{blue}{9. Is the MSM model explicitly constrained to obey detailed balance? The numbers that come out seem consistent with detailed balance - forward and backward trajectories at the same T essentially look the same (they're equally likely to go through intermediates). This should be a feature of the MSM, but it's unclear to me whether it's hard-coded or just arises naturally in the coarse-graining.}

\textbf{Author Reply:}    The reviewer raises a good point regarding the assumptions made during MSM construction. We use the reversible Bayesian MSM framework described here \url{http://www.emma-project.org/latest/api/generated/pyemma.msm.BayesianMSM.html} which enforces detailed balance. This is an important point which we have made explicit in section 3.4:

``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms. \rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.''}

\textcolor{blue}{Relatedly, some of the discussion of the hybridization and de-hybridization pathways in non-repetitive sequences seems to ignore this fact. In these equilibrium simulations, hybridization trajectories played in reverse should be statistically indistinguishable from de-hybridization trajectories. I don't think anything that's said is wrong, per se, but it might be worth considering how the contrast between "nucleation and zippering" and "fraying-peeling" is consistent with the symmetry in the pathways.}

\textbf{Author Reply:}   \noteg{Do we want to change this terminology? or clarify? Fraying-peeling is effectively the reverse of nucleation-zippering. If we imply that they are distinct processes, then the reviewer is correct that it undercuts the reversible MSM assumption we make when looking at the other sequences. A proposed addition is below:}

`` In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs. Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes. \rood{These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} which can be thought of as the analogous reverse process to nucleation-zippering. Indeed, the qualitative similarity between the forward and reverse process suggest detailed balance is obeyed and that our reversible MSM analyses are well-founded.}'' \noteg{I know this doesn't sound very good, but want to make sure this is what the reviewer is getting at.  Also, should we just get rid of citations when quoting from the main text?}

\textcolor{blue}{10. I think some of the characterisation of the "all or nothing" picture is a bit misleading. The two state model of duplex formation is really a thermodynamic model, based on the assumption that the ensemble of hybridized and de-hybridized states look very different and are separated by a large barrier. This model is perfectly consistent with the transition between the two ensembles involving a complex network of metastable states. Indeed, as the authors note on line 3 of p4, a number of previous authors have considered what such states might look like, particularly for repetitive sequences. The authors' primary contribution here is not proposing that these states exist for the first time, but trying to tie together evidence from simulation and experiment to describe them.}

\textbf{Author Reply:}   This is very helpful point, and we have updated some introductory language (in response to reviewer \#2) to better communicate the novelty of the kinetic network over the states themselves:

\rood{same change as used above in response to \#2}

\noteg{Not sure the best way to address the ``all-or-nothing model is fine'' portion of this comment.}

\textcolor{blue}{11. Related to the above, the references on line 3 of p4 should probably also include Ref 52, where a similar analysis of metastable states in a repetitive sequence is performed (for a different coarse-grained model) and [Flamm, W., Fontana,J.M., Hofacker, I.L. and Schuster,P. (2000) RNA folding at elementary step resolution. RNA, 6, 325–338] where a defect diffusion mechanism is proposed.
}

\textbf{Author Reply:}   We agree with the reviewer's suggested additions in this section, and we have included both references in the following section. \noteg{need to read through these}

\citep{Flamm2000RNAResolution}

\textcolor{blue}{12. The reasoning behind why the broader relaxation spectrum is indicative of misaligned states was not that clear to me. Why would "elevated" fraying cause a broader response? Wouldn't it also, then, be broader for CG-core, which has elevated fraying? It seems to me that the argument is more along the lines of the heterogeneity of the configurations leading to heterogeneous dynamics, and hence a broader relaxation profile. But I am no expert on this technique.}

\textbf{Author Reply:}   Make some references to the relaxation experiments of reviewer \#1 is we decide to include this.

% Should not use "elevated" response here but rather put the emphasis on broadening the response. (borrow lanaguage)

The latter interpretation here is correct. The key is that the response does not come from shifting itself, so it’s fraying in the shifted state that would cause the change in response. We have have updated the language in the section to better express this point. (configurational entropy)

``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying. Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns. Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix. It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.''

\textcolor{blue}{13. How much of the code is being made available? Ideally, the initialization files for the simulations and the code used to perform the MSM analysis would be publicly available. Additionally, it would be good to publish the raw experimental data and the processing tools used to obtain the final results.}

\textbf{Author Reply:}   We agree that the data and analysis code should be made accessible for future analysis. We have uploaded input files and analysis scripts to Github ( \url{https://github.com/mrjoness/DNA_SRV-MSMs}) and  \noteg{45 GB} of featurized trajectories to Zenodo (\noteg{Zenodo link}).

\begin{shaded}
\textbf{Additional editorial edits}
\end{shaded}

\textcolor{blue}{1. Please include at least the first page for each reference citation for the following incomplete journal reference 48}

Updated

\textcolor{blue}{2. Please update the publication status of reference 70, if possible.  Providing the DOI is acceptable.}

\noteg{Does not seem like there is a submission history or DOI for this. Actually listed as a book chapter, but unclear from where. Could just remove this since most info is contained in ref 69 anyway?}

\textcolor{blue}{3. Please include publisher for the following incomplete book reference: 79.}

Updated

\textcolor{blue}{4. The TOC graphic should fit in an area no larger than 3.25 in. × 1.75 in. (approx. 8.25 cm × 4.45  cm) and should have adequate resolution and clarity. Confirm that all text is legible at this size.}

SRV axis labels were enlarged in the bottom right panel and the loss function was removed from the top right panel. \noteg{perhaps the flattened version is more appropriate for the aspect ratio? Should we provide both?}

\bibliographystyle{achemso}
\bibliography{references_210514_abv}
\end{document}

Brennan comments:
    \begin{enumerate}
        \item Using new fig1 plot -- is the fluct correction appropriate here?
        \item Accounting for NN macrostates included in the F4 configuration -- need percentages of each frayed state
    \end{enumerate}

### 2021-07-14 manuscript_marked_up.tex
"response letter figures"

--- 2021-05-20 manuscript.tex
@@
 
-%\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\rood}[1]{{\color{red}{#1}}}
 %\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
@@
 \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to high-resolution Markovian macrostate MSMs.
-We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
-We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
+We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits \rood{such that random combinations of 20 trajectories were partitioned into each set}.
+\citep{Sidky2019High-ResolutionVAMPnets} We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
 A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
@@
 First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
-This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
+This cutoff was selected as the distance beyond which the strands are effectively non-interacting and defines the dissociated state.
 We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
@@
 We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
-In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
+\rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole.} In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
 By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F).

### 2021-07-15 cover_letter/response_letter.tex
fixed figure size and latex bugs

@@
 
-Brennan comments:
-\begin{enumerate}
-    \item Get rid of "elevated" fraying response
-    \item Rev 1, either view to show
-    \item Figure out if the conc correction makes sense -- how much does it actually improve things?
-    \item Kd hybridization does not actually increase with temp
-\end{enumerate}
-
 \begin{shaded}
@@
 
-When defining macrostates, we used both visual inspection of configurations and plots along a variety of structural metrics.
-In order to demonstrate that configuration in a given macrostates are well-defined by the representations in figure 2, we show the distribution of several structural metrics for each macrostate and sequence combination.
+When defining macrostates, we both visually inspected configurations and compared distributions along a variety of structural metrics.
+During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate.
+Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions.
+We have visualized these features below (Figure \ref{fig:macrostate_colors_FES}), and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation.
+Each color map corresponds to a different macrostate according to the legend, and lighter regions correspond to higher probability at equilibrium (lower free energy).
+We agree with the reviewer that it is important readers to understand the physical definitions and distributions of each state, and we have added this figure to the supplemental information.
+
+%In order to demonstrate that configuration in a given macrostates are well-defined by the representations in figure 2, we show the distribution of several structural metrics for each macrostate and sequence combination.
 These metrics are a collection of mean intermolecular distance that we have found work well to differentiate between states.
 
-During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate.
-Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5\prime and 3\prime directions.
-We have visualized these features below, and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation.
-Each color map corresponds to a different macrostate according to the legend, and lighter regions correspond to higher probability at equilibrium (lower free energy).
-
-We agree with the reviewer that it is important readers to understand the physical definitions and distributions of each state, and we have added this figure to the supplemental information.
-
-\noteg{state distribution figure}
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=200mm]{cover_letter/revision_figures/macrostate_colors_FES.png}
+        \caption{The free energy surface of each macrostate and sequence combination.
+The x and y axes correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions, respectively.
+These distances are visualized by the corresponding red arrow representation.
+Color maps define both the macrostate identity and relative free energy of a region within that macrostate, lighter regions correspond to lower free energy.}
+        \label{fig:macrostate_colors_FES}
+	\end{center}
+\end{figure}
 
@@
 \noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering.
-I'be tried to address both aspect above, but I think}
+I'be tried to address both aspect above and can add more detail the first figure or choose a different kind of representation.}
 
@@
 	\begin{center}
-        \includegraphics[width=100mm]{cover_letter/revision_figures/AT-all_added_modes.png}
+        \includegraphics[width=75mm]{cover_letter/revision_figures/AT-all_added_modes.png}
         \caption{Cross-validation for the AT-all sequence with two additional modes to show the presence of a kink in the VAMP-2 score after the fifth mode.}
@@
 
-\textbf{Author Reply:}     We have updated all Figure S4 plots to include goodness-of-fit values determined by the $R^2$ values of log fits (Figure \ref{goodness-of-fit}).
+\textbf{Author Reply:}     We have updated all Figure S4 plots to include goodness-of-fit values determined by the $R^2$ values of log fits (Figure \ref{fig:goodness-of-fit}).
 We have also re-plotted these on the log axis for easier comparisons between the simulated data and fits.
@@
 As the reviewer suggests, this may be a contributing factor to the weaker experimental agreement for GC-core as well, however it is difficult for us to make that claim definitely given the data.
-\noteg{Not sure if we really want to go into this or update the main text?}
+As we note in the original text, this discrepancy ``may be a result of its high propensity to fray'' \noteg{Not sure if we really want to go into this or update the main text?}
 
@@
 	\begin{center}
-        \includegraphics[width=100mm]{cover_letter/revision_figures/goodness_of_fit.png}
+        \includegraphics[width=120mm]{cover_letter/revision_figures/goodness_of_fit.png}
         \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
 
+\noteg{This paragraph may not be necessary}
 We have since experimented with some of these techniques using built-in PyEMMA functions, and although the results are interesting, they do not provide the most informative analysis of these particular dynamics or a strong experimental connection.
@@
 	\begin{center}
-        \includegraphics[width=100mm]{cover_letter/revision_figures/responses_all.png}
+        \includegraphics[width=120mm]{cover_letter/revision_figures/responses_all.png}
         \caption{a) Experimental rate representations for each sequence.
@@
 On the other hand, multi-base fraying observed in GC-core is slower and corresponds to the second leading slow mode for that sequence.
-As shown in Figure \ref{fig:responses_all})b, we only see a distinct fast response for GC-core despite the fact that termini fraying is occurring in each sequence (as demonstrated by Figure 1 in the main text).
+As shown in Figure \ref{fig:responses_all}b, we only see a distinct fast response for GC-core despite the fact that termini fraying is occurring in each sequence (as demonstrated by Figure 1 in the main text).
 
@@
 For the slow response in particular,  timescales have an exponential dependence on temperature which could not be sufficiently standardized between experiment and simulation.
-Furthermore, generating equilibrium simulation data over a range of temperatures for each sequence was computationally prohibitive.
+Furthermore, generating equilibrium simulation data over a range of temperatures for each sequence was computationally prohibitive, prompting our decision to run an ensemble of shorter simulations at each temperature.
 
@@
 Indeed, although similar states have been identified previously, the transition probabilities between states and sequence-dependent role in hybridization had not been identified.
-As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states \noteg{how easy will this be to do...} reflects the main significance of this work rather than the discovery of the states themselves.
+As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states reflects the main significance of this work rather than the discovery of the states themselves.
 
-\rood{update intro to focus on transition probabilities and not just states}
+\noteg{focus more on transition probabilities and not just states? See Reviewer \#3.10 comment:}
+
+\noteg{I think we do a pretty good job of this here:}
+
+``Out-of-register ``shifted'' base paired structures \citep{Flamm2000RNAResolution, Romano2013DNADependence, Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.''
+
+\noteg{Add some more changes below?}
+
+``Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of out-of-register shifted states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
+Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states.
+Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states.''
 
@@
 
-\rood{``A couple sentences on why we chose unbiased simulation''}
+\noteg{Expand on this here:}
+
+``The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.''
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_4K_shift.png}
+        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_no_correction_4K_shift.png}
         \caption{Updated figure 1 of the main text. We added units to both plots and displayed the slow response on a log axis as proposed by reviewers \#1 and \#3, respectively}
-        \label{fig:conc_correction}
+        \label{fig:conc_no_correction}
 	\end{center}
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{slow_response_compare_cutoff.png}
+        \includegraphics[width=\textwidth]{cover_letter/revision_figures/slow_response_compare_cutoff.png}
         \caption{Comparing the experimental slow response when increasing the cutoff parameter from 1.3 nm to 2.0 nm}
@@
 
-\textbf{Author Reply:}   \noteg{ask Brennan about this...} NN calculations we only explicitly considered configurations when we have four unbound on either side, and the rest are fully bound.
-We could also add configurations with partial fraying on one side but this is unlikely to change the values substantially
+\textbf{Author Reply:}   \noteg{Brennan working on this}  NN calculations only explicitly considered configurations where four base pairs are entirely unbound on either side, and the rest are fully bound.
+The analysis below adds configurations with partial fraying on one side to better treat the diversity of microstates present in F4 (Figure \ref{fig:F4_microstates}).
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{cover_letter/revision_figures/F4_microstates.png}
+        \caption{Microstates distribution within the F4 macrostate.
+Fraying in the permutable up/down direction is symmetrized such that all frayed ends are on the left and all bound ends on are the right.
+Panel B shows the percentage of distances greater than a 1.3 nm cutoff, where 21\% of states have at least one base paired frayed and 11\% of states have at least two base pairs frayed.
+\noteg{Brennan calculating how the inclusion of these states affects the estimated free energy.}}
+        \label{fig:F4_microstates}
+	\end{center}
+\end{figure}
 
@@
 
-\textbf{Author Reply:}   We appreciate this comment, and we feel that it illustrates some confusion regarding the log population plot shown in Figure 2.
+\textbf{Author Reply:}   We appreciate this comment, and we feel that it highlights some possible confusion regarding the log population plot shown in Figure 2.
 Indeed, it may appear that the F4 state is >50\% that of H for the GC-core, however it is only about 1\% of the of H.
@@
 
-``(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
+``Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
 Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
-\rood{Values are reported on a log y-axis to properly visualize less low-population states}''
+\rood{Values are reported on a log y-axis to properly visualize less low-population states}'' \noteg{maybe this is too condescending}
 
@@
 
-\textbf{Author Reply:}   This is very helpful point, and we have updated some introductory language (in response to reviewer \#2) to better communicate the novelty of the kinetic network over the states themselves:
+\noteg{The ``all-or-nothing'' comment is an important point.
+Brennan had cautioned using this term necessarily synonym for ``two-state'' and the reviewer seems to be making the same point.
+There are 11 instances of ``all-or-nothing'' in the main text, and it is often used in conjunction with ``two-state all-or-nothing.'' I think we may want to consider removing most of them, but we should probably word this carefully.}
 
-\rood{same change as used above in response to \#2}
+\textbf{Author Reply:}   We appreciate the reviewer's helpful suggestion to highlight the role of macrostates rather than the identification of states themselves.
+We have updated some introductory language (in response comments from reviewer \#2 as well) to better communicate the novelty of the kinetic network:
 
-\noteg{Not sure the best way to address the ``all-or-nothing model is fine'' portion of this comment.}
+\noteg{Same change as used above in response to \#2 wrt to better introductory language}
 
@@
 
-\textbf{Author Reply:}   We agree with the reviewer's suggested additions in this section, and we have included both references in the following section. \noteg{need to read through these}
+\textbf{Author Reply:}   We agree that the reviewer's suggested additions are pertinent and provide the interested reader more context into shifted states and their associated mechanisms.
+We have included both references in the following section: \noteg{do we actually need to paste the complete sentence if we're just adding references?}
 
-\citep{Flamm2000RNAResolution}
+``\rood{Out-of-register ``shifted'' base paired structures \citep{Flamm2000RNAResolution, Romano2013DNADependence, Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.''
 
@@
 
-\textbf{Author Reply:}   Make some references to the relaxation experiments of reviewer \#1 is we decide to include this.
+\textbf{Author Reply:}   The reviewer makes a good comment on our choice of language used to described the stretched AT-all response.
+
+In its current form, the section may mislead the reading into thinking that an elevated fraying response alone would cause a broader fast response.
+As the reviewer indicates, this would would suggest that GC-core might have the most stretched response.
+Our hypothesis, rather, is that differential fraying behavior from an ensemble of shifted states will stretch the fast response relative to fraying from only intact duplexes.
+We believe that the reviewer's apt suggestion ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into the final paragraph of section 3.7.
+We have also
 
@@
 ``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
 It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.''
@@
 The TOC graphic should fit in an area no larger than 3.25 in.
-× 1.75 in.
+by 1.75 in.
 (approx.
-8.25 cm × 4.45  cm) and should have adequate resolution and clarity.
+8.25 cm by 4.45  cm) and should have adequate resolution and clarity.
 Confirm that all text is legible at this size.}
 
-SRV axis labels were enlarged in the bottom right panel and the loss function was removed from the top right panel.
+SRV axis labels were enlarged in the bottom right panel and the loss function was removed from the top right panel (Figure \ref{fig:TOC_square}).
 \noteg{perhaps the flattened version is more appropriate for the aspect ratio?
-Should we provide both?}
+Should we provide both?
+(Figure \ref{fig:TOC_long})}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{cover_letter/revision_figures/TOC_updated.png}
+        \caption{}
+        \label{fig:TOC_square}
+	\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \includegraphics[width=\textwidth]{cover_letter/revision_figures/TOC_updated_long.png}
+        \caption{}
+        \label{fig:TOC_long}
+	\end{center}
+\end{figure}
+
+\clearpage
+\newpage
 
@@
 \end{document}
-
-Brennan comments:
-    \begin{enumerate}
-        \item Using new fig1 plot -- is the fluct correction appropriate here?
-        \item Accounting for NN macrostates included in the F4 configuration -- need percentages of each frayed state
-    \end{enumerate}

### 2021-07-15 manuscript.tex
fixed figure size and latex bugs

@@
 A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved.
-For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
+For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Flamm2000RNAResolution, Romano2013DNADependence, Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
 The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.

### 2021-07-15 cover_letter/response_letter.tex
"fixed syntax and format"

@@
 
-\noteg{update below}
-
 Thank you for your 23-Jul-2020 message regarding our article “Molecular Latent Space Simulators” assigned ID SC-EDG-07-2020-003635.
@@
 \newpage
+
+\noteg{My notes and questions are all in green.
+The rest of the color scheme is the same as the example you sent me.
+For most comments I have attempted some change in this doc but have not changed anything major in the actual manuscript or tracked version (manuscript\_marked\_up.tex) yet.
+In some places I have proposed a couple options but did not want to delve to deep in case I'm going down the wrong path.}
 
@@
 
-\textbf{Author Reply:}   We acknowledge that this information in crucial to the core conclusions of the manuscript and that some information regarding state definitions and representative structures was unclear.
+\textbf{Author Reply:}   We acknowledge that the designation of states is crucial to the core conclusions of the manuscript and that some information regarding state definitions and representative structures was unclear.
 
-When defining macrostates, we both visually inspected configurations and compared distributions along a variety of structural metrics.
+\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering.
+I've tried to address both aspects below and can add more detail the first figure or choose a different kind of representation.}
+
+When defining macrostates, we both visually inspected configurations and compared distributions along a variety of structural features.
 During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate.
 Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions.
+These features highlight not only shifted states in both directions, but also a distinct frayed regions between the H and D wells.
 We have visualized these features below (Figure \ref{fig:macrostate_colors_FES}), and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation.
-Each color map corresponds to a different macrostate according to the legend, and lighter regions correspond to higher probability at equilibrium (lower free energy).
-We agree with the reviewer that it is important readers to understand the physical definitions and distributions of each state, and we have added this figure to the supplemental information.
+Each color map corresponds to a different macrostate according to the legend, and lighter regions indicate a higher probability at equilibrium (lower free energy).
+We agree with the reviewer that it is important for readers to understand the physical definitions and distributions of each macrostate and its associated microstates, and we have added this figure to the supplemental information.
 
@@
 	\begin{center}
-        \includegraphics[width=200mm]{cover_letter/revision_figures/macrostate_colors_FES.png}
-        \caption{The free energy surface of each macrostate and sequence combination.
+        \includegraphics[width=150mm]{cover_letter/revision_figures/macrostate_colors_FES.png}
+        \caption{Free energy surface of each macrostate and sequence combination.
 The x and y axes correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions, respectively.
-These distances are visualized by the corresponding red arrow representation.
+These distances are visualized by the corresponding red arrow representations.
 Color maps define both the macrostate identity and relative free energy of a region within that macrostate, lighter regions correspond to lower free energy.}
@@
 
-With regard the representative states shown in figure 2A, these were chosen from microstates with the highest probability of being assigned to the macrostates they represent.
-Accordingly, they should correspond to structures  that are representative of that macrostate as whole.
-As the reviewer points out, it is unclear how these selected microstates were defined and we include the following addition to section 3.2 to make this point explicit:
+With regard the molecular renderings shown in figures 2A and 2C, these were chosen from microstates with the highest probability of being assigned to their corresponding macrostates.
+Accordingly, they should correspond to structures  that are most representative of that macrostate as whole.
+As the reviewer points out, it is unclear how these selected microstates were defined in the main text, and we include the following addition to section 3.2 to make this point explicit:
 
@@
 2a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
-\rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole''}
-
-\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering.
-I'be tried to address both aspect above and can add more detail the first figure or choose a different kind of representation.}
+\rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole.
+A more detailed representation of macrostate membership is shown in supplemental figure \noteg{add supplemental reference here}''}
 
@@
 
-\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence.
-As suggested by the reviewer, we added two additional modes to the cross-validation procedure showing this kink after five modes and have updated the AT-all subplot of Figure S1 as follows in Figure \ref{fig:AT-all_crossval}.
+\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences.
+As suggested by the reviewer, we added two additional modes to the cross-validation procedure showing this kink after five modes and have updated the AT-all subplot of Figure S1 with the panel shown in Figure \ref{fig:AT-all_crossval}.
 
@@
 \textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14.
-We agree that this sentence in unclear in its current form and have made the following change:
+We agree that the sentence on (P.
+14 L.
+44) in unclear in its current form and have made the following change:
 
@@
 As the reviewer suggests, this may be a contributing factor to the weaker experimental agreement for GC-core as well, however it is difficult for us to make that claim definitely given the data.
-As we note in the original text, this discrepancy ``may be a result of its high propensity to fray'' \noteg{Not sure if we really want to go into this or update the main text?}
+As we note in the original text, this discrepancy may also be a result of its high propensity to fray.
+\noteg{Not sure if we really want to go into this or update the main text?}
 
@@
 	\begin{center}
-        \includegraphics[width=120mm]{cover_letter/revision_figures/responses_all.png}
+        \includegraphics[width=150mm]{cover_letter/revision_figures/responses_all.png}
         \caption{a) Experimental rate representations for each sequence.
@@
 The additional references cited by the reviewer highlight important related work and should be included \citep{Noe2011DynamicalExperiments, Pinamonti2017, Remington2019MolecularKinetics}.
-Although we discussed the connection to (J.
+Although we discussed our work's connection to (J.
 Chem.
 Theory Comput.
-2017;108(2):926–934)\citep{Pinamonti2017} in the introduction, we failed to reference it with respect to relaxation rate simulations.
+2017;108(2):926–934)\citep{Pinamonti2017} in the introduction, we did not originally reference it with respect to relaxation rate simulations.
 We do believe that our work is quite distinct from these previous studies, given that much smaller RNA oligonucleotide systems were investigated.
 Furthermore, the RNA studies were either lacking experimental data or compared against older data which suffered some experimental flaws.
-This highlights the difficulty in making strong connections uses MSM relaxation techniques alone.
+This highlights the difficulty in making strong experimental comparions for oligonucleotide systems using MSM relaxation techniques alone.
 
@@
 \rood{"Previous studies have used a dynamical fingerprinting approach \citep{Noe2011DynamicalExperiments} to compute rates from MSMs.
-These techniques were used to study base stacking dynamics in 2-4 base RNA oligomers \citep{Pinamonti2017, Remington2019MolecularKinetics} and showed some reasonable agreement to experiment.
-We tried this techniques with our SRV-MSMs as well, but found that i) termini fraying timescales were too fast for a Markovian lag time and ii) coarse-graining effects could not be properly standardized between experiment and simulations at single temperature.
+These techniques were used to study base stacking dynamics in 2-4 base RNA oligomers \citep{Pinamonti2017, Remington2019MolecularKinetics} and some attempts were made to compare against experiment.
+However, Limitations in all-atom force fields and potential experimental artefacts have made strong comparisons difficult to achieve for oligonucleotide systems \citep{Pinamonti2017}.
+We tried applying these techniques with our SRV-MSMs as well, but found that i) termini fraying timescales were too fast for a Markovian lag time and ii) coarse-graining effects could not be properly standardized between experiment and simulations at a single temperature.
 This motivated us to directly analyze responses from simulation data over a range of temperatures."}
@@
 
-\textbf{Author Reply:}   We thank the reviewer... \noteg{need to say anything here?}
+\textbf{Author Reply:}   We thank the reviewer for their comments. \noteg{need to say anything here?}
 
@@
 
-\textbf{Author Reply:}  We are glad that the reviewer finds our results interesting, and we acknowledge that there is precedent in the literature for mechanisms that we discuss.
-Indeed, although similar states have been identified previously, the transition probabilities between states and sequence-dependent role in hybridization had not been identified.
-As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states reflects the main significance of this work rather than the discovery of the states themselves.
+\textbf{Author Reply:}  We are glad that the reviewer finds our results interesting, and we acknowledge that there is precedent in the literature for the states and mechanisms that we discuss.
+Indeed, although similar dynamics have been identified previously, the transition probabilities between states and sequence-dependent affect on these transitions has not been identified and quantified.
+As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states reflects is significance of this work rather than the discovery of the states or mechanisms themselves
 
@@
 
-\noteg{Expand on this here:}
+\noteg{Expand on this here?}
 
@@
 \textbf{Author Reply:}   We greatly appreciate this comment and suggestion, especially in the context of questions from other reviewers regarding temperature and concentration.
+
+\noteg{Add a small methods paragraph to describe the temp-dependent fits? I think this would also help address reviewer \#3 confusions about the various temps used }
 
@@
         \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_no_correction_4K_shift.png}
-        \caption{Updated figure 1 of the main text. We added units to both plots and displayed the slow response on a log axis as proposed by reviewers \#1 and \#3, respectively}
+        \caption{Updated figure 1 of the main text.
+We added units to both plots and displayed the slow response on a log axis as proposed by reviewers \#1 and \#3, respectively \noteg{This is with no correction and the results do not look all the different (homodimer correction is less than hetero).
+I'm leaning towards just using this given that the temp-dependent simulations are not performed in equlibrium and the correction appears unncessary}}
         \label{fig:conc_no_correction}
@@
 
+\noteg{Everything below would only be if we choose to include the fluctuation corrections}
+
 The following was added to the methods section to describe the fluctuation correction:
-
-\noteg{Add a small methods paragraph to describe the temp-dependent fits? I think this would also help address reviewer \#3 confusions about the various temps used }
-
-\noteg{Everything below would only be if we choose to include the fluctuation corrections}
 
@@
 
-\textbf{Author Reply:}   We appreciate ...
+\textbf{Author Reply:}   We appreciate the reviewer's detailed summary of our work
 
@@
 
-\textbf{Author Reply:}   Good comment, add some clarification to the text.
-Be more explicit here.
-The “main” data is all taken at Tm, but maybe we should make more not of the temperature range in the intro/methods?
+\textbf{Author Reply:}   This is a good comment that requires some clarification in the text.
+We should be more explicit that the MSM data was all acquired at sequence-specific melting temperatures, but the experimental validatoin in 3.1 was performed over a range of temperature in experiment and simulation.
+
+\noteg{we don't really have a methods section on the temp-dependent simulations, maybe this should be added to 2.1.1 address this comment and a question from reviewer \#2?
+Alternatively we could just add something like this to the intro in addition to another sentence or two?}
+
+"We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data \rood{over a range of temperatures} and reinterpret our prior experimental observations in light of the new computational understanding."
 
@@
 
-\textbf{Author Reply:}    We share the reviewers concerns with using ``slithering'' to describe a process that would be more appropriately described as ``defect diffusion'' or the ``inchworm mechanism.`` In the manuscript, we made the effort to only use this term in reference to studies that were focused on this processes and described it as such -- namely refs 17 and 49 \noteg{update these after adding ref}.
+\textbf{Author Reply:}    We share the reviewer's concerns with using ``slithering'' to describe a process that would be more appropriately called ``defect diffusion'' or the ``inchworm mechanism.`` In the manuscript, we made the effort to only use this term in reference to studies that were focused on this processes and described it as such -- namely refs 17 and 49 \noteg{update these after adding ref}.
 In our revisions, we have removed the term from the introduction so that it now reads as follows:
@@
 
-\noteg{Same change as used above in response to \#2 wrt to better introductory language}
+\noteg{Same change as used above in response to \#2 wrt to better introductory language. Maybe some addtions in the main text or conclusions as well?}
 

### 2021-07-27 cover_letter/response_letter_MJ_210721.tex
just working on 3.7 and 3.8

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 1.0 (28/12/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
% with extensive modifications by Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% may not be compatible with rest of template
% \documentclass[journal=jpcbfk,manuscript=article]{achemso}

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

%\documentclass[11pt,a4paper,draft]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\usepackage{letterbib}
\usepackage{graphicx} % Required for including pictures
\usepackage{microtype} % Improves typography
\usepackage{fancyhdr}
\usepackage[T1]{fontenc} % Required for accented characters
\usepackage[sort&compress,numbers,super]{natbib}
\usepackage{caption}
\usepackage{newfloat}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{soul}
\usepackage[svgnames]{xcolor}
\usepackage{framed}
\usepackage{calligra}
\usepackage{fp}
\FPset\marginWidth{0.7}	% margin width (all sides) in inches
\usepackage[margin=\marginWidth in]{geometry}
\usepackage{marginfix}
\usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}
\usepackage[font={small,sf}, singlelinecheck=false]{caption}
\usepackage{capt-of}

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter

\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%more new commands and definitions
%\newcommand*{\rood}[1]{\color{red}{#1}}
\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\noteg}[1]{\textcolor{green}{[[#1]]}}		% notes 
\newcommand*{\noter}[1]{{\color{red}{[[#1]]}}}

\newcommand*{\blauw}[1]{\color{blue}{#1}}
\newcommand*{\groen}[1]{\color{green}{#1}}
\newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
\newcommand{\scripty}[1]{\ensuremath{\mathcalligra{#1}}}

% code for custom comments
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{mathtools}

% markers (black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, olive, orange, pink, purple, red, teal, violet, white, yellow, more: https://en.wikibooks.org/wiki/LaTeX/Colors)
\newcommand{\marker}[3]{
  \tikz[baseline=(X.base)]{
    \node [fill=#1!40,rounded corners] (X) {#2:};
  }
  {\color{#1!80!black}#3}
}
\newcommand{\khb}[1]{\marker{teal}{KHB}{#1}}  % Kathy Breen
\newcommand{\replyhead}{\textbf{Author Reply: }}

\newfloat{figure}{h}{lot}
\floatname{figure}{Figure}
\newfloat{sfigure}{h}{lot}
\floatname{sfigure}{Figure}
\newfloat{lfigure}{h}{lot}
\floatname{lfigure}{Figure}

\definecolor{shadecolor}{named}{LightGray}

\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\DeclareFontShape{T1}{calligra}{m}{n}{<->s*[2.2]callig15}{}

\renewcommand{\thesfigure}{S\arabic{sfigure}}
\renewcommand{\thelfigure}{\Alph{lfigure}}

%-----------------------------------------------------------------------------
% HEADER/FOOTER INFO
%-----------------------------------------------------------------------------
%\pagestyle{fancy}
\fancyhf{}
%\cfoot{\textit{telephone} 217-333-1441 $\bullet$ \textit{fax} 217-333-2736 $\bullet$ \textit{email} matse@illinois.edu}
\cfoot{\thepage}

%----------------------------------------------------------------------------------------
%	SENDER INFORMATION
%----------------------------------------------------------------------------------------

\def\Who{Andrew L. Ferguson} % Your name
\def\What{, PhD} % Your title
\def\Where{Pritzker School of Molecular Engineering\\University of Chicago} % Your department/institution
\def\department{Pritzker School of Molecular Engineering}
\def\institution{University of Chicago}
\def\Address{5640 South Ellis Avenue} % Your address
\def\CityZip{Chicago, IL 60637} % Your city, zip code, country, etc
\def\Email{e:~\url{andrewferguson@uchicago.edu}} % Your email address
\def\TEL{p:~(773)~702-5950} % Your phone number
\def\URL{w:~\url{www.ferglab.com}} % Your URL
\def\titleone{Associate Professor\\ Vice Dean for Equity, Diversity, and Inclusion}

%----------------------------------------------------------------------------------------
%	HEADER AND FROM ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing institution logo indent based on margin width
\FPsub\logoIndent{7.55}{\tmp}

\address{
\includegraphics[width=3.5in]{UChicago_PME_logo.png} % Include the logo of your institution
%\hspace{\logoIndent in} % Position of the institution logo, increase to move left, decrease to move right
\hspace{2.8in} % Position of the institution logo, increase to move left, decrease to move right
\vskip -0.82in~\\ % Position of the text in relation to the institution logo, increase to move down, decrease to move up
\Large\hspace{0.75in} \hfill ~\\[0.05in] % First line of institution name, adjust hspace if your logo is wide
\hspace{0.75in} \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
\makebox[0ex][r]{\bf \Who \What }\hspace{0.65in} % Print your name and title with a little whitespace to the right
~\\[-0.11in] % Reduce the whitespace above the horizontal rule
\hspace{0in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
\hspace{\fill}\parbox[t]{4in}{ % Create a box for your details underneath the horizontal rule on the right
\footnotesize % Use a smaller font size for the details
%\Who \\  % Your name, all text after this will be italicized
%\\
\titleone
%\titletwo\\
%\titlethree\\
%\titlefour\\
\vskip 0.05in
\department\\ % Your department
\institution\\ % Your institution
\Address\\ % Your address
\CityZip
\vskip 0.05in
\TEL\\ % Your phone number
\Email\\ % Your email address
\URL\\ % Your URL
}
\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
\vspace{-1in} % Move the letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	TO ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing date indent based on margin width
\FPsub\dateIndent{4.15}{\tmp}

\def\opening#1{\thispagestyle{empty}
{\centering\fromaddress \vspace{1.1in} \\ % Print the header and from address here, add whitespace to move date down
%
\hspace*{4.15 in}\today\hspace*{\fill}\par
} % Print today's date, remove \today to not display it
{\raggedright \toname \\ \toaddress \par} % Print the to name and address
\vspace{0.05in} % White space after the to address
\noindent #1 % Print the opening line
% Uncomment the 4 lines below to print a footnote with custom text
%\def\thefootnote{}
%\def\footnoterule{\hrule}
%\footnotetext{\hspace*{0.18\textwidth}{Telephone 217-333-1441 $\bullet$ Fax 217-333-2736 $\bullet$ email matse@illinois.edu\footnotesize \hspace{3cm} \thepage}}
%\def\thefootnote{\arabic{footnote}}
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who \What} % The signature is a combination of your name and title

\long\def\closing#1{
\vspace{0.1in} % Some whitespace after the letter content and before the signature
\noindent % Stop paragraph indentation
%\hspace*{\longindentation} % Move the signature right
%\parbox{\indentedwidth}{\raggedright
#1 % Print the signature text
\vskip 0.0in
\includegraphics[width=0.2\textwidth]{signature_720}
%\vskip 0.65in % Whitespace between the signature text and your name
\\
\noindent
\fromsig
\vskip -0.05in
\titleone\\
\department\\
\institution\\
\vskip -0.1in
\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
}

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}
{
Prof.\ Hanadi Sleiman \\
Associate Editor, \textit{The Journal of the American Chemical Society}
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Prof.\ Sleiman}

Thank you for your 6-Jul-2021 message regarding our article ``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy'' assigned manuscript ID ja-2021-05219p. We thank you for handling our submission and the anonymous reviewers for their careful reading and critical assessment of our work.

We reproduce below the full text of the reviewer comments in blue along with our responses, reproduction of the changes made to our manuscript, and the locations of these edits. The modifications within our revised submission in response to the reviewer comments are indicated in red text.

In light of these changes, we hope that our revised submission is suitable for publication in \textit{The Journal of the American Chemical Society}.

\closing{Yours Sincerely}

\end{letter}

%----------------------------------------------------------------------------------------

\clearpage
\newpage

\begin{shaded}
\textbf{Response to Reviewer \# 1}
\end{shaded}

\textcolor{blue}{In the manuscript ``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy'', Jones et al. provide a well-designed and compelling study of DNA hybridization kinetics and dynamics that succinctly encompasses both experiment and theory. They demonstrate that the coarse-grained model of DNA, 3SPN.2, can not only well-reproduce the relative kinetic rates of sequence-dependent hybridization but also can reveal critical insights into their physical underpinnings. These insights are driven by the use of state-of-the-art applications of machine learning to Markov state model construction and well-validated with temperature jump infrared spectroscopy. In particular, the demonstration of how the placement of two G:C base pairs within the sequence can dramatically affect the energy landscape is truly impactful for the design of DNA sequences with desired properties. We believe this work is well suited for JACS publication after the Authors address a few major and minor points.}

\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased that they find the work impactful and suitable for publication in JACS.

\textcolor{blue}{Major point.
1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D. While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ``representative'' structures shown in Figures 2A, C). A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern. This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations. At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
}

\textbf{Author Reply:} The designation and structural representation of the macrostates is crucial to the core conclusions of the manuscript and we concur that we can and should better expose for the reader how the representative microstates were selected, how representative they are of the microstate ensemble within the macrostate, and our analysis of the base-pairing patterns within each macrostate that motivated our naming scheme.

On p.~19, we have added the following text:

``...each of these macrostates. \rood{The representative microstates emblematic of each macrostate were selected randomly from those possessing high ($>$99\%) macrostate membership probabilities. Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution. In Fig.~S{\blauw{X}} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble. In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig. 2b we present...''

We also added the new Fig.~S7 \noteg{is this order fine?} to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer. The figure and its caption are reproduced below.

\clearpage
\newpage

\begin{figure}[ht!]
	\centering
    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First. each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
        %\label{fig:macrostate_colors_FES}
	\end{center}
\end{figure}

\clearpage
\newpage

\textcolor{blue}{1)      It is unclear how five datasets were selected used for cross-validation in the SRVs (P. 10 L. 43). One would naively think a 50:50 split of a single trajectory would result in only one dataset. A bit more detail could be provided here.}

\textbf{Author Reply:}   Five-fold cross-validation was conducted by constructing five random partitions of the 40 independent trajectories into a training set of 20 trajectories and a validation set of 20 trajectories. This manner of splitting the data at the level of complete trajectories -- as opposed to splitting a single trajectory into multiple subtrajectories -- is desirable as it allows our SRV/MSM approach to learn over long 25 $\mu$s trajectories and infer slowly relaxing collective modes that may not be accessible within short sub-trajectories~\citep{Sidky2019High-ResolutionVAMPnets}. We have better clarified this procedure for the reader by revising the text on p.~11 to read:

We guarded against overfitting using five-fold cross-validation in which we \rood{constructed five random partitions of the 40 independent simulation trajectories into a training set of 20 trajectories and a validation set of 20 trajectories}.\citep{Sidky2019High-ResolutionVAMPnets}

\textcolor{blue}{2)      For the validation of embedding dimensions (P. 11 L. 3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Fig.~S1. Could a few additional dimensions be provided?}

textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences. As suggested by the reviewer, we added two additional modes to the cross-validation procedure that does indeed better expose this kink at five slow modes. We have updated the corresponding panel Fig.~S1 and also reproduce it below.

% Fig. S1
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=1.0\textwidth]{FigS1.pdf}
         \caption*{\textbf{Figure S1.}Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence. A knee in the VAMP-2 plot was identified at the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively. An embedding of corresponding dimensionality was then used to cluster frames into discrete states. The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
        \label{fig:SIFig1}
	\end{center}
\end{figure}

\textcolor{blue}{3)      The values associated with the ``slow'' and ``fast'' rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}

\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14. We agree that the sentence on (P. 14 L. 44) is unclear in its current form and have made the following change:

``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively''}

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Thank you for catching this error. Fig.~1 has been updated to include units in s$^{-1}$.

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ goodness-of-fit values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits. The revised figure is reproduced below. In all cases except the lowest temperature (T$_m$-5) slow response we report good fits with $R^2$ $>$ 0.88. We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good \noteg{not sure what we're reporting here? \rood{X.XX} $<$ $R^2$ $<$ \rood{X.XX})}.

\clearpage
\newpage

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.75\textwidth]{FigS4.pdf}
   \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
    \label{fig:SIFig4}
\end{figure}

\clearpage
\newpage

\textcolor{blue}{6)       The authors chose to compute the temperature jump relaxation rates from fits to quantities obtained directly from the MD simulation. However, it is not clear why the authors did not also compute the relaxation of observables from estimated Markov state models following (P. N. A. S. USA. 2011;108(12):4822-4827). This approach has also been previously used to compare kinetics obtained from Markov state models with other spectroscopic temperature-jump based experiments in DNA (J. Phys. Chem. B. 2019;123(10):2291-2304.) and RNA (J. Chem. Theory Comput. 2017;108(2):926-934) oligomers which should be mentioned. If possible, it would be interesting to compare or discuss the differences between their approach and these previous approaches.}

\textbf{Author Reply:}   The reviewer raises an important point regarding the dynamical fingerprinting methodology that have been developed to compare MSM predictions and relaxation experiments. We are familiar with these techniques, are currently applying them in our current work on abasic DNA, and did actually attempt to apply them in the present work. These analyses, however, were less informative than we had hoped due to three primary factors.

First, the experimental data contains fast responses occurring on time scales of $\sim$100 ns. As detailed in Section 3.1 and Figure 1, these fast motions occur in our simulations at an acceleration factor of approximately 120$\times$ and so proceed on a time scale of $\sim$1 ns. As such, these motions lie below the $\tau$ = 1.2 ns lag time used to construct our MSMs that is required to converge the leading slow modes and produce a Markovian model capable of passing the Chapman-Kolmogorov test (cf. Figs.~S\rood{2-3}). As such, we cannot draw comparisons of our MSM against the fast experimental data.

Second, the dynamical fingerprinting method provides an elegant means to compare experimental data against an MSM constructed at the same state point (e.g., temperature, pressure, slat conditions). We constructed only a single MSM for each sequence at the melting temperature of the 3SPN.2 model that required the collection of 1 ms of simulation data for each sequence in order to observe sufficiently many hybridization/dehybridization events for robust fitting of the model parameters (Section 2.1.1). The experimental relaxation data, however, was collected over a range of temperatures and we wished to make comparisons across this entire range. Fitting additional MSMs was too computationally burdensome due to the high volume of data required and the cost of these simulations, so instead we conducted 120 $\times$ 1 $\mu$s simulations for each DNA sequence at each temperature. These trajectory data were too sparse to fit a robust MSM, but did allow us to make a direct comparison between the MD simulations and the experimental measurements over the entire temperature range.

We therefore elected to structure our manuscript such that we first test the MD simulation predictions against experimental data in Section 3.1, then, after having validated the simulation data upon which the MSMs are constructed, perform MSM construction and analysis in Sections 3.2-6. Finally, in Section 3.7 we close with a test of a specific prediction of our MSM (the existence of long-lived metastable shifted states) that was amenable to comparison against T-jump IR.

We concur with the reviewer that we can and should explicitly mention dynamical fingerprinting approach and provide some discussion for the reader why we did not employ this approach in the present study. To this end, we have added the following text to the end of Section 3.1:

``fast kinetics of the four DNA oligomers\rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence. We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}. These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}. We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected. Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}''

\clearpage
\newpage

\begin{shaded}
\textbf{Response to Reviewer \# 2}
\end{shaded}

\textcolor{blue}{This paper reports simulation results on the hybridization and dehybridization mechanisms near to the melting temperature of a number of short DNA sequences, and accompanying temperature jump IR experiments. The sequences are chosen to have a repeating AT motif that potentially allows relatively stable out-of register duplex states, with a focus being on how these states are disrupted, by the presence of GC base pairs. The authors show that the introduction of just 2 appropriately-positioned base pairs can change the kinetics to a simple two-state behaviour from one in which multiple states contribute.}

\textcolor{blue}{I found this an interesting study that nicely illustrated some of the potential effects of sequence on (de)hybridization mechanisms. However, I found the results unsurprising, and these types of effects have been previously explored in the literature. Although I can't therefore recommend publication in JACS, I think it would be very appropriate for one of the more specialized ACS journals.}

\textbf{Author Reply:}  We are pleased that the reviewer finds our results interesting and suitable for publication. We concur with the reviewer that this work builds upon prior studies in the literature on sequence effects upon hybridization/dehybridization mechanisms, but goes beyond these studies in presenting a mean to rigorously resolve and quantify these dynamics. In response to comments from Reviewer \#2 and Reviewer \#3, we have updated several introductory sentences to clarify the novelty and impact of this work like less in the discovery of the particular metastable states, but in furnishing quantitative and predictive kinetic models of the dynamical transition network between these states that furnishes experimentally testable predictions and new understanding and quantification of the influence of sequence upon thermodynamics and kinetics.

We better contextualize the scope and impact of our work in the final sentence of the Abstract:

``Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization \rood{by furnishing quantitative and predictive kinetic models of the dynamical transition network between metastable states}, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation for DNA nanotechnology applications.''

and we state in the Introduction:

``Out-of-register ``shifted'' base paired structures \citep{Flamm2000RNAResolution, Romano2013DNADependence, Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.''

``We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding. Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of out-of-register shifted states -- leads to richer dynamics populated by a diversity of long-lived metastable states. Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states. Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states.''

``Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to control the dynamical behaviors of the system. We anticipate that this new foundational understanding, and the extension of our approach to more extensive families of DNA sequences, can guide the rational design of DNA oligomers with tailored kinetic properties engineered for DNA nanotechnology applications such as DNA-PAINT. \citep{Shah2019, Strauss2020UpDNA-PAINT}''

\textcolor{blue}{One limitation of the study is its reliance on ``brute-force'' simulations (rather than using advanced sampling/rare-event techniques) to probe the transition. This means that systems of very high concentration (7 M) and temperatures close to the melting point have to be used, which are probably not the conditions of most interest.}

\textbf{Author Reply:}    (I) We agree with the reviewer that it may be possible to use biasing techniques in order to efficiently collect the data required to construct the MSMs. We are very familiar with the use of such techniques: indeed, another major interest of our research group is iterative strategies to perform data-driven determination of collective variables from simulation trajectories, enhanced sampling in these variables, and then estimation of new and improved collective variables from the enhanced sampling trajectories \citep{Sidky2020MolecularSimulators, Sidky2020MachineSimulation, Chen2019CapabilitiesSystems, Chen2019NonlinearVAMPnets, Phys2018CollectiveDesign, Chen2018MolecularExploration}. In the present work, we chose not to employ biased simulations and instead collect unbiased trajectories for three main reasons.

-- First, the 3SPN.2 coarse-grained DNA model is sufficiently inexpensive that it was possible to collect adequate unbiased simulation data to parameterize robust MSMs at reasonable costs. As we state on p.~8:

``Each simulation was conducted for 26 $\mu$s and frames saved to disc every 100 ps. Each simulation required $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.''

-- Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1, Fig.~S\rood{4}) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S\rood{1-3}). This indicates that the data are of sufficient quality and volume to present favorable experimental comparisons and learn reliable long-time kinetic models.

-- Third, although it is straightforward to correct biased simulation data to recover the unbiased thermodynamics, it is more challenging to recover the unbiased kinetics. Although a number of approaches do exist, it has been our experience that the kinetic corrections can present difficulties in numerical convergence. That is not to say that we could not have successfully applied this methods, but just that we did not have cause to do so since the unbiased data were not unreasonably expensive to collect and showed favorable experimental comparisons and produced robust MSMs.

We did note the possibility of performing biased simulations in the Introduction on p.~4-5, and have now elaborated these remarks to better explain our choice to proceed with unbiased simulations:

``The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}. \rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable Markov state models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling. A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}. The application of these methods under the conditions of high bias necessary for good sampling can, however, present challenges for numerical convergence.} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways. These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}. \rood{In this work, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}''

(II) Our specification of a single strand concentration of 7 mol/L in Section 2.1.1 on p.~7 was an (embarrassing) typo -- the true concentration is actually 7 mmol/L (i.e., 7 mM). This is only 3.5$\times$ higher than the 2 mmol/L experimental concentration, and so is actually quite a good approximation for the experimental conditions. We regret our error, and have corrected this text to now read:

``A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of \rood{7 mM}. \rood{This concentration is only 3.5$\times$ larger than the 2 mM concentration employed in our experimental analyses (Section 2.2).}''

(III) Finally, we deliberately chose to explore the melting temperature as the state point at which the hybridized and dehybridized populations are in balance and sought to probe the mechanism and kinetics of the hybridization/dehybridization transitions under these equilibrium conditions corresponding to no net thermodynamic preference for either state (i.e., minimally driven conditions). A favorable side effect of this choice was indeed that we observe approximately equal numbers of hybridization and dehybridization events in a single (sufficiently long) simulation, whereas temperatures away from $T_\mathrm{melt}$ would favor one over the other. It is an interesting question to consider how the mechanism and kinetics of hybridization/dehybridization may change with temperature. Exploring small excursions either side of $T_\mathrm{melt}$ may still be possible using unbiased simulations, but we concur with the reviewer that biased sampling would be required to observe sufficient numbers of hybridization (dehybridization) events at large excursions above (below) $T_\mathrm{melt}$.

In response to this point, we have modified text in the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting point for follow-up study:

``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons. \rood{We also suggest that the MSM approach followed in this work, possible coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels} , may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''

\textcolor{blue}{Very small points:
+ p16 line 3: "Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results". This sentence confused me at first. I presume you are just shifting your simulated kd slow curves to better match experiment, rather than shifting the data so that the melting temperatures agree. You may want to be more explicit as I read it as the latter at first. (Note one shouldn't expect the melting temperatures to agree as very different concentrations are being used in simulation and experiment, and one also has to correct for finite-size effects when estimating melting points from single duplex simulations (https://doi.org/10.1088/0953-8984/22/10/104102)}

\textbf{Author Reply:}   (I) We regret our lack of clarity in our choice of wording. Yes, we simply shifted the $k_d$ curves by a universal temperature shift, determined to be +4 K, that produced the optimal match with the experimental data.

We have also reworded the unclear wording regarding the temperature correction in Section 3.1 p.~16 to now read:

``Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.''

(II) We regret our erroneous reporting of a 7 M rather than 7 mM concentration as detailed in our previous response. As we have now clarified, our simulations and experiments are conducted at similar concentrations (7mM and 2 mM, respectively). We agree that we still cannot expect perfect agreement in the melting temperatures due to the residual small  difference in concentrations, but likely this is small in comparison to discrepancies introduced by the approximations inherent in the coarse-grained molecular model.

(III) We greatly appreciate this comment and suggestion to apply a fluctuation correction to our data in order to correct for artifacts associated with finite-size effects following the approach detailed in Ref.~\citep{Ouldridge2010ExtractingSimulations}. We have reanalyzed our slow repsonse data under the homodimer correction and this has pleasingly resulted in a slightly improved agreement between our simulations and experiment. The revised figure is reproduced below. \noteg{We note that it is unclear how this correction should be applied to a periodic system that is not in thermodynamics equilibrium, however we choose to apply the correction so that we will observe asymptotically correct behavior at long timescales.}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Fig1.pdf}
        \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses. Results are reported in terms of the final T-jump temperature. (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales. \rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales. The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics. The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

In light of this modification, we have also added the following text to the Section 3.1 on p.~16:

\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to correct our predictions for $f_\mathrm{hybridized}$ bulk yields. We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}

    \begin{equation}\label{}
    \phi = \frac{\mathrm{f_\mathrm{hybridized}^\mathrm{sim}}}{\mathrm{1 - f_\mathrm{hybridized}^\mathrm{sim}}}
    \end{equation}

    \begin{equation}\label{}
	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
	\end{equation}

Finally, we have also edited the following text in Section 3.1 on p.~17 to clarify that the fluctuation correction is only applied for the purposes of the experimental comparison in Section 3.1 and is not propagated throughout the remainder of the paper:

``The computational time scales and melting temperatures reported in the remainder of the paper are not corrected \rood{by these calibration corrections or finite-size fluctuation corrections} since we are only interested in the relative trends in the behaviors of the four sequences.''

\textcolor{blue}{+ p15 line 20 "beyond with" -> "beyond which"}

\textbf{Author Reply:}    This typo has been corrected.

\clearpage
\newpage

\begin{shaded}
\textbf{Response to Reviewer \# 3}
\end{shaded}

\textcolor{blue}{Jones et al present a combined experimental and computational paper that probes the dynamics of the DNA hybridization/de-hybridization transition. Through a Markov state model analysis of coarse-grained simulation, the authors identify states that contribute to the process, and present experimental evidence to support the role of these states. In particular, repetitive sequences are observed in simulations to have a number of misaligned states that are both appreciably stable and contribute to hybridization and de-hybridization pathways as intermediates. The role of these states is diminished by base pairs that disrupt the repetitive sequence. In the special case of a sequence with a strong core and weaker extremities, metastable "frayed" states are also observed. The presence of these states is assessed experimentally trough the analysis of relaxation timescales observed in T-jump IR spectroscopy; fast relaxation is attributed to fraying equilibria, and a broader fast relaxation to an increased heterogeniety due to the presence of mis-aligned intermediates.}

\textcolor{blue}{The paper is generally well-written, and, I believe, represents a good attempt to tackle this important process. I personally have had a manuscript "desk-rejected" by JACS on the basis that it is "too focussed on DNA biotechnology", and "primarily of interest to RNA and DNA chemists". The same statements could be made about this paper. However, I disagree with the conclusion that such papers are not appropriate for JACS. I would therefore recommend publication, subject to satisfactory responses to the comments below.  }

\textbf{Author Reply:}   We are delighted by the reviewer's warm reception of our paper, and pleased that they find it to be appropriate for publication in JACS.

\textcolor{blue}{1.   I was frequently confused as to the temperatures used in the experiments and analysis. At one point, it is noted that simulations are performed at melting temperatures. However, simulations are actually performed at a range of temperatures and it isn't always clear when looking at a figure which temperatures were used for simulation, NN model comparison or experiments (see eg. Figs 2 and 3 and 5). I would recommend noting the temperature used to obtain the data in each case.}

\textbf{Author Reply:}   We thank the reviewer for this suggestion and regret our lack of clarity in this matter. All simulations performed in this work were conducted at the melting temperature, with the exception of the simulations discussed in Section 3.1 where we consider a range of temperatures in order to draw comparisons to experimental measurements over this temperature range. We have added a number of clarifications to the text in order to more clearly convey this important point to the reader.

In Section 2.1.1 of the Methods on p.~7, we have added:

``\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.''

In Section 2.1.2 of the Methods on p.~8, we have added:

``In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective melting temperatures} from the 40$\times$25 simulation trajectories following a four-step protocol...''

In the first sentence of Section 3.3 on p.~21, we have added:

``We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.''

In the first sentence of the caption of Fig.~2, we have added:

``Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures} for AT-all, GC-end, GC-core, and GC-mix.''

In the first sentence of the caption of Fig.~2, we have added:

``Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.''

In the first sentence of Section 3.4 on p.~26, we have added:

``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.''

In the second sentence of the Conclusions in Section 4 on p.~36, we have added:

``We conducted 1 ms of unbiased coarse-grained molecular dynamics simulations \rood{at the melting temperature of each sequence} and...''

\textcolor{blue}{2. P7 line 13 - surely these simulations were not really done at 7M? That sounds impossibly dense.}

\textbf{Author Reply:}   We regret our (embarrassing) typo -- the concentration is in fact 7 mM, not 7 M. The calculation is below and we have made the correction to Section 2.1.1 on p.~7. We regret our error.

\begin{equation}\label{}
\frac{2 \mathrm{\:molecules}}{\mathrm{box\:volume}} = \frac{2 \mathrm{\:molecules}}{(7.80 \times 7.80 \times 7.80) \mathrm{\:nm}^3} = \frac{3.32 \times 10^{-24} \mathrm{\:moles}}{4.75 \times 10^{-25} \mathrm{\:m}^3} = 7.0 \frac{\mathrm{moles}}{\mathrm{m}^3} = 7.0 \frac{\mathrm{millimoles}}{\mathrm{L}} = 7 \; \mathrm{mM}
\end{equation}

\textcolor{blue}{3.  The authors repeatedly refer to "slithering" mechanisms for rearrangement. I believe this is unfortunate. In the context of the 3SPN model, the "slithering" term was first introduced in Ref. 49 by Sambriski et al, and describes two helical single strands sliding along each other. As noted by Hinckley et al in Ref 18, that mechanism is largely an artifact of the form of the potential in 3SPN.1. The rearrangement mechanism in 3SPN.2 is more of a "defect diffusion", analogous to the "inchworm mechanism" previously identified in the oxDNA model (ref 52 - note, the author list for this reference is wrong). Given this context, it seems unnecessarily confusing to talk about "slithering".}

\textbf{Author Reply:}    (I) We share the reviewer's concerns with using ``slithering'' to describe a process that is more appropriately referred to as ``defect diffusion'', ``inchworm mechanism'', or ``out-of-register diffusion''.

In Section 2.1.2 on p.~9, we have modified the text to replace ``slithering'' with ``out-of-register diffusion'':

``An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms.~\citep{Xiao2019}''}

The only other place we use the term is in Section 3.4 on p.~29 where we refer to work conducted in Ref.~\citep{Xiao2019}. Here we choose to maintain the term ``slithering'' to maintain consistency with the mechanism nomenclature adopted in this paper, but place it in quotation marks to indicate that we are quoting from that paper and it is not a term that we would necessarily favor in the context of our own work.

\rood{``Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism. (In contrast, out-of-register slithering and in-register zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.)'' }

(II) We also appreciate the reviewer pointing out the incorrect author list in ref 52, and we have fixed the issue.

\textcolor{blue}{4. P15: I am a bit surprised that a separation of 1.3nm between the central base pairs is enough to guarantee strand separation, given that frayed end can separate by 1.3nm when the core is intact. Can the authors verify that all their duplexes had truly separated at this point?}

\textbf{Author Reply:}   This a good point of clarification. We motivated our choice of this distance as that beyond which the strands are effectively non-interacting, but also verified that our estimates of $k_d^{slow}$ are robust to choices of this cutoff in the range 0.8--2.5 nm. We illustrate this in \textbf{Figure R1} below where we show that our extracted values of $k_d^{slow}$ are robust in changing the cutoff from 1.3 nm to 2.0 nm.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{FigR1.pdf}
        \caption*{\textbf{Figure R1.} Comparing the experimental slow response when increasing the cutoff parameter from 1.3 nm to 2.0 nm}
        \label{fig:compare_cutoffs}
	\end{center}
\end{figure}

We have modified the relevant text in Section 3.1 on p.~15 to read:

``First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex. This cutoff was selected as the distance beyond with the strands are effectively non-interacting and define the dissociated state. \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.}''

\textcolor{blue}{5. The authors refer to "exponential" increase in Fig. 1. However, it's hard to see that an increase is exponential by eye unless the data is plotted on a log scale.}

\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1 and Fig.~S\rood{4} on log-linear axes and reproduce these below.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Fig1.pdf}
        \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses. Results are reported in terms of the final T-jump temperature. (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales. \rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales. The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics. The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

\clearpage
\newpage

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.75\textwidth]{FigS4.pdf}
   \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
    \label{fig:SIFig4}
\end{figure}

\clearpage
\newpage

\textcolor{blue}{6. Do the authors have any data on the scaling of k\_hybridization with T? In experiment, rates appear to increase with temperature (albeit with a much weaker dependence than de-hybridization rates). Simple "nucleation and zipper" models would predict the opposite, as does the oxDNA model (see ref 52).}

\textbf{Author Reply:}   This is an interesting question and are, in fact, considering embarking upon some additional work in this vein. We do not, however, have any data on this scaling at the current time. Rigorous engagement of this question would require us to fit MSMs at a variety of temperatures in order to compare the pathways, mechanisms, and rates for hybridization and dehybridization. In the present work, we have only constructed MSMs at the melting temperature of each sequence and the collection of sufficient data to parameterize MSMs at other temperatures would require both (i) a substantial investment of time to collect the $\sim$1 ms of data required to fit robust MSMs and (ii) the use of biasing techniques in order to adequately sample hybridization (dehybridization) events at large temperature excursions above (below) the melting temperature. As such, we do not have the necessary data to engage this question in the present work, but would like to explore it in the future.

Reviewer \#2 raised a related point in their Comment \#2 and we have made the following modification to the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting follow-up study:

``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons. \rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''

\textcolor{blue}{7. The authors could explain a bit more carefully (perhaps in the SI) how they handle the multiple nearest-neighbour states that correspond to their microstates (eg. F4 clearly involves more than one NN configuration).}

We have modified the text in the ``Nearest neighbor model of duplex thermodynamics'' text in the Supporting Information to better describe for the reader how we performed this calculation. Specifically, we have appended the following text at the end of this discussion:

\noteg{Have not incorporated this yet, but this would be one way to address it without making the FE calculations significantly (and unnecessarily) more complicated}
``\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a. Indeed, each macrostate is composed of an ensemble of partially frayed stayed with higher free energy contributions than the intact pattern. However, when looking at the difference between states, we have found that incorporating frayed contributions changes 
$\Delta$ F values by $<$ 1 kJ/mol when compared to used a single macrostate. Given the extent to which a single NN representation captures each macrostates, we have elected to maintain a single microstate free energy estimate for each macrostate}''

\noteg{Can also include a figure for the F4 or shifted states showing a minimal difference.}

\textcolor{blue}{8. I don't understand how the population of F4 could be >50\% that of H for the GC-core system, when the free energy of F4 is reported as > 5kJ/mol. My back-of-the-envelope maths predicts a ratio of 8:1 for a 5kJ/mol gap. Incidentally, 8:1 is roughly the ratio I got from running the configurations through NUPACK.}

\textbf{Author Reply:}  \noteg{I added everything below, not sure if this is adequately adressing the mistake} We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence. Specifically, we did not convert the $\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta$ F (13.7 kJ/mol) relative to the hybridized state.

We recognize that this is a significant miscalculation, and we have re-checked all free energy results to ensure it was an error for the GC-core free energy calculations only and did not affect the reported values for GC-end and AT-all. Furthermore, we have verified that the error does not propagate to any other relevant results in the manuscript.

We have updated Figure 3 with the correct $\Delta$ F  as shown below.

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$. We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model. (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model. The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
        \label{fig:NN_table}
	\end{center}
\end{figure}

We have also updated the second paragraph of section 3.3 as follows:

As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.

\noteg{The line above is replacing the following: ``As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 relative to the H state compared to the NN model. The NN and MSM predictions for the frayed state F4 are in good agreement.''}

\noteg{Do we still want to include these calculations or just say that it was wrong?}

Our back-of-the-envelope calculations yield a similar result based on the $\Delta F$ values reported in Fig.~3. In brief, assuming:

\begin{itemize}
\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
\item $G_F$ = 5 kJ/mol = 2 kT
\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
\end{itemize}

we can solve this system of equations to yield $f_H$ + $f_D$ = 0.47 and $f_F$ = 0.06, such that $f_H$ / $f_F$ = 7.8. A more rigorous accounting of the state populations by analyzing the MSM predictions yields a slightly lower prediction of $f_F$ = 0.0028 that we report in Fig.~2b.

We believe that the root of this comment may be our choice of a logarithmic y-axis for the bar plots in Fig.~2b that make the bar associated with the F4 state appear nearly 50\% of the height of the H and D bars for GC-core. We made this plotting choice in order to make visible on these bar plots the small state populations associated with the shifted and frayed states and noted the percentage occupancies on the bars themselves. To help eliminate this confusion, we have modified the portion of the Fig.~2 caption referring to this panel to read:

``(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. \rood{Values are reported on a log y-axis to make the small populations of the shifted and frayed states visible.}''

\textcolor{blue}{9. Is the MSM model explicitly constrained to obey detailed balance? The numbers that come out seem consistent with detailed balance - forward and backward trajectories at the same T essentially look the same (they're equally likely to go through intermediates). This should be a feature of the MSM, but it's unclear to me whether it's hard-coded or just arises naturally in the coarse-graining.}

\textbf{Author Reply:}    The reviewer raises a good point regarding the assumptions made during MSM construction. We do indeed use the reversible Bayesian MSM framework described here \url{http://www.emma-project.org/latest/api/generated/pyemma.msm.BayesianMSM.html} which enforces detailed balance. This is an important point which we have now made explicit by modifying the text in Section 2.1.2 on p.~11 to read:

``The 200 microstates comprising each system were coarsened into our terminal macrostate MSM. \rood{Since we are interested in the construction of models of the equilibrium kinetics, we explicitly enforce detailed balance in the construction of the MSM. \citep{Wehmeyer2019IntroductionSoftware}}''

and the text in Section 3.4 on p.~26 to read:

``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms. \rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.''}

\textcolor{blue}{Relatedly, some of the discussion of the hybridization and de-hybridization pathways in non-repetitive sequences seems to ignore this fact. In these equilibrium simulations, hybridization trajectories played in reverse should be statistically indistinguishable from de-hybridization trajectories. I don't think anything that's said is wrong, per se, but it might be worth considering how the contrast between "nucleation and zippering" and "fraying-peeling" is consistent with the symmetry in the pathways.}

\textbf{Author Reply:}  This is a nice observation, and we would do well to note this symmetry for the reader. We have added a sentence to this effect in Section 3.6 on p.~31:

``In dehybridization, we observe fraying \rood{of one half of the duplex with the strands remaining associated until the loss of the final A:T pairs and ultimately the last G:C pair}. Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes. These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility \cite{McCully2008MicroscopicHomeodomain}. A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics. This is indeed borne out by our mechanistic observations for GC-mix wherein the early stages of ``nucleation-zippering'' proceed by the formation of one G:C contact followed by one or more additional A:T pairs and the late stages by the formation of all remaining WC pairs, which we compare with the early stages ``fraying-peeling'' wherein one half of the duplex frays and the late stages wherein dissociation finally completes by the dissolution of the last few A:T contacts and the final G:C pair.}''

\textcolor{blue}{10. I think some of the characterisation of the "all or nothing" picture is a bit misleading. The two state model of duplex formation is really a thermodynamic model, based on the assumption that the ensemble of hybridized and de-hybridized states look very different and are separated by a large barrier. This model is perfectly consistent with the transition between the two ensembles involving a complex network of metastable states. Indeed, as the authors note on line 3 of p4, a number of previous authors have considered what such states might look like, particularly for repetitive sequences. The authors' primary contribution here is not proposing that these states exist for the first time, but trying to tie together evidence from simulation and experiment to describe them.}

\textbf{Author Reply:}   We thank the reviewer for this important observation and illuminating our need to better clarify this potentially ambiguous concept. We have modified the text in Section 3.6 on p.~29 to read:

\noteg{Added sentence starting with ``Indeed, the sequence...'' Not totally sure about this, but I don't think we need to remove all references to all-or-nothing}
``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM. \rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time. (We note that ``all-or-nothing'' can also be used, somewhat ambiguously, to refer to a two-state thermodynamic model where the hybridized and dehybridized are separated by a large free energy barrier but that the transitions between them may proceed through a network of metastable states.)}''

\noteg{Brennan's recommendation:}
We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state.''

``FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a \rood{two-state model} to determine the fraction of intact duplex $\theta$ as a function of temperature.''

\textcolor{blue}{11. Related to the above, the references on line 3 of p4 should probably also include Ref 52, where a similar analysis of metastable states in a repetitive sequence is performed (for a different coarse-grained model) and [Flamm, W., Fontana,J.M., Hofacker, I.L. and Schuster,P. (2000) RNA folding at elementary step resolution. RNA, 6, 325-338] where a defect diffusion mechanism is proposed.
}

\textbf{Author Reply:}   We have added this reference.

\textcolor{blue}{12. The reasoning behind why the broader relaxation spectrum is indicative of misaligned states was not that clear to me. Why would "elevated" fraying cause a broader response? Wouldn't it also, then, be broader for CG-core, which has elevated fraying? It seems to me that the argument is more along the lines of the heterogeneity of the configurations leading to heterogeneous dynamics, and hence a broader relaxation profile. But I am no expert on this technique.}

\textbf{Author Reply:}   The reviewer makes a good comment on our choice of language used to described the stretched AT-all response. It is indeed the heterogeneity of the configurations in the initial ensemble, and the fraying dynamics exhibited in each of these states, that is the origin of the broadening of the relaxation profile, but our wording here was insufficiently clear. We believe that the reviewer's apt suggestion ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into our edits of Section 3.7 on p.~33:

``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying. Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying. This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix. It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.''

\textcolor{blue}{13. How much of the code is being made available? Ideally, the initialization files for the simulations and the code used to perform the MSM analysis would be publicly available. Additionally, it would be good to publish the raw experimental data and the processing tools used to obtain the final results.}

\textbf{Author Reply:}   As a computationally-focused paper, we agree that the data and analysis code should be made available to the reader. We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}). The links to both of these resources are now noted in the Supporting Information statement on p.~38:

``\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).}''

\clearpage
\newpage

\begin{shaded}
\textbf{Additional editorial edits}
\end{shaded}

\textcolor{blue}{1. Please include at least the first page for each reference citation for the following incomplete journal reference 48}

Updated.

\textcolor{blue}{2. Please update the publication status of reference 70, if possible.  Providing the DOI is acceptable.}

We are unable to locate any additional citation information beyond the arXiv link.

\textcolor{blue}{3. Please include publisher for the following incomplete book reference: 79.}

Updated.

\textcolor{blue}{4. The TOC graphic should fit in an area no larger than 3.25 in. by 1.75 in. (approx. 8.25 cm by 4.45  cm) and should have adequate resolution and clarity. Confirm that all text is legible at this size.}

Updated and confirmed. \noteg{Pick a TOC that will fit in those dimensions}

\clearpage
\newpage

\bibliographystyle{achemso}
\bibliography{references_210514_abv}
\end{document}

### 2021-07-27 SI_marked_up.tex
just working on 3.7 and 3.8

--- 2021-05-20 SI.tex
@@
 \newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\newcommand*{\rood}[1]{{\color{red}{#1}}}
 
-%\newcommand*{\rood}[1]{{\color{red}{#1}}}
-%\newcommand*{\rood}[1]{#1}
 %\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
@@
 	\centering
-    \includegraphics[width=.70\textwidth]{FigS4.pdf}
+    \includegraphics[width=.6\textwidth]{FigS4.pdf}
     \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
 We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
-Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.}
+Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
+In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
     \label{fig:SIFig4}
@@
     \label{fig:SIFig6}
+\end{figure}
+
+\clearpage
+\newpage
+
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
+        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis.
+The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
+(a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying.
+We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction.
+This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction.
+We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands.
+It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
+We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift.
+We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
+(b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence.
+The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
+(c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
+This panel conveys two important pieces of information.
+First.
+each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
+These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
+Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex.
+Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
+A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green).
+A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
+Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
+The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
+(d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments.
+The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
+        %\label{fig:macrostate_colors_FES}
+	\end{center}
 \end{figure}
@@
 
-\begin{comment}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=450, scale=1]{Figs/figs_imp/GC-core_tracking_modes.pdf}
-        \caption{Diffusion map representation configurations sampled from the F4 state.
-The x and y axes represent the the first two non-trivial diffusion map eigenfunctions.
-Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
-        \label{fig:GC-core_dmaps}
-	\end{center}
-\end{figure}
-
-\end{comment}
-
-% shift to include this in the overall diffusion map section
-%\subsubsection{Diffusion maps show ensemble of frayed states}
-
-%We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate.
-We build diffusion maps using 10000 frames sampled from the macrostate (Figure \ref{fig:GC-core_dmaps}).
-Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes.
-We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate.
-This distance also correlated closely with the second diffusion map mode.
-Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in ``overlap'' distance between adjacent A or T bases and the GC core.
-We defined overlap as the distance between a G/C and the A/T adjacent to its native complement.
-At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place.
-Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation.
-Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.
-%\subsection{Explicit model comparisons}
-
-%We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation.
-In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3SPN2 explicit implementation \citep{Hinckley2015}.
-These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities.
-Comparisons were limited by the higher computational demand required to simulate 5x the beads in the box, a shorter integration timestep, and reliance on Ewald summation to account for electrostatics.
-We caution over-interpretation of these results as the explicit model has not been as widely adopted or thoroughly validated compared to the implicit model.
-
-%\begin{figure}[ht!]
-%	\begin{center}
-%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.pdf}
-%        \caption{Full 5$^\prime$ diffusion maps including degenerate third diffusion map mode.}
-%        \label{fig:GC-end_dmaps_full}
-%	\end{center}
-%\end{figure}
-
-%\begin{figure}[ht!]
-%	\begin{center}
-%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_3prime.pdf}
-%        \caption{Full 3$^\prime$ diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct ``looping'' region for intact GC bonds}
-%        \label{fig:GC-end_dmaps_3prime}
-%	\end{center}
-%\end{figure}
-
-%add anymore figures for the time analysis: As shown bleow:
-\begin{comment}
-
-\subsubsection{\label{sec:Results}Code to provide on GitHub} 
-\begin{enumerate}
-    \item Cross val data
-    \item lammps run script and each sequence input files
-    \item Featurization with mdtraj (reindexing and sparse saving)
-    \item Dmaps for GC-core and GC-end
-    \item timescales comparison
-    \item SRV cross valss
-    \item MSM construction
-
-\end{enumerate}
-\end{comment}
-
-\begin{comment}
-
-\section{\label{sec:Discussion}Discussion}
-
-\citep{Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}
-Long simulations shows that terminal fraying does not converge on the picosecond timescale
-
-\citep{Nonin1995TerminalFraying}
-Most simulations suppress fraying by placing GC pairs at termini.
-
-\citep{Andreatta2006UltrafastHelix}
-- Base opening process is slower than 40 ns but faster than 1ms
-- Open termini state shows new fraying mode of 5ps (I think shows free base motions)
-\end{comment}
-
-\begin{comment}
-
-%%%%%% Other MSM and timescales approaches  %%%%%%%%%%%
-
-McGibbon, R.
-T., & Pande, V.
-S.
-(2015).
-Variational cross-validation of slow dynamical modes in molecular kinetics : Importance of not generating too many microstates and employting cross-validation to avoid artificially high scores or artificially slow dynamics.
-Both of these lead to models fitting statistical noise which is the root of boosted scores.
-
-Pinamonti, G., Paul, F., Rodriguez, A., & Bussi, G.
-(n.d.).
-analyzed with core-set Markov state models, 43: Good ideas on core-based clustering.
-Assigning microstates based on physical characteristics is interesting but didn't seem to work for me in practice (could try doing this based on sim.log data with number of base pairs bound or some energy metric).
-Lot's of notes on this paper regarding comparisons to timescales.
-Good notes on higher resolution fraying interactions that we likely can't resolve.
-
-Sua, E., Adelman, J. L., & Zuckerman, D. M. (2016). Accurate Estimation of Protein Folding and Unfolding Times : Beyond Markov State Models, 1. https://doi.org/10.1021/acs.jctc.6b00339:
-More accurate ways to adapt MFPT into a rate
-
-Refs for DNA fraying:
-\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro).
-Use TPS to focus on kinetic pathway by which singl base pair bind/unbind.
-Focus on end base of 3-bp oligomer in explicit solvent.
-Finds order parameters tracking base pair binding as well as intra-strand stacking.
-(energy and distance for each).
-
-\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,
-
-6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
-37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
-38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding
-
-\end{comment}
-
 \clearpage

### 2021-07-27 manuscript_marked_up.tex
just working on 3.7 and 3.8

--- 2021-05-20 manuscript.tex
@@
 Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical ``all-or-nothing'' two-state model of hybridization/dehybridization with no intermediate metastable states.
-Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation for DNA nanotechnology applications.
+Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization \rood{by furnishing quantitative and predictive kinetic models of the dynamical transition network between metastable states}, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation for DNA nanotechnology applications.
 
@@
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}.
+\rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable Markov state models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling.
+A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}.
+The application of these methods under the conditions of high bias necessary for good sampling can, however, present challenges for numerical convergence.} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
+\rood{In this work, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}
 
@@
 All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model.
-\citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L.
-Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of 9.94 $\times$ 10$^{-11}$ m$^2$/s.
+\citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of \rood{7 mM}.
+\rood{This concentration is only 3.5$\times$ larger than the 2 mM concentration employed in our experimental analyses (Section 2.2).} Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of 9.94 $\times$ 10$^{-11}$ m$^2$/s.
 \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius.
@@
 \citep{Schneider1978Molecular-dynamicsTransitions}.
-Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
 Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
@@
 In this manner, they provide an interpretable and predictive model of the system thermodynamics and kinetics.
-MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms}.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective melting temperatures} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to high-resolution Markovian macrostate MSMs.
-We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits.
-We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
+We guarded against overfitting using five-fold cross-validation in which we \rood{constructed five random partitions of the 40 independent simulation trajectories into a training set of 20 trajectories and a validation set of 20 trajectories}.\citep{Sidky2019High-ResolutionVAMPnets} We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card.
 A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.
@@
 \textbf{Macrostate clustering.} The 200 microstates comprising each system were coarsened into our terminal macrostate MSM.
-The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionSoftware} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
+\rood{Since we are interested in the construction of models of the equilibrium kinetics, we explicitly enforce detailed balance in the construction of the MSM.
+\citep{Wehmeyer2019IntroductionSoftware}} The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionSoftware} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}.
 We observe rapid convergence of the implied time scales $t_i$ with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns.
@@
 
-FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex $\theta$ as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
+FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a \rood{two-state model} to determine the fraction of intact duplex $\theta$ as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
 	\begin{equation}\label{e3a}
@@
 We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments.
-We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
+We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively}.
 The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
@@
 First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
-This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
-We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
+\rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to correct our predictions for $f_\mathrm{hybridized}$ bulk yields.
+We correct the values observed in our simulations -- $f_\mathrm{hybridized}$ = $f_\mathrm{hybridized}^\mathrm{sim}$ -- to the bulk corrected result -- $f$ = $f_\mathrm{hybridized}$  -- using the following relations:}
+
+    \begin{equation}\label{}
+    \phi = \frac{\mathrm{f_\mathrm{hybridized}^\mathrm{sim}}}{\mathrm{1 - f_\mathrm{hybridized}^\mathrm{sim}}}
+    \end{equation}
+
+    \begin{equation}\label{}
+	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
+	\end{equation}
+
 Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction.
-We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
+We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_{\mathrm{unfrayed}}(t) = \exp(-k_d^\mathrm{fast}t)$.
 
@@
 We apply these sequence-independent scaling factors to our reported computational values in Fig.~\ref{fig:relaxation-comparison}.
-Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective \rood{temperature shift} to our computational results.
 We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
-The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.
+The computational time scales and melting temperatures reported in the remainder of the paper are not corrected \rood{by these calibration corrections or finite-size fluctuation corrections} since we are only interested in the relative trends in the behaviors of the four sequences.
 
@@
 (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-(b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
+\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
 The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
@@
 Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements.
-The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.
+The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers \rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence.
+We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}.
+These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}.
+We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected.
+Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}.
 
@@
 We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
-In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
-By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F).
+\rood{The representative microstates emblematic of each macrostate were selected randomly from those possessing high ($>$99\%) macrostate membership probabilities.
+Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution.
+In Fig.~S{\blauw{X}} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
+In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
+By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F\rood{4}).
 In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates.
@@
 \begin{figure}[ht!]
-        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix.
+        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures} for AT-all, GC-end, GC-core, and GC-mix.
 (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D).
@@
 Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
-(c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
+\rood{Values are reported on a log y-axis to make the small populations of the shifted and frayed states visible.} (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence.
 The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a.
@@
 
-We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
 The NN model predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
@@
         \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
 We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
@@
 
-In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms.
-We now proceed to analyze these predictions for each of the four sequences and illuminate the relationship between sequence and dynamics.
+In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.
+\rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.} We now proceed to analyze these predictions for each of the four sequences and illuminate the relationship between sequence and dynamics.
 Two of our sequences, AT-all and GC-end, support out-of-register metastable states, and we commence our analysis with the role of these shifted states.
@@
 
-The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
-The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+\rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
+(We note that ``all-or-nothing'' can also be used, somewhat ambiguously, to refer to a two-state thermodynamic model where the hybridized and dehybridized are separated by a large free energy barrier but that the transitions between them may proceed through a network of metastable states.)} The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
 We note that we do observe substantial transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
@@
 During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another and the formation of key native WC contacts immediately prior to duplex formation: specifically, one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
-This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
+This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying \rood{of one half of the duplex with the strands remaining associated until the loss of the final A:T pairs and ultimately the last G:C pair}.
 Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
-These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}
+These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility \noter{Add ref: 10.1021/bi800118b}.
+A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics.
+This is indeed borne out by our mechanistic observations for GC-mix wherein the early stages of ``nucleation-zippering'' proceed by the formation of one G:C contact followed by one or more additional A:T pairs and the late stages by the formation of all remaining WC pairs, which we compare with the early stages ``fraying-peeling'' wherein one half of the duplex frays and the late stages wherein dissociation finally completes by the dissolution of the last few A:T contacts and the final G:C pair.}
 
@@
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
 It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
@@
 We have conducted an integrated computational and experimental study of the sequence-dependent kinetic mechanisms for the hybridization and dehybridization dynamics of a family of four self-complementary 10-mer DNA oligomers: ATATATATAT (AT-all), GATATATATC (GC-end), ATATGCATAT (GC-core), and ATGATATCAT (GC-mix).
-We conducted 1 ms of unbiased coarse-grained molecular dynamics simulations and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
+We conducted 1 ms of unbiased coarse-grained molecular dynamics simulations \rood{at the melting temperature of each sequence} and employed deep learning techniques to construct high-resolution Markov state models as predictive and interpretable models of the sequence dependent dynamics.
 T-jump IR spectroscopy was used to calibrate the kinetic time scales of the coarse-grained molecular model and validate the kinetic prediction of the Markov state models that the AT-all sequence should possess long-lived out-of-register shifted states that are detectable within T-jump IR t-HDVE time traces.
@@
 Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
+\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}
+, may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.} We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
 
@@
 Details of nearest neighbor (NN) model calculations, supplementary figures illustrating selection of optimal number of SRV slow modes for each sequence, convergence of MSM implied time scales, MSM Chapman-Kolmogorov tests, ``computational T-jump'' calculations, and GC-mix hybridization and dehybridization trajectories.
+\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).}
 

### 2021-07-28 cover_letter/response_letter_MJ_210721.tex
updated figS7

@@
 The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
+Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively.
 (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
@@
 In all cases except the lowest temperature (T$_m$-5) slow response we report good fits with $R^2$ $>$ 0.88.
-We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good \noteg{not sure what we're reporting here?
-\rood{X.XX} $<$ $R^2$ $<$ \rood{X.XX})}.
+We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.
 
@@
 ``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possible coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels} , may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
+\rood{We also suggest that the MSM approach followed in this work, possible coupled with biased sampling and reweighting techniques, \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels} may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
 
@@
 The revised figure is reproduced below.
-\noteg{We note that it is unclear how this correction should be applied to a periodic system that is not in thermodynamics equilibrium, however we choose to apply the correction so that we will observe asymptotically correct behavior at long timescales.}
 
@@
 
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to correct our predictions for $f_\mathrm{hybridized}$ bulk yields.
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
 We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}
@@
     \begin{equation}\label{}
-    \phi = \frac{\mathrm{f_\mathrm{hybridized}^\mathrm{sim}}}{\mathrm{1 - f_\mathrm{hybridized}^\mathrm{sim}}}
+    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
     \end{equation}
@@
 
-\noteg{Have not incorporated this yet, but this would be one way to address it without making the FE calculations significantly (and unnecessarily) more complicated}
 ``\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-Indeed, each macrostate is composed of an ensemble of partially frayed stayed with higher free energy contributions than the intact pattern.
-However, when looking at the difference between states, we have found that incorporating frayed contributions changes
-$\Delta$ F values by $<$ 1 kJ/mol when compared to used a single macrostate.
-Given the extent to which a single NN representation captures each macrostates, we have elected to maintain a single microstate free energy estimate for each macrostate}''
+Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
+However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes
+$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate.
+Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}''
 
-\noteg{Can also include a figure for the F4 or shifted states showing a minimal difference.}
+\noteg{Can add some more text here include a figure for the F4 or shifted states showing a minimal difference.}
 
@@
 
-\textbf{Author Reply:}  \noteg{I added everything below, not sure if this is adequately adressing the mistake} We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence.
-Specifically, we did not convert the $\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta$ F (13.7 kJ/mol) relative to the hybridized state.
+\textbf{Author Reply:}  We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence.
+Specifically, we did not convert the $\Delta\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta\Delta$ F (13.7 kJ/mol) relative to the hybridized state.
+Using this correct $\Delta\Delta$ F value we see agreement between the reported free energy in Fig~3 and populations reported in Fig.~2.
+In brief, we perform the following back-of-the-envelope verification:
 
-We recognize that this is a significant miscalculation, and we have re-checked all free energy results to ensure it was an error for the GC-core free energy calculations only and did not affect the reported values for GC-end and AT-all.
+\begin{itemize}
+\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
+\item $G_{F4}$ = 5.1 KT
+\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
+\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
+\end{itemize}
+
+We can solve this system of equations to yield $f_{F4}$ = 0.0030.
+A more rigorous accounting of the state populations by analyzing the MSM predictions yields a slightly lower prediction of $f_{F4}$ = 0.0028 that we report in Fig.~2b.
+
+We recognize that this was a significant miscalculation, and we have re-checked all free energy results to ensure it was an error for the GC-core free energy calculations only and did not affect the reported values for GC-end and AT-all.
 Furthermore, we have verified that the error does not propagate to any other relevant results in the manuscript.
@@
 
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
-
-\noteg{The line above is replacing the following: ``As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 relative to the H state compared to the NN model.
-The NN and MSM predictions for the frayed state F4 are in good agreement.''}
-
-\noteg{Do we still want to include these calculations or just say that it was wrong?}
-
-Our back-of-the-envelope calculations yield a similar result based on the $\Delta F$ values reported in Fig.~3. In brief, assuming:
-
-\begin{itemize}
-\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
-\item $G_F$ = 5 kJ/mol = 2 kT
-\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
-\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
-\end{itemize}
-
-we can solve this system of equations to yield $f_H$ + $f_D$ = 0.47 and $f_F$ = 0.06, such that $f_H$ / $f_F$ = 7.8.
-A more rigorous accounting of the state populations by analyzing the MSM predictions yields a slightly lower prediction of $f_F$ = 0.0028 that we report in Fig.~2b.
+As illustrated in Fig.~3, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
 
@@
 \noteg{Added sentence starting with ``Indeed, the sequence...'' Not totally sure about this, but I don't think we need to remove all references to all-or-nothing}
-``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
 \rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
@@
 
-\noteg{Brennan's recommendation:}
 We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state.''

### 2021-07-28 SI_marked_up.tex
updated figS7

@@
 
+\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
+However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes
+$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate.
+Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}
+
 \vspace{+0.5in}

### 2021-07-28 manuscript_marked_up.tex
updated figS7

@@
 \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to correct our predictions for $f_\mathrm{hybridized}$ bulk yields.
-We correct the values observed in our simulations -- $f_\mathrm{hybridized}$ = $f_\mathrm{hybridized}^\mathrm{sim}$ -- to the bulk corrected result -- $f$ = $f_\mathrm{hybridized}$  -- using the following relations:}
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
+We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}
 
     \begin{equation}\label{}
-    \phi = \frac{\mathrm{f_\mathrm{hybridized}^\mathrm{sim}}}{\mathrm{1 - f_\mathrm{hybridized}^\mathrm{sim}}}
+    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
     \end{equation}

### 2021-07-29 SI_marked_up.tex
Andy edits on response letter through p. 10

@@
 \usepackage{algpseudocode}
+\usepackage[font={small,sf}, singlelinecheck=false]{caption}
+\usepackage{capt-of}
 
@@
 	\centering
-    \includegraphics[width=.6\textwidth]{FigS4.pdf}
+	\includegraphics[width=.97\textwidth]{FigS4.pdf}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
     	\caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
     \label{fig:SIFig4}
+    	\end{center}
 \end{figure}
@@
 	\begin{center}
-        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis.
+        \caption*{Figure S7.
+\rood{Macrostate interpretation and structural analysis.
 The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
@@
 This panel conveys two important pieces of information.
-First.
-each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
+First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
 These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
@@
 Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
-A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green).
+A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
 A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
 Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
-The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
 The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.

### 2021-07-29 manuscript_marked_up.tex
Andy edits on response letter through p. 10

@@
 
-%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
-%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
 %\newcommand*{\noter}[1]{}					% notes off
-%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
+\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
 
@@
 Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements.
-The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers \rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence.
+The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers\rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence.
 We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}.

### 2021-07-29 response_letter/response_letter.tex
created response dir

--- 2021-07-28 cover_letter/response_letter_MJ_210721.tex
@@
 
-We also added the new Fig.~S7 \noteg{is this order fine?} to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer.
+We also added the new Fig.~S7 to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer.
 The figure and its caption are reproduced below.
@@
 
-\noteg{Can add some more text here include a figure for the F4 or shifted states showing a minimal difference.}
-
 \textcolor{blue}{8.
@@
 
-\noteg{Added sentence starting with ``Indeed, the sequence...'' Not totally sure about this, but I don't think we need to remove all references to all-or-nothing}
 ``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
@@
 
-Updated and confirmed. \noteg{Pick a TOC that will fit in those dimensions}
+Updated and confirmed.
 

### 2021-07-30 response_letter/response_letter.tex
new directory, reordered SI

@@
 Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution.
-In Fig.~S{\blauw{X}} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
+In Fig.~S5 we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
 In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig.
@@
 
-We also added the new Fig.~S7 to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer.
+We also added the new Fig.~S5 to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer.
 The figure and its caption are reproduced below.
@@
 	\centering
-    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
+    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS5.pdf}
 \end{figure}
@@
 	\begin{center}
-        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis.
+        \caption*{\textbf{Figure S5.}\rood{Macrostate interpretation and structural analysis.
 The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
@@
 
-textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences.
+\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences.
 As suggested by the reviewer, we added two additional modes to the cross-validation procedure that does indeed better expose this kink at five slow modes.
@@
 
-\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14.
+\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 15.
 We agree that the sentence on (P.
-14 L.
-44) is unclear in its current form and have made the following change:
+15 L.
+6) is unclear in its current form and have made the following change:
 
@@
 The revised figure is reproduced below.
-In all cases except the lowest temperature (T$_m$-5) slow response we report good fits with $R^2$ $>$ 0.88.
+In all cases except the lowest temperature (T$_m$-5) slow response fits (where dissociation data is sparse) we report good fits with $R^2$ $>$ 0.88.
 We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.
@@
 Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
+\rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
 In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
@@
 As such, these motions lie below the $\tau$ = 1.2 ns lag time used to construct our MSMs that is required to converge the leading slow modes and produce a Markovian model capable of passing the Chapman-Kolmogorov test (cf.
-Figs.~S\rood{2-3}).
+Figs.~S2-3).
 As such, we cannot draw comparisons of our MSM against the fast experimental data.
@@
 We concur with the reviewer that we can and should explicitly mention dynamical fingerprinting approach and provide some discussion for the reader why we did not employ this approach in the present study.
-To this end, we have added the following text to the end of Section 3.1:
+To this end, we have added the following text to the end of Section 3.1 (p.18):
 
@@
 
--- Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1, Fig.~S\rood{4}) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S\rood{1-3}).
+-- Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1, Fig.~S4) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S1-3).
 This indicates that the data are of sufficient quality and volume to present favorable experimental comparisons and learn reliable long-time kinetic models.
@@
 
-\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1 and Fig.~S\rood{4} on log-linear axes and reproduce these below.
+\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1 and Fig.~S4 on log-linear axes and reproduce these below.
 
@@
 The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
-        \label{fig:relaxation-comparison}
+        \label{fig:relaxation-comparison_2}
 	\end{center}
@@
 In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
-    \label{fig:SIFig4}
+    \label{fig:SIFig4_2}
 \end{figure}
@@
 
-``\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+``\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
 Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
-However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes
-$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate.
-Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}''
+However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that in all cases incorporating additional NN configurations changes $\Delta$ F values (relative the corresponding partially frayed H state) by $<$ 1 kJ/mol when compared to using a single microstate.
+Furthermore, treating termini fraying as complete removal of a nearest neighbor pair is a crude approximation since the model was not intended to incorporate fraying.
+Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate}''
 
@@
 \textbf{Author Reply:}  We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence.
-Specifically, we did not convert the $\Delta\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta\Delta$ F (13.7 kJ/mol) relative to the hybridized state.
-Using this correct $\Delta\Delta$ F value we see agreement between the reported free energy in Fig~3 and populations reported in Fig.~2.
-In brief, we perform the following back-of-the-envelope verification:
+Specifically, we did not convert the $\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta$ F (13.7 kJ/mol) relative to the hybridized state.
+Using this correct $\Delta$ F value we see agreement between the reported free energy in Fig~3 and populations reported in Fig.~2. In brief, we perform the following back-of-the-envelope verification:
 

### 2021-07-30 SI_marked_up.tex
new directory, reordered SI

@@
 
-% Fig. S5
 \begin{figure}[ht!]
 	\centering
-    \includegraphics[width=0.8\textwidth]{FigS5a.pdf}
-\end{figure}
-\begin{figure}[ht!]
-	\centering
-    \includegraphics[width=0.8\textwidth]{FigS5b.pdf}
+	\includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS5.pdf}
 \end{figure}
@@
 \begin{figure}[ht!]
-	\centering
-    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
-Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events.
-Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
-Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
-Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
-Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
-    \label{fig:SIFig5}
+	\begin{center}
+        \caption{\rood{Macrostate interpretation and structural analysis.
+The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
+(a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying.
+We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction.
+This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction.
+We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands.
+It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
+We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift.
+We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
+(b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence.
+The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
+(c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
+This panel conveys two important pieces of information.
+First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
+These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
+Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex.
+Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
+A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
+A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
+Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
+The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
+(d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments.
+The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
+        %\label{fig:macrostate_colors_FES}
+	\end{center}
 \end{figure}
@@
 
-\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
 Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
-However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes
-$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate.
-Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}
+However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that in all cases incorporating additional NN configurations changes $\Delta$ F values (relative the corresponding partially frayed H state) by $<$ 1 kJ/mol when compared to using a single microstate.
+Furthermore, treating termini fraying as complete removal of a nearest neighbor pair is a crude approximation since the model was not intended to incorporate fraying.
+Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate}
 
@@
     \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
-    \label{fig:SIFig6}
+    \label{fig:SIFig7}
 \end{figure}
@@
 
+% Fig. S7
 \begin{figure}[ht!]
 	\centering
-	\includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
+    \includegraphics[width=0.8\textwidth]{FigS7a.pdf}
+\end{figure}
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=0.8\textwidth]{FigS7b.pdf}
 \end{figure}
@@
 \begin{figure}[ht!]
-	\begin{center}
-        \caption*{Figure S7.
-\rood{Macrostate interpretation and structural analysis.
-The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
-(a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying.
-We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction.
-This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction.
-We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands.
-It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
-We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift.
-We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
-(b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence.
-The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
-(c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
-This panel conveys two important pieces of information.
-First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
-These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
-Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex.
-Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
-A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
-A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
-Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
-The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
-The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
-(d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments.
-The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
-        %\label{fig:macrostate_colors_FES}
-	\end{center}
+	\centering
+    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
+Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events.
+Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
+Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
+Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
+    \label{fig:SIFig6}
 \end{figure}

### 2021-07-30 manuscript_marked_up.tex
new directory, reordered SI

@@
 We apply these sequence-independent scaling factors to our reported computational values in Fig.~\ref{fig:relaxation-comparison}.
-Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective \rood{temperature shift} to our computational results.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.
 We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
@@
 Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution.
-In Fig.~S{\blauw{X}} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
+In \blauw{Fig.~S5} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
 In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
@@
 We apply the NN model to predict the free energy $F^\mathrm{NN}$ of each of the macrostates occupied by each of the four sequences.
-A full accounting of our application of the NN model is provided in the \blauw{Supporting Information}.
+A full accounting of our application of the NN model is provided in \blauw{Fig.~S6} of the \blauw{Supporting Information}.
 The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
@@
 To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events.
-We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more in \blauw{Fig.~S5}.
+We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more in \blauw{Fig.~S7}.
 During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another and the formation of key native WC contacts immediately prior to duplex formation: specifically, one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
@@
 Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
-These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility \noter{Add ref: 10.1021/bi800118b}.
-A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics.
+These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility.
+\cite{McCully2008MicroscopicHomeodomain} A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics.
 This is indeed borne out by our mechanistic observations for GC-mix wherein the early stages of ``nucleation-zippering'' proceed by the formation of one G:C contact followed by one or more additional A:T pairs and the late stages by the formation of all remaining WC pairs, which we compare with the early stages ``fraying-peeling'' wherein one half of the duplex frays and the late stages wherein dissociation finally completes by the dissolution of the last few A:T contacts and the final G:C pair.}
@@
 Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.
-Four additional hybridization events and four additional dehybridization events are presented in \blauw{Fig.~S5}.
+Four additional hybridization events and four additional dehybridization events are presented in \blauw{Fig.~S7}.
         }

### 2021-07-30 SI_marked_up.tex
punctuation

@@
 Furthermore, treating termini fraying as complete removal of a nearest neighbor pair is a crude approximation since the model was not intended to incorporate fraying.
-Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate}
+Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate.}
 

### 2021-07-31 response_letter/response_letter.tex
Andy edits

@@
 	\centering
-    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS5.pdf}
+    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{../FigS5.pdf}
 \end{figure}
@@
 It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
-We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift.
+We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift, represented by green arrows.
 We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
@@
 Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
-A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green).
+A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
 A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
 Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
-The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
 The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
@@
 	\begin{center}
-        \includegraphics[width=1.0\textwidth]{FigS1.pdf}
+        \includegraphics[width=0.8\textwidth]{../FigS1.pdf}
          \caption*{\textbf{Figure S1.}Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
@@
 \textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 15.
-We agree that the sentence on (P.
-15 L.
-6) is unclear in its current form and have made the following change:
+We agree that the sentence on is unclear in its current form and have made the following change to that text on p.~15:
 
-``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively''}
+``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes proceeding on 10-30 $\mu$s and 70-100 ns timescales, respectively.''}
 
@@
 
-\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ goodness-of-fit values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits.
+\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits.
 The revised figure is reproduced below.
-In all cases except the lowest temperature (T$_m$-5) slow response fits (where dissociation data is sparse) we report good fits with $R^2$ $>$ 0.88.
+In all cases except the lowest temperature T$_m$-5 K slow response fits where dissociation data is sparse, we report good fits with $R^2$ $>$ 0.88.
 We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.
@@
 	\centering
-    \includegraphics[width=.75\textwidth]{FigS4.pdf}
+    \includegraphics[width=.75\textwidth]{../FigS4.pdf}
    \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
 \rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
-In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
+In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse.}}
     \label{fig:SIFig4}
@@
 \textbf{Author Reply:}   The reviewer raises an important point regarding the dynamical fingerprinting methodology that have been developed to compare MSM predictions and relaxation experiments.
-We are familiar with these techniques, are currently applying them in our current work on abasic DNA, and did actually attempt to apply them in the present work.
-These analyses, however, were less informative than we had hoped due to three primary factors.
+We are familiar with these techniques and, in fact, are currently applying them in our current work on abasic DNA.
+We did attempt to apply them in the present work, but these analyses proved less informative than we had hoped due to two primary factors.
 
@@
 
-Second, the dynamical fingerprinting method provides an elegant means to compare experimental data against an MSM constructed at the same state point (e.g., temperature, pressure, slat conditions).
+Second, the dynamical fingerprinting method provides an elegant means to compare experimental data against an MSM constructed at the same state point (e.g., temperature, pressure, salt concentration).
 We constructed only a single MSM for each sequence at the melting temperature of the 3SPN.2 model that required the collection of 1 ms of simulation data for each sequence in order to observe sufficiently many hybridization/dehybridization events for robust fitting of the model parameters (Section 2.1.1).
-The experimental relaxation data, however, was collected over a range of temperatures and we wished to make comparisons across this entire range.
+The experimental relaxation data, however, was collected over a range of temperatures and we wished to make comparisons across this entire experimental range.
 Fitting additional MSMs was too computationally burdensome due to the high volume of data required and the cost of these simulations, so instead we conducted 120 $\times$ 1 $\mu$s simulations for each DNA sequence at each temperature.
-These trajectory data were too sparse to fit a robust MSM, but did allow us to make a direct comparison between the MD simulations and the experimental measurements over the entire temperature range.
+These trajectory data were too sparse to fit a robust MSM, but did allow us to make a direct comparison between the MD simulations and the experimental measurements over the full temperature range.
 
@@
 We concur with the reviewer that we can and should explicitly mention dynamical fingerprinting approach and provide some discussion for the reader why we did not employ this approach in the present study.
-To this end, we have added the following text to the end of Section 3.1 (p.18):
+To this end, we have added the following text to the end of Section 3.1 on p.~18:
 
@@
 \textbf{Author Reply:}  We are pleased that the reviewer finds our results interesting and suitable for publication.
-We concur with the reviewer that this work builds upon prior studies in the literature on sequence effects upon hybridization/dehybridization mechanisms, but goes beyond these studies in presenting a mean to rigorously resolve and quantify these dynamics.
-In response to comments from Reviewer \#2 and Reviewer \#3, we have updated several introductory sentences to clarify the novelty and impact of this work like less in the discovery of the particular metastable states, but in furnishing quantitative and predictive kinetic models of the dynamical transition network between these states that furnishes experimentally testable predictions and new understanding and quantification of the influence of sequence upon thermodynamics and kinetics.
+We concur with the reviewer that this work builds upon prior studies in the literature on sequence effects upon hybridization/dehybridization mechanisms, but we contend that it goes beyond these studies in presenting a means to rigorously resolve and quantify these dynamics.
+The primary novelty and impact of this work lies less in the discovery of the particular metastable states, but in furnishing quantitative and predictive kinetic models of the dynamical transition network between these states that furnishes experimentally testable predictions and new understanding and quantification of the influence of sequence upon thermodynamics and kinetics.
 
@@
 
-and we state in the Introduction:
+and we go on to state in the Introduction:
 
@@
 \textbf{Author Reply:}    (I) We agree with the reviewer that it may be possible to use biasing techniques in order to efficiently collect the data required to construct the MSMs.
-We are very familiar with the use of such techniques: indeed, another major interest of our research group is iterative strategies to perform data-driven determination of collective variables from simulation trajectories, enhanced sampling in these variables, and then estimation of new and improved collective variables from the enhanced sampling trajectories \citep{Sidky2020MolecularSimulators, Sidky2020MachineSimulation, Chen2019CapabilitiesSystems, Chen2019NonlinearVAMPnets, Phys2018CollectiveDesign, Chen2018MolecularExploration}.
+We are very familiar with the use of such techniques and another major interest of our research group is iterative strategies to perform data-driven determination of collective variables from simulation trajectories, enhanced sampling in these variables, and then estimation of new and improved collective variables from the enhanced sampling trajectories \citep{Sidky2020MolecularSimulators, Sidky2020MachineSimulation, Chen2019CapabilitiesSystems, Chen2019NonlinearVAMPnets, Phys2018CollectiveDesign, Chen2018MolecularExploration}.
 In the present work, we chose not to employ biased simulations and instead collect unbiased trajectories for three main reasons.
 
--- First, the 3SPN.2 coarse-grained DNA model is sufficiently inexpensive that it was possible to collect adequate unbiased simulation data to parameterize robust MSMs at reasonable costs.
+First, the 3SPN.2 coarse-grained DNA model is sufficiently inexpensive that it was possible to collect adequate unbiased simulation data to parameterize robust MSMs at reasonable costs.
 As we state on p.~8:
@@
 
--- Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1, Fig.~S4) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S1-3).
+Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S1-3).
 This indicates that the data are of sufficient quality and volume to present favorable experimental comparisons and learn reliable long-time kinetic models.
 
--- Third, although it is straightforward to correct biased simulation data to recover the unbiased thermodynamics, it is more challenging to recover the unbiased kinetics.
+Third, although it is straightforward to correct biased simulation data to recover the unbiased thermodynamics, it is more challenging to recover the unbiased kinetics.
 Although a number of approaches do exist, it has been our experience that the kinetic corrections can present difficulties in numerical convergence.
@@
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
-\rood{In this work, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}''
+\rood{In this study, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}''
 
@@
 ``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possible coupled with biased sampling and reweighting techniques, \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels} may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
+\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
 
@@
 \textbf{Author Reply:}   (I) We regret our lack of clarity in our choice of wording.
-Yes, we simply shifted the $k_d$ curves by a universal temperature shift, determined to be +4 K, that produced the optimal match with the experimental data.
+Yes, we simply shifted the $k_d$ curves by a universal temperature shift, determined empirically to be +4 K, that produced the optimal match with the experimental data.
 
-We have also reworded the unclear wording regarding the temperature correction in Section 3.1 p.~16 to now read:
+We have also reworded the unclear wording regarding the temperature correction in Section 3.1 p.~17 to now read:
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Fig1.pdf}
+        \includegraphics[width=0.8\textwidth]{../Fig1.pdf}
         \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
@@
 (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
+\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction at long times.} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
 The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
@@
 
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
-We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
+We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ as,}
 
@@
 
-In Section 2.1.2 of the Methods on p.~8, we have added:
+In Section 2.1.2 of the Methods on p.~9, we have added:
 
@@
 
-In the first sentence of the caption of Fig.~2, we have added:
+In the first sentence of the caption of Fig.~3, we have added:
 
@@
 The only other place we use the term is in Section 3.4 on p.~29 where we refer to work conducted in Ref.~\citep{Xiao2019}.
-Here we choose to maintain the term ``slithering'' to maintain consistency with the mechanism nomenclature adopted in this paper, but place it in quotation marks to indicate that we are quoting from that paper and it is not a term that we would necessarily favor in the context of our own work.
+Here we choose to maintain the term ``slithering'' to maintain consistency with the mechanism nomenclature adopted in this paper, but place it in quotation marks to indicate that we are quoting from that paper and it is not a term that we would necessarily favor.
 
-\rood{``Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
-(In contrast, out-of-register slithering and in-register zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.)'' }
+``Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
+(In contrast, out-of-register slithering and in-register zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.)''
 
@@
 
-\textbf{Author Reply:}   This a good point of clarification.
-We motivated our choice of this distance as that beyond which the strands are effectively non-interacting, but also verified that our estimates of $k_d^{slow}$ are robust to choices of this cutoff in the range 0.8--2.5 nm.
-We illustrate this in \textbf{Figure R1} below where we show that our extracted values of $k_d^{slow}$ are robust in changing the cutoff from 1.3 nm to 2.0 nm.
+\textbf{Author Reply:}   We motivated our choice of this distance as that beyond which the strands are effectively non-interacting, but also verified that our estimates of $k_d^{slow}$ are robust to choices of this cutoff in the range 0.8--2.5 nm.
+As an illustrative example, we demonstrate in \textbf{Figure R1} below that our extracted values of $k_d^{slow}$ are robust to changing the cutoff from 1.3 nm to 2.0 nm.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{FigR1.pdf}
+        \includegraphics[width=\textwidth]{../FigR1.pdf}
         \caption*{\textbf{Figure R1.} Comparing the experimental slow response when increasing the cutoff parameter from 1.3 nm to 2.0 nm}
@@
 ``First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
-This cutoff was selected as the distance beyond with the strands are effectively non-interacting and define the dissociated state.
+This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and define the dissociated state.
 \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.}''
@@
 
-\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1 and Fig.~S4 on log-linear axes and reproduce these below.
+\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1a and Fig.~S4 on log-linear axes and reproduce these below.
 
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{Fig1.pdf}
+        \includegraphics[width=1.0\textwidth]{../Fig1.pdf}
         \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
@@
 	\centering
-    \includegraphics[width=.75\textwidth]{FigS4.pdf}
+    \includegraphics[width=.8\textwidth]{../FigS4.pdf}
    \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
 \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
-In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
+In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse.}}
     \label{fig:SIFig4_2}
@@
 \textbf{Author Reply:}   This is an interesting question and are, in fact, considering embarking upon some additional work in this vein.
-We do not, however, have any data on this scaling at the current time.
 Rigorous engagement of this question would require us to fit MSMs at a variety of temperatures in order to compare the pathways, mechanisms, and rates for hybridization and dehybridization.
 In the present work, we have only constructed MSMs at the melting temperature of each sequence and the collection of sufficient data to parameterize MSMs at other temperatures would require both (i) a substantial investment of time to collect the $\sim$1 ms of data required to fit robust MSMs and (ii) the use of biasing techniques in order to adequately sample hybridization (dehybridization) events at large temperature excursions above (below) the melting temperature.
-As such, we do not have the necessary data to engage this question in the present work, but would like to explore it in the future.
+As such, we do not have the necessary data to engage this question at the present time.
 
-Reviewer \#2 raised a related point in their Comment \#2 and we have made the following modification to the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting follow-up study:
+Reviewer \#2 raised a related point in their Comment \#2 and in response to both of these points, we have made the following modification to the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting follow-up study:
 
 ``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
+\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
 
@@
 ``\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
-However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that in all cases incorporating additional NN configurations changes $\Delta$ F values (relative the corresponding partially frayed H state) by $<$ 1 kJ/mol when compared to using a single microstate.
-Furthermore, treating termini fraying as complete removal of a nearest neighbor pair is a crude approximation since the model was not intended to incorporate fraying.
-Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate}''
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
+Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}''
 
@@
 
-\textbf{Author Reply:}  We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence.
-Specifically, we did not convert the $\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta$ F (13.7 kJ/mol) relative to the hybridized state.
-Using this correct $\Delta$ F value we see agreement between the reported free energy in Fig~3 and populations reported in Fig.~2. In brief, we perform the following back-of-the-envelope verification:
+\textbf{Author Reply:}  We apologize for the confusion and have resolved the origin of this apparent inconsistency to two factors: (i) a unit conversion error on our part and (ii) our choice to present the bar plots in Fig.~2b on a log axis.
+
+(I) First, we made an (embarrassing) unit conversion error in reporting the $\Delta F$ value for the F4 state of the GC-core sequence from our MSM model.
+Specifically, we did not convert the $\Delta$F from units of $kT$ at $T_m$ = 324 K to kJ/mol, resulting us to erroneously report a value of 5.1 kJ/mol instead of 5.1 $kT$ = 13.7 kJ/mol.
+Using this correct $\Delta$F value we see agreement between the reported free energy in Fig.~3 and populations reported in Fig.~2.
+We can demonstrate this by performing the same back-of-the-envelope verification proposed by the reviewer:
 
@@
 \item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
-\item $G_{F4}$ = 5.1 KT
+\item $G_{F4}$ = 5.1 $kT$ at $T_m$ = 324 K = 13.7 kJ/mol
 \item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
@@
 
-We can solve this system of equations to yield $f_{F4}$ = 0.0030.
-A more rigorous accounting of the state populations by analyzing the MSM predictions yields a slightly lower prediction of $f_{F4}$ = 0.0028 that we report in Fig.~2b.
+We can solve this system of equations to yield $f_{F4}$ = 0.0030 that is in excellent agreement with the MSM predictions of $f_{F4}$ = 0.0028 that we report in Fig.~2b.
 
-We recognize that this was a significant miscalculation, and we have re-checked all free energy results to ensure it was an error for the GC-core free energy calculations only and did not affect the reported values for GC-end and AT-all.
-Furthermore, we have verified that the error does not propagate to any other relevant results in the manuscript.
-
-We have updated Figure 3 with the correct $\Delta$ F  as shown below.
+We have updated Figure 3 with the correct $\Delta$F as shown below.
 
@@
 	\begin{center} 
-        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+        \includegraphics[width=140mm, scale=1]{../Fig3.pdf}
+        \caption*{Figure 3.
+Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
 We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
 (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model.
-The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
+The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) relative to the hybridized state H compared to the NN model.}
         \label{fig:NN_table}
@@
 
-We have also updated the second paragraph of section 3.3 as follows:
+We have also updated the second paragraph of Section 3.3 on p.~23 as follows:
 
-As illustrated in Fig.~3, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
+``As illustrated in Fig.~3, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
+Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?''
 
-We believe that the root of this comment may be our choice of a logarithmic y-axis for the bar plots in Fig.~2b that make the bar associated with the F4 state appear nearly 50\% of the height of the H and D bars for GC-core.
+We recognize that this was a significant miscalculation, and we have re-checked all free energy results to ensure that (i) we did not make a similar error for any of the other reported values in Fig.~3 or anywhere else and (ii) that this error did not propagate through to any subsequent calculations, results, or conclusions within the remainder of the manuscript.
+
+(II) We also note that our choice of a logarithmic y-axis for the bar plots in Fig.~2b that make the bar associated with the F4 state appear nearly 50\% of the height of the H and D bars for GC-core.
 We made this plotting choice in order to make visible on these bar plots the small state populations associated with the shifted and frayed states and noted the percentage occupancies on the bars themselves.
@@
 \textbf{Author Reply:}    The reviewer raises a good point regarding the assumptions made during MSM construction.
-We do indeed use the reversible Bayesian MSM framework described here \url{http://www.emma-project.org/latest/api/generated/pyemma.msm.BayesianMSM.html} which enforces detailed balance.
+We do indeed use a reversible MSM framework that enforces detailed balance.
 This is an important point which we have now made explicit by modifying the text in Section 2.1.2 on p.~11 to read:
@@
 
-``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms.
-\rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.''}
+``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.
+\rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.}''
 
@@
 
-\textbf{Author Reply:}  This is a nice observation, and we would do well to note this symmetry for the reader. We have added a sentence to this effect in Section 3.6 on p.~31:
+\textbf{Author Reply:}  This is a nice observation, and we would do well to note this symmetry for the reader. We have added a sentence to this effect in Section 3.6 on p.~32:
 
@@
 
-We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state.''
+We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state'' so that the relevant text in Section 2.2.3 on p.~14 now reads:
 
@@
 
-\textbf{Author Reply:}   We have added this reference.
+\textbf{Author Reply:}   We have added these references to those at the top of p.~4.
 
@@
 \textbf{Author Reply:}   The reviewer makes a good comment on our choice of language used to described the stretched AT-all response.
-It is indeed the heterogeneity of the configurations in the initial ensemble, and the fraying dynamics exhibited in each of these states, that is the origin of the broadening of the relaxation profile, but our wording here was insufficiently clear.
-We believe that the reviewer's apt suggestion ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into our edits of Section 3.7 on p.~33:
+It is indeed our contention that the heterogeneity of the configurations in the initial ensemble and the fraying dynamics exhibited in each of these states that is the origin of the broadening of the relaxation profile, but our wording here was insufficiently clear.
+We believe that the reviewer's phrasing ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into our edits of Section 3.7 on p.~33:
 
@@
 
-\textbf{Author Reply:}   As a computationally-focused paper, we agree that the data and analysis code should be made available to the reader.
-We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).
+\textbf{Author Reply:}   We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).
 The links to both of these resources are now noted in the Supporting Information statement on p.~38:
@@
 \bibliographystyle{achemso}
-\bibliography{references_210514_abv}
+\bibliography{../references_210514_abv}
 \end{document}

### 2021-07-31 SI_marked_up.tex
Andy edits

--- 2021-07-30 SI_marked_up.tex
@@
 Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
-The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
 The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
@@
 \rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern.
-However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that in all cases incorporating additional NN configurations changes $\Delta$ F values (relative the corresponding partially frayed H state) by $<$ 1 kJ/mol when compared to using a single microstate.
-Furthermore, treating termini fraying as complete removal of a nearest neighbor pair is a crude approximation since the model was not intended to incorporate fraying.
-Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S5), we have elected to maintain a single microstate free energy estimate for each macrostate}
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
+Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}
 

### 2021-07-31 manuscript_marked_up.tex
Andy edits

@@
 A comprehensive understanding of the full dynamical landscape of hybridization/dehybridization accounting for the sequence-dependent metastable states and association/dissociation pathways remains lacking and fundamental questions remain unresolved.
-For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
+For example, it remains unclear the extent to which hybridization of short DNA oligomers largely proceeds in a conventionally assumed ``all-or-nothing'' fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Wyer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization, Xiao2019} Out-of-register ``shifted'' base paired structures \citep{Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019, Romano2013DNADependence,Flamm2000RNAResolution} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.
 The development of predictive models and design rules with which to engineer DNA strands with tailored hybridization/dehybridization kinetics and pathways is vital to advancing rational design of DNA strands for nanotechnology applications and is also of importance in understanding fundamental biological processes such as transcription and gene regulation.
@@
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
-\rood{In this work, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}
+\rood{In this study, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}
 
@@
 We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments.
-We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively}.
+We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes proceeding on 10-30 $\mu$s and 70-100 ns timescales, respectively}.
 The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
@@
 \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
-We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
+We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ as,}
 
@@
 We apply the NN model to predict the free energy $F^\mathrm{NN}$ of each of the macrostates occupied by each of the four sequences.
-A full accounting of our application of the NN model is provided in \blauw{Fig.~S6} of the \blauw{Supporting Information}.
+A full accounting of our application of the NN model is provided in \blauw{Fig.~S6} \rood{and supporting text within} the \blauw{Supporting Information}.
 The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
@@
 
-As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for the out-of-register shifted macrostates 5S2, 3S2, 5S4, and 3S4 relative to the H state compared to the NN model.
-The NN and MSM predictions for the frayed state F4 are in good agreement.
-Although the shifted state trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?
+As illustrated in Fig.~\ref{fig:NN_table}, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
+Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?
 First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data.
@@
 (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model.
-The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
+The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) relative to the hybridized state H compared to the NN model.}
         \label{fig:NN_table}
@@
 Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}
-, may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.} We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
+\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.} We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
 

### 2021-08-01 response_letter/response_letter.tex
Brennan edits

@@
 
-``\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.''
+``\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.''
 
@@
 
-``In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective melting temperatures} from the 40$\times$25 simulation trajectories following a four-step protocol...''
+``In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective 3SPN.2 melting temperatures} from the 40$\times$25 simulation trajectories following a four-step protocol...''
 
@@
 
-``We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.''
+``We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence 3SPN.2 melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.''
 
@@
 
-``Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures} for AT-all, GC-end, GC-core, and GC-mix.''
+``Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, and GC-mix: 324 K}.''
 
@@
 
-``Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.''
+``Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence 3SPN.2 melting temperatures}.''
 
@@
 ``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.''
+
+We have modified the opening of the caption of Fig.~5 to read:
+
+``\rood{The final temperature for each measurement is AT-all: 320 K, GC-end: 319 K, GC-core: 327 K, and GC-mix: 326 K.
+(a) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
+Normalized time traces for each sequence are shown at (b) 1600 cm$^{-1}$ and (c) 1660 cm$^{-1}$.}''
 
@@
 \textbf{Author Reply:}   We thank the reviewer for this important observation and illuminating our need to better clarify this potentially ambiguous concept.
-We have modified the text in Section 3.6 on p.~29 to read:
+We have modified the text in Section 3.6 on p.~31 to read:
 
-``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
 \rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
@@
 ``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying.
 This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
@@
 
+=====
+
+We have also corrected a small number of typos throughout the main text that we have indicated in red.
+
 \clearpage

### 2021-08-01 manuscript_marked_up.tex
Brennan edits

@@
 \citep{Schneider1978Molecular-dynamicsTransitions}.
-\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
 Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
@@
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms}.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective melting temperatures} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective 3SPN.2 melting temperatures} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 \begin{figure}[ht!]
-        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures} for AT-all, GC-end, GC-core, and GC-mix.
+        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, and GC-mix: 324 K}.
 (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D).
@@
 
-We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence 3SPN.2 melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
 The NN model predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
@@
         \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence 3SPN.2 melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
 We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
@@
 
-The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
 \rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
@@
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying.
 This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
@@
         \caption{T-Jump IR responses reflect sequence-dependent conformational heterogeneity.
-(a) MSM predictions of equilibrium macrostate life times at the sequence melting temperature.
-(b) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
-Normalized time traces for each sequence are shown at (c) 1600 cm$^{-1}$ and (d) 1660 cm$^{-1}$.
-The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
+\rood{The final temperature for each measurement is AT-all: 320 K, GC-end: 319 K, GC-core: 327 K, and GC-mix: 326 K.
+(a) Mid-IR t-HDVE difference spectra for GC-core at time delays from 5 ns to 560 $\mu$s.
+Normalized time traces for each sequence are shown at (b) 1600 cm$^{-1}$ and (c) 1660 cm$^{-1}$.} The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations.
 Each time trace is fit to the sum of stretched exponential with two exponentials (solid lines): $S(t) = A\exp{(-(t/\tau_{\mathrm{fast}})^{\beta_\mathrm{fast}})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$.

### 2021-08-04 response_letter/response_letter.tex
JACS resubmission

@@
 	\begin{center}
-        \caption*{\textbf{Figure S5.}\rood{Macrostate interpretation and structural analysis.
+        \caption*{\textbf{Figure S5.} \rood{Macrostate interpretation and structural analysis.
 The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
@@
         \includegraphics[width=0.8\textwidth]{../FigS1.pdf}
-         \caption*{\textbf{Figure S1.}Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
+         \caption*{\textbf{Figure S1.} Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
 A knee in the VAMP-2 plot was identified at the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively.
@@
 
+We have also update the y-axis label in Fig.~1 to specify units in s$^{-1}$.
+
 \textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}
@@
 The revised figure is reproduced below.
-In all cases except the lowest temperature T$_m$-5 K slow response fits where dissociation data is sparse, we report good fits with $R^2$ $>$ 0.88.
+In all cases except the lowest temperature T$_m$ - 5 K slow response fits where dissociation data is sparse, we report good fits with $R^2$ $>$ 0.88.
 We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.
@@
 \rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
-In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse.}}
+In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$ - 5 K where dissociation events are sparse.}}
     \label{fig:SIFig4}
@@
 ``The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}.
-\rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable Markov state models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling.
+\rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable kinetic models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling.
 A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}.
@@
 \textbf{Author Reply:}   We thank the reviewer for this suggestion and regret our lack of clarity in this matter.
-All simulations performed in this work were conducted at the melting temperature, with the exception of the simulations discussed in Section 3.1 where we consider a range of temperatures in order to draw comparisons to experimental measurements over this temperature range.
+All simulations performed in this work were conducted at the melting temperatures determined within the 3SPN.2 model, with the exception of the simulations discussed in Section 3.1 where we consider a range of temperatures in order to draw comparisons to experimental measurements over this temperature range.
 We have added a number of clarifications to the text in order to more clearly convey this important point to the reader.
@@
 
-\textbf{Author Reply:}   This is an interesting question and are, in fact, considering embarking upon some additional work in this vein.
+\textbf{Author Reply:}   This is an interesting question and we are, in fact, already considering embarking upon some additional work in this vein.
 Rigorous engagement of this question would require us to fit MSMs at a variety of temperatures in order to compare the pathways, mechanisms, and rates for hybridization and dehybridization.
@@
         \includegraphics[width=140mm, scale=1]{../Fig3.pdf}
-        \caption*{Figure 3.
-Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+        \caption*{\textbf{Figure 3.} Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
 We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
@@
 ``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote a \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying.
 This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
@@
 
-\textbf{Author Reply:}   We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).
+\textbf{Author Reply:}   We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5148949}).
 The links to both of these resources are now noted in the Supporting Information statement on p.~38:
 
-``\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).}''
+``\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5148949}).}''
 
@@
 
-=====
+============================================================
 

### 2021-08-04 SI.tex
JACS resubmission

@@
 
-%\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\rood}[1]{{\color{red}{#1}}}
 %\newcommand*{\rood}[1]{#1}
 %\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
-\newcommand*{\blauw}[1]{#1}
+%\newcommand*{\blauw}[1]{#1}
 %\newcommand*{\groen}[1]{\textcolor{green}{#1}}
@@
 	\centering
-    \includegraphics[width=.70\textwidth]{FigS4.pdf}
+	\includegraphics[width=.97\textwidth]{FigS4.pdf}
+\end{figure}
+
+\begin{figure}[ht!]
+	\begin{center}
     	\caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
@@
 We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
-Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.}
+Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
+In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$ - 5 K where dissociation events are sparse.}}
     \label{fig:SIFig4}
+    	\end{center}
 \end{figure}
@@
 \begin{figure}[ht!]
-	\centering
-    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
-Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events.
-Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
-Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
-Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
-Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
-    \label{fig:SIFig5}
+	\begin{center}
+        \caption{\rood{Macrostate interpretation and structural analysis.
+The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
+(a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying.
+We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction.
+This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction.
+We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands.
+It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
+We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift, represented by green arrows.
+We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
+(b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence.
+The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
+Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively.
+(c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
+This panel conveys two important pieces of information.
+First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
+These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
+Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex.
+Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
+A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
+A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
+Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
+The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
+The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
+(d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments.
+The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
+        %\label{fig:macrostate_colors_FES}
+	\end{center}
 \end{figure}
@@
 $\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TheMotifs}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy.
-A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig6}.
+A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig7}.
 $\Delta S_{NN}^{\circ}$ is constructed in an analogous manner from NN, dangling end, and initiation terms, plus a symmetry term, $\Delta S_{sym}^{\circ}$, that applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.
@@
 
+\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
+Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}
+
 \vspace{+0.5in}
@@
     \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
+    \label{fig:SIFig7}
+\end{figure}
+
+\clearpage
+\newpage
+
+% Fig. S7
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=0.8\textwidth]{FigS7a.pdf}
+\end{figure}
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=0.8\textwidth]{FigS7b.pdf}
+\end{figure}
+
+\begin{figure}[ht!]
+	\centering
+    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling.
+Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events.
+Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot.
+Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange).
+Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex.
+Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
     \label{fig:SIFig6}

### 2021-08-04 manuscript.tex
just working on 3.7 and 3.8
JACS resubmission

--- 2021-05-20 manuscript.tex
@@
 From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} Recently, single molecule localization techniques have exploited the rapid and transient binding of short DNA oligomers in order to achieve super-resolution microscopy and optical multiplexing.
-\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnology applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the dynamical, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation For Imaging In Nanoscale Topography), but these technologies have outpaced our fundamental understanding of the dynamics themselves.
+\cite{Schnitzbauer2017Super-resolutionDNA-PAINT, Shah2019, Strauss2020UpDNA-PAINT} Predictive understanding of the sequence-dependent thermodynamics of DNA hybridization/dehybridization -- the assembly/disassembly of a DNA duplex from two single strands -- has underpinned the rational design of DNA oligomer sequences for nanotechnology applications, where sequence-dependent nearest-neighbor models can accurately account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Predictive models of the dynamical, as opposed to purely thermodynamical, behaviors of DNA have become increasingly important in developing technologies such as DNA-PAINT (DNA Points Accumulation \rood{f}or Imaging \rood{i}n Nanoscale Topography), but these technologies have outpaced our fundamental understanding of the dynamics themselves.
 \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} Many experimental and computational studies have investigated DNA dynamical phenomena over picosecond to millisecond time scales.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition} Kinetic models have been developed for particular DNA processes such as toehold exchanges and optical barcoding \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting} and supervised machine learning techniques have been combined with experimental measurements to predict the on/off rates as a function of sequence.
@@
 Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}.
-The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
+The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Hinckley2014Coarse-grainedEffects} but they can be observed by employing enhanced sampling techniques\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}.
+\rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable kinetic models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling.
+A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}.
+The application of these methods under the conditions of high bias necessary for good sampling can, however, present challenges for numerical convergence.} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
 These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
+\rood{In this study, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}
 
@@
 \citep{Schneider1978Molecular-dynamicsTransitions}.
-Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
 Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
@@
 In this manner, they provide an interpretable and predictive model of the system thermodynamics and kinetics.
-MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms}.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective 3SPN.2 melting temperatures} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments.
-We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response.
+We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes proceeding on 10-30 $\mu$s and 70-100 ns timescales, respectively}.
 The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}
@@
 First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
-This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state.
-We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
+\rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
+We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ as,}
+
+    \begin{equation}\label{}
+    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
+    \end{equation}
+
+    \begin{equation}\label{}
+	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
+	\end{equation}
+
 Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction.
-We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
+We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_{\mathrm{unfrayed}}(t) = \exp(-k_d^\mathrm{fast}t)$.
 
@@
 We apply these sequence-independent scaling factors to our reported computational values in Fig.~\ref{fig:relaxation-comparison}.
-Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results.
+Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.
 We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
-The computational time scales and melting temperatures reported in the remainder of the paper are not corrected since we are only interested in the relative trends in the behaviors of the four sequences.
+The computational time scales and melting temperatures reported in the remainder of the paper are not corrected \rood{by these calibration corrections or finite-size fluctuation corrections} since we are only interested in the relative trends in the behaviors of the four sequences.
 
@@
 (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-(b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
+\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction at long times.} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
 The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
@@
 Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements.
-The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers.
+The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers\rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence.
+We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}.
+These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}.
+We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected.
+Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}.
 
@@
 We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates.
-In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
-By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F).
+\rood{The representative microstates emblematic of each macrostate were selected randomly from those possessing high ($>$99\%) macrostate membership probabilities.
+Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution.
+In \blauw{Fig.~S5} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
+In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium.
+By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F\rood{4}).
 In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates.
@@
 \begin{figure}[ht!]
-        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix.
+        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, and GC-mix: 324 K}.
 (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D).
@@
 
-We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
+We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence 3SPN.2 melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.
 The NN model predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends.
 We apply the NN model to predict the free energy $F^\mathrm{NN}$ of each of the macrostates occupied by each of the four sequences.
-A full accounting of our application of the NN model is provided in the \blauw{Supporting Information}.
+A full accounting of our application of the NN model is provided in \blauw{Fig.~S6} \rood{and supporting text within} the \blauw{Supporting Information}.
 The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies.
@@
         \includegraphics[width=160mm, scale=1]{Fig3.pdf}
-        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
+        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence 3SPN.2 melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
 We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
 (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model.
-The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
+The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) relative to the hybridized state H compared to the NN model.}
         \label{fig:NN_table}
@@
 
-The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
-The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
+The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
+\rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
+(We note that ``all-or-nothing'' can also be used, somewhat ambiguously, to refer to a two-state thermodynamic model where the hybridized and dehybridized are separated by a large free energy barrier but that the transitions between them may proceed through a network of metastable states.)} The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states.
 We note that we do observe substantial transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM.
@@
 To do so, we followed all 10 intermolecular distances between native WC base pairs and tracked their evolution through a number of hybridization and dehybridization events.
-We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more in \blauw{Fig.~S5}.
+We present one representative example of each event in Fig.~\ref{fig:GC-mix_transitions} and four more in \blauw{Fig.~S7}.
 During the hybridization process, we observe a global decrease in all 10 distances as the strands approach one another and the formation of key native WC contacts immediately prior to duplex formation: specifically, one of the G:C WC pairs and at least one neighboring A:T pair or 2-3 central A:T pairs.
-This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying on the two-base AT-tails on one or both sides of the duplex followed by rapid dissociation of the central WC base pairs.
+This behavior is consistent with a ``nucleation-zippering'' mechanism as has been reported in previous studies.\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization} In dehybridization, we observe fraying \rood{of one half of the duplex with the strands remaining associated until the loss of the final A:T pairs and ultimately the last G:C pair}.
 Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
-These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}
+These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility.
+\cite{McCully2008MicroscopicHomeodomain} A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics.
+This is indeed borne out by our mechanistic observations for GC-mix wherein the early stages of ``nucleation-zippering'' proceed by the formation of one G:C contact followed by one or more additional A:T pairs and the late stages by the formation of all remaining WC pairs, which we compare with the early stages ``fraying-peeling'' wherein one half of the duplex frays and the late stages wherein dissociation finally completes by the dissolution of the last few A:T contacts and the final G:C pair.}
 
@@
 Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an elevated fraying response over the course of the relaxation, and should therefore be manifest in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.
-Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
+Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote a \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying.
+This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions (Fig.~\ref{fig:allseq_table}b) reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
 It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.
@@
 Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
+\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.} We anticipate that these insights may provide foundational design rules by which to improve understanding of \emph{in vivo} hybridization processes and rationally engineer optimized sequences for DNA nanotechnology applications such as DNA-PAINT~\citep{Strauss2020UpDNA-PAINT} and DNA barcoding.~\citep{Shah2019}
 
@@
 Details of nearest neighbor (NN) model calculations, supplementary figures illustrating selection of optimal number of SRV slow modes for each sequence, convergence of MSM implied time scales, MSM Chapman-Kolmogorov tests, ``computational T-jump'' calculations, and GC-mix hybridization and dehybridization trajectories.
+\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5148949}).}
 

### 2021-09-08 response_letter_2/response_letter.tex
Draft second round of reveiew responses

--- 2021-07-31 response_letter/response_letter.tex
@@
 \newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\roodr}[1]{{\color{red}{#1}}}
 \newcommand*{\noteg}[1]{\textcolor{green}{[[#1]]}}		% notes 
@@
 
-Thank you for your 6-Jul-2021 message regarding our article ``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy'' assigned manuscript ID ja-2021-05219p.
+Thank you for your 7-Sep-2021 message regarding our article ``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy'' assigned manuscript ID ja-2021-05219p.R1.
 We thank you for handling our submission and the anonymous reviewers for their careful reading and critical assessment of our work.
@@
 
-\textcolor{blue}{In the manuscript ``Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy'', Jones et al.
-provide a well-designed and compelling study of DNA hybridization kinetics and dynamics that succinctly encompasses both experiment and theory.
-They demonstrate that the coarse-grained model of DNA, 3SPN.2, can not only well-reproduce the relative kinetic rates of sequence-dependent hybridization but also can reveal critical insights into their physical underpinnings.
-These insights are driven by the use of state-of-the-art applications of machine learning to Markov state model construction and well-validated with temperature jump infrared spectroscopy.
-In particular, the demonstration of how the placement of two G:C base pairs within the sequence can dramatically affect the energy landscape is truly impactful for the design of DNA sequences with desired properties.
-We believe this work is well suited for JACS publication after the Authors address a few major and minor points.}
-
-\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased that they find the work impactful and suitable for publication in JACS.
-
-\textcolor{blue}{Major point.
-1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D.
-While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ``representative'' structures shown in Figures 2A, C).
-A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern.
-This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations.
-At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
+\textcolor{blue}{Recommendation: Publish in JACS without change.
 }
 
-\textbf{Author Reply:} The designation and structural representation of the macrostates is crucial to the core conclusions of the manuscript and we concur that we can and should better expose for the reader how the representative microstates were selected, how representative they are of the microstate ensemble within the macrostate, and our analysis of the base-pairing patterns within each macrostate that motivated our naming scheme.
+\textcolor{blue}{Comments:
+Most of my comments were properly addressed. The manuscript, as far as I understand, is suitable for a journal like JACS.
+}
 
-On p.~19, we have added the following text:
+\textbf{Author Reply:} We are pleased that the reviewer is satisfied with our modifications and responses and that they are pleased to support publication in JACS.
+We would like to thank them again for their time, effort, and considered and useful expert comments.
 
-``...each of these macrostates.
-\rood{The representative microstates emblematic of each macrostate were selected randomly from those possessing high ($>$99\%) macrostate membership probabilities.
-Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution.
-In Fig.~S5 we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble.
-In all cases, we find these microstate distributions to be relatively narrowly focused within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig.
-2b we present...''
-
-We also added the new Fig.~S5 to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer.
-The figure and its caption are reproduced below.
-
-\clearpage
-\newpage
-
-\begin{figure}[ht!]
-	\centering
-    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{../FigS5.pdf}
-\end{figure}
-
-\begin{figure}[ht!]
-	\begin{center}
-        \caption*{\textbf{Figure S5.}\rood{Macrostate interpretation and structural analysis.
-The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble.
-(a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying.
-We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction.
-This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction.
-We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands.
-It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands.
-We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift, represented by green arrows.
-We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns.
-(b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence.
-The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate.
-Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively.
-(c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$.
-This panel conveys two important pieces of information.
-First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble.
-These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail.
-Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex.
-Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$.
-A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green).
-A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black).
-Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple).
-The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red).
-The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends.
-(d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments.
-The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
-        %\label{fig:macrostate_colors_FES}
-	\end{center}
-\end{figure}
-
-\clearpage
-\newpage
-
-\textcolor{blue}{1)      It is unclear how five datasets were selected used for cross-validation in the SRVs (P.
-10 L.
-43).
-One would naively think a 50:50 split of a single trajectory would result in only one dataset.
-A bit more detail could be provided here.}
-
-\textbf{Author Reply:}   Five-fold cross-validation was conducted by constructing five random partitions of the 40 independent trajectories into a training set of 20 trajectories and a validation set of 20 trajectories.
-This manner of splitting the data at the level of complete trajectories -- as opposed to splitting a single trajectory into multiple subtrajectories -- is desirable as it allows our SRV/MSM approach to learn over long 25 $\mu$s trajectories and infer slowly relaxing collective modes that may not be accessible within short sub-trajectories~\citep{Sidky2019High-ResolutionVAMPnets}.
-We have better clarified this procedure for the reader by revising the text on p.~11 to read:
-
-We guarded against overfitting using five-fold cross-validation in which we \rood{constructed five random partitions of the 40 independent simulation trajectories into a training set of 20 trajectories and a validation set of 20 trajectories}.\citep{Sidky2019High-ResolutionVAMPnets}
-
-\textcolor{blue}{2)      For the validation of embedding dimensions (P.
-11 L.
-3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Fig.~S1.
-Could a few additional dimensions be provided?}
-
-\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences.
-As suggested by the reviewer, we added two additional modes to the cross-validation procedure that does indeed better expose this kink at five slow modes.
-We have updated the corresponding panel Fig.~S1 and also reproduce it below.
-
-% Fig. S1
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.8\textwidth]{../FigS1.pdf}
-         \caption*{\textbf{Figure S1.}Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence.
-A knee in the VAMP-2 plot was identified at the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively.
-An embedding of corresponding dimensionality was then used to cluster frames into discrete states.
-The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
-        \label{fig:SIFig1}
-	\end{center}
-\end{figure}
-
-\textcolor{blue}{3)      The values associated with the ``slow'' and ``fast'' rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}
-
-\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 15.
-We agree that the sentence on is unclear in its current form and have made the following change to that text on p.~15:
-
-``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes proceeding on 10-30 $\mu$s and 70-100 ns timescales, respectively.''}
-
-\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}
-
-\textbf{Author Reply:}   Thank you for catching this error. Fig.~1 has been updated to include units in s$^{-1}$.
-
-\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P.
-17 L.
-9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system?
-It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}
-
-\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits.
-The revised figure is reproduced below.
-In all cases except the lowest temperature T$_m$-5 K slow response fits where dissociation data is sparse, we report good fits with $R^2$ $>$ 0.88.
-We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.
-
-\clearpage
-\newpage
-
-% Fig. S4
-\begin{figure}[ht!]
-	\centering
-    \includegraphics[width=.75\textwidth]{../FigS4.pdf}
-   \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
-From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time.
-We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm.
-We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
-Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
-In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse.}}
-    \label{fig:SIFig4}
-\end{figure}
-
-\clearpage
-\newpage
-
-\textcolor{blue}{6)       The authors chose to compute the temperature jump relaxation rates from fits to quantities obtained directly from the MD simulation.
-However, it is not clear why the authors did not also compute the relaxation of observables from estimated Markov state models following (P.
-N.
-A.
-S.
-USA.
-2011;108(12):4822-4827).
-This approach has also been previously used to compare kinetics obtained from Markov state models with other spectroscopic temperature-jump based experiments in DNA (J.
-Phys.
-Chem.
-B.
-2019;123(10):2291-2304.) and RNA (J.
-Chem.
-Theory Comput.
-2017;108(2):926-934) oligomers which should be mentioned.
-If possible, it would be interesting to compare or discuss the differences between their approach and these previous approaches.}
-
-\textbf{Author Reply:}   The reviewer raises an important point regarding the dynamical fingerprinting methodology that have been developed to compare MSM predictions and relaxation experiments.
-We are familiar with these techniques and, in fact, are currently applying them in our current work on abasic DNA.
-We did attempt to apply them in the present work, but these analyses proved less informative than we had hoped due to two primary factors.
-
-First, the experimental data contains fast responses occurring on time scales of $\sim$100 ns.
-As detailed in Section 3.1 and Figure 1, these fast motions occur in our simulations at an acceleration factor of approximately 120$\times$ and so proceed on a time scale of $\sim$1 ns.
-As such, these motions lie below the $\tau$ = 1.2 ns lag time used to construct our MSMs that is required to converge the leading slow modes and produce a Markovian model capable of passing the Chapman-Kolmogorov test (cf.
-Figs.~S2-3).
-As such, we cannot draw comparisons of our MSM against the fast experimental data.
-
-Second, the dynamical fingerprinting method provides an elegant means to compare experimental data against an MSM constructed at the same state point (e.g., temperature, pressure, salt concentration).
-We constructed only a single MSM for each sequence at the melting temperature of the 3SPN.2 model that required the collection of 1 ms of simulation data for each sequence in order to observe sufficiently many hybridization/dehybridization events for robust fitting of the model parameters (Section 2.1.1).
-The experimental relaxation data, however, was collected over a range of temperatures and we wished to make comparisons across this entire experimental range.
-Fitting additional MSMs was too computationally burdensome due to the high volume of data required and the cost of these simulations, so instead we conducted 120 $\times$ 1 $\mu$s simulations for each DNA sequence at each temperature.
-These trajectory data were too sparse to fit a robust MSM, but did allow us to make a direct comparison between the MD simulations and the experimental measurements over the full temperature range.
-
-We therefore elected to structure our manuscript such that we first test the MD simulation predictions against experimental data in Section 3.1, then, after having validated the simulation data upon which the MSMs are constructed, perform MSM construction and analysis in Sections 3.2-6.
-Finally, in Section 3.7 we close with a test of a specific prediction of our MSM (the existence of long-lived metastable shifted states) that was amenable to comparison against T-jump IR.
-
-We concur with the reviewer that we can and should explicitly mention dynamical fingerprinting approach and provide some discussion for the reader why we did not employ this approach in the present study.
-To this end, we have added the following text to the end of Section 3.1 on p.~18:
-
-``fast kinetics of the four DNA oligomers\rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence.
-We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}.
-These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}.
-We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected.
-Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}''
-
-\clearpage
-\newpage
-
-\begin{shaded}
-\textbf{Response to Reviewer \# 2}
-\end{shaded}
-
-\textcolor{blue}{This paper reports simulation results on the hybridization and dehybridization mechanisms near to the melting temperature of a number of short DNA sequences, and accompanying temperature jump IR experiments.
-The sequences are chosen to have a repeating AT motif that potentially allows relatively stable out-of register duplex states, with a focus being on how these states are disrupted, by the presence of GC base pairs.
-The authors show that the introduction of just 2 appropriately-positioned base pairs can change the kinetics to a simple two-state behaviour from one in which multiple states contribute.}
-
-\textcolor{blue}{I found this an interesting study that nicely illustrated some of the potential effects of sequence on (de)hybridization mechanisms.
-However, I found the results unsurprising, and these types of effects have been previously explored in the literature.
-Although I can't therefore recommend publication in JACS, I think it would be very appropriate for one of the more specialized ACS journals.}
-
-\textbf{Author Reply:}  We are pleased that the reviewer finds our results interesting and suitable for publication.
-We concur with the reviewer that this work builds upon prior studies in the literature on sequence effects upon hybridization/dehybridization mechanisms, but we contend that it goes beyond these studies in presenting a means to rigorously resolve and quantify these dynamics.
-The primary novelty and impact of this work lies less in the discovery of the particular metastable states, but in furnishing quantitative and predictive kinetic models of the dynamical transition network between these states that furnishes experimentally testable predictions and new understanding and quantification of the influence of sequence upon thermodynamics and kinetics.
-
-We better contextualize the scope and impact of our work in the final sentence of the Abstract:
-
-``Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization \rood{by furnishing quantitative and predictive kinetic models of the dynamical transition network between metastable states}, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation for DNA nanotechnology applications.''
-
-and we go on to state in the Introduction:
-
-``Out-of-register ``shifted'' base paired structures \citep{Flamm2000RNAResolution, Romano2013DNADependence, Hinckley2014Coarse-grainedEffects, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} and frayed structures \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} stand as candidates for metastable states with the potential to mediate substantial deviations from all-or-nothing behavior, but the degree to which these states are kinetically relevant is difficult to determine experimentally and is likely to be highly sequence-dependent.''
-
-``We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data and reinterpret our prior experimental observations in light of the new computational understanding.
-Consistent with previous studies, \citep{Hinckley2014Coarse-grainedEffects,Romano2013DNADependence,Araque2016LatticeCooperativity} we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of out-of-register shifted states -- leads to richer dynamics populated by a diversity of long-lived metastable states.
-Our data-driven modeling and analysis rigorously quantifies these behaviors and furnishes accurate predictive models of the hybridization/dehybridization rates, dynamical pathways, and metastable states.
-Specifically, we demonstrate that disrupting repetitive stretches of A:T bases by placement of interrupting G:C base pairs enables us to tune the landscape from rich six-state to simple two-state ``all-or-nothing'' behavior, and the specific location of the interrupting pair can be used to modulate the stability of long-lived frayed states.''
-
-``Taken together, our analyses establish new molecular-level understanding of the sequence-dependent kinetics and pathways through quantitative predictive models for the long-time system dynamics, resolution of the dynamical folding pathways and metastable states, and elementary design rules with which to control the dynamical behaviors of the system.
-We anticipate that this new foundational understanding, and the extension of our approach to more extensive families of DNA sequences, can guide the rational design of DNA oligomers with tailored kinetic properties engineered for DNA nanotechnology applications such as DNA-PAINT.
-\citep{Shah2019, Strauss2020UpDNA-PAINT}''
-
-\textcolor{blue}{One limitation of the study is its reliance on ``brute-force'' simulations (rather than using advanced sampling/rare-event techniques) to probe the transition.
-This means that systems of very high concentration (7 M) and temperatures close to the melting point have to be used, which are probably not the conditions of most interest.}
-
-\textbf{Author Reply:}    (I) We agree with the reviewer that it may be possible to use biasing techniques in order to efficiently collect the data required to construct the MSMs.
-We are very familiar with the use of such techniques and another major interest of our research group is iterative strategies to perform data-driven determination of collective variables from simulation trajectories, enhanced sampling in these variables, and then estimation of new and improved collective variables from the enhanced sampling trajectories \citep{Sidky2020MolecularSimulators, Sidky2020MachineSimulation, Chen2019CapabilitiesSystems, Chen2019NonlinearVAMPnets, Phys2018CollectiveDesign, Chen2018MolecularExploration}.
-In the present work, we chose not to employ biased simulations and instead collect unbiased trajectories for three main reasons.
-
-First, the 3SPN.2 coarse-grained DNA model is sufficiently inexpensive that it was possible to collect adequate unbiased simulation data to parameterize robust MSMs at reasonable costs.
-As we state on p.~8:
-
-``Each simulation was conducted for 26 $\mu$s and frames saved to disc every 100 ps.
-Each simulation required $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores.
-The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.''
-
-Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S1-3).
-This indicates that the data are of sufficient quality and volume to present favorable experimental comparisons and learn reliable long-time kinetic models.
-
-Third, although it is straightforward to correct biased simulation data to recover the unbiased thermodynamics, it is more challenging to recover the unbiased kinetics.
-Although a number of approaches do exist, it has been our experience that the kinetic corrections can present difficulties in numerical convergence.
-That is not to say that we could not have successfully applied this methods, but just that we did not have cause to do so since the unbiased data were not unreasonably expensive to collect and showed favorable experimental comparisons and produced robust MSMs.
-
-We did note the possibility of performing biased simulations in the Introduction on p.~4-5, and have now elaborated these remarks to better explain our choice to proceed with unbiased simulations:
-
-``The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}.
-\rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable Markov state models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling.
-A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}.
-The application of these methods under the conditions of high bias necessary for good sampling can, however, present challenges for numerical convergence.} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways.
-These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.
-\rood{In this study, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}''
-
-(II) Our specification of a single strand concentration of 7 mol/L in Section 2.1.1 on p.~7 was an (embarrassing) typo -- the true concentration is actually 7 mmol/L (i.e., 7 mM).
-This is only 3.5$\times$ higher than the 2 mmol/L experimental concentration, and so is actually quite a good approximation for the experimental conditions.
-We regret our error, and have corrected this text to now read:
-
-``A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of \rood{7 mM}.
-\rood{This concentration is only 3.5$\times$ larger than the 2 mM concentration employed in our experimental analyses (Section 2.2).}''
-
-(III) Finally, we deliberately chose to explore the melting temperature as the state point at which the hybridized and dehybridized populations are in balance and sought to probe the mechanism and kinetics of the hybridization/dehybridization transitions under these equilibrium conditions corresponding to no net thermodynamic preference for either state (i.e., minimally driven conditions).
-A favorable side effect of this choice was indeed that we observe approximately equal numbers of hybridization and dehybridization events in a single (sufficiently long) simulation, whereas temperatures away from $T_\mathrm{melt}$ would favor one over the other.
-It is an interesting question to consider how the mechanism and kinetics of hybridization/dehybridization may change with temperature.
-Exploring small excursions either side of $T_\mathrm{melt}$ may still be possible using unbiased simulations, but we concur with the reviewer that biased sampling would be required to observe sufficient numbers of hybridization (dehybridization) events at large excursions above (below) $T_\mathrm{melt}$.
-
-In response to this point, we have modified text in the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting point for follow-up study:
-
-``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
-
-\textcolor{blue}{Very small points:
-+ p16 line 3: "Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective calibration to our computational results".
-This sentence confused me at first.
-I presume you are just shifting your simulated kd slow curves to better match experiment, rather than shifting the data so that the melting temperatures agree.
-You may want to be more explicit as I read it as the latter at first.
-(Note one shouldn't expect the melting temperatures to agree as very different concentrations are being used in simulation and experiment, and one also has to correct for finite-size effects when estimating melting points from single duplex simulations (https://doi.org/10.1088/0953-8984/22/10/104102)}
-
-\textbf{Author Reply:}   (I) We regret our lack of clarity in our choice of wording.
-Yes, we simply shifted the $k_d$ curves by a universal temperature shift, determined empirically to be +4 K, that produced the optimal match with the experimental data.
-
-We have also reworded the unclear wording regarding the temperature correction in Section 3.1 p.~17 to now read:
-
-``Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.''
-
-(II) We regret our erroneous reporting of a 7 M rather than 7 mM concentration as detailed in our previous response.
-As we have now clarified, our simulations and experiments are conducted at similar concentrations (7mM and 2 mM, respectively).
-We agree that we still cannot expect perfect agreement in the melting temperatures due to the residual small  difference in concentrations, but likely this is small in comparison to discrepancies introduced by the approximations inherent in the coarse-grained molecular model.
-
-(III) We greatly appreciate this comment and suggestion to apply a fluctuation correction to our data in order to correct for artifacts associated with finite-size effects following the approach detailed in Ref.~\citep{Ouldridge2010ExtractingSimulations}.
-We have reanalyzed our slow repsonse data under the homodimer correction and this has pleasingly resulted in a slightly improved agreement between our simulations and experiment.
-The revised figure is reproduced below.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=0.8\textwidth]{../Fig1.pdf}
-        \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
-Results are reported in terms of the final T-jump temperature.
-(a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction at long times.} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
-The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
-The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
-        \label{fig:relaxation-comparison}
-	\end{center}
-\end{figure}
-
-In light of this modification, we have also added the following text to the Section 3.1 on p.~16:
-
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
-We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ as,}
-
-    \begin{equation}\label{}
-    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
-    \end{equation}
-
-    \begin{equation}\label{}
-	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
-	\end{equation}
-
-Finally, we have also edited the following text in Section 3.1 on p.~17 to clarify that the fluctuation correction is only applied for the purposes of the experimental comparison in Section 3.1 and is not propagated throughout the remainder of the paper:
-
-``The computational time scales and melting temperatures reported in the remainder of the paper are not corrected \rood{by these calibration corrections or finite-size fluctuation corrections} since we are only interested in the relative trends in the behaviors of the four sequences.''
-
-\textcolor{blue}{+ p15 line 20 "beyond with" -> "beyond which"}
-
-\textbf{Author Reply:}    This typo has been corrected.
-
-\clearpage
-\newpage
+%%%
 
@@
 
-\textcolor{blue}{Jones et al present a combined experimental and computational paper that probes the dynamics of the DNA hybridization/de-hybridization transition.
-Through a Markov state model analysis of coarse-grained simulation, the authors identify states that contribute to the process, and present experimental evidence to support the role of these states.
-In particular, repetitive sequences are observed in simulations to have a number of misaligned states that are both appreciably stable and contribute to hybridization and de-hybridization pathways as intermediates.
-The role of these states is diminished by base pairs that disrupt the repetitive sequence.
-In the special case of a sequence with a strong core and weaker extremities, metastable "frayed" states are also observed.
-The presence of these states is assessed experimentally trough the analysis of relaxation timescales observed in T-jump IR spectroscopy; fast relaxation is attributed to fraying equilibria, and a broader fast relaxation to an increased heterogeniety due to the presence of mis-aligned intermediates.}
-
-\textcolor{blue}{The paper is generally well-written, and, I believe, represents a good attempt to tackle this important process.
-I personally have had a manuscript "desk-rejected" by JACS on the basis that it is "too focussed on DNA biotechnology", and "primarily of interest to RNA and DNA chemists".
-The same statements could be made about this paper.
-However, I disagree with the conclusion that such papers are not appropriate for JACS.
-I would therefore recommend publication, subject to satisfactory responses to the comments below.
+\textcolor{blue}{Recommendation: Publish in JACS after minor revisions.
 }
 
-\textbf{Author Reply:}   We are delighted by the reviewer's warm reception of our paper, and pleased that they find it to be appropriate for publication in JACS.
-
-\textcolor{blue}{1.
-I was frequently confused as to the temperatures used in the experiments and analysis.
-At one point, it is noted that simulations are performed at melting temperatures.
-However, simulations are actually performed at a range of temperatures and it isn't always clear when looking at a figure which temperatures were used for simulation, NN model comparison or experiments (see eg.
-Figs 2 and 3 and 5).
-I would recommend noting the temperature used to obtain the data in each case.}
-
-\textbf{Author Reply:}   We thank the reviewer for this suggestion and regret our lack of clarity in this matter.
-All simulations performed in this work were conducted at the melting temperature, with the exception of the simulations discussed in Section 3.1 where we consider a range of temperatures in order to draw comparisons to experimental measurements over this temperature range.
-We have added a number of clarifications to the text in order to more clearly convey this important point to the reader.
-
-In Section 2.1.1 of the Methods on p.~7, we have added:
-
-``\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC- core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.''
-
-In Section 2.1.2 of the Methods on p.~9, we have added:
-
-``In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective melting temperatures} from the 40$\times$25 simulation trajectories following a four-step protocol...''
-
-In the first sentence of Section 3.3 on p.~21, we have added:
-
-``We first compare the thermodynamic predictions for the equilibrium macrostate probabilities \rood{made by the MSM models fitted at the sequence melting temperatures} to those of the nearest neighbor (NN) model as a popular empirical model of DNA hybridization thermodynamics.''
-
-In the first sentence of the caption of Fig.~2, we have added:
-
-``Thermodynamic and kinetic predictions of the sequence-specific MSMs \rood{fitted at the sequence melting temperatures} for AT-all, GC-end, GC-core, and GC-mix.''
-
-In the first sentence of the caption of Fig.~3, we have added:
-
-``Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.''
-
-In the first sentence of Section 3.4 on p.~26, we have added:
-
-``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.''
-
-In the second sentence of the Conclusions in Section 4 on p.~36, we have added:
-
-``We conducted 1 ms of unbiased coarse-grained molecular dynamics simulations \rood{at the melting temperature of each sequence} and...''
-
-\textcolor{blue}{2. P7 line 13 - surely these simulations were not really done at 7M? That sounds impossibly dense.}
-
-\textbf{Author Reply:}   We regret our (embarrassing) typo -- the concentration is in fact 7 mM, not 7 M.
-The calculation is below and we have made the correction to Section 2.1.1 on p.~7.
-We regret our error.
-
-\begin{equation}\label{}
-\frac{2 \mathrm{\:molecules}}{\mathrm{box\:volume}} = \frac{2 \mathrm{\:molecules}}{(7.80 \times 7.80 \times 7.80) \mathrm{\:nm}^3} = \frac{3.32 \times 10^{-24} \mathrm{\:moles}}{4.75 \times 10^{-25} \mathrm{\:m}^3} = 7.0 \frac{\mathrm{moles}}{\mathrm{m}^3} = 7.0 \frac{\mathrm{millimoles}}{\mathrm{L}} = 7 \; \mathrm{mM}
-\end{equation}
-
-\textcolor{blue}{3.
-The authors repeatedly refer to "slithering" mechanisms for rearrangement.
-I believe this is unfortunate.
-In the context of the 3SPN model, the "slithering" term was first introduced in Ref.
-49 by Sambriski et al, and describes two helical single strands sliding along each other.
-As noted by Hinckley et al in Ref 18, that mechanism is largely an artifact of the form of the potential in 3SPN.1.
-The rearrangement mechanism in 3SPN.2 is more of a "defect diffusion", analogous to the "inchworm mechanism" previously identified in the oxDNA model (ref 52 - note, the author list for this reference is wrong).
-Given this context, it seems unnecessarily confusing to talk about "slithering".}
-
-\textbf{Author Reply:}    (I) We share the reviewer's concerns with using ``slithering'' to describe a process that is more appropriately referred to as ``defect diffusion'', ``inchworm mechanism'', or ``out-of-register diffusion''.
-
-In Section 2.1.2 on p.~9, we have modified the text to replace ``slithering'' with ``out-of-register diffusion'':
-
-``An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms.~\citep{Xiao2019}''}
-
-The only other place we use the term is in Section 3.4 on p.~29 where we refer to work conducted in Ref.~\citep{Xiao2019}.
-Here we choose to maintain the term ``slithering'' to maintain consistency with the mechanism nomenclature adopted in this paper, but place it in quotation marks to indicate that we are quoting from that paper and it is not a term that we would necessarily favor.
-
-``Out-of-register states for 5$^\prime$-GCGCGC-3$^\prime$ hexamers were identified as deep kinetic traps along the hybridization pathway and ``slithering'' through these states did not provide a significant hybridization pathway compared to an alternative ``zippering'' mechanism.
-(In contrast, out-of-register slithering and in-register zippering served as two parallel pathways for hybridization of 5$^\prime$-GGGGGG-3$^\prime$.)''
-
-(II) We also appreciate the reviewer pointing out the incorrect author list in ref 52, and we have fixed the issue.
-
-\textcolor{blue}{4.
-P15: I am a bit surprised that a separation of 1.3nm between the central base pairs is enough to guarantee strand separation, given that frayed end can separate by 1.3nm when the core is intact.
-Can the authors verify that all their duplexes had truly separated at this point?}
-
-\textbf{Author Reply:}   We motivated our choice of this distance as that beyond which the strands are effectively non-interacting, but also verified that our estimates of $k_d^{slow}$ are robust to choices of this cutoff in the range 0.8--2.5 nm.
-As an illustrative example, we demonstrate in \textbf{Figure R1} below that our extracted values of $k_d^{slow}$ are robust to changing the cutoff from 1.3 nm to 2.0 nm.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=\textwidth]{../FigR1.pdf}
-        \caption*{\textbf{Figure R1.} Comparing the experimental slow response when increasing the cutoff parameter from 1.3 nm to 2.0 nm}
-        \label{fig:compare_cutoffs}
-	\end{center}
-\end{figure}
-
-We have modified the relevant text in Section 3.1 on p.~15 to read:
-
-``First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
-This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and define the dissociated state.
-\rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.}''
-
-\textcolor{blue}{5. The authors refer to "exponential" increase in Fig. 1. However, it's hard to see that an increase is exponential by eye unless the data is plotted on a log scale.}
-
-\textbf{Author Reply:}   We agree that a log-linear plot would better expose these trends. We have replotted Fig.~1a and Fig.~S4 on log-linear axes and reproduce these below.
-
-\begin{figure}[ht!]
-	\begin{center}
-        \includegraphics[width=1.0\textwidth]{../Fig1.pdf}
-        \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses.
-Results are reported in terms of the final T-jump temperature.
-(a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction at long times.} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
-The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
-The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
-        \label{fig:relaxation-comparison_2}
-	\end{center}
-\end{figure}
-
-\clearpage
-\newpage
-
-% Fig. S4
-\begin{figure}[ht!]
-	\centering
-    \includegraphics[width=.8\textwidth]{../FigS4.pdf}
-   \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments.
-From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time.
-We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm.
-We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.
-Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits.
-In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse.}}
-    \label{fig:SIFig4_2}
-\end{figure}
-
-\clearpage
-\newpage
-
-\textcolor{blue}{6.
-Do the authors have any data on the scaling of k\_hybridization with T?
-In experiment, rates appear to increase with temperature (albeit with a much weaker dependence than de-hybridization rates).
-Simple "nucleation and zipper" models would predict the opposite, as does the oxDNA model (see ref 52).}
-
-\textbf{Author Reply:}   This is an interesting question and are, in fact, considering embarking upon some additional work in this vein.
-Rigorous engagement of this question would require us to fit MSMs at a variety of temperatures in order to compare the pathways, mechanisms, and rates for hybridization and dehybridization.
-In the present work, we have only constructed MSMs at the melting temperature of each sequence and the collection of sufficient data to parameterize MSMs at other temperatures would require both (i) a substantial investment of time to collect the $\sim$1 ms of data required to fit robust MSMs and (ii) the use of biasing techniques in order to adequately sample hybridization (dehybridization) events at large temperature excursions above (below) the melting temperature.
-As such, we do not have the necessary data to engage this question at the present time.
-
-Reviewer \#2 raised a related point in their Comment \#2 and in response to both of these points, we have made the following modification to the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting follow-up study:
-
-``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons.
-\rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to expose changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''
-
-\textcolor{blue}{7.
-The authors could explain a bit more carefully (perhaps in the SI) how they handle the multiple nearest-neighbour states that correspond to their microstates (eg.
-F4 clearly involves more than one NN configuration).}
-
-We have modified the text in the ``Nearest neighbor model of duplex thermodynamics'' text in the Supporting Information to better describe for the reader how we performed this calculation.
-Specifically, we have appended the following text at the end of this discussion:
-
-``\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
-Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
-As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}''
-
-\textcolor{blue}{8.
-I don't understand how the population of F4 could be >50\% that of H for the GC-core system, when the free energy of F4 is reported as > 5kJ/mol.
-My back-of-the-envelope maths predicts a ratio of 8:1 for a 5kJ/mol gap.
-Incidentally, 8:1 is roughly the ratio I got from running the configurations through NUPACK.}
-
-\textbf{Author Reply:}  We apologize for the confusion and have resolved the origin of this apparent inconsistency to two factors: (i) a unit conversion error on our part and (ii) our choice to present the bar plots in Fig.~2b on a log axis.
-
-(I) First, we made an (embarrassing) unit conversion error in reporting the $\Delta F$ value for the F4 state of the GC-core sequence from our MSM model.
-Specifically, we did not convert the $\Delta$F from units of $kT$ at $T_m$ = 324 K to kJ/mol, resulting us to erroneously report a value of 5.1 kJ/mol instead of 5.1 $kT$ = 13.7 kJ/mol.
-Using this correct $\Delta$F value we see agreement between the reported free energy in Fig.~3 and populations reported in Fig.~2.
-We can demonstrate this by performing the same back-of-the-envelope verification proposed by the reviewer:
-
-\begin{itemize}
-\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
-\item $G_{F4}$ = 5.1 $kT$ at $T_m$ = 324 K = 13.7 kJ/mol
-\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
-\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
-\end{itemize}
-
-We can solve this system of equations to yield $f_{F4}$ = 0.0030 that is in excellent agreement with the MSM predictions of $f_{F4}$ = 0.0028 that we report in Fig.~2b.
-
-We have updated Figure 3 with the correct $\Delta$F as shown below.
-
-\begin{figure}[ht!]
-	\begin{center} 
-        \includegraphics[width=140mm, scale=1]{../Fig3.pdf}
-        \caption*{Figure 3.
-Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$.
-We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model.
-(b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model.
-The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) relative to the hybridized state H compared to the NN model.}
-        \label{fig:NN_table}
-	\end{center}
-\end{figure}
-
-We have also updated the second paragraph of Section 3.3 on p.~23 as follows:
-
-``As illustrated in Fig.~3, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.
-Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stabilities?''
-
-We recognize that this was a significant miscalculation, and we have re-checked all free energy results to ensure that (i) we did not make a similar error for any of the other reported values in Fig.~3 or anywhere else and (ii) that this error did not propagate through to any subsequent calculations, results, or conclusions within the remainder of the manuscript.
-
-(II) We also note that our choice of a logarithmic y-axis for the bar plots in Fig.~2b that make the bar associated with the F4 state appear nearly 50\% of the height of the H and D bars for GC-core.
-We made this plotting choice in order to make visible on these bar plots the small state populations associated with the shifted and frayed states and noted the percentage occupancies on the bars themselves.
-To help eliminate this confusion, we have modified the portion of the Fig.~2 caption referring to this panel to read:
-
-``(b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities.
-Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts.
-\rood{Values are reported on a log y-axis to make the small populations of the shifted and frayed states visible.}''
-
-\textcolor{blue}{9.
-Is the MSM model explicitly constrained to obey detailed balance?
-The numbers that come out seem consistent with detailed balance - forward and backward trajectories at the same T essentially look the same (they're equally likely to go through intermediates).
-This should be a feature of the MSM, but it's unclear to me whether it's hard-coded or just arises naturally in the coarse-graining.}
-
-\textbf{Author Reply:}    The reviewer raises a good point regarding the assumptions made during MSM construction.
-We do indeed use a reversible MSM framework that enforces detailed balance.
-This is an important point which we have now made explicit by modifying the text in Section 2.1.2 on p.~11 to read:
-
-``The 200 microstates comprising each system were coarsened into our terminal macrostate MSM.
-\rood{Since we are interested in the construction of models of the equilibrium kinetics, we explicitly enforce detailed balance in the construction of the MSM.
-\citep{Wehmeyer2019IntroductionSoftware}}''
-
-and the text in Section 3.4 on p.~26 to read:
-
-``In addition to thermodynamic stabilities, the macrostate MSM also furnishes quantitative and interpretable predictions of hybridization and dehybridization pathways and mechanisms \rood{at the sequence melting temperatures}.
-\rood{It should be noted that because we are using a reversible MSM framework, detailed balance is enforced by construction.}''
-
-\textcolor{blue}{Relatedly, some of the discussion of the hybridization and de-hybridization pathways in non-repetitive sequences seems to ignore this fact.
-In these equilibrium simulations, hybridization trajectories played in reverse should be statistically indistinguishable from de-hybridization trajectories.
-I don't think anything that's said is wrong, per se, but it might be worth considering how the contrast between "nucleation and zippering" and "fraying-peeling" is consistent with the symmetry in the pathways.}
-
-\textbf{Author Reply:}  This is a nice observation, and we would do well to note this symmetry for the reader. We have added a sentence to this effect in Section 3.6 on p.~32:
-
-``In dehybridization, we observe fraying \rood{of one half of the duplex with the strands remaining associated until the loss of the final A:T pairs and ultimately the last G:C pair}.
-Qualitatively, we observed some short-lived states composed of two to four native WC base pair contacts immediately before full dissociation occurs, but, in contrast to the F4 state we observe in GC-core, these conformations do not constitute a metastable state within our MSM nor do they tend to reform intact duplexes.
-These dissociation dynamics are consistent with a ``fraying-peeling'' dehybridization mechanism.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA} \rood{We observe that the principle of microscopic reversibility for a molecular system at thermodynamic equilibrium imposes the conditions of detailed balance and symmetry of the classical equations of motion under time reversibility \cite{McCully2008MicroscopicHomeodomain}.
-A consequence of these considerations is that if our system is indeed at equilibrium, then the ``fraying-peeling'' dehybridization mechanism can be considered a reversal of the ``nucleation-zippering'' hybridization pathway since the time-reversed simulation trajectories represent an equally valid sampling of the system dynamics.
-This is indeed borne out by our mechanistic observations for GC-mix wherein the early stages of ``nucleation-zippering'' proceed by the formation of one G:C contact followed by one or more additional A:T pairs and the late stages by the formation of all remaining WC pairs, which we compare with the early stages ``fraying-peeling'' wherein one half of the duplex frays and the late stages wherein dissociation finally completes by the dissolution of the last few A:T contacts and the final G:C pair.}''
-
-\textcolor{blue}{10.
-I think some of the characterisation of the "all or nothing" picture is a bit misleading.
-The two state model of duplex formation is really a thermodynamic model, based on the assumption that the ensemble of hybridized and de-hybridized states look very different and are separated by a large barrier.
-This model is perfectly consistent with the transition between the two ensembles involving a complex network of metastable states.
-Indeed, as the authors note on line 3 of p4, a number of previous authors have considered what such states might look like, particularly for repetitive sequences.
-The authors' primary contribution here is not proposing that these states exist for the first time, but trying to tie together evidence from simulation and experiment to describe them.}
-
-\textbf{Author Reply:}   We thank the reviewer for this important observation and illuminating our need to better clarify this potentially ambiguous concept.
-We have modified the text in Section 3.6 on p.~29 to read:
-
-``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM.
-\rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time.
-(We note that ``all-or-nothing'' can also be used, somewhat ambiguously, to refer to a two-state thermodynamic model where the hybridized and dehybridized are separated by a large free energy barrier but that the transitions between them may proceed through a network of metastable states.)}''
-
-We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state'' so that the relevant text in Section 2.2.3 on p.~14 now reads:
-
-``FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a \rood{two-state model} to determine the fraction of intact duplex $\theta$ as a function of temperature.''
-
-\textcolor{blue}{11.
-Related to the above, the references on line 3 of p4 should probably also include Ref 52, where a similar analysis of metastable states in a repetitive sequence is performed (for a different coarse-grained model) and [Flamm, W., Fontana,J.M., Hofacker, I.L.
-and Schuster,P.
-(2000) RNA folding at elementary step resolution.
-RNA, 6, 325-338] where a defect diffusion mechanism is proposed.
+\textcolor{blue}{Comments: Jones et al. have updated their manuscript in response to reviewers' comments and improved it significantly. I am essentially happy to proceed towards publication.
 }
 
-\textbf{Author Reply:}   We have added these references to those at the top of p.~4.
+\textbf{Author Reply:} We are pleased that the reviewer finds the manuscript improved and suitable for publication in JACS subject to minor revision.
 
-\textcolor{blue}{12.
-The reasoning behind why the broader relaxation spectrum is indicative of misaligned states was not that clear to me.
-Why would "elevated" fraying cause a broader response?
-Wouldn't it also, then, be broader for CG-core, which has elevated fraying?
-It seems to me that the argument is more along the lines of the heterogeneity of the configurations leading to heterogeneous dynamics, and hence a broader relaxation profile.
-But I am no expert on this technique.}
+\textcolor{blue}{
+Some remaining issues that I would like to bring to the author's attention:
+}
 
-\textbf{Author Reply:}   The reviewer makes a good comment on our choice of language used to described the stretched AT-all response.
-It is indeed our contention that the heterogeneity of the configurations in the initial ensemble and the fraying dynamics exhibited in each of these states that is the origin of the broadening of the relaxation profile, but our wording here was insufficiently clear.
-We believe that the reviewer's phrasing ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into our edits of Section 3.7 on p.~33:
+\textcolor{blue}{
+(1) The authors have applied the finite size correction of Ouldridge et al.
+following a comment by reviewer 2, but it hasn't been applied in the correct way.
+It would be correct to apply it to the melting curves discussed at the bottom of p7/top of p8.
+Since the sequences are homodimers, the correction to Tm at the simulation concentration would actually be zero; instead, the widths of transitions would change, as noted by Ouldridge et al.
+}
 
-``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying.
-Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote an \rood{elevated fraying response over the course of the relaxation that is distinct from that of in-register fraying.
-This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix.
-It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.''
+\textcolor{blue}{
+However, the authors have actually simulated at 3.5x the experimental conc.
+If the Ouldridge correction - which assumes ideal behaviour - applies, then it should be relatively simple to correct the curves for the concentration difference.
+Simply divide the phi ratio by 3.5, and then calculate f\_hybridised using the modified phi.
+If the Tms were previously too low, this change will make them lower still (perhaps by a couple of K), but it is the fairest comparison.
+It shouldn't affect the rest of the analysis, although it is worth noting that the simulations were always performed "at the melting temperature for a conc of 7mM" - there is no well-defined, concentration-independent melting temperature.
+}
 
-\textcolor{blue}{13.
-How much of the code is being made available?
-Ideally, the initialization files for the simulations and the code used to perform the MSM analysis would be publicly available.
-Additionally, it would be good to publish the raw experimental data and the processing tools used to obtain the final results.}
+\textcolor{blue}{
+However, it is not appropriate to apply the Ouldridge correction to data measuring the dissociation time.
+The authors effectively measure the statistics of dissociation events in infinite dilution - these are unaffected by the factors that lead to the Ouldridge correction.
+More specifically, the Ouldridge correction accounts for concentration fluctuations neglected in small, canonical simulations.
+These concentration fluctuations contribute to the hybridisation rates, not the dissociation rates.
+}
 
-\textbf{Author Reply:}   We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).
-The links to both of these resources are now noted in the Supporting Information statement on p.~38:
+\textbf{Author Reply:} XXX
 
-``\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5140040}).}''
+1.
+Removal of correction.
+Report updated text here and in main text and SI using roodr code.
+Maybe something like ``(We note that although the melting temperature of the homodimer is not affected by the Ouldridge et al.
+finite size corrections [REF], this empirical correction does account for the higher DNA concentration employed in simulation (7 mM) relative to experiment (2 mM).
+Although we do not do so here, Owczarzy et al.
+present a prescription to apply concentration corrections to Tm using knowledge of the enthalpy and entropy of duplex formation and helix nucleation [REF; ``Predicting Sequence- Dependent Melting Stability of Short Duplex DNA Oligomers'' Biopoly 44: 217?239, 1997].''.
+In your response you can also elaborate on your point of using Eqn.
+11 in Ouldridge would fix conc effects at same T but not at new Tm so would need new T ramp or knowledge of dH and dS.
+Close with saying in any case our empirical correction accounts for these effects plus other uncontrolled effects such as model accuracy etc.
 
-\clearpage
-\newpage
+2. Adding ``at 7 mM'' at various places in text where necessary in defining Tm.)
+
+\textcolor{blue}{
+(2) I realise now that the criterion for dissociation is that both of the central pairs separate to 1.3nm, which is more reasonable than my previous understanding.
+However, although the authors say that the estimates of kd\^slow are robust to a range of choices for this cutoff, the yellow, red and green curves all have a noticeable shift between 1.3 and 2.0nm (unsurprisingly blue does not, since the GC pairs are in the core).
+So what exactly does robust mean?
+Why not just report the data for larger cutoffs?
+}
+
+\textbf{Author Reply:} XXX (Must draw some cutoff, our approach was to select physically motivated cutoff and show that results are insensitive to choice.
+Show that 1.3 nm satisfies these criteria -- results change by some small \% amount in switching cutoff to 2.0 nm so using either value is equally valid.)
+
+\textcolor{blue}{
+(3) In their response to point (7) in my original review, the authors talk about "averaging" over all microstates. Do they mean estimating F as
+}
+
+\textcolor{blue}{
+F = -kT ln sum\_i  exp(-F\_i/kT)
+}
+
+\textcolor{blue}{
+where i runs over all microstates in the macrostate? This is very different from averaging F\_i, as might be inferred from the text.
+}
+
+\textbf{Author Reply:} XXX (Modify SI to describe this procedure as weighted average.)
+
+%%%
 
 \begin{shaded}
-\textbf{Additional editorial edits}
+\textbf{Response to Reviewer \# 4}
 \end{shaded}
 
-\textcolor{blue}{1. Please include at least the first page for each reference citation for the following incomplete journal reference 48}
+\textcolor{blue}{Recommendation: In between publish elsewhere and accept without change.
+}
 
-Updated.
+\textcolor{blue}{Comments:
+The authors have addressed the comments of all the reviewers very thoughtfully and thoroughly.
+Although I'm still of the opinion that it's probably just below the level for JACS, I do think it is a very good paper, and wouldn't argue against acceptance if that's the majority view of the reviewers.
+}
 
-\textcolor{blue}{2. Please update the publication status of reference 70, if possible.  Providing the DOI is acceptable.}
+\textbf{Author Reply:} We thank the reviewer again for their time and effort in reviewing our work, and are pleased that they found all comments to have been thoroughly addressed.
 
-We are unable to locate any additional citation information beyond the arXiv link.
+\textcolor{blue}{
+PS I'm glad the 7M was just a typo!
+}
 
-\textcolor{blue}{3. Please include publisher for the following incomplete book reference: 79.}
-
-Updated.
-
-\textcolor{blue}{4.
-The TOC graphic should fit in an area no larger than 3.25 in.
-by 1.75 in.
-(approx.
-8.25 cm by 4.45  cm) and should have adequate resolution and clarity.
-Confirm that all text is legible at this size.}
-
-Updated and confirmed.
+\textbf{Author Reply:} Us too!
 

### 2021-09-09 response_letter_2/response_letter.tex
round 2 review comments

@@
 
-\textbf{Author Reply:} XXX
-
-1.
-Removal of correction.
+%Remove of correction.
 Report updated text here and in main text and SI using roodr code.
@@
 
-2. Adding ``at 7 mM'' at various places in text where necessary in defining Tm.)
+\textbf{Author Reply:} We appreciate the reviewer pointing out that the fluctuation correction should not be applied at an effective infinite dilution, and we have removed the fluctuation correction discussion and text from section 3.1.
+We have also updated the Figure 1 plot and caption in line with the uncorrected (original) data.
+As mentioned in the previous response letter, the uncorrected rates maintain high qualitative similarity to the corrected rates, and the slight deviation does not change our interpretation of the data.
+
+We have also edited the following line on page 17. which erroneously suggested that the homodimer fluctuation correction would affect melting temperature predictions.
+
+`` The computational time scales and melting temperatures reported in the remainder of the paper are not \roodr{corrected by these calibration corrections} since we are only interested in the relative trends in the behaviors of the four sequences.''
+
+We agree with the reviewer's suggestion to specify that melting temperatures are only valid under the corresponding experimental conditions.
+As such we have revised the text in several places to note we are using the melting temperature specific to the 7 mM concentration:
+
+On page 7:
+
+``each sequence was simulated at its \rood{respective melting temperature at 7mM concentration as dictated by the 3SPN.2 model}''
+
+On page 9:
+
+`` MSMs were constructed for each of the four DNA sequences \roodr{at their respective 3SPN.2 melting temperatures at 7 mM concentration} from the 40$\times$25 $\mu$s simulation trajectories''
+
+On page 20:
+
+``Since each system is simulated at the \roodr{same 7 mM concentration} and at its respective melting temperature''
+
+The reviewer helpfully points out that equation (7) in Ouldridge et al.
+\cite{Ouldridge2010ExtractingSimulations} may be used to determine the effective dimer fraction at the experimental concentration.
+Furthermore, a new melting curve could be refit using this concentration correction in order to determine the effective melting temperature at the experimental concentration.
+Although this relation would not changed our approach -- given our explicit goal to ``maximize the number of spontaneous transitions between dissociated and hybridized states'' -- we agree the reader should be aware that the predicted melting temperatures would in fact be slightly lower than our simulated temperatures.
+We have specified this below:
+
+On page 17:
+``Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.
+\roodr{We note that we would expect the model to slightly over-predict melting temperature due to the 3.5$\times$ higher concentration relative to experiment, however the systematic shift shows that relative sequence-dependent trends are well maintained.
+Furthermore,} We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.''
 
@@
 
-\textbf{Author Reply:} XXX (Must draw some cutoff, our approach was to select physically motivated cutoff and show that results are insensitive to choice.
-Show that 1.3 nm satisfies these criteria -- results change by some small \% amount in switching cutoff to 2.0 nm so using either value is equally valid.)
+\textbf{Author Reply:} Our intention was to maintain a consistent cutoff between the fraying and dissociation analysis, however the reviewer points out that this is not a necessary constraint given that we are measuring fundamentally difference dynamical processes.
+When comparing the 1.3 nm cutoff to a 2.0 nm, we found an average of only 7\% deviation between slow rates, and we see similar deviation up to 2.5 nm.
+Although this change may be visually perceptible it does not change our interpretation of the data, and we consider the sequence-dependent rates ``robust'' against cutoff selection.
+Still, we agree with the reviewer that the choice of a larger cutoff reduces sequence-dependent bias, and we have we have elected to update the Figure 1 slow response with the higher 2.0 nm cutoff value.
+We have also updated the text as follows:
+
+On page 15:
+
+``First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of \roodr{2.0} nm starting from an initial fully hybridized duplex.
+This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
+\roodr{We verified that our results were robust to choices of this cutoff over the range 1.3-2.5 nm.
+as measured by a <7\% average deviation in rates over that range.}''
 
@@
 
-\textbf{Author Reply:} XXX (Modify SI to describe this procedure as weighted average.)
+\textbf{Author Reply:} We understand the confusion regarding our free energy calculation, and we have added the following text to page S10 of the Supporting Information for clarification:
+
+\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.} \roodr{To evaluate the extent to which incorporating microstate ensembles would change our free energy estimates, we calculated a NN microstate free energy by taking a simple average of the NN free energy associated with each configurations in a given microstate.
+This is a reasonable assumption given the structural similarity of configurations within a microstate.
+Macrostate free energies were then calculated by weighting each microstate free energy according the its corresponding probability in the MSM stationary distribution.
+This can be represented as follows, where i indexes configurations, j indexes microstates, and $p_j$ represents the microstate probability.}
+
+\begin{equation}
+    F^{micro} = \frac{1}{N} \sum_{i \in micro} F_i^{NN} 
+\end{equation}
+
+\begin{equation}  
+    F^{macro} = \sum_{j \in macro} F_j^{micro} p_j
+\end{equation}
+
+\roodr{Recalculating the NN predictions in this way leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}
 

### 2021-09-09 SI.tex
Draft second round of reveiew responses
round 2 review comments

@@
 
-\newcommand*{\rood}[1]{{\color{red}{#1}}}
-%\newcommand*{\rood}[1]{#1}
+%\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\rood}[1]{#1}
 %\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
@@
 %\newcommand*{\blauwr}[1]{#1}
+\newcommand*{\roodr}[1]{{\color{red}{#1}}}
 
@@
 \rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
-Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.} \roodr{To evaluate the extent to which incorporating microstate ensembles would change our free energy estimates, we calculated a NN microstate free energy by taking a simple average of the NN free energy associated with each configurations in a given microstate.
+This is a reasonable assumption given the structural similarity of configurations within a microstate.
+Macrostate free energies were then calculated by weighting each microstate free energy according the its corresponding probability in the MSM stationary distribution.
+This can be represented as follows, where i indexes configurations, j indexes microstates, and $p_j$ represents the microstate probability.}
+
+\begin{equation}
+    F^{micro} = \frac{1}{N} \sum_{i \in micro} F_i^{NN} 
+\end{equation}
+
+\begin{equation}  
+    F^{macro} = \sum_{j \in macro} F_j^{micro} p_j
+\end{equation}
+
+\roodr{Recalculating the NN predictions in this way leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
 As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}

### 2021-09-09 manuscript.tex
Draft second round of reveiew responses
round 2 review comments

@@
 
-\newcommand*{\rood}[1]{{\color{red}{#1}}}
+%\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\rood}[1]{{#1}}
 %\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
 \newcommand*{\blauw}[1]{{#1}}
+\newcommand*{\roodr}[1]{{\color{red}{#1}}}
 
-\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
-\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
+%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
+%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
 %\newcommand*{\noter}[1]{}					% notes off
-\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
+%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes
 
@@
 \citep{Schneider1978Molecular-dynamicsTransitions}.
-\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature at 7mM concentration as dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
 Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
@@
 MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and \rood{out-of-register diffusion mechanisms}.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima using a transition state theory approximation.
-A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences \rood{at their respective 3SPN.2 melting temperatures} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
+A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work,  MSMs were constructed for each of the four DNA sequences \roodr{at their respective 3SPN.2 melting temperatures at 7 mM concentration} from the 40$\times$25 $\mu$s simulation trajectories following a four-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference.
 Calculations were performed using the PyEMMA software package.
@@
 We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}).
-First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex.
+First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of \roodr{2.0} nm starting from an initial fully hybridized duplex.
 This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
-\rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
-\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
+\roodr{We verified that our results were robust to choices of this cutoff over the range 1.3-2.5 nm.
+as measured by a <7\% average deviation in rates over that range.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+
+%\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long times, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields.
 We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ as,}
 
-    \begin{equation}\label{}
-    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
-    \end{equation}
+%    \begin{equation}\label{}
+%    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
+%    \end{equation}
 
-    \begin{equation}\label{}
-	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
-	\end{equation}
+%    \begin{equation}\label{}
+%	f_\mathrm{hybridized} = (1+\frac{1}{4\phi}) - \sqrt{(1+\frac{1}{4\phi})^2 - 1}
+%	\end{equation}
 
@@
 Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.
-We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
-The computational time scales and melting temperatures reported in the remainder of the paper are not corrected \rood{by these calibration corrections or finite-size fluctuation corrections} since we are only interested in the relative trends in the behaviors of the four sequences.
+\roodr{We note that we would expect the model to slightly over-predict melting temperature due to the 3.5$\times$ higher concentration relative to experiment, however the systematic shift shows that relative sequence-dependent trends are well maintained.
+Furthermore,} We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
+The computational time scales and melting temperatures reported in the remainder of the paper are not \roodr{corrected by these calibration corrections} since we are only interested in the relative trends in the behaviors of the four sequences.
 
@@
 (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales.
-\rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction at long times.} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
+(b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales.
 The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics.
@@
 The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization.
-Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization.
+Since each system is simulated at the \roodr{same 7 mM concentration} and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization.
 The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways.

### 2021-09-11 response_letter_2/response_letter.tex
Response 2 Andy edits

--- 2021-09-08 response_letter_2/response_letter.tex
@@
 %\newcommand*{\rood}[1]{\color{red}{#1}}
-\newcommand*{\rood}[1]{{\color{red}{#1}}}
+\newcommand*{\rood}[1]{{#1}}
 \newcommand*{\roodr}[1]{{\color{red}{#1}}}
@@
 \textbf{Author Reply:} We are pleased that the reviewer is satisfied with our modifications and responses and that they are pleased to support publication in JACS.
-We would like to thank them again for their time, effort, and considered and useful expert comments.
+We would like to thank them again for their time, effort, and expert comments.
 
@@
 
-2. Adding ``at 7 mM'' at various places in text where necessary in defining Tm.)
+\textbf{Author Reply:} We appreciate these valuable comments and have made several corrections and modifications to the revised manuscript in light of these recommendations.
+
+(I) We thank the reviewer for pointing out that the Ouldridge finite-size fluctuation corrections do not affect the dissociation rates and are therefore not relevant for this comparison.
+We have eliminated our discussion of this correction at the top of p.~16 in Section 3.1 (``To correct for finite-size effects...'') and modified the plot and caption in Figure 1 to report the original uncorrected data.
+As mentioned in our previous response letter, the fluctuation correction led to only a slight change in the reported rates and so this modification does not change any of our analysis or interpretations of the data.
+
+It is true that the homodimer finite-size fluctuation correction does not affect the melting temperature but only widens the transition, and this was indeed what we saw when we applied the correction.
+We have also edited the text on p.~17 that erroneously suggested that this correction would change the dissociation rate or melting temperature:
+
+``The computational time scales and melting temperatures reported in the remainder of the paper are not corrected by these calibration corrections \roodr{\st{or finite-size fluctuations}} since we are only interested in the relative trends in the behaviors of the four sequences.''
+
+(II) We appreciate reviewer's suggestion to be more explicit that melting temperatures are a function of concentration, and that our definition applies to a particular model at a particular concentration.
+As such we have revised the text in several places to note we are using the melting temperature specific to the 7 mM concentration employed in our simulations:
+
+On page 7:
+
+``each sequence was simulated at its respective melting temperature \roodr{at 7mM concentration as} dictated by the 3SPN.2 model''
+
+On page 9:
+
+`` MSMs were constructed for each of the four DNA sequences \roodr{at their respective 3SPN.2 melting temperatures at 7 mM concentration} from the 40$\times$25 $\mu$s simulation trajectories''
+
+On page 20:
+
+``Since each system is simulated at the \roodr{same 7 mM concentration} and at its respective melting temperature''
+
+(III) We thank the reviewer for the suggestion to correct our simulation prediction of the melting temperature to the experimental concentration.
+The proposed strategy is to assume ideal behavior, reduce the computed $\phi$ ratio per Eqn.~7 in Ouldridge et al.\ \cite{Ouldridge2010ExtractingSimulations} by the 3.5$\times$ concentration difference, and then recompute $f_{hybridized}$.
+We believe that this procedure would work under the assumption that the equilibrium constant does not change with temperature, but in general this may not be a good assumption since $-R T_m K_{eq} = \Delta G^\circ = \Delta H^\circ - T_m \Delta S$.
+Owczarzy et al.\ \cite{Owczarzy1997predicting} engage this issue to develop a van't Hoff-like concentration correction to $T_m$ based on the enthalpies of duplex formation and helix nucleation.
+In the absence of knowledge of these standard enthalpies, we could instead calculate the melting temperature by performing multiple simulations at a range of temperatures, applying the proposed concentration correction to predict $f_{hybridized}$ at that temperature, then fitting a melting curve from which we could extract the concentration corrected $T_m$.
+This additional suite of required calculations would be exclusively in service of nailing down the concentration correction and would not change our approach since we intentionally selected the 7 mM 3SPN.2 melting temperature in order to maximize the number of spontaneous transitions between dissociated and hybridized states to provide a rich data set of these events over which to train our MSM.
+
+The approach we take to the concentration correction in the manuscript is an empirical one, wherein we draw a comparison of the computational predictions against experimental measurements.
+The observed calibration corrections therefore account for net effect of both the 3.5$\times$ higher concentration in simulation and the assumptions and approximations inherent in the 3SPN.2 model.
+As the reviewer observes, the (+4) K calibration correction encompassing both of these effects is overly optimistic since the concentration corrected $T_m$ will be depressed and controlling for this would require a larger positive correction for the error due to the 3SPN.2 model approximations.
+
+In light of the reviewer's suggestion and the discussion above, we have modified the text on p.~16 to alert the reader to these important issues:
+
+``Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a universal (+4) K corrective temperature shift to our computational results.
+\roodr{We note that our simulations are expected to slightly over-predict the melting temperature since they are conducted at 3.5$\times$ higher concentration relative to experiment, and the concentration-adjusted corrective shift accounting for the approximations inherent in the 3SPN.2 model would be slightly larger.
+Owczarzy et al.\ present an analytical prescription to apply concentration corrections to the melting temperature using knowledge of the enthalpy changes associated with duplex formation and helix nucleation \cite{Owczarzy1997predicting}.
+In the absence of these quantities, one could instead empirically estimate concentration corrected melting temperatures by conducting a suite of simulations over a range of temperatures and assume ideal molecular behavior to apply concentration corrections to the observed hybridized fractions \cite{Ouldridge2010ExtractingSimulations}.
+(As demonstrated by Ouldridge et al., finite-size corrections do not affect the melting temperature for homodimers \cite{Ouldridge2010ExtractingSimulations}.) Furthermore, we} note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.''
 
@@
 
-\textbf{Author Reply:} XXX (Must draw some cutoff, our approach was to select physically motivated cutoff and show that results are insensitive to choice.
-Show that 1.3 nm satisfies these criteria -- results change by some small \% amount in switching cutoff to 2.0 nm so using either value is equally valid.)
+\textbf{Author Reply:} Our original motivation for selecting a 1.3 nm cutoff was to maintain consistency between that for fraying and dissociation.
+This is, however, an unnecessary constraint since these are fundamentally different processes.
+As the reviewer recommends, we now report the data in Figure 1 using the larger 2.0 nm cutoff.
+We do indeed see relatively small changes in our results upon changing the cutoff from 1.3 nm to 2.0 nm, producing an average change in the calculated values of $k_d^\mathrm{slow}$ of only 7\%, demonstrating that our calculation is robust to the particular choice of cutoff within this range.
+This modification does not affect any of our subsequent analysis or interpretation.
+
+We have updated the text on p.~15 as follows:
+
+``First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of \roodr{2.0} nm starting from an initial fully hybridized duplex.
+This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
+We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\roodr{We verified that our cutoff was sufficiently large by observing that our calculated values for $k_d^\mathrm{slow}$ changed by an average of only 7\% by adopting a 1.3 nm cutoff.}''
 
@@
 
-\textbf{Author Reply:} XXX (Modify SI to describe this procedure as weighted average.)
+\textbf{Author Reply:} Thank you for pointing out our lack of clarity on this point.
+We have modified the text on p.~S10 of the Supporting Information to more clearly convey the details of this calculation:
+
+\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.} \roodr{To evaluate the extent to which the microstate ensembles would change our free energy estimates, we calculated a NN microstate free energy by taking a simple average over the NN free energies predicted for each configuration within a given microstate -- a reasonable assumption given the structural similarity of configurations within a microstate -- then computing macrostate free energies via a weighted average according to the MSM equilibrium occupancy probabilities for each microstate.
+These operations can be represented as follows, where $i$ indexes over the $N_j$ configurations in microstate $j$, $j$ indexes over microstates, and $p_j$ represents the MSM equilibrium occupancy of microstate $j$,}
+\begin{equation}
+    F^{micro} = \frac{1}{N_j} \sum_{i \in micro} F_i^{NN} 
+\end{equation}
+\begin{equation}  
+    F^{macro} = \sum_{j \in macro} p_j F_j^{micro}
+\end{equation}
+Recalculating the NN predictions \roodr{in this way} leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.
 

### 2021-09-11 SI.tex
Response 2 Andy edits

--- 2021-08-04 SI.tex
@@
 \rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a.
-In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.
-Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
-As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}
+In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model.} \roodr{To evaluate the extent to which the microstate ensembles would change our free energy estimates, we calculated a NN microstate free energy by taking a simple average over the NN free energies predicted for each configuration within a given microstate -- a reasonable assumption given the structural similarity of configurations within a microstate -- then computing macrostate free energies via a weighted average according to the MSM equilibrium occupancy probabilities for each microstate.
+These operations can be represented as follows, where $i$ indexes over the $N_j$ configurations in microstate $j$, $j$ indexes over microstates, and $p_j$ represents the MSM equilibrium occupancy of microstate $j$,}
+\begin{equation}
+    F^{micro} = \frac{1}{N_j} \sum_{i \in micro} F_i^{NN} 
+\end{equation}
+\begin{equation}  
+    F^{macro} = \sum_{j \in macro} p_j F_j^{micro}
+\end{equation}
+Recalculating the NN predictions \roodr{in this way} leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol.
+As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.
 

### 2021-09-11 manuscript.tex
Response 2 Andy edits

@@
 \citep{Schneider1978Molecular-dynamicsTransitions}.
-\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its \rood{respective melting temperature at 7mM concentration as dictated by the 3SPN.2 model} -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
+\rood{With the exception of the simulation data reported in Section 3.1 where we draw comparisons against experimental data collected over a range of temperatures,} each sequence was simulated at its respective melting temperature \roodr{at 7mM concentration as} dictated by the 3SPN.2 model -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states.
 Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature,~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} and selecting the temperature at which the hybridized and dissociated populations were approximately equal.
@@
 This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state.
-\roodr{We verified that our results were robust to choices of this cutoff over the range 1.3-2.5 nm.
-as measured by a <7\% average deviation in rates over that range.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.
+\roodr{We verified that our cutoff was sufficiently large by observing that our calculated values for $k_d^\mathrm{slow}$ changed by an average of only 7\% by adopting a 1.3 nm cutoff.}
 
@@
 Although the 3SPN.2 model reproduces melting temperatures relatively well, we observed a systematic 4 K under-prediction relative to experiment and so we apply a \rood{universal (+4) K corrective temperature shift} to our computational results.
-\roodr{We note that we would expect the model to slightly over-predict melting temperature due to the 3.5$\times$ higher concentration relative to experiment, however the systematic shift shows that relative sequence-dependent trends are well maintained.
-Furthermore,} We note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
-The computational time scales and melting temperatures reported in the remainder of the paper are not \roodr{corrected by these calibration corrections} since we are only interested in the relative trends in the behaviors of the four sequences.
+\roodr{We note that our simulations are expected to slightly over-predict the melting temperature since they are conducted at 3.5$\times$ higher concentration relative to experiment, and the concentration-adjusted corrective shift accounting for the approximations inherent in the 3SPN.2 model would be slightly larger.
+Owczarzy et al.\ present an analytical prescription to apply concentration corrections to the melting temperature using knowledge of the enthalpy changes associated with duplex formation and helix nucleation \cite{Owczarzy1997predicting}.
+In the absence of these quantities, one could instead empirically estimate concentration corrected melting temperatures by conducting a suite of simulations over a range of temperatures and assume ideal molecular behavior to apply concentration corrections to the observed hybridized fractions \cite{Ouldridge2010ExtractingSimulations}.
+(As demonstrated by Ouldridge et al., finite-size corrections do not affect the melting temperature for homodimers \cite{Ouldridge2010ExtractingSimulations}.) Furthermore, we} note that these empirical calibration factors to the 3SPN.2 predictions are applied only for the purposes of making an experimental comparison, but acknowledge that there are uncertainties introduced by assuming that the equilibrium dynamics at fixed temperature can be compared directly to relaxation kinetics following a 15$^\circ$C T-jump.
+The computational time scales and melting temperatures reported in the remainder of the paper are not corrected by these calibration corrections since we are only interested in the relative trends in the behaviors of the four sequences.
 

### 2021-09-27 response_letter_2/response_letter.tex
Improved resolution
JACS resubmission

@@
 (III) We thank the reviewer for the suggestion to correct our simulation prediction of the melting temperature to the experimental concentration.
-The proposed strategy is to assume ideal behavior, reduce the computed $\phi$ ratio per Eqn.~7 in Ouldridge et al.\ \cite{Ouldridge2010ExtractingSimulations} by the 3.5$\times$ concentration difference, and then recompute $f_{hybridized}$.
+The proposed strategy is to assume ideal behavior, reduce the computed $\phi$ ratio per Eqn.~7 in Ouldridge et al.\ \cite{Ouldridge2010ExtractingSimulations} by the 3.5$\times$ concentration difference, and then recompute $f_\mathrm{hybridized}$.
 We believe that this procedure would work under the assumption that the equilibrium constant does not change with temperature, but in general this may not be a good assumption since $-R T_m K_{eq} = \Delta G^\circ = \Delta H^\circ - T_m \Delta S$.
 Owczarzy et al.\ \cite{Owczarzy1997predicting} engage this issue to develop a van't Hoff-like concentration correction to $T_m$ based on the enthalpies of duplex formation and helix nucleation.
-In the absence of knowledge of these standard enthalpies, we could instead calculate the melting temperature by performing multiple simulations at a range of temperatures, applying the proposed concentration correction to predict $f_{hybridized}$ at that temperature, then fitting a melting curve from which we could extract the concentration corrected $T_m$.
-This additional suite of required calculations would be exclusively in service of nailing down the concentration correction and would not change our approach since we intentionally selected the 7 mM 3SPN.2 melting temperature in order to maximize the number of spontaneous transitions between dissociated and hybridized states to provide a rich data set of these events over which to train our MSM.
+In the absence of knowledge of these standard enthalpies, we could instead calculate the melting temperature by performing simulations over a range of temperatures, applying the proposed concentration correction to predict $f_\mathrm{hybridized}$ at that temperature, then fitting a melting curve from which we could extract the concentration-corrected $T_m$.
+This additional suite of calculations would be exclusively in service of nailing down the concentration correction and would not change our approach since we intentionally selected the 7 mM 3SPN.2 melting temperature in order to maximize the number of spontaneous transitions between dissociated and hybridized states to provide a rich data set of these events over which to train our MSM.
 
@@
 The observed calibration corrections therefore account for net effect of both the 3.5$\times$ higher concentration in simulation and the assumptions and approximations inherent in the 3SPN.2 model.
-As the reviewer observes, the (+4) K calibration correction encompassing both of these effects is overly optimistic since the concentration corrected $T_m$ will be depressed and controlling for this would require a larger positive correction for the error due to the 3SPN.2 model approximations.
+As the reviewer observes, the (+4) K calibration correction encompassing both of these effects is overly optimistic since the concentration-corrected $T_m$ will be depressed and controlling for this would require a larger positive correction for the error due to the 3SPN.2 model approximations.
 
